### 2009-05-04 doc/report/msc-report.tex
added report to docs.

\documentclass[a4paper,10pt]{book}

\author{Per Andersson, Fredrik M{\"o}llerstrand}
\title{OHM -- OHM HBase Masseuse [DRAFT]}

\begin{document}

\maketitle

\noindent \Large{\textbf{Abstract}}

\vspace{12pt}

\noindent This is where we put down two to four sentences about:
Introduction, problem, existing tools, our solution and implementation,
conclusions.

\tableofcontents

\vfill

\pagebreak

\chapter{Introduction}

As large scale web services grow they generate vast volumes of log data,
for instance Apache web logs. The ability to accurately analyse gathered
data is important to the evaluation, planning, and success of the web
service. Storing and analysing massive volumes of log data requires
special measures to be taken. From an end-user perspective the requirement
is an easy, complete and fast data analysing procedure. The technical
aspects of this requirement boils down to the following key points:
Visualising, storing and importing the data. The result is an
multidimensional information system which is subject to
OLAP\footnote{On-Line Analytical Processing}. \cite{olap_solutions}

All three requirements are depend on one another, what data and how it is
stored affects what can be visualised. Also, what data is imported--and
how it is analysed--affects how it can be stored and hence visualised.

It is assumed that the audience of this paper has at least a basic
understanding of how one models data in relational databases. As will
quickly become apparent, a BigTable-like \cite{bigtable} data store is
drastically different from the relational model, meaning that those who
know their table schemas will have some un-learning to do before being able
to fully grasp the leaner BigTable-model.

% TODO: necessity of this paragraph?
The reader is also assumed to possess basic understanding about distributed
systems, with focus on high level abstractions rather than detailed
knowledge about practical implementation details.

\section{BigTable}

The data model is a simple one: Rows of data are stored in named tables.
Each row consists of a sortable row key and an arbitrary number of columns.
The columns are of the form "column-family:qualifier", where the column
family is one of a number of fixed column families defined by the table,
and the qualifier is an arbitrary string specified by the application. In
effect, column families are pre-announced while qualifiers are not. The
contents of each column family are stored together, so the user will want
to store items that have similar characteristics in the same column family.
Each data point is called a cell. A cell is an uninterpreted array of
bytes-- there are no data types, and it is up to the application to
interpret the data correctly. The table can store any number of timestamped
versions of each cell, allowing versioning of data.

The model above can indeed be thought of as a sparse distributed
multidimensional sorted map. In concept, this multidimensional map is
identical to a nested hash map from your favourite programming language. The
dimensions of this multidimensional map is mapped out on the row key, the
column family and the column qualifier, and possibly also of the version
timestamp.

In actual practice, the main difference between the two models is that
querying is highly restricted in a BigTable-like data store--there are no
joins, for example--whereas the relational database was born to be queried.
On the other hand, the restrictions placed upon the BigTable models are
what allows it to scale to such a large dataset. Indeed, BigTable was born
to scale.

\textbf{TODO: -- elaborate on reasons to why starsql does not scale--}

Many applications have the behaviour where they write data once and read it
a lot more, log analysers for instance. This writing-once-reading-many
behaviour is precisely what BigTable is designed for. BigTable is designed
for sequential, as opposed to random, data access. A typical use case is
the scan, where a large number of rows are read in sequence and a certain
column or column-family are picked up and returned. Certainly, you will
want to store those data points that are likely to be needed simultaneously
close to each other. \cite{bigtable}

\section{MapReduce}

Analysing the massive amounts of log data raises high demand for computing
power. The MapReduce \cite{mapreduce} pattern suits the analysing issue
well. The great benefit of MapReduce is that it runs on a computer cluster
of commodity hardware where computation takes place independently in
parallel. All results computed by each node are then reduced into one
single result.

% ONE DOES NOT SIMPLY PROCESS THE DATA ON ONE COMPUTER INTO MORDOR.
Basically MapReduce is a two phase computation. First the input data is
partitioned and each of these partitions is sent to a \textit{mapper} which
performs computations on each row of the data. The mapper then outputs each
resulting computation on the format \texttt{key,value}. These key-value
pairs are received by a \textit{reducer} which performs the second stage of
the computation.

\section{Summary}

Regarding the infrastructure, both BigTable and MapReduce has high
availability, scalability and reliability. Since BigTable runs on top of
GFS\footnote{Google File System}, the data stored in BigTable is
replicated, usually three times, onto other nodes in the cluster.
\cite{gfs} Likewise, if a node in a MapReduce job goes down or fails to
complete the job it is reassigned to another node. \cite{mapreduce}

\chapter{Problem description}

\section{Scope}

Time-series based analysis is the most important one when analysing web
logs, it comes natural since all log entries are timestamped.
\cite{discoveringweb} It is also the most common form we have seen for logs
in general.

Assuming that time-series analysis is the most important, we have imposed
the restriction on the system that is built so that it only handles
time-series. Because of this restriction the data store obviously becomes
less general purpose. Although since most log data is time-series based
this restriction does not cripple the system much--if at all.

\section{Data size}

The sheer size of the data makes it unfeasible to use a single computer.
Eventually a single machine would be fed so much data that it cannot cope
with it. For instance a single web log file for one day can be up to
several gigabytes. It is problematic for a single machine to handle this
because of the limitations of memory, loading the entire file in memory.
Distributed computing solves this by spreading the computations out to
several machines in parallel. The obvious gain is that the waiting time
before a computation is done is divided by the amount of nodes used,
compared to running the computation on one machine. There is also a gain
related to the actual machines, it is possible to use commodity hardware.
Because MapReduce and BigTable has high availability and reliability a node
in the cluster can safely be replaced on the fly if it fails.

Scalability is also very important. Using single machines for computing is
problematic when the dataset grows, more powerful machines has to be
obtained. More power contained in a single machine results in higher price,
eventually it is too expensive to perform computations on very large
datasets. Both MapReduce and BigTable scale close to linearly, adding one
machine to the cluster adds equal amount of processing power. Since
MapReduce and BigTable can use commodity hardware it is reasonable to buy
extra machines to gain extra computing power or storage space.

\chapter{Data model}

\section{Sorted data}

\subsection{Rows}

Each row in a BigTable is associated with a unique key. All the data is
stored sorted by rowkey. Because of this it is important to have a thorough
analysis of how the data is going to be accessed in order to store it
efficiently.

\subsection{Regions}

BigTable use the concept of tablets, HBase in turn uses regions. In order
to be consistent with the rest of this paper we henceforth use the term
region.

A region is a partition of the data stored on a node in the computer
cluster. A region server (referred to as tablet server in \cite{mapreduce})
holds information on which regions are located on which nodes. More
precisely the region server knows what range of rowkeys are in each region
on each node. This behaviour makes it fast to access data while keeping
reliability and scalability features.

\section{Dimensions}

\subsection{Column-families}

Each column-family is stored as a single map-file in hbase.

\section{Hierarchies}

Depending on how the rowkeys are formatted different hierarchies are
created. The order in which the data is stored is very important, it
defines how effective scanning for data will be. For instance, let's say we
have thousand rows and want ten out of those. In this case only one per
cent of the data is interesting to us, if a scan would be required to visit
all the thousand rows a lot of rows are skipped. Skipping rows is expensive
in that sense that we have to read them but throw away the result. The goal
is to push this cost to a bare minimum, using as many as possible of the
rows we scan.

To minimize the cost of scanning for data we want it to be ordered so that
a scan would use every row it visits. In the case with thousand rows of
which we want ten, it would mean that we just scan ten rows and use all of
them.

\chapter{OHM}

We have built a system--henceforth called OHM--which has the goal of being
a large scale data warehouse for time-series based log-like data sources.

\section{Deployment}

A manage script is used to create tables from a pre-defined schema.

\subsection{Query language}

\section{Import}

The user writes MapReduce programs which imports log-like data to the data
store.

After the data is imported a metadata scanner is run which creates metadata
about the imported data.

\section{Store data}

Data is stored in a BigTable like database. A natural consequence of this
that the data is stored sorted by rowkey allowing fast access to an
arbitrary (sequential) range.

\section{Export}

It is possible to query OHM about metadata it has on each of the imported
data sets.

Data is served to the user either over HTTP in JSON format or over Thrift.
\cite{thrift}

\chapter{Implementation}

OHM is built upon existing tools and software. It uses Apache Hadoop and
HBase, free implementations of Googles' MapReduce and BigTable
respectively, both implemented in Java.

When data is exchanged between OHM and the user either
JSON\footnote{JavaScript Object Notation} is transported over a RESTful
interface or data is transported with Thrift.

OHM is implemented in Python.

\section{Administration}

\subsection{YAML}

\section{Apache Hadoop}

\subsection{Hadoop Distributed File System}

\subsection{Dumbo}

\section{Apache HBase}

\section{Data interchange}

\subsection{Input and output formats}

\subsection{Thrift}

\subsection{REST}

\subsection{JSON}

\chapter{Examples}

\section{Web log analysis}

\chapter{Conclusion}

\chapter{Related work}

HDW is system which goal is to create a high performance large scale data
warehouse based on MapReduce and Bigtable. \cite{hdw}

\pagebreak

\bibliographystyle{plain} \bibliography{references}

\end{document}

### 2009-05-04 doc/report/msc-report.tex
minor additions and changes to report.

@@
 \author{Per Andersson, Fredrik M{\"o}llerstrand}
-\title{OHM -- OHM HBase Masseuse [DRAFT]}
+\title{Zonotoop--a large scale aggregating data warehouse for
+time-series-based data. [DRAFT]}
 
@@
 
-\chapter{OHM}
+\chapter{Usage}
 
 We have built a system--henceforth called OHM--which has the goal of being
-a large scale data warehouse for time-series based log-like data sources.
+a large scale aggregating data warehouse for time-series-based log-like
+data sources.
 
@@
 
-\subsection{Query language}
-
 \section{Import}
 
-The user writes MapReduce programs which imports log-like data to the data
-store.
+The user writes the map part of the MapReduce programs which imports
+log-like data to the data store. The reduce step is taken care of by OHM,
+which uses sum reducer that simply sums up values with the same keys.
 
-After the data is imported a metadata scanner is run which creates metadata
-about the imported data.
+\subsection{User mappers}
 
-\section{Store data}
+Import jobs use mappers defined by the user, either in Python or Java.
+Mappers written by the user import each line from the input file and set
+the arguments to the map function, key and value, as line number and line
+contents, respectively. The user can then perform computations on these
+entities and finally yield the results as a dictionary. The resulting
+dictionary will be interpreted by OHM and stored accordingly.
+
+\section{Storing data}
 
@@
 
-When data is exchanged between OHM and the user either
+Importing data to OHM is as simple as the user defines a mapper--the first
+half of a MapReduce program--which is run over a file that resides on
+HDFS\footnote{Hadoop Distributed Filesystem}.
+
+Data is stored on Apache HBase
+
+Exporting data
+
+and the user 
 JSON\footnote{JavaScript Object Notation} is transported over a RESTful
@@
 
+The user's datasets are defined using YAML--a human friendly data
+serialization standard.
+
 \section{Apache Hadoop}
@@
 
+\subsection{Thrift}
+
 \section{Data interchange}
@@
 \subsection{Input and output formats}
-
-\subsection{Thrift}
 

### 2009-05-05 doc/report/msc-report.tex
usage additions.

@@
 
+\begin{verbatim}
+   zohmg create project
+\end{verbatim}
+
+\noindent Edit the configuration files.
+
+\begin{verbatim}
+   zohmg setup
+\end{verbatim}
+
+\noindent The above command executes the configuration made and creates the
+infrastructure (HBase tables) for the project.
+
 \section{Import}
@@
 
+\begin{verbatim}
+   zohmg import mappers/somemapper.py path/on/hdfs
+\end{verbatim}
+
 \section{Storing data}
@@
 \cite{thrift}
+
+\begin{verbatim}
+   zohmg serve
+\end{verbatim}
+
+\noindent The above serves on default port 8086, with the --port argument it is
+possible to change server port.
+
+To extract data, visit the server URL (e.g. http://localhost:8086/) and
+give a request and query.
+
+\subsection{Data server}
+
+TODO
+
+URLParser
+
+Client
+
+Data
+
+Transform
 
@@
 
-\section{Administration}
+\section{Configuration}
 
@@
 The user's datasets are defined using YAML--a human friendly data
-serialization standard.
+serialization standard. The datasets reside in the config directory.
+
+\subsection{Environment}
+
+Certain paths need to be known by OHM in order to function, these are set
+in the config/environment.py file.
 

### 2009-05-05 doc/report/msc-report.tex
contents rearrangement and minor additions.

@@
 
+\subsection{Column-qualifiers}
+
 \section{Hierarchies}
@@
 them.
+
+\chapter{Theory [TODO: needs better name]}
+
+\section{Time-series}
+
+\section{Multidimensional arrays}
+
+\subsection{Data cubes}
+
+\subsection{Cube lattice}
+
+\section{Dimensions}
+
+\section{Projections}
+
+\section{Granularity}
+
+\chapter{Tools}
+
+\section{Apache Hadoop}
+
+\subsection{Hadoop Distributed File System}
+
+\section{Apache HBase}
+
+\subsection{Thrift}
+
+\subsection{REST}
+
+\subsection{JSON}
+
+\section{Python}
+
+\subsection{Dumbo}
+
+\subsection{Paste}
+
+\subsection{PyYAML}
+
+\subsection{Setuptools}
+
+\subsection{simplejson}
+
+\chapter{Implementation}
+
+OHM is built upon existing tools and software. It uses Apache Hadoop and
+HBase, free implementations of Google's MapReduce and BigTable
+respectively, both implemented in Java.
+
+Importing data to OHM is as simple as the user defines a mapper--the first
+half of a MapReduce program--which is run over a file that resides on
+HDFS\footnote{Hadoop Distributed Filesystem}.
+
+Data is stored on Apache HBase
+
+Exporting data
+
+JSON\footnote{JavaScript Object Notation} is transported over a RESTful
+interface or data is transported with Thrift.
+
+OHM is implemented in Python.
+
+\section{Configuration}
+
+\subsection{YAML}
+
+The user's datasets are defined using YAML--a human friendly data
+serialization standard. The dataset files reside in the config directory.
+
+\subsection{Environment}
+
+Certain paths need to be known by OHM in order to function, these are set
+in the config/environment.py file.
+
+\section{Importing}
+
+The user defines a mapper which OHM wraps, interpreting the output from the
+mapper program and storing it in HBase.
+
+Support exists for writing the mapper in Java or Python.
+
+\section{Data interchange}
+
+\subsection{Input formats}
+
+\subsection{Output formats}
+
+\subsubsection{Darling}
 
@@
 
-\noindent The above serves on default port 8086, with the --port argument it is
-possible to change server port.
+\noindent The above serves on default port 8086, with the --port argument
+it is possible to change server port.
 
@@
 
-TODO
+Built with Python Paste--a middleware framework--the data server dispatches
+to the requested application depending on the URI it was fed.
 
-URLParser
+\section{URLParser}
 
-Client
+Dispatching.
 
-Data
+\section{Clients}
 
-Transform
+Dispatches to static files in client directory in the project directory.
+Custom clients can be built with HTML and JavaScript, see the examples.
 
-\chapter{Implementation}
+\section{Data}
 
-OHM is built upon existing tools and software. It uses Apache Hadoop and
-HBase, free implementations of Googles' MapReduce and BigTable
-respectively, both implemented in Java.
+Serves data from HBase, dumps it as JSONP.
 
-Importing data to OHM is as simple as the user defines a mapper--the first
-half of a MapReduce program--which is run over a file that resides on
-HDFS\footnote{Hadoop Distributed Filesystem}.
+\section{Transformers}
 
-Data is stored on Apache HBase
-
-Exporting data
-
-and the user 
-JSON\footnote{JavaScript Object Notation} is transported over a RESTful
-interface or data is transported with Thrift.
-
-OHM is implemented in Python.
-
-\section{Configuration}
-
-\subsection{YAML}
-
-The user's datasets are defined using YAML--a human friendly data
-serialization standard. The datasets reside in the config directory.
-
-\subsection{Environment}
-
-Certain paths need to be known by OHM in order to function, these are set
-in the config/environment.py file.
-
-\section{Apache Hadoop}
-
-\subsection{Hadoop Distributed File System}
-
-\subsection{Dumbo}
-
-\section{Apache HBase}
-
-\subsection{Thrift}
-
-\section{Data interchange}
-
-\subsection{Input and output formats}
-
-\subsection{REST}
-
-\subsection{JSON}
+Fetches data from HBase and transforms it with the requested transform
+program in the transformers directory in the project directory, output is
+dumped as JSONP.
 

### 2009-05-05 doc/report/msc-report.tex
added Future work.
added Hadoop streaming.
msc-report.tex: changed title.
msc-report.tex: rearrange contents and minor additions.
added three paragraphs to the introduction and sections about hadoop, hbase and dumbo plus future work.

@@
 \author{Per Andersson, Fredrik M{\"o}llerstrand}
-\title{Zonotoop--a large scale aggregating data warehouse for
-time-series-based data. [DRAFT]}
+\title{Zohmg--a large scale data warehouse for aggregated time-series-based
+data. [DRAFT]}
 
@@
 \chapter{Introduction}
+
+This paper talks about a data store for aggregated multi-dimensional time-series-based data.
+The data is aggregated over many units of measurements and across multiple dimensions.
+One of these dimensions is always time.
+
+Zohmg solves the problem of summing and storing measurements taken from log files or similar. It stores solely summarized data, not atomic data. Zohmg is optimized for speed of data retrieval.
+
+The paper is presented in three major sections: problem statement, the theoretical solution and the implementation details.
+
+Says Wikipedia: "Data warehouse is a repository of an organization's electronically stored data.
+Data warehouses are designed to facilitate reporting and analysis." The main actions of a data warehouse is extract, transform and load data, although not necessarily in that order.
+
+The classic thinking in data warehousing is the divide between facts and dimensons.
+The facts are measures categorized by dimensions.
+In OLAP, a sort of data warehouse, the OLAP cube is typically modelled in fact tables and dimension tables in a relational database.
+The novelty of our approach is to model OLAP cube-like aggregation data store on top of a bigtable-like data store.
 
@@
 knowledge about practical implementation details.
+
+\section{Problem description}
+
+\subsection{Scope}
+
+Time-series based analysis is the most important one when analysing web
+logs, it comes natural since all log entries are timestamped.
+\cite{discoveringweb} It is also the most common form we have seen for logs
+in general.
+
+Assuming that time-series analysis is the most important, we have imposed
+the restriction on the system that is built so that it only handles
+time-series. Because of this restriction the data store obviously becomes
+less general purpose. Although since most log data is time-series based
+this restriction does not cripple the system much--if at all.
+
+\subsection{Data size}
+
+The sheer size of the data makes it unfeasible to use a single computer.
+Eventually a single machine would be fed so much data that it cannot cope
+with it. For instance a single web log file for one day can be up to
+several gigabytes. It is problematic for a single machine to handle this
+because of the limitations of memory, loading the entire file in memory.
+Distributed computing solves this by spreading the computations out to
+several machines in parallel. The obvious gain is that the waiting time
+before a computation is done is divided by the amount of nodes used,
+compared to running the computation on one machine. There is also a gain
+related to the actual machines, it is possible to use commodity hardware.
+Because MapReduce and BigTable has high availability and reliability a node
+in the cluster can safely be replaced on the fly if it fails.
+
+Scalability is also very important. Using single machines for computing is
+problematic when the dataset grows, more powerful machines has to be
+obtained. More power contained in a single machine results in higher price,
+eventually it is too expensive to perform computations on very large
+datasets. Both MapReduce and BigTable scale close to linearly, adding one
+machine to the cluster adds equal amount of processing power. Since
+MapReduce and BigTable can use commodity hardware it is reasonable to buy
+extra machines to gain extra computing power or storage space.
+
+\section{Overview}
+
+\chapter{Theory [TODO: needs better name]}
+
+\section{Time-series}
+
+\section{Multidimensional arrays}
+
+\subsection{Data cubes}
+
+\subsection{Cube lattice}
+
+\section{Dimensions}
+
+\section{Projections}
+
+\section{Granularity}
+
+\chapter{Tools}
+
+Regarding the infrastructure, both BigTable and MapReduce has high
+availability, scalability and reliability. Since BigTable runs on top of
+GFS\footnote{Google File System}, the data stored in BigTable is
+replicated, usually three times, onto other nodes in the cluster.
+\cite{gfs} Likewise, if a node in a MapReduce job goes down or fails to
+complete the job it is reassigned to another node. \cite{mapreduce}
+
+\section{MapReduce}
+
+Analysing the massive amounts of log data raises high demand for computing
+power. The MapReduce \cite{mapreduce} pattern suits the analysing issue
+well. The great benefit of MapReduce is that it runs on a computer cluster
+of commodity hardware where computation takes place independently in
+parallel. All results computed by each node are then reduced into one
+single result.
+
+% ONE DOES NOT SIMPLY PROCESS THE DATA ON ONE COMPUTER INTO MORDOR.
+Basically MapReduce is a two phase computation. First the input data is
+partitioned and each of these partitions is sent to a \textit{mapper} which
+performs computations on each row of the data. The mapper then outputs each
+resulting computation on the format \texttt{key,value}. These key-value
+pairs are received by a \textit{reducer} which performs the second stage of
+the computation.
 
@@
 
-\section{MapReduce}
+\subsection{Data model}
 
-Analysing the massive amounts of log data raises high demand for computing
-power. The MapReduce \cite{mapreduce} pattern suits the analysing issue
-well. The great benefit of MapReduce is that it runs on a computer cluster
-of commodity hardware where computation takes place independently in
-parallel. All results computed by each node are then reduced into one
-single result.
+Sorted distributed multidimensional sparse hashtable.
 
-% ONE DOES NOT SIMPLY PROCESS THE DATA ON ONE COMPUTER INTO MORDOR.
-Basically MapReduce is a two phase computation. First the input data is
-partitioned and each of these partitions is sent to a \textit{mapper} which
-performs computations on each row of the data. The mapper then outputs each
-resulting computation on the format \texttt{key,value}. These key-value
-pairs are received by a \textit{reducer} which performs the second stage of
-the computation.
-
-\section{Summary}
-
-Regarding the infrastructure, both BigTable and MapReduce has high
-availability, scalability and reliability. Since BigTable runs on top of
-GFS\footnote{Google File System}, the data stored in BigTable is
-replicated, usually three times, onto other nodes in the cluster.
-\cite{gfs} Likewise, if a node in a MapReduce job goes down or fails to
-complete the job it is reassigned to another node. \cite{mapreduce}
-
-\chapter{Problem description}
-
-\section{Scope}
-
-Time-series based analysis is the most important one when analysing web
-logs, it comes natural since all log entries are timestamped.
-\cite{discoveringweb} It is also the most common form we have seen for logs
-in general.
-
-Assuming that time-series analysis is the most important, we have imposed
-the restriction on the system that is built so that it only handles
-time-series. Because of this restriction the data store obviously becomes
-less general purpose. Although since most log data is time-series based
-this restriction does not cripple the system much--if at all.
-
-\section{Data size}
-
-The sheer size of the data makes it unfeasible to use a single computer.
-Eventually a single machine would be fed so much data that it cannot cope
-with it. For instance a single web log file for one day can be up to
-several gigabytes. It is problematic for a single machine to handle this
-because of the limitations of memory, loading the entire file in memory.
-Distributed computing solves this by spreading the computations out to
-several machines in parallel. The obvious gain is that the waiting time
-before a computation is done is divided by the amount of nodes used,
-compared to running the computation on one machine. There is also a gain
-related to the actual machines, it is possible to use commodity hardware.
-Because MapReduce and BigTable has high availability and reliability a node
-in the cluster can safely be replaced on the fly if it fails.
-
-Scalability is also very important. Using single machines for computing is
-problematic when the dataset grows, more powerful machines has to be
-obtained. More power contained in a single machine results in higher price,
-eventually it is too expensive to perform computations on very large
-datasets. Both MapReduce and BigTable scale close to linearly, adding one
-machine to the cluster adds equal amount of processing power. Since
-MapReduce and BigTable can use commodity hardware it is reasonable to buy
-extra machines to gain extra computing power or storage space.
-
-\chapter{Data model}
-
-\section{Sorted data}
-
-\subsection{Rows}
+\subsubsection{Rows}
 
@@
 
-\subsection{Regions}
+\subsubsection{Regions}
 
@@
 
-\section{Dimensions}
+\subsection{Dimensions}
 
-\subsection{Column-families}
+\subsubsection{Column-families}
 
@@
 
-\subsection{Column-qualifiers}
+\subsubsection{Column-qualifiers}
 
-\section{Hierarchies}
+\subsection{Hierarchies}
 
@@
 
-\chapter{Theory [TODO: needs better name]}
+\section{Apache Hadoop}
 
-\section{Time-series}
+Hadoop is an open-source implementation of Google's MapReduce-framework described in [mapreduce].
+Hadoop is written in Java and is part of the Apache Software Foundation.
+It is a fairly mature piece of software and is used in production at numerous companies.
 
-\section{Multidimensional arrays}
+MapReduce is a framework for running jobs in parallel across multiple machines.
+As the name implies, MapReduce is made up of two part or phases: map and reduce.
+The map phase consumes one piece of input at a time (a piece of input can be a line from a log file, for example), and emits an intermediate key and value.
+These keys and values are then processed by the reduce phase, which emits the final key-value pairs.
 
-\subsection{Data cubes}
+MapReduce works by splitting the input, which commonly is in the size range of many gigabytes, into n parts and lets each node of the computing cluster work on one or more parts.
+The mapper on each node emits key-value pairs which are fed into the reduce phase.
+The final output is collated, usually on a network file system.
 
-\subsection{Cube lattice}
-
-\section{Dimensions}
-
-\section{Projections}
-
-\section{Granularity}
-
-\chapter{Tools}
-
-\section{Apache Hadoop}
+All Zohmg programs are executed on Hadoop.
 
@@
 
+\subsection{Hadoop streaming}
+
 \section{Apache HBase}
+
+HBase is an open-source implementation of Google's BigTable data store described in [bigtable].
+It is a sub-project of Hadoop, written in Java and a part of the Apache Software Foundation.
+HBase is still in its infancy and few organizations use it for mission-critical applications.
+
+BigTable is a data store designed for traversal of very large data sets.
+Time-stamped cells inside column-familes in rows.
+The row has a key and any number of cells attached to it.
+The rows in a table are sorted by the row key.
+The typical use-case of BigTable is to scan a range of rows.
+It is therefor important to set up one's keys so that related data is close to each other.
+For example, if it makes sense to traverse the data in chronological order, the key might contain a representation of the data's timestamp.
+The classic example is to have the reversed domain name (i.e.
+com.google.www) as the row key, which means that all sub-domains of any domain are next to each other.
+
+Sparseness is a big thing. Null values are free of charge. *Describe how we leverage sparsity*.
+
+The typical use case for Zohmg is to read one or more qualifiers, which correspond to one or more points in n-space, over a range of time.
+Therefor, the keys are made up of the time and unit of the measurement, and the qualifier queried for represents the point in n-space.
+More on this in the data model section.
+
+All Zohmg data is stored in HBase.
+
+\section{Dumbo}
+
+Dumbo is a framework for writing Hadoop programs in languages other than Java, in a so-called Streaming Mode.
+Dumbo is written in Python.
+It is used extensively at last.fm for writing short prototypes.
+
+All Zohmg programs written in Python use Dumbo to execute on Hadoop.
 
@@
 
-\subsection{REST}
-
-\subsection{JSON}
+\section{Java}
 
 \section{Python}
-
-\subsection{Dumbo}
 
@@
 \subsection{simplejson}
+
+\section{REST}
+
+\section{JSON}
 
@@
 
-\subsection{Input formats}
+\subsection{Internal}
 
-\subsection{Output formats}
+\subsubsection{Input formats}
 
-\subsubsection{Darling}
+\subsubsection{Output formats}
+
+\subsection{External}
+
+\subsubsection{JSON}
 
@@
 
-\section{URLParser}
+\subsubsection{URLParser}
 
@@
 
-\section{Clients}
+\subsubsection{Clients}
 
@@
 
-\section{Data}
+\subsubsection{Data}
 
@@
 
-\section{Transformers}
+\subsubsection{Transformers}
 
@@
 
-\chapter{Conclusion}
+\chapter{Discussion}
 
-\chapter{Related work}
+\section{Related work}
 
@@
 warehouse based on MapReduce and Bigtable. \cite{hdw}
+
+\section{Future work}
+
+Says Wikipedia: "Organizations generally start off with relatively simple use of data warehousing.
+Over time, more sophisticated use of data warehousing evolves." Zohmg is that relatively simple data warehouse; the seed from which a more sophisticated data warehouse may evolve.
+
+There is as of yet no support for specifying granularities. The only supported resolution of time is the day.
+
+The single aggregation performed is summing of the measurements. Future work may include average, max and min, etc.
+
+There is no metadata store. It would be very helpful for the application writer to know what the known values are for each dimension.
+
+\subsection{Multiple datasets}
+
+\subsection{Metadata}
+
+\subsection{Updating dataset files}
+
+Update tables on configuration--dataset files--changes.
+
+\subsection{Importing without MapReduce}
+
+Importing content without firing of a MapReduce job, e.g. via a Thrift
+service.
+
+\chapter{Conclusion}
 

### 2009-05-05 doc/report/msc-report.tex
doc/report/msc-report.tex: + merge of Hadoop, HBase and Dumbo sections. + additions to Apache/Thrift, Serializing, Java and Python.

@@
 \author{Per Andersson, Fredrik M{\"o}llerstrand}
-\title{Zohmg--a large scale data warehouse for aggregated time-series-based
+\title{Zohmg---a large scale data warehouse for aggregated time-series-based
 data. [DRAFT]}
@@
 All three requirements are depend on one another, what data and how it is
-stored affects what can be visualised. Also, what data is imported--and
-how it is analysed--affects how it can be stored and hence visualised.
+stored affects what can be visualised. Also, what data is imported---and
+how it is analysed---affects how it can be stored and hence visualised.
 
@@
 less general purpose. Although since most log data is time-series based
-this restriction does not cripple the system much--if at all.
+this restriction does not cripple the system much---if at all.
 
@@
 Each data point is called a cell. A cell is an uninterpreted array of
-bytes-- there are no data types, and it is up to the application to
+bytes---there are no data types, and it is up to the application to
 interpret the data correctly. The table can store any number of timestamped
@@
 In actual practice, the main difference between the two models is that
-querying is highly restricted in a BigTable-like data store--there are no
-joins, for example--whereas the relational database was born to be queried.
+querying is highly restricted in a BigTable-like data store---there are no
+joins, for example---whereas the relational database was born to be queried.
 On the other hand, the restrictions placed upon the BigTable models are
@@
 A region is a partition of the data stored on a node in the computer
-cluster. A region server (referred to as tablet server in \cite{mapreduce})
+cluster. A region server (referred to as tablet server in \cite{bigtable})
 holds information on which regions are located on which nodes. More
@@
 
-Hadoop is an open-source implementation of Google's MapReduce-framework described in [mapreduce].
+Hadoop is an free software implementation of Google's MapReduce-framework described in \cite{mapreduce}.
 Hadoop is written in Java and is part of the Apache Software Foundation.
@@
 
-All Zohmg programs are executed on Hadoop.
+All Zohmg import programs are executed on Hadoop.
 
@@
 
-HBase is an open-source implementation of Google's BigTable data store described in [bigtable].
+HBase is an free software implementation of Google's BigTable data store described in \cite{bigtable}.
 It is a sub-project of Hadoop, written in Java and a part of the Apache Software Foundation.
@@
 
-Sparseness is a big thing. Null values are free of charge. *Describe how we leverage sparsity*.
+Sparseness is a big thing. Null values are free of charge. TODO: *Describe how we leverage sparsity*.
 
@@
 
+\subsection{Thrift}
+
+A remote procedure call framework for building scalable cross-language
+services. Thrift combines a software stack with a code generation engine to
+build services that work seamlessly between a wide range of programming and
+scripting languages. \cite{thrift}
+
+Thrift was originally developed at Facebook and in April 2007 it was
+released as free software. It entered the Apache Incubator in May 2008.
+
+\subsubsection{Python}
+
+HBase has a Thrift server which serves as the interface to languages other
+than Java--for instance Python.
+
 \section{Dumbo}
@@
 Dumbo is written in Python.
-It is used extensively at last.fm for writing short prototypes.
+It is used extensively at Last.fm for writing short prototypes.
 
@@
 
-\subsection{Thrift}
+\section{Serializing}
+
+Serializing data is the task of converting an object into a sequence of
+bits so that it can be stored on a storage medium or transmitted over a
+network link. Rereading the sequence of bits should restore the original
+state of the object it was in before the serialization.
+
+\subsection{JSON}
+
+JSON\footnote{JavaScript Object Notation} is a lightweight text-based
+human-readable data interchange format. It is used for representing simple
+data structures and associative arrays (called objects). JSON is often used
+for serializing structured data and transmitting it over network
+connection.
+
+\subsection{YAML}
+
+YAML\footnote{YAML Ain't a Markup Language} is a human-readable data
+serialization format available for a wide range of programming and
+scripting languages.
+
+The outline and markup of YAML documents makes it well suited for data
+which humans are likely to view or edit, such as configuration files.
 
@@
 
+Both Hadoop and HBase are written in Java---a popular high-level
+object-oriented programming language.
+
+Most Hadoop users write their MapReduce programs in Java, since this is the
+native language of Hadoop.
+
 \section{Python}
+
+Python is a popular general-purpose high-level multi-paradigm---most
+notably: object-oriented, imperative and functional---scripting language.
 
@@
 
+Middleware. WSGI. HTTP server.
+
 \subsection{PyYAML}
+
+PyYAML is a complete YAML 1.1 compliant parser and emitter for Python.
 
@@
 
-\section{JSON}
+REST\footnote{Representational State Transfer}
 
@@
 
-OHM is built upon existing tools and software. It uses Apache Hadoop and
-HBase, free implementations of Google's MapReduce and BigTable
-respectively, both implemented in Java.
+As stated in chapter 3, Zohmg is built upon existing tools and software. It
+uses Apache Hadoop and HBase, free implementations of Google's MapReduce
+and BigTable respectively, both implemented in Java.
 
-Importing data to OHM is as simple as the user defines a mapper--the first
-half of a MapReduce program--which is run over a file that resides on
-HDFS\footnote{Hadoop Distributed Filesystem}.
+Importing data to Zohmg is as simple as the user defines a mapper---the
+first half of a MapReduce program---which is run over a file that resides
+on HDFS\footnote{Hadoop Distributed Filesystem}.
 
-Data is stored on Apache HBase
+Data is stored in Apache HBase.
 
-Exporting data
-
-JSON\footnote{JavaScript Object Notation} is transported over a RESTful
-interface or data is transported with Thrift.
+Exporting data from Zohmg is done via a RESTful interface which dumps JSON.
+It is possible to transform the stored data with user defined transformers.
 
@@
 
-The user's datasets are defined using YAML--a human friendly data
-serialization standard. The dataset files reside in the config directory.
+The user's datasets are defined using YAML. The dataset files reside in the
+config directory.
 
@@
 
-We have built a system--henceforth called OHM--which has the goal of being
+We have built a system---henceforth called OHM---which has the goal of being
 a large scale aggregating data warehouse for time-series-based log-like
@@
 
-Built with Python Paste--a middleware framework--the data server dispatches
+Built with Python Paste---a middleware framework---the data server dispatches
 to the requested application depending on the URI it was fed.
@@
 
-Update tables on configuration--dataset files--changes.
+Update tables on configuration---dataset files---changes.
 

### 2009-05-05 doc/report/msc-report.tex
doc/report/msc-report.tex: added info about Python modules. removed REST.

@@
 
-Middleware. WSGI. HTTP server.
+Python Paste is a framework for building WSGI\footnote{Web Server Gateway
+Interface, a HTTP extension where a CGI-like environment is passed around.}
+applications and middleware.
+
+WSGI applications receive a WSGI request and returns a response with the
+built-in web server.
+
+Middleware is software that acts as an intermediary. WSGI middleware
+receives a WSGI request and then performs logic upon this request, before
+passing the request to a WSGI application.
 
@@
 
+Python Setuptools is a framework---built on top of the Python module
+distutils---for basic package management of Python software. Amongst other
+things it creates and installs Python Eggs\footnote{A Zip file which
+bundles Python software.}.
+
 \subsection{simplejson}
 
-\section{REST}
-
-REST\footnote{Representational State Transfer}
+A simple, fast, complete and extensible JSON encoder and decoder for
+Python. The Python module simplejson is implemented entirely in Python and
+has no external dependencies. However a C extension which provides a
+serious speed boost is included.
 

### 2009-05-05 doc/report/msc-report.tex
added 'theory'.

@@
 
-\chapter{Theory [TODO: needs better name]}
+\chapter{Theory}
 
-\section{Time-series}
+\section{Data cubes}
 
-\section{Multidimensional arrays}
+A data cube is a conceptual n-dimensional array of values.
+The values are numerical facts belonging to some dimension.
+The data cube is also known as a hypercube.
+In this context, each dimension of the cube corresponds to an attribute of the data in question and each point along the dimensions axis is a value of the corresponding attribute.
+The mental model of a data cube aids in reasoning about dimensions.
+Data cubes are traditionally modeled as star schemas in relational database systems.
 
-\subsection{Data cubes}
+\subsection{Dimensions}
 
-\subsection{Cube lattice}
+The dimensions of the data cube are..
 
-\section{Dimensions}
+\subsection{Projections}
 
-\section{Projections}
+One of the most common operations on a data cube is the projection.
+A projection is a dimensionality reduction, where one or more dimensions are discarded from the cube with the measurements along each axis summed up, leaving a data cube of a lesser dimension.
 
-\section{Granularity}
+A typical projection is that from an n-dimensional cube down to two dimensions, which is the case when plotting an image on screen or paper.
+
+\subsection{Slice and dice}
+
+The slice operation fixes the values of one or several dimensions in the cube.
+The data omitted from this operation would be any data associated with the values of the dimension that were not fixed.
+The dice operation is a slice on more than two dimensions.
+[source: http://www.cs.ualberta.ca/~zaiane/courses/cmput690/glossary.html]
+
+\section{Time series}
+
+Time is the most common of the dimensions in the data cube constructed for Zohmg.
+The time stamp is a common property for all data handled by Zohmg; every measurement is expected to have time as one of its dimensions.
+TODO: really, this belongs in the implementation section.
+I find nothing interesting to say about time series in this section.
+
+\section{Aggregates}
+
+An aggregate is a composite of several values. Typical examples of aggregate functions are average, count and sum.
+
+In a standard OLAP setup, a number of aggregates are pre-computed and stored in the cube.
+These base aggregates represent only a fraction of the very many possible aggregations.
+The remaining aggregates are computed from the base aggregates as they are demanded.
 
@@
 
-Hadoop is an free software implementation of Google's MapReduce-framework described in \cite{mapreduce}.
+Hadoop is an open source software implementation of Google's MapReduce-framework described in \cite{mapreduce}.
 Hadoop is written in Java and is part of the Apache Software Foundation.
@@
 MapReduce is a framework for running jobs in parallel across multiple machines.
-As the name implies, MapReduce is made up of two part or phases: map and reduce.
+As the name implies, MapReduce is made up of two parts or phases: map and reduce.
 The map phase consumes one piece of input at a time (a piece of input can be a line from a log file, for example), and emits an intermediate key and value.
@@
 
-HBase is an free software implementation of Google's BigTable data store described in \cite{bigtable}.
+HBase is an open source implementation of Google's BigTable data store described in \cite{bigtable}.
 It is a sub-project of Hadoop, written in Java and a part of the Apache Software Foundation.

### 2009-05-05 doc/report/msc-report.tex
doc/report/msc-report.tex: + changed "paper" to "thesis". + added overview section.

--- 2009-05-05 doc/report/msc-report.tex
@@
 
-This paper talks about a data store for aggregated multi-dimensional time-series-based data.
+This thesis talks about a data store for aggregated multi-dimensional time-series-based data.
 The data is aggregated over many units of measurements and across multiple dimensions.
@@
 
-The paper is presented in three major sections: problem statement, the theoretical solution and the implementation details.
+The thesis is presented in three major sections: problem statement, the theoretical solution and the implementation details.
 
@@
 \section{Overview}
+
+This thesis, fully, describes the Zohmg system and how to use it.
+
+Chapter 2 the theoretical base of the problem statement is established.
+
+In chapter 3 the tools used for the implementation are declared, moving
+over to chapter 4 where the details of the implementation are presented.
+
+How to use the Zohmg system is presented in chapter 5, while chapter 6
+shows an example work flow.
+
+Chapters 7 and 8 discusses the results, related, and future work, and
+presents the conclusion, respectively.
 

### 2009-05-05 doc/report/msc-report.tex
doc/report/msc-report.tex: + minor corrections in Data cubes section. + added content to Dimensions section.

@@
 
-A data cube is a conceptual n-dimensional array of values.
-The values are numerical facts belonging to some dimension.
-The data cube is also known as a hypercube.
-In this context, each dimension of the cube corresponds to an attribute of the data in question and each point along the dimensions axis is a value of the corresponding attribute.
-The mental model of a data cube aids in reasoning about dimensions.
-Data cubes are traditionally modeled as star schemas in relational database systems.
+A data cube is a conceptual n-dimensional array of values. The values are
+numerical facts belonging to some dimension. The data cube is also known as a
+hypercube, it relates to multidimensional analogues of squares and cubes.
+In this context, each dimension of the cube corresponds to an attribute of
+the data in question and each point along the dimension's axis is a value of
+the corresponding attribute. The mental model of a data cube aids in
+reasoning about dimensions. Data cubes are traditionally modeled as star
+schemas in relational database systems. \cite{olap_solutions}
 
@@
 
-The dimensions of the data cube are..
+The school book example of a data cube is three dimensional where each of
+the three dimensions correspond to one of the attributes time, products,
+and stores. Adding another attribute adds another dimension to the data
+cube, because of the one to one relationship of attributes and dimensions.
 

### 2009-05-05 doc/report/msc-report.tex
doc/report/msc-report.tex: + added section on HDFS. + %s/OHM/Zohmg/g + minor changes here and there.

@@
 
-Each column-family is stored as a single map-file in hbase.
+Each column-family is stored as a single map-file in HBase.
 
 \subsubsection{Column-qualifiers}
+
+Each column-family is further divided into one or more column-qualifiers.
 
@@
 created. The order in which the data is stored is very important, it
-defines how effective scanning for data will be. For instance, let's say we
-have thousand rows and want ten out of those. In this case only one per
-cent of the data is interesting to us, if a scan would be required to visit
-all the thousand rows a lot of rows are skipped. Skipping rows is expensive
-in that sense that we have to read them but throw away the result. The goal
-is to push this cost to a bare minimum, using as many as possible of the
-rows we scan.
+defines how effective scanning for data will be. For instance, let's say
+we have thousand rows and want ten out of those. In this case only one per
+cent of the data is interesting to us, if a scan would be required to
+visit all the thousand rows a lot of rows are skipped. Skipping rows is
+expensive in that sense that we have to read them but throw away the
+result. The goal is to push this cost to a bare minimum, using as many as
+possible of the rows we scan.
 
@@
 
-Hadoop is an open source software implementation of Google's MapReduce-framework described in \cite{mapreduce}.
-Hadoop is written in Java and is part of the Apache Software Foundation.
-It is a fairly mature piece of software and is used in production at numerous companies.
+Hadoop is a free software implementation of Google's MapReduce-framework
+described in \cite{mapreduce}. Hadoop is written in Java and is part of
+the Apache Software Foundation.  It is a fairly mature piece of software
+and is used in production at numerous companies.
 
-MapReduce is a framework for running jobs in parallel across multiple machines.
-As the name implies, MapReduce is made up of two parts or phases: map and reduce.
-The map phase consumes one piece of input at a time (a piece of input can be a line from a log file, for example), and emits an intermediate key and value.
-These keys and values are then processed by the reduce phase, which emits the final key-value pairs.
+MapReduce is a framework for running jobs in parallel across multiple
+machines. As the name implies, MapReduce is made up of two parts or
+phases: map and reduce. The map phase consumes one piece of input at a
+time (a piece of input can be a line from a log file, for example), and
+emits an intermediate key and value. These keys and values are then
+processed by the reduce phase, which emits the final key-value pairs.
 
-MapReduce works by splitting the input, which commonly is in the size range of many gigabytes, into n parts and lets each node of the computing cluster work on one or more parts.
-The mapper on each node emits key-value pairs which are fed into the reduce phase.
-The final output is collated, usually on a network file system.
+MapReduce works by splitting the input, which commonly is in the size
+range of many gigabytes, into n parts and lets each node of the computing
+cluster work on one or more parts. The mapper on each node emits key-value
+pairs which are fed into the reduce phase. The final output is collated,
+usually on a distributed file system.
 
@@
 
+HDFS is a free software implementation of the GFS \cite{gfs}.
+
+The goal of HDFS is to store large data sets while being able to handle
+hardware failure and being deployed in heterogenous hardware and software
+eco systems. Additionally the Hadoop and HDFS interaction is designed with
+the notion that it is cheaper to move the computation, MapReduce programs,
+to the data than the other way around.
+
+Every Hadoop node can be formatted with HDFS, this reserves disk space on
+the node to be used for HDFS. The HDFS container resides on the nodes
+general purpose file system. It is also possible to use HDFS as a
+stand-alone general purpose distributed file system, without Hadoop.
+
+The default block size on HDFS 64 MB, significantly larger than
+file systems used for hard drives. The motivation is that applications
+that use HDFS are not general purpose applications which run on general
+purpose file systems, but batch applications which read, write, or both,
+large amounts of data.
+
 \subsection{Hadoop streaming}
+
+Any program that reads and outputs to the file pointers \texttt{stdin} and
+\texttt{stdout}, respectively, can be used as MapReduce programs with
+Hadoop Streaming. This makes it possible to use any shell script or
+program which inputs and outputs this way as a MapReduce program.
 
@@
 
-HBase is an open source implementation of Google's BigTable data store described in \cite{bigtable}.
-It is a sub-project of Hadoop, written in Java and a part of the Apache Software Foundation.
-HBase is still in its infancy and few organizations use it for mission-critical applications.
+HBase is a free software implementation of Google's BigTable data store
+described in \cite{bigtable}. It is a sub-project of Hadoop, written in
+Java and a part of the Apache Software Foundation. HBase is still in its
+infancy and few organizations use it for mission-critical applications.
 
 BigTable is a data store designed for traversal of very large data sets.
-Time-stamped cells inside column-familes in rows.
-The row has a key and any number of cells attached to it.
-The rows in a table are sorted by the row key.
-The typical use-case of BigTable is to scan a range of rows.
-It is therefor important to set up one's keys so that related data is close to each other.
-For example, if it makes sense to traverse the data in chronological order, the key might contain a representation of the data's timestamp.
-The classic example is to have the reversed domain name (i.e.
-com.google.www) as the row key, which means that all sub-domains of any domain are next to each other.
+Time-stamped cells inside column-familes in rows. The row has a key and
+any number of cells attached to it. The rows in a table are sorted by the
+row key.  The typical use-case of BigTable is to scan a range of rows. It
+is therefor important to set up one's keys so that related data is close
+to each other. For example, if it makes sense to traverse the data in
+chronological order, the key might contain a representation of the data's
+timestamp. The classic example is to have the reversed domain name (i.e.
+com.google.www) as the row key, which means that all sub-domains of any
+domain are next to each other.
 
-Sparseness is a big thing. Null values are free of charge. TODO: *Describe how we leverage sparsity*.
+Sparseness is a big thing. Null values are free of charge. TODO: *Describe
+how we leverage sparsity*.
 
-The typical use case for Zohmg is to read one or more qualifiers, which correspond to one or more points in n-space, over a range of time.
-Therefor, the keys are made up of the time and unit of the measurement, and the qualifier queried for represents the point in n-space.
-More on this in the data model section.
+TODO: *this should go into implementation.*
+The typical use case for Zohmg is to read one or more column-qualifiers,
+which correspond to one or more points in n-space, over a range of time.
+Therefor, the keys are made up of the time and unit of the measurement,
+and the qualifier queried for represents the point in n-space. More on
+this in the data model section.
 
@@
 released as free software. It entered the Apache Incubator in May 2008.
-
-\subsubsection{Python}
 
@@
 things it creates and installs Python Eggs\footnote{A Zip file which
-bundles Python software.}.
+bundles Python software files.}.
 
@@
 
-As stated in chapter 3, Zohmg is built upon existing tools and software. It
-uses Apache Hadoop and HBase, free implementations of Google's MapReduce
-and BigTable respectively, both implemented in Java.
+As stated in chapter 3, Zohmg is built upon existing tools and software.
+It uses Apache Hadoop and Apache HBase, free software implementations of
+Google's MapReduce and BigTable respectively, both implemented in Java.
 
@@
 
-Data is stored in Apache HBase.
+The resulting aggregated data, from the mapper, is stored in Apache HBase.
 
-Exporting data from Zohmg is done via a RESTful interface which dumps JSON.
-It is possible to transform the stored data with user defined transformers.
+Exporting data from Zohmg is done via a web interface which dumps JSON.
+It is possible to transform the stored data before exporting it with user
+defined transformers.
 
-OHM is implemented in Python.
+Zohmg is implemented in Python.
 
@@
 
-The user's datasets are defined using YAML. The dataset files reside in the
-config directory.
+The user's datasets are defined using YAML. The dataset files reside in
+the config directory.
 
@@
 
-Certain paths need to be known by OHM in order to function, these are set
+Certain paths need to be known by Zohmg in order to function, these are set
 in the config/environment.py file.
@@
 
-The user defines a mapper which OHM wraps, interpreting the output from the
+The user defines a mapper which Zohmg wraps, interpreting the output from the
 mapper program and storing it in HBase.
@@
 
-We have built a system---henceforth called OHM---which has the goal of being
+We have built a system---henceforth called Zohmg---which has the goal of being
 a large scale aggregating data warehouse for time-series-based log-like
@@
 The user writes the map part of the MapReduce programs which imports
-log-like data to the data store. The reduce step is taken care of by OHM,
+log-like data to the data store. The reduce step is taken care of by Zohmg,
 which uses sum reducer that simply sums up values with the same keys.
@@
 entities and finally yield the results as a dictionary. The resulting
-dictionary will be interpreted by OHM and stored accordingly.
+dictionary will be interpreted by Zohmg and stored accordingly.
 
 \begin{verbatim}
-   zohmg import mappers/somemapper.py path/on/hdfs
+   zohmg import mappers/somemapper.py hdfs/path
 \end{verbatim}
@@
 
-Data is stored in a BigTable like database. A natural consequence of this
-that the data is stored sorted by rowkey allowing fast access to an
-arbitrary (sequential) range.
+Data is stored in Apache HBase. A natural consequence of this that the
+data is stored sorted by rowkey allowing fast access to an arbitrary
+(sequential) range.
 
@@
 
-It is possible to query OHM about metadata it has on each of the imported
+It is possible to query Zohmg about metadata it has on each of the imported
 data sets.
@@
 Data is served to the user either over HTTP in JSON format or over Thrift.
-\cite{thrift}
 
@@
 
-\noindent The above serves on default port 8086, with the --port argument
-it is possible to change server port.
+\noindent The above serves on default port 8086, with the \texttt{--port}
+argument it is possible to change server port.
 
 To extract data, visit the server URL (e.g. http://localhost:8086/) and
-give a request and query.
+give a query request.
 
@@
 
-Built with Python Paste---a middleware framework---the data server dispatches
-to the requested application depending on the URI it was fed.
+Built with Python Paste---a middleware framework---the data server
+dispatches to the requested application depending on the URI it was fed.
 
@@
 
-Dispatching.
+Middleware which dispatches to applications based on the URI.
 
@@
 
-Dispatches to static files in client directory in the project directory.
+Dispatches request to serve static files from the project's client directory.
 Custom clients can be built with HTML and JavaScript, see the examples.
@@
 
-Serves data from HBase, dumps it as JSONP.
+Serves data from HBase as JSONP\footnote{JSON with padding: JSON extension which adds name of callback function as an argument to the function itself.}.
 
@@
 Fetches data from HBase and transforms it with the requested transform
-program in the transformers directory in the project directory, output is
+program from the project's transformers directory, output is
 dumped as JSONP.
@@
 
-Update tables on configuration---dataset files---changes.
+Update tables on changes in configuration---dataset files.
 

### 2009-05-06 doc/report/msc-report.tex
doc/report/msc-report.tex: + changed "data warehouse" to "data store". + added citations to tools.
doc/report/references.bib:
+ added references to tools.
+ cosmetic changes.
doc/report/msc-report.tex: * changed "dataset" to "data set". * added text in sections about data server (implementation and usage). * added brief outline to web log analysis section. * added one sentence to conclusion.

@@
 \author{Per Andersson, Fredrik M{\"o}llerstrand}
-\title{Zohmg---a large scale data warehouse for aggregated time-series-based
+\title{Zohmg---a large scale data store for aggregated time-series-based
 data. [DRAFT]}
@@
 Scalability is also very important. Using single machines for computing is
-problematic when the dataset grows, more powerful machines has to be
+problematic when the data set grows, more powerful machines has to be
 obtained. More power contained in a single machine results in higher price,
 eventually it is too expensive to perform computations on very large
-datasets. Both MapReduce and BigTable scale close to linearly, adding one
+data sets. Both MapReduce and BigTable scale close to linearly, adding one
 machine to the cluster adds equal amount of processing power. Since
@@
 
-A typical projection is that from an n-dimensional cube down to two dimensions, which is the case when plotting an image on screen or paper.
+A typical projection is that from an n-dimensional cube down to two
+dimensions, which is the case when plotting an object---with three or more
+dimensions---on screen or paper.
 
@@
 On the other hand, the restrictions placed upon the BigTable models are
-what allows it to scale to such a large dataset. Indeed, BigTable was born
+what allows it to scale to such a large data set. Indeed, BigTable was born
 to scale.
@@
 the Apache Software Foundation.  It is a fairly mature piece of software
-and is used in production at numerous companies.
+and is used in production at numerous companies. \cite{hadoop}
 
@@
 infancy and few organizations use it for mission-critical applications.
+\cite{hbase}
 
@@
 
-Dumbo is a framework for writing Hadoop programs in languages other than Java, in a so-called Streaming Mode.
-Dumbo is written in Python.
-It is used extensively at Last.fm for writing short prototypes.
+Dumbo is a framework for writing Hadoop programs in languages other than
+Java, in a so-called Streaming Mode. Dumbo is written in Python. It is used
+extensively at Last.fm for writing short prototypes. \cite{dumbo}
 
@@
 Most Hadoop users write their MapReduce programs in Java, since this is the
-native language of Hadoop.
+native language of Hadoop. \cite{java}
 
@@
 notably: object-oriented, imperative and functional---scripting language.
+\cite{python}
+
+Zohmg is written mostly in Python. Zohmg uses the following Python modules
+internally; it depends on them to be available on the system.
 
@@
 Interface, a HTTP extension where a CGI-like environment is passed around.}
-applications and middleware.
+applications and middleware. \cite{definitive_guide_to_pylons}
 
@@
 receives a WSGI request and then performs logic upon this request, before
-passing the request to a WSGI application.
+passing the request to a WSGI application. \cite{paste}
 
@@
 PyYAML is a complete YAML 1.1 compliant parser and emitter for Python.
+\cite{pyyaml}
 
@@
 things it creates and installs Python Eggs\footnote{A Zip file which
-bundles Python software files.}.
+bundles Python software files.}. \cite{setuptools}
 
@@
 has no external dependencies. However a C extension which provides a
-serious speed boost is included.
+serious speed boost is included. \cite{simplejson}
 
@@
 
-\subsection{YAML}
+\subsection{Data sets}
 
-The user's datasets are defined using YAML. The dataset files reside in
+The user's data sets are defined using YAML. The data set files reside in
 the config directory.
@@
 
+The Darling output format for HBase. Between Zohmg and Hadoop JSON is used
+for serializing data.
+
 \subsection{External}
 
-\subsubsection{JSON}
+The data export server is uilt with Python Paste---a middleware
+framework---the data server dispatches to the requested application
+depending on the URI it was fed.
+
+\subsubsection{URLParser}
+
+Middleware which dispatches to applications based on the URI. The URLParser
+removes the first part of the URI and tries to find a matching application
+to relay the request to. For instance, consider the URI
+\texttt{/client/file.html}. The first part---\texttt{client}---will be
+removed and the request \texttt{/file.html} will be sent to the application
+named \texttt{client.py}.
+
+\subsubsection{Clients}
+
+Dispatches request to serve static files from the project's
+\texttt{clients} directory. Custom clients---which interact with Zohmg---can
+be built with HTML and JavaScript, see the examples.
+
+\subsubsection{Data}
+
+Fetches data from HBase via its Thrift interface. The data is served to the
+user as JSONP\footnote{JSON with padding: JSON extension which adds name of
+callback function as an argument to the function itself.}.
+
+\subsubsection{Transformers}
+
+Fetches data from HBase and transforms it with the requested transform
+program from the project's transformers directory, output is
+dumped as JSONP.
 
@@
 
-To extract data, visit the server URL (e.g. http://localhost:8086/) and
-give a query request.
+The data server have three methods of serving data: raw, transformed, and
+static. The raw and transformed data are aggregates served from HBase,
+transformed in the latter case, while the static method just serve static
+files from the project's \textit{clients} directory. Aggregates served from
+HBase are returned to the user as JSONP.
 
-\subsection{Data server}
+\subsection{Data}
 
-Built with Python Paste---a middleware framework---the data server
-dispatches to the requested application depending on the URI it was fed.
+Based on the query---which the user submits---the corresponding data is
+served from HBase as JSONP. The query is a simple definition of wanted row
+range, dimensions, and values for these dimensions. The query is submitted
+to the data server via HTTP, an example query could look like (the URL is
+wrapped to fit the page) \\
 
-\subsubsection{URLParser}
+\texttt{http://localhost:8086/data/?t0=20090101 \\
+\&t1=20090420\&unit=pageviews\&d0=country\&d0v=SE} \\
 
-Middleware which dispatches to applications based on the URI.
+The above query would return the pageviews for SE between 1 January 2009
+and 4 April 2009.
 
-\subsubsection{Clients}
-
-Dispatches request to serve static files from the project's client directory.
-Custom clients can be built with HTML and JavaScript, see the examples.
-
-\subsubsection{Data}
-
-Serves data from HBase as JSONP\footnote{JSON with padding: JSON extension which adds name of callback function as an argument to the function itself.}.
-
-\subsubsection{Transformers}
+\subsection{Transformers}
 
 Fetches data from HBase and transforms it with the requested transform
-program from the project's transformers directory, output is
+program from the project's \texttt{transformers} directory, output is
 dumped as JSONP.
+
+\subsection{Clients}
+
+It is possible to build static web pages out of HTML and JavaScript, which
+can interact with the data server. These static pages can for instance
+present some sort of query building mechanism (for insntance with drop down
+boxes) and then send this query to the data server and graph the resulting
+data.
+
+The data server serves static files from the project's \texttt{clients}
+directory at the server URL appended with \texttt{client/filename}.
 
@@
 \section{Web log analysis}
+
+Configure the dataset.yaml.
+
+Write a custom mapper, which imports data from web logs.
+
+Start data server and extract data from it.
+
+Possibly write a transformer which performs some transformation on the
+aggregated data.
+
+Possibly write a client which will render a graph.
 
@@
 
-\subsection{Multiple datasets}
+\subsection{Multiple data sets}
 
@@
 
-\subsection{Updating dataset files}
+\subsection{Updating data set files}
 
-Update tables on changes in configuration---dataset files.
+Update tables on changes in configuration---data set files.
 
@@
 
+Last.fm is looking at using Zohmg for multiple application areas, such as
+radio stats and web page stats.
+
 \pagebreak

### 2009-05-06 scripts/install_hadoop_hbase.sh
scripts/install_hadoop_hbase.sh: fixed patch hang.

@@ -35,7 +35,8 @@ function exec_and_log() {
 
     # execute and log to intermediate.
     if [ "x" = "x$screenoutput" ]; then
-        $1 &>$install_tmplog
+        $1
+#        $1 &>$install_tmplog
     else
         $1
     fi
@@ -199,14 +200,14 @@ else
     printf "Extracting Apache Hadoop... "
     exec_and_log "tar zxf $files/$hadoop_tar -C $prefix"
     echo "done."
+    exec_and_log "echo ... Patching Apache Hadoop ..."
     for patch in "$patch_1722" "$patch_5450"; do
         num=$(echo $patch | sed 's/.patch$//')
         printf "Applying patch $num... "
-        exec_and_log "echo ... Patching Apache Hadoop ..."
-        cd $hadoop
-        exec_and_log "patch -p0 <$files/patches/$patch" "Error: Could not apply patch $num."
+        exec_and_log "cd $hadoop ; patch -p0 <$files/patches/$patch" "Error: Could not apply patch $num."
         echo "done."
     done
+    exec_and_log "echo ok"
     printf "Compiling Apache Hadoop... "
     exec_and_log "echo ... Compiling Apache Hadoop ..."
     cd $hadoop

### 2009-05-06 scripts/install_hadoop_hbase.sh
scripts/install_hadoop_hbase.sh: * fixed logging function. * removed permission check from script, it was not necessary. * added checks for downloading hadoop/hbase only. * changed EOF to EOconfigfiles for config file emit. * added escaping of $ in variables emitted to config files.

@@ -35,8 +35,7 @@ function exec_and_log() {
 
     # execute and log to intermediate.
     if [ "x" = "x$screenoutput" ]; then
-        $1
-#        $1 &>$install_tmplog
+        $1 &>$install_tmplog
     else
         $1
     fi
@@ -160,14 +159,25 @@ if [ "x" = "x$files" ]; then
     echo "done."
     echo "Downloading files... "
     # printing progress to screen, without logging.
-    # release files.
     cd $files
-    exec_and_log "wget $hadoop_release" "Error: Could not download $hadoop_release" "true"
-    exec_and_log "wget $hbase_release" "Error: Could not download $hbase_release" "true"
-    # download patches.
-    cd patches
-    exec_and_log "wget $hadoop_1722" "Error: Could not download $hadoop_1722" "true"
-    exec_and_log "wget $hadoop_5450" "Error: Could not download $hadoop_5450" "true"
+    # empty hadoop
+    if [ ! "x" = "x$hbase_only" ]; then
+        echo "Skipping download of Apache Hadoop..."
+    else
+        # hadoop release file.
+        exec_and_log "wget $hadoop_release" "Error: Could not download $hadoop_release" "true"
+        # download patches.
+        cd patches
+        exec_and_log "wget $hadoop_1722" "Error: Could not download $hadoop_1722" "true"
+        exec_and_log "wget $hadoop_5450" "Error: Could not download $hadoop_5450" "true"
+    fi
+    if [ ! "x" = "x$hadoop_only" ]; then
+        echo "Skipping download of Apache HBase..."
+    else
+        cd $files
+        # hbase release file.
+        exec_and_log "wget $hbase_release" "Error: Could not download $hbase_release" "true"
+    fi
     echo "Done downloading files."
 else
     printf "Using previously downloaded files in $files... "
@@ -187,11 +197,6 @@ fi
 
 # install.
 echo "Installing..."
-
-# check permissions.
-for dir in "$hadoop" "$hbase"; do
[…]

### 2009-05-07 doc/report/msc-report.tex
moved intro to abstract, added TODO thoughts to intro plus cosmetic changes.

@@
 \begin{document}
-
 \maketitle
@@
 \noindent \Large{\textbf{Abstract}}
-
 \vspace{12pt}
 
-\noindent This is where we put down two to four sentences about:
-Introduction, problem, existing tools, our solution and implementation,
-conclusions.
+\noindent 
+This thesis describes a data store for multi-dimensional time-series-based data.
+The data is aggregated over many units of measurements and across multiple dimensions.
+One of these dimensions is always time.
+
+Zohmg solves the problem of summing, storing and serving measurements.
+These measurements may be taken from log files or similar sources.
+Zohmg stores only the summarized data, not the atomic measurements.
+Zohmg is optimized for speed of data retrieval.
+
+The thesis is presented in three major sections: problem statement, theoretical solution and implementation details.
 
 \tableofcontents
-
 \vfill
-
 \pagebreak
@@
 
-This thesis talks about a data store for aggregated multi-dimensional time-series-based data.
-The data is aggregated over many units of measurements and across multiple dimensions.
-One of these dimensions is always time.
-
-Zohmg solves the problem of summing and storing measurements taken from log files or similar. It stores solely summarized data, not atomic data. Zohmg is optimized for speed of data retrieval.
-
-The thesis is presented in three major sections: problem statement, the theoretical solution and the implementation details.
-
+% TODO: den har texten passar battre i ett eget stycke/sektion, background reading style.
 Says Wikipedia: "Data warehouse is a repository of an organization's electronically stored data.
@@
 how it is analysed---affects how it can be stored and hence visualised.
-
-It is assumed that the audience of this paper has at least a basic
-understanding of how one models data in relational databases. As will
-quickly become apparent, a BigTable-like \cite{bigtable} data store is
-drastically different from the relational model, meaning that those who
-know their table schemas will have some un-learning to do before being able
-to fully grasp the leaner BigTable-model.
-
-% TODO: necessity of this paragraph?
-The reader is also assumed to possess basic understanding about distributed
-systems, with focus on high level abstractions rather than detailed
-knowledge about practical implementation details.
 
@@
 
-This thesis, fully, describes the Zohmg system and how to use it.
+This thesis describes the Zohmg system and how to use it.
 
-Chapter 2 the theoretical base of the problem statement is established.
+In chapter 2, the theoretical base of the problem statement is established.
 
-In chapter 3 the tools used for the implementation are declared, moving
+In chapter 3, the tools used for the implementation are declared, moving
 over to chapter 4 where the details of the implementation are presented.
@@
 numerical facts belonging to some dimension. The data cube is also known as a
-hypercube, it relates to multidimensional analogues of squares and cubes.
+hypercube. The data cube relates to multidimensional analogues of squares and cubes.
 In this context, each dimension of the cube corresponds to an attribute of
@@
 We have built a system---henceforth called Zohmg---which has the goal of being
-a large scale aggregating data warehouse for time-series-based log-like
+a large scale aggregating data store for time-series-based log-like
 data sources.

### 2009-05-07 doc/report/msc-report.tex
added a pair of TODOs and a future work idea.
doc/report/msc-report.tex: * corrections of sentences and misspellings. * added to and polished Aggregates section. * changed "defined" to "written" for a more clear discussion of mappers.

--- 2009-05-06 doc/report/msc-report.tex
@@
 
-This thesis, fully, describes the Zohmg system and how to use it.
+This thesis fully describes the Zohmg system and how to use it.
 
-Chapter 2 the theoretical base of the problem statement is established.
+Chapter 2 establish the theoretical base of the problem statement.
 
@@
 
-Chapters 7 and 8 discusses the results, related, and future work, and
+Chapters 7 and 8 discuss the results, related, and future work, and
 presents the conclusion, respectively.
@@
 the three dimensions correspond to one of the attributes time, products,
-and stores. Adding another attribute adds another dimension to the data
-cube, because of the one to one relationship of attributes and dimensions.
+and stores. Adding another attribute, for instance sales, adds another
+dimension to the data cube, because of the one to one relationship of
+attributes and dimensions.
 
@@
 
-An aggregate is a composite of several values. Typical examples of aggregate functions are average, count and sum.
+An aggregate is a composite of several values. Typical examples of
+aggregate functions are average, count, and sum.
 
-In a standard OLAP setup, a number of aggregates are pre-computed and stored in the cube.
-These base aggregates represent only a fraction of the very many possible aggregations.
-The remaining aggregates are computed from the base aggregates as they are demanded.
+In a standard OLAP setup, a number of aggregates are pre-computed and
+stored in the cube. These base aggregates represent only a fraction of the
+many possible aggregations. The remaining aggregates are computed from the
+base aggregates as they are demanded.
+
+The main reason for pre-computing aggregates is to reduce access times, the
+time it takes to compute an aggregate may be unacceptable to a waiting
+user. Pre-computing frequently requested aggregates is therefore regularly
+done in the background, giving the user fast access to these.
+\cite{olap_solutions}
+
+The classic trade-off between time and space applys to wether or not
+pre-compute an aggregate. Pre-computing to many aggregates might render an
+unfeasible of data amount to store, while pre-computing to few aggregates
+renders longer access times for aggregates which are not pre-computed. This
+problem is further discussed in the thesis, but left to the user of the
+Zohmg system to solve.
 
@@
 Depending on how the rowkeys are formatted different hierarchies are
-created. The order in which the data is stored is very important, it
-defines how effective scanning for data will be. For instance, let's say
-we have thousand rows and want ten out of those. In this case only one per
-cent of the data is interesting to us, if a scan would be required to
-visit all the thousand rows a lot of rows are skipped. Skipping rows is
-expensive in that sense that we have to read them but throw away the
-result. The goal is to push this cost to a bare minimum, using as many as
-possible of the rows we scan.
+created. The order in which the data is stored is important, it defines how
+effective scanning data will be. In the case of having thousand rows and
+requesting ten out of these, then only one por cent of the data is
+interesting. If a scan would be required to visit all the thousand rows a
+lot of rows are skipped. Skipping rows is expensive in that sense that they
+are but the result is thrown away. The goal is to push this cost to a bare
+minimum, using as many of the rows as possible.
 
-To minimize the cost of scanning for data we want it to be ordered so that
-a scan would use every row it visits. In the case with thousand rows of
-which we want ten, it would mean that we just scan ten rows and use all of
-them.
+The minimum cost of scanning data is achieved by ordering the data so that
+a scan uses every visited row. In the case with thousand rows of which ten
+are wanted, would mean that ten rows scan are scanned of which all are
+used.
 
@@
 
-Both Hadoop and HBase are written in Java---a popular high-level
-object-oriented programming language.
+Both Hadoop and HBase are written in Java---a high-level object-oriented
+programming language.
 
@@
 
-Python is a popular general-purpose high-level multi-paradigm---most
-notably: object-oriented, imperative and functional---scripting language.
+Python is a general-purpose high-level multi-paradigm---amongst others:
+object-oriented, imperative and functional---scripting language.
 \cite{python}
@@
 has no external dependencies. However a C extension which provides a
-serious speed boost is included. \cite{simplejson}
+serious speed boost is included in simplejson. \cite{simplejson}
 
@@
 
+Using Zohmg the user first creates a project directory. This is further
+explained in chapter 5.
+
 \section{Configuration}
@@
 The user's data sets are defined using YAML. The data set files reside in
-the config directory.
+the project's \texttt{config} directory.
 
@@
 
-The user defines a mapper which Zohmg wraps, interpreting the output from the
-mapper program and storing it in HBase.
+The user writes a mapper which Zohmg wraps, interpreting the output from
+the mapper program and storing it in HBase.
 
@@
 
-The data export server is uilt with Python Paste---a middleware
+The data export server is built with Python Paste---a middleware
 framework---the data server dispatches to the requested application
@@
 
-\noindent Edit the configuration files.
+\noindent Edit the configuration files, which are text files in YAML and
+Python format.
 
@@
 
-Import jobs use mappers defined by the user, either in Python or Java.
+Import jobs use mappers written by the user, either in Python or Java.
 Mappers written by the user import each line from the input file and set
@@
 
-The data server have three methods of serving data: raw, transformed, and
+The data server have three methods of serving data: raw, transformed, or
 static. The raw and transformed data are aggregates served from HBase,
 transformed in the latter case, while the static method just serve static
-files from the project's \textit{clients} directory. Aggregates served from
+files from the project's \texttt{clients} directory. Aggregates served from
 HBase are returned to the user as JSONP.
@@
 can interact with the data server. These static pages can for instance
-present some sort of query building mechanism (for insntance with drop down
-boxes) and then send this query to the data server and graph the resulting
+present some sort of query building mechanism---for instance with drop down
+boxes---and then send this query to the data server and graph the resulting
 data.

### 2009-05-07 doc/report/msc-report.tex

@@
 
-This thesis fully describes the Zohmg system and how to use it.
+This thesis describes the Zohmg system and how to use it.
 
@@
 
-In chapter 3 the tools used for the implementation are declared, moving
+In chapter 3, the tools used for the implementation are declared, moving
 over to chapter 4 where the details of the implementation are presented.
@@
 
+% TODO> I argue that we should have 'hadoop' and 'hbase' as the main sections,
+% and let mapreduce be a subsection of hadooop, bigtable a dito of hbase.
+
 \section{MapReduce}
@@
 
-% ONE DOES NOT SIMPLY PROCESS THE DATA ON ONE COMPUTER INTO MORDOR.
 Basically MapReduce is a two phase computation. First the input data is
@@
 
+% TODO> why is this Tool & BigTable-subsection called Dimensions? Disambiguation needed. -fredrik
 \subsection{Dimensions}
@@
 
+It'd be interesting to think a bit about how the data store could be fully elastic, i.e.
+not require any data set definition from the user and instead create column-families on the fly based on the dimensions output from the user's map.
+
 \subsection{Multiple data sets}

### 2009-05-07 doc/report/msc-report.tex
added a rather nice section on, I think, usage.

@@
 
+% COMMENT: do we really need to repeat this?
 As stated in chapter 3, Zohmg is built upon existing tools and software.
-It uses Apache Hadoop and Apache HBase, free software implementations of
-Google's MapReduce and BigTable respectively, both implemented in Java.
+It uses Dumbo to run jobs on Apache Hadoop, and uses Apache HBase to persist the data.
 
-Importing data to Zohmg is as simple as the user defines a mapper---the
-first half of a MapReduce program---which is run over a file that resides
-on HDFS\footnote{Hadoop Distributed Filesystem}.
+% TODO: not sure this is precisely a story about the implementation. putting it here for now.
 
-The resulting aggregated data, from the mapper, is stored in Apache HBase.
+-- A truly magnetic data store - the idea of enabling the user --
 
-Exporting data from Zohmg is done via a web interface which dumps JSON.
-It is possible to transform the stored data before exporting it with user
-defined transformers.
+From a user's perspective, storing aggregates in Zohmg is as simple as writing a map (in the MapReduce sense) and letting Zohmg run that map over the data set.
+The user's map is called 'usermap' from here on.
+In order to let Zohmg know what output to expect from the mapper, the user also defines the dimensions and units of his data.
+After all, the user most likely knows his own data better than anyone else.
 
-Zohmg is implemented in Python.
+A standard MapReduce map emits key-value pairs.
+Not so in the case of the Zohmg mapper: the usermap emits triples (that is, a tuple of three values) consisting of a timestamp, a list of dimensions and their respective values, and a list of measurements.
+In effect, the dimensions and the timestamp specify a point in n-space and the measurements represent the observations made at that point.
 
-Using Zohmg the user first creates a project directory. This is further
-explained in chapter 5.
+Going back to the running example of apache query logs, a usermap might look something like this:
+
+\begin{verbatim}
+# assumes Apache common log format.
+def map(key, value):
+  apachelog = parse_apache_log(value)
+  dimensions = {'useragent' : apachelog.useragent,
+                'path'      : apachelog.path,
+                'status'    : apachelog.statuscode}
+  values = {'pageviews' : 1,
+            'bytes'     : apachelog.size}
+
+ yield parsed.timestamp, dimensions, measurements
+\end{verbatim}
+
+The actual parsing of the log is abstracted; it is the emitting of dimensionality data I want to showcase in this example.
+We assume that parse_apache_log() accepts a single line from the log file and returns an object with the appropriate fields, such as useragent ant statuscode.
+
+For every line of the log file, we emit a point in n-space (the dimensions and the timestamp) together with a count of pageviews and bytes; one pageview per request and  the number of bytes served in the request.
+It is likely that you will want to add business logic here, perhaps only counting requests from useragents that aren not spiders and that resulted in a status code of 200.
+
+The last line 'yields' the tuple. This is a python construct called a generator that lets the map emit any number of tuples (or none!) for a single invocation.
+
+The user will import a day worth of logs like so:
+
+\begin{verbatim}
+  zohmg import mappers/apache.py /data/weblogs/2009/05/10
+\end{verbatim}
+
+This will instruct zohmg to apply the map function found in apache.py to all logs that it can find in the /data/weblogs/2009/05/10 directory on HDFS.
+The emitted output is duly collected and projections are carefully stored in the data store.
 
@@
 \subsection{Data sets}
+
+% TODO: define what a data set is, and what makes it different from a project.
 
@@
 
+% TODO: not truly correct
 The Darling output format for HBase. Between Zohmg and Hadoop JSON is used
@@
 
+% ?
 Fetches data from HBase via its Thrift interface. The data is served to the
@@
 
-We have built a system---henceforth called Zohmg---which has the goal of being
-a large scale aggregating data store for time-series-based log-like
-data sources.
+This is the user manual.
 
@@
 
+% COMMENT: I don't think we should mention the underlying store in the usage section.
 Data is stored in Apache HBase. A natural consequence of this that the

### 2009-05-07 doc/report/msc-report.tex
doc/report/msc-report.tex: * fixed underscores so tex source compiles.

@@
 
-The actual parsing of the log is abstracted; it is the emitting of dimensionality data I want to showcase in this example.
-We assume that parse_apache_log() accepts a single line from the log file and returns an object with the appropriate fields, such as useragent ant statuscode.
+The actual parsing of the log is abstracted; it is the emitting of
+dimensionality data I want to showcase in this example. We assume that
+parse\_apache\_log() accepts a single line from the log file and returns an
+object with the appropriate fields, such as useragent ant statuscode.
 

### 2009-05-07 doc/report/msc-report.tex
wrapped the lines of that text I wrote earlier today.
added a 500 word section on the implementation of dimensionality reduction, etc

@@
 % COMMENT: do we really need to repeat this?
-As stated in chapter 3, Zohmg is built upon existing tools and software.
-It uses Dumbo to run jobs on Apache Hadoop, and uses Apache HBase to persist the data.
+As stated in chapter 3, Zohmg is built upon existing tools and software. It
+uses Dumbo to run jobs on Apache Hadoop, and uses Apache HBase to persist the
+data.
 
@@
 
-From a user's perspective, storing aggregates in Zohmg is as simple as writing a map (in the MapReduce sense) and letting Zohmg run that map over the data set.
-The user's map is called 'usermap' from here on.
-In order to let Zohmg know what output to expect from the mapper, the user also defines the dimensions and units of his data.
-After all, the user most likely knows his own data better than anyone else.
+From a user's perspective, storing aggregates in Zohmg is as simple as writing
+a map (in the MapReduce sense) and letting Zohmg run that map over the data
+set. The user's map is called 'usermap' from here on. In order to let Zohmg
+know what output to expect from the mapper, the user also defines the
+dimensions and units of his data. After all, the user most likely knows his
+own data better than anyone else.
 
-A standard MapReduce map emits key-value pairs.
-Not so in the case of the Zohmg mapper: the usermap emits triples (that is, a tuple of three values) consisting of a timestamp, a list of dimensions and their respective values, and a list of measurements.
-In effect, the dimensions and the timestamp specify a point in n-space and the measurements represent the observations made at that point.
+A standard MapReduce map emits key-value pairs. Not so in the case of the
+Zohmg mapper: the usermap emits triples (that is, a tuple of three values)
+consisting of a timestamp, a list of dimensions and their respective values,
+and a list of measurements. In effect, the dimensions and the timestamp
+specify a point in n-space and the measurements represent the observations
+made at that point.
 
-Going back to the running example of apache query logs, a usermap might look something like this:
+Going back to the running example of apache query logs, a usermap might look
+something like this:
 
@@
 
-For every line of the log file, we emit a point in n-space (the dimensions and the timestamp) together with a count of pageviews and bytes; one pageview per request and  the number of bytes served in the request.
-It is likely that you will want to add business logic here, perhaps only counting requests from useragents that aren not spiders and that resulted in a status code of 200.
+For every line of the log file, we emit a point in n-space (the dimensions and
+the timestamp) together with a count of pageviews and bytes; one pageview per
+request and the number of bytes served in the request. It is likely that you
+will want to add business logic here, perhaps only counting requests from
+useragents that aren not spiders and that resulted in a status code of 200.
 
-The last line 'yields' the tuple. This is a python construct called a generator that lets the map emit any number of tuples (or none!) for a single invocation.
+The last line 'yields' the tuple. This is a python construct called a
+generator that lets the map emit any number of tuples (or none!) for a single
+invocation.
 
@@
 
-This will instruct zohmg to apply the map function found in apache.py to all logs that it can find in the /data/weblogs/2009/05/10 directory on HDFS.
-The emitted output is duly collected and projections are carefully stored in the data store.
+This will instruct zohmg to apply the map function found in apache.py to all
+logs that it can find in the /data/weblogs/2009/05/10 directory on HDFS. The
+emitted output is duly collected and projections are carefully stored in the
+data store.
+
+- Something about cardinality -
+
+When reasoning about dimensions, the issue of cardinality will come up sooner
+or later.
+
+The cardinality of a dimension is the number of distinct values of the
+dimension. For example, the dimension of 'logged in' has a cardinality of two
+since 'logged in' is a boolean variable; the user is either logged in or not.
+The 'country' dimension has a much higher cardinality: there are some 300
+countries in the world. Certainly, there could be dimension whose cardinality
+is infinite: the 'path' dimension in the web log example, for example, could
+in theory have a limitless number of values.
+The number of possible combinations of dimensions of your data -- the number
+of points in n-space -- is directly related to the cardinality of each
+dimension. It is likely that your data is of such a nature that it is
+physically impossible to store information about all dimension combinations.
+This is why we ask you, the user, to define what projections you are
+interested in: storing aggregates for the slices of the cube that interest you
+is both a time and space save, and in many cases the only feasible solution.
+
+-- Limits on cardinality --
+
+[What are the limits of our system? How high cardinality do we support? (This
+is closely tied to hbase and the qualifier-limit.) If Martin is going to have
+'normalized_path' as one of his dimensions, we might run into trouble.]
+
+-- The implementation of all that --
+
+Since we wish to store projections rather than all combinations of all
+dimensions, we need to mangle the data slightly before sending it down the
+pipe to the underlying data store.
+
+Dumbo expects a key-value pair from the mapper. Hadoop will collect all keys
+and give the reducer a key and an iterator of values. If we are to store
+projections, we need to encode both the desired dimensions and the unit in
+question in the key and emit the value of the unit in the 'value' field. This
+way we can use a simple SumReducer to sum the values for each unit at any
+point in n-space and be done.
+
+Remember, the usermapper gives us a point in n-space and a list of
+measurements. We need to break out the measurements and sum over each such
+measurement. Zohmg solves this by wrapping the usermapper in a cleverly named
+devicie, namely a mapperwrapper.
+
+% TODO: this para sounds as if it was written by a drunken monkey. The
+mapperwrapper takes each tuple emitted by the usermapper and performs a
+dimensionality reduction, ending up with the projections requested by the
+user. For every projection and unit, the dimensions remaining after the
+reduction are appended to the string representation of the unit and emitted as
+the key. The value is simply the value of the unit. The Hadoop framework will
+do the heavy work of collecting all keys and summing the values.
+
+An alternative approach would have been to perform the dimensionality
+reduction at a later stage, namely in the reducer of a multi-pass MapReduce
+job.
+
+The reducer, then, is very simple: it's a simple sum of the values. The output
+of the reducer will not be stored on disk as is usually the case, but will
+instead be interpreted by a custom OutputReader that persists the data to
+HBase.
+
+The reducer needs to communicate to the OutputReader under what rowkey,
+column-family and qualifier to store the computed sum. It does this by
+emitting the row key as the key and a json-serialized map of the form
+{"column-family:qualifier" : value}.
+
+The OutputReader interprets the json and persists the data. Et voila!
 

### 2009-05-07 doc/report/msc-report.tex
doc/report/msc-report.tex: * added \chapterquote and quotes in the beginning of each chapter. * altered flow in Dimension section.

--- 2009-05-07 doc/report/msc-report.tex
@@
 \author{Per Andersson, Fredrik M{\"o}llerstrand}
+
 \title{Zohmg---a large scale data store for aggregated time-series-based
@@
 
+\newcommand{\chapterquote}[2] {
+\begin{quote}
+\textit{"{#1}"}
+
+--- {#2}
+\end{quote}
+
+\vspace{24pt}
+}
+
 \begin{document}
+
 \maketitle
@@
 \noindent \Large{\textbf{Abstract}}
+
 \vspace{12pt}
@@
 \tableofcontents
+
 \vfill
+
 \pagebreak
@@
 % TODO: den har texten passar battre i ett eget stycke/sektion, background reading style.
-Says Wikipedia: "Data warehouse is a repository of an organization's electronically stored data.
-Data warehouses are designed to facilitate reporting and analysis." The main actions of a data warehouse is extract, transform and load data, although not necessarily in that order.
+
+\chapterquote{Data warehouse is a repository of an organization's
+electronically stored data. Data warehouses are designed to facilitate
+reporting and analysis.}{Wikipedia}
+
+The main actions of a data warehouse is extract, transform and load data,
+although not necessarily in that order.
 
@@
 
+\chapterquote{Einstein explained his theory to me every day, and on my
+arrival I was fully convinced that he understood it.}{Chaim Weizmann}
+
 \section{Data cubes}
 
-A data cube is a conceptual n-dimensional array of values. The values are
-numerical facts belonging to some dimension. The data cube is also known as a
-hypercube. The data cube relates to multidimensional analogues of squares and cubes.
-In this context, each dimension of the cube corresponds to an attribute of
-the data in question and each point along the dimension's axis is a value of
-the corresponding attribute. The mental model of a data cube aids in
-reasoning about dimensions. Data cubes are traditionally modeled as star
-schemas in relational database systems. \cite{olap_solutions}
+A data cube is a conceptual n-dimensional array of values, in which the values
+are numerical facts belonging to some dimension. The data cube is also known as
+a hypercube, relating to multidimensional analogues of squares and cubes. In
+this context, each dimension of the cube corresponds to an attribute of the
+data in question and each point along the dimension's axis is a value of the
+corresponding attribute. The mental model of a data cube aids in reasoning
+about dimensions. Data cubes are traditionally modeled as star schemas in
+relational database systems. \cite{olap_solutions}
 
@@
 \chapter{Tools}
+
+\chapterquote{Arrange whatever pieces come your way.}{Virgina Wolf}
+
+%\chapterquote{In short, intelligence, considered in what seems to be its
+%original feature, is the faculty of manufacturing artificial objects,
+%especially tools to make tools, and of indefinitely urging the
+%manufacture.}{Henri Bergson}
 
@@
 
+%\chapterquote{A good idea is about ten percent and implementation and hard
+%work, and luck is 90 percent.}{Guy Kawasaki}
+
+\chapterquote{It is not always what we know or analyzed before we make a
+decision that makes it a great decision. It is what we do after we make the
+decision to implement and execute it that makes it a good
+decision.}{William Pollard}
+
 % COMMENT: do we really need to repeat this?
@@
 
+\chapterquote{There comes a time when all people break down and do the
+unthinkable: They read the manual.}{Unknown}
+
 This is the user manual.
@@
 
+\chapterquote{Example is not the main thing in influencing others, it's
+the only thing}{Albert Schweitzer}
+
+%\chapterquote{There is nothing so annoying as a good example!!}{Mark Twain}
+
+%\chapterquote{Example isn't another way to teach, it is the only way to
+%teach}{Albert Einstein}
+
 \section{Web log analysis}
@@
 \chapter{Discussion}
+
+\chapterquote{In discussion comes the light!}{Unknown}
 
@@
 
+\chapterquote{When we can't dream any longer we die.}{Emma Goldman}
+
 Last.fm is looking at using Zohmg for multiple application areas, such as

### 2009-05-07 doc/report/msc-report.tex
zohmg/process.py: fixed 'zohmg.util' import error.

@@
 a map (in the MapReduce sense) and letting Zohmg run that map over the data
-set. The user's map is called 'usermap' from here on. In order to let Zohmg
+set. The user's map is called \textit{usermap} from here on. In order to let Zohmg
 know what output to expect from the mapper, the user also defines the
@@
 
-Going back to the running example of apache query logs, a usermap might look
+Going back to the running example of apache access logs, a usermap might look
 something like this:
@@
 \begin{verbatim}
-# assumes Apache common log format.
 def map(key, value):
-  apachelog = parse_apache_log(value)
-  dimensions = {'useragent' : apachelog.useragent,
-                'path'      : apachelog.path,
-                'status'    : apachelog.statuscode}
+  log = parse_apache_log(value)
+  dimensions = {'useragent' : log.useragent,
+                'path'      : log.path,
+                'status'    : log.statuscode}
   values = {'pageviews' : 1,
-            'bytes'     : apachelog.size}
+            'bytes'     : log.size}
 
@@
 
+\vspace{12pt}
+
 The actual parsing of the log is abstracted; it is the emitting of
 dimensionality data I want to showcase in this example. We assume that
-parse\_apache\_log() accepts a single line from the log file and returns an
-object with the appropriate fields, such as useragent ant statuscode.
+\texttt{parse\_apache\_log()} accepts a single line from the log file and
+returns an object with the appropriate fields, such as useragent ant
+statuscode.
 
@@
 
-The last line 'yields' the tuple. This is a python construct called a
+The last line \textit{yields} the tuple. This is a python construct called a
 generator that lets the map emit any number of tuples (or none!) for a single
@@
 
-This will instruct zohmg to apply the map function found in apache.py to all
-logs that it can find in the /data/weblogs/2009/05/10 directory on HDFS. The
-emitted output is duly collected and projections are carefully stored in the
-data store.
+This will instruct zohmg to apply the map function found in \texttt{apache.py}
+to all logs that it can find in the \texttt{/data/weblogs/2009/05/10} directory
+on HDFS. The emitted output is duly collected and projections are carefully
+stored in the data store.
 
@@
 The cardinality of a dimension is the number of distinct values of the
-dimension. For example, the dimension of 'logged in' has a cardinality of two
-since 'logged in' is a boolean variable; the user is either logged in or not.
-The 'country' dimension has a much higher cardinality: there are some 300
-countries in the world. Certainly, there could be dimension whose cardinality
-is infinite: the 'path' dimension in the web log example, for example, could
-in theory have a limitless number of values.
-The number of possible combinations of dimensions of your data -- the number
-of points in n-space -- is directly related to the cardinality of each
+dimension. For example, the dimension of \textit{logged in} has a cardinality
+of two since \textit{logged in} is a boolean variable; the user is either
+logged in or not. The \textit{country} dimension has a much higher
+cardinality: there are some 300 countries in the world. Certainly, there could
+be dimension whose cardinality is infinite: the \textit{path} dimension in the
+web log example, for example, could in theory have a limitless number of
+values.
+
+The number of possible combinations of dimensions of your data---the number
+of points in n-space---is directly related to the cardinality of each
 dimension. It is likely that your data is of such a nature that it is
@@
 is closely tied to hbase and the qualifier-limit.) If Martin is going to have
-'normalized_path' as one of his dimensions, we might run into trouble.]
+\textit{normalized\_path} as one of his dimensions, we might run into trouble.]
 
@@
 projections, we need to encode both the desired dimensions and the unit in
-question in the key and emit the value of the unit in the 'value' field. This
-way we can use a simple SumReducer to sum the values for each unit at any
+question in the key and emit the value of the unit in the \textit{value} field.
+This way we can use a simple SumReducer to sum the values for each unit at any
 point in n-space and be done.

### 2009-05-07 zohmg/process.py

@@ -1,5 +1,5 @@
 from zohmg.config import Config, Environ
-from zohmg.util import fail
+from zohmg.utils import fail
 import os, re
 
 class Process(object):

### 2009-05-07 zohmg/config.py
zohmg/config.py: * changed error messages to be consistent. * changed exception handling in Environ.
zohmg/create.py:
* changed default paths for jars in CLASSPATH.

@@ -64,24 +64,27 @@ class Config(object):
             ps = self.projections()
         except:
             # might as well return straight away; nothing else will work.
-            sys.stderr.write("hey!, you must define 'dataset', 'dimensions', 'units' and 'projections'.\n")
+            print >>sys.stderr, "[%s] Error: Data set: Missing definition of dataset, dimensions, units, projections." \
+                                % time.asctime()
             return False
 
         # dimensions, projections and units must be non-empty.
         if ds == None or us == None or ps == None or \
-           len(ds) == 0 or len(us) == 0 or len(ps) == 0:
-            sys.stderr.write("hey!, dimensions, projections and units must be non-empty\n")
-            return False
+            len(ds) == 0 or len(us) == 0 or len(ps) == 0:
+                print >>sys.stderr, "[%s] Error: Data set: dimensions, projections and units must be non-empty." \
+                                    % time.asctime()
+                return False
 
         # also, the configuration may not reference unknown dimensions.
         for p in ps:
             if ps[p] == None or len(ps[p]) == 0:
-                sys.stderr.write("hey!, you may not specify empty projections.\n")
+                print >>sys.stderr, "[%s] Error: Data set: Empty projections are not allowed." % time.asctime()
                 return False
             for d in ps[p]:
                 if d not in ds:
-                    sys.stderr.write("hey!, '%s' is a reference to an unknown dimension.\n" % d)
-                    sane = False
+                    print >>sys.stderr, "[%s] Error: Data set: %s is a reference to an unkown dimension." \
+                                    % (time.asctime(),d)
+                    sane = False # TODO: return False ?
 
         # also, there must be no funny characters in
         # the name of the dataset, the dimensions or units.
@@ -89,8 +92,8 @@ class Config(object):
             for x in xs:
                 m = re.match('^[a-zA-Z0-9]+$', x)
                 if m == None:
-                    sys.stderr.write("hey!, '%s' is an invalid name.\n" % x)
-                    sane = False
+                    print >>sys.stderr, "[%s] Error: Data set: '%s' is an invalid name." % (time.asctime(),x)
+                    sane = False # TODO: return False ?
 
         return sane
 
@@ -107,9 +110,9 @@ class Environ(object):
         sys.path.append("") # add cwd so we can import from it.
         try:
             env = __import__(config_path+"environment")
-        except ImportError, ioe:
[…]

### 2009-05-07 zohmg/create.py

@@ -13,9 +13,9 @@ HADOOP_HOME = ''
 # Jars for Hadoop Core, Hadoop Streaming and HBase
 CLASSPATH = (
   # be sure to set these correctly!
-  '/home/hadoop/hadoop-0.19/build/hadoop-0.19.0-dev-core.jar',
-  '/home/hadoop/hadoop-0.19/build/contrib/streaming/hadoop-0.19.0-dev-streaming.jar',
-  '/home/hadoop/hbase-0.19/build/hbase-0.19.0-dev.jar'
+  '/opt/hadoop-0.19.1/build/hadoop-0.19.0-dev-core.jar',
+  '/opt/hadoop-0.19.1/build/contrib/streaming/hadoop-0.19.0-dev-streaming.jar',
+  '/opt/hbase-0.19.1/build/hbase-0.19.0-dev.jar'
 )
 """
 

### 2009-05-07 doc/report/msc-report.tex
doc/report/msc-report.tex: * changed tense to passive in Implementation sections. * minor cosmetic changes.

@@
 HBase has a Thrift server which serves as the interface to languages other
-than Java--for instance Python.
+than Java---for instance Python.
 
@@
 From a user's perspective, storing aggregates in Zohmg is as simple as writing
-a map (in the MapReduce sense) and letting Zohmg run that map over the data
-set. The user's map is called \textit{usermap} from here on. In order to let Zohmg
-know what output to expect from the mapper, the user also defines the
-dimensions and units of his data. After all, the user most likely knows his
-own data better than anyone else.
+a map function (in the MapReduce sense) and letting Zohmg run that map function
+over the data set. The user's map function is called \textit{usermap function}
+from here on. In order to let Zohmg know what output to expect from the mapper,
+the user also defines the dimensions and units of his data. After all, the user
+most likely knows his own data better than anyone else.
 
-A standard MapReduce map emits key-value pairs. Not so in the case of the
-Zohmg mapper: the usermap emits triples (that is, a tuple of three values)
-consisting of a timestamp, a list of dimensions and their respective values,
-and a list of measurements. In effect, the dimensions and the timestamp
-specify a point in n-space and the measurements represent the observations
-made at that point.
+A standard MapReduce map function emits key-value pairs. Not so in the case of
+the Zohmg mapper: the usermap function emits triples (that is, a tuple of three
+values) consisting of a timestamp, a list of dimensions and their respective
+values, and a list of measurements. In effect, the dimensions and the timestamp
+specify a point in n-space and the measurements represent the observations made
+at that point.
 
-Going back to the running example of apache access logs, a usermap might look
-something like this:
+Going back to the running example of apache access logs, a usermap function
+might look something like this:
 
@@
 
-The actual parsing of the log is abstracted; it is the emitting of
-dimensionality data I want to showcase in this example. We assume that
+The actual parsing of the log is abstracted; in the above example the
+emitting of dimensionality is showcased. It is assumed that
 \texttt{parse\_apache\_log()} accepts a single line from the log file and
-returns an object with the appropriate fields, such as useragent ant
+returns an object with the appropriate fields, such as useragent and
 statuscode.
 
-For every line of the log file, we emit a point in n-space (the dimensions and
-the timestamp) together with a count of pageviews and bytes; one pageview per
-request and the number of bytes served in the request. It is likely that you
-will want to add business logic here, perhaps only counting requests from
-useragents that aren not spiders and that resulted in a status code of 200.
+For every line of the log file, a point in n-space (the dimensions and the
+timestamp) is emitted together with a count of pageviews and bytes; one
+pageview per request and the number of bytes served in the request. It is
+possible to add business logic here, perhaps only counting requests from
+useragents that are not web spiders and that resulted in a HTTP 200 status
+code.
 
-The last line \textit{yields} the tuple. This is a python construct called a
-generator that lets the map emit any number of tuples (or none!) for a single
-invocation.
+The last line \textit{yields} the tuple. This is a Python construct called a
+generator that lets the map function emit any number of tuples (or none!)
+for a single invocation.
 
@@
 \begin{verbatim}
-  zohmg import mappers/apache.py /data/weblogs/2009/05/10
+   zohmg import mappers/apache.py \
+   /data/weblogs/2009/05/10
 \end{verbatim}
@@
 This will instruct zohmg to apply the map function found in \texttt{apache.py}
-to all logs that it can find in the \texttt{/data/weblogs/2009/05/10} directory
-on HDFS. The emitted output is duly collected and projections are carefully
-stored in the data store.
+to all logs that it can find in the directory
+\texttt{/data/weblogs/2009/05/10} on HDFS. The emitted output is duly
+collected and projections are carefully stored in the data store.
 
@@
 
-When reasoning about dimensions, the issue of cardinality will come up sooner
-or later.
+TODO?: When reasoning about dimensions, the issue of cardinality will come
+up sooner or later.
 
 The cardinality of a dimension is the number of distinct values of the
-dimension. For example, the dimension of \textit{logged in} has a cardinality
-of two since \textit{logged in} is a boolean variable; the user is either
-logged in or not. The \textit{country} dimension has a much higher
-cardinality: there are some 300 countries in the world. Certainly, there could
-be dimension whose cardinality is infinite: the \textit{path} dimension in the
-web log example, for example, could in theory have a limitless number of
-values.
+dimension, see it as number of members in a set where the dimension is
+viewed as a set. For example, the dimension of \textit{logged in} has a
+cardinality of two since \textit{logged in} is a boolean variable; the user
+is either logged in or not. The \textit{country} dimension has a much
+higher cardinality: there are some 300 countries in the world. Certainly,
+there could be dimensions whose cardinality is infinite: the \textit{path}
+dimension in the web log example, for example, could in theory have a
+limitless number of values.
 
-The number of possible combinations of dimensions of your data---the number
+The number of possible combinations of dimensions of the data---the number
 of points in n-space---is directly related to the cardinality of each
-dimension. It is likely that your data is of such a nature that it is
-physically impossible to store information about all dimension combinations.
-This is why we ask you, the user, to define what projections you are
-interested in: storing aggregates for the slices of the cube that interest you
-is both a time and space save, and in many cases the only feasible solution.
+dimension. It is likely that the user's data is of such a nature that it is
+physically impossible to store information about all dimension
+combinations.  This is why the user is asked to define what projections are
+interesting: storing aggregates for the slices of the cube that is
+interesting is both a time and space save, and in many cases the only
+feasible solution.
 
@@
 
-Since we wish to store projections rather than all combinations of all
-dimensions, we need to mangle the data slightly before sending it down the
-pipe to the underlying data store.
+Since the wish is to store projections rather than all combinations of all
+dimensions, the data needs to be mangled slightly before sending it down
+the pipe to the underlying data store.
 
 Dumbo expects a key-value pair from the mapper. Hadoop will collect all keys
-and give the reducer a key and an iterator of values. If we are to store
-projections, we need to encode both the desired dimensions and the unit in
-question in the key and emit the value of the unit in the \textit{value} field.
-This way we can use a simple SumReducer to sum the values for each unit at any
-point in n-space and be done.
+and give the reducer a key and an iterator of values.
+
+If projections are to be stored, both the desired dimensions and the unit
+in question needs to be encoded in the key, and the value of the unit
+emitted in the \textit{value} field.
+
+By doing this it is possible to use a simple SumReducer to sum the values
+for each unit at any point in n-space.
 
@@
 
-% TODO: this para sounds as if it was written by a drunken monkey. The
-mapperwrapper takes each tuple emitted by the usermapper and performs a
+% TODO: this para sounds as if it was written by a drunken monkey.
+The mapperwrapper takes each tuple emitted by the usermapper and performs a
 dimensionality reduction, ending up with the projections requested by the
 user. For every projection and unit, the dimensions remaining after the
-reduction are appended to the string representation of the unit and emitted as
-the key. The value is simply the value of the unit. The Hadoop framework will
-do the heavy work of collecting all keys and summing the values.
+reduction are appended to the string representation of the unit and emitted
+as the key. The value is simply the value of the unit. The Hadoop framework
+will do the heavy work of collecting all keys and summing the values.
 
@@
 column-family and qualifier to store the computed sum. It does this by
-emitting the row key as the key and a json-serialized map of the form
-{"column-family:qualifier" : value}.
+emitting the row key as the key and a JSON-serialized map of the form
+\texttt{\{"column-family:qualifier" : value\}}.
 
-The OutputReader interprets the json and persists the data. Et voila!
+The OutputReader interprets the JSON and persists the data. Et voila!
 

### 2009-05-08 zohmg/create.py
zohmg/create.py: * added default value for HADOOP_HOME. * commented out variables.

@@ -8,14 +8,14 @@ DIRS = ["clients","config","lib","mappers","transformers"]
 ENV_SCRIPT = """# Please define the following environment variables.
 
 # Path to Hadoop directory.
-HADOOP_HOME = ''
+#HADOOP_HOME = '/opt/hadoop-0.19.1'
 
 # Jars for Hadoop Core, Hadoop Streaming and HBase
 CLASSPATH = (
   # be sure to set these correctly!
-  '/opt/hadoop-0.19.1/build/hadoop-0.19.0-dev-core.jar',
-  '/opt/hadoop-0.19.1/build/contrib/streaming/hadoop-0.19.0-dev-streaming.jar',
-  '/opt/hbase-0.19.1/build/hbase-0.19.0-dev.jar'
+  #'/opt/hadoop-0.19.1/build/hadoop-0.19.0-dev-core.jar',
+  #'/opt/hadoop-0.19.1/build/contrib/streaming/hadoop-0.19.0-dev-streaming.jar',
+  #'/opt/hbase-0.19.1/build/hbase-0.19.0-dev.jar'
 )
 """
 

### 2009-05-08 zohmg/create.py
zohmg/create.py: * version numbers of jars correspond to what install_hadoop_hbase.sh installs.

@@ -13,9 +13,9 @@ ENV_SCRIPT = """# Please define the following environment variables.
 # Jars for Hadoop Core, Hadoop Streaming and HBase
 CLASSPATH = (
   # be sure to set these correctly!
-  #'/opt/hadoop-0.19.1/build/hadoop-0.19.0-dev-core.jar',
-  #'/opt/hadoop-0.19.1/build/contrib/streaming/hadoop-0.19.0-dev-streaming.jar',
-  #'/opt/hbase-0.19.1/build/hbase-0.19.0-dev.jar'
+  #'/opt/hadoop-0.19.1/build/hadoop-0.19.2-dev-core.jar',
+  #'/opt/hadoop-0.19.1/build/contrib/streaming/hadoop-0.19.2-dev-streaming.jar',
+  #'/opt/hbase-0.19.1/build/hbase-0.19.1-dev.jar'
 )
 """
 

### 2009-05-08 zohmg/create.py
zohmg/create.py: * corrected name of hbase jar.

@@ -15,7 +15,7 @@ CLASSPATH = (
   # be sure to set these correctly!
   #'/opt/hadoop-0.19.1/build/hadoop-0.19.2-dev-core.jar',
   #'/opt/hadoop-0.19.1/build/contrib/streaming/hadoop-0.19.2-dev-streaming.jar',
-  #'/opt/hbase-0.19.1/build/hbase-0.19.1-dev.jar'
+  #'/opt/hbase-0.19.1/build/hbase-0.19.1.jar'
 )
 """
 

### 2009-05-08 setup.py
is safe for zippin

@@ -7,7 +7,7 @@ setup(
     author_email = '{fredrik,per}@last.fm',
     license = 'GNU Fearsome Dude License',
     packages = ['zohmg'],
-    zip_safe = False,
+    zip_safe = True,
     entry_points = {
         'console_scripts': [
             'zohmg = zohmg.cmd:zohmg',

### 2009-05-08 scripts/install_hadoop_hbase.sh
scripts/install_hadoop_hbase.sh: * changed log filename. * quoted log paths.

@@ -19,7 +19,7 @@ hadoop_release="http://mirrors.ukfast.co.uk/sites/ftp.apache.org/hadoop/core/had
 hadoop_1722="https://issues.apache.org/jira/secure/attachment/12401426/$patch_1722"
 hadoop_5450="https://issues.apache.org/jira/secure/attachment/12401846/$patch_5450"
 hbase_release="http://mirrors.ukfast.co.uk/sites/ftp.apache.org/hadoop/hbase/hbase-0.19.1/$hbase_tar"
-install_log="$(pwd)/hadoop_hbase_install.log"
+install_log="$(pwd)/install_hadoop_hbase.log"
 install_tmplog=$(mktemp /tmp/zohmg-log.XXXXXXXX)
 
 
@@ -35,7 +35,7 @@ function exec_and_log() {
 
     # execute and log to intermediate.
     if [ "x" = "x$screenoutput" ]; then
-        $1 &>$install_tmplog
+        $1 &>"$install_tmplog"
     else
         $1
     fi
@@ -43,8 +43,8 @@ function exec_and_log() {
 
     # log if not requested not to.
     if [ "x" = "x$3" ]; then
-        cat $install_log $install_tmplog >"$install_log.new"
-        mv "$install_log.new" $install_log
+        cat "$install_log" "$install_tmplog" >"$install_log.new"
+        mv "$install_log.new" "$install_log"
     fi
 
     # check exit code.
@@ -129,7 +129,7 @@ done
 
 
 # truncate logs.
->$install_log
+>"$install_log"
 >"$install_log.new"
 
 

### 2009-05-08 scripts/install_hadoop_hbase.sh
scripts/install_hadoop_hbase.sh: fixed hadoop patching.

@@ -53,7 +53,6 @@ function exec_and_log() {
         echo $msg
         exit 1
     fi
-
 }
 
 
@@ -209,7 +208,22 @@ else
     for patch in "$patch_1722" "$patch_5450"; do
         num=$(echo $patch | sed 's/.patch$//')
         printf "Applying patch $num... "
-        exec_and_log "cd $hadoop ; patch -p0 <$files/patches/$patch" "Error: Could not apply patch $num."
+        cd $hadoop
+        # inlined exec_and_log because of sh -c "command".
+        # execute and log to intermediate.
+        sh -c "patch -p0 <$files/patches/$patch" &>"$install_tmplog"
+        ret=$?
+
+        # log.
+        cat "$install_log" "$install_tmplog" >"$install_log.new"
+        mv "$install_log.new" "$install_log"
+
+        # check exit code.
+        if [ $ret -ne 0 ]; then
+            echo
+            echo "Error: Could not apply patch $num."
+            exit 1
+        fi
         echo "done."
     done
     exec_and_log "echo ok"

### 2009-05-08 doc/report/msc-report.tex
hard-wrapped a bunch of long lines. no content change.
improved abstract and introduction.

@@
 \noindent
-This thesis describes a data store for multi-dimensional time-series-based data.
-The data is aggregated over many units of measurements and across multiple dimensions.
-One of these dimensions is always time.
 
-Zohmg solves the problem of summing, storing and serving measurements.
-These measurements may be taken from log files or similar sources.
-Zohmg stores only the summarized data, not the atomic measurements.
-Zohmg is optimized for speed of data retrieval.
+This thesis describes a data store for multi-dimensional time-series-based
+data. The data is aggregated over many units of measurements and across
+multiple dimensions. One of these dimensions is always time.
 
-The thesis is presented in three major sections: problem statement, theoretical solution and implementation details.
+In this thesis we present a data store called Zohmg. Zohmg is built on top of
+HBase and uses Hadoop to run import jobs in parallel. It solves the problem of
+summing, storing and serving measurements. These measurements may be taken
+from log files or similar sources. Zohmg stores only the summarized data, not
+the atomic measurements. The data store is optimized for speed of data
+retrieval: one of the design goals was to serve data at mouse-click rate and
+promote real-time data exploration.
+
+Similar data stores exist: they generally use a relational database system as
+their underlying data store. The novelty of our approach is to model OLAP
+cubes on top of a bigtable-like data store.
+
+The thesis is presented in three major sections: problem statement,
+theoretical solution and implementation details.
 
@@
 
-% TODO: den har texten passar battre i ett eget stycke/sektion, background reading style.
+Any organization will eventually wish to evaluate and analyze various
+characteristics of the service and operations they manage. These services can
+be anything from web sites to theme parks, and the evaluation might include
+characteristics such as the number of visitors per day, and the average time a
+theme park guest spends queueing for a particular ride.
 
-\chapterquote{Data warehouse is a repository of an organization's
-electronically stored data. Data warehouses are designed to facilitate
-reporting and analysis.}{Wikipedia}
+Inevitably, these services and operations will generate a log for later
+analysis. As the services grow in size, so will the data sets they generate grow.
 
-The main actions of a data warehouse is extract, transform and load data,
-although not necessarily in that order.
-
-The classic thinking in data warehousing is the divide between facts and dimensons.
-The facts are measures categorized by dimensions.
-In OLAP, a sort of data warehouse, the OLAP cube is typically modelled in fact tables and dimension tables in a relational database.
-The novelty of our approach is to model OLAP cube-like aggregation data store on top of a bigtable-like data store.
-
-As large scale web services grow they generate vast volumes of log data,
-for instance Apache web logs. The ability to accurately analyse gathered
-data is important to the evaluation, planning, and success of the web
-service. Storing and analysing massive volumes of log data requires
-special measures to be taken. From an end-user perspective the requirement
-is an easy, complete and fast data analysing procedure. The technical
-aspects of this requirement boils down to the following key points:
-Visualising, storing and importing the data. The result is an
-multidimensional information system which is subject to
-OLAP\footnote{On-Line Analytical Processing}. \cite{olap_solutions}
-
-All three requirements are depend on one another, what data and how it is
-stored affects what can be visualised. Also, what data is imported---and
-how it is analysed---affects how it can be stored and hence visualised.
+An important property of these data sets is that they generally have a regular
+structure and are represented in independent records. The records can be
+broken up without losing any of their meaning and are as such very well suited
+for parallel analysis. The data sets are not fit to store using traditional
+database techniques, if only for the fact that they are too large to fit in a
+relational database.
 
@@
 
+\section{Other data warehouses}
+
+A data warehouse is a repository of an organization's
+electronically stored data. Data warehouses are designed to facilitate
+reporting and analysis. \cite{wikipedia}
+
+The core actions of a data warehouse is to load, transform and extract data.
+The classic thinking in modelling data warehouses is to divide data into facts
+and dimensons. The facts are measures categorized by dimensions.
+
+In OLAP\footnote{On-Line Analytical Processing}, which is a kind of data
+warehouse, the data is thought of as being structured as a cube. The cube is
+typically modelled in a realational database with fact tables and dimension
+tables, upon which the system performs joins to extract data.
+\cite{olap_solutions}
+
 \section{Data cubes}
@@
 
-One of the most common operations on a data cube is the projection.
-A projection is a dimensionality reduction, where one or more dimensions are discarded from the cube with the measurements along each axis summed up, leaving a data cube of a lesser dimension.
+One of the most common operations on a data cube is the projection. A
+projection is a dimensionality reduction, where one or more dimensions are
+discarded from the cube with the measurements along each axis summed up,
+leaving a data cube of a lesser dimension.
 
@@
 
-% TODO> why is this Tool & BigTable-subsection called Dimensions? Disambiguation needed. -fredrik
+% TODO> why is this Tool & BigTable-subsection called Dimensions?
+% Disambiguation needed. -fredrik
 \subsection{Dimensions}
@@
 
-% TODO: not sure this is precisely a story about the implementation. putting it here for now.
+% TODO: not sure this is precisely a story about the implementation.
+% putting it here for now.
 
@@
 
-Says Wikipedia: "Organizations generally start off with relatively simple use of data warehousing.
-Over time, more sophisticated use of data warehousing evolves." Zohmg is that relatively simple data warehouse; the seed from which a more sophisticated data warehouse may evolve.
+Says Wikipedia: "Organizations generally start off with relatively simple use
+of data warehousing. Over time, more sophisticated use of data warehousing
+evolves." Zohmg is that relatively simple data warehouse; the seed from which
+a more sophisticated data warehouse may evolve.
 
-There is as of yet no support for specifying granularities. The only supported resolution of time is the day.
+There is as of yet no support for specifying granularities. The only supported
+resolution of time is the day.
 
-The single aggregation performed is summing of the measurements. Future work may include average, max and min, etc.
+The single aggregation performed is summing of the measurements. Future work
+may include average, max and min, etc.
 
-There is no metadata store. It would be very helpful for the application writer to know what the known values are for each dimension.
+There is no metadata store. It would be very helpful for the application
+writer to know what the known values are for each dimension.
 
-It'd be interesting to think a bit about how the data store could be fully elastic, i.e.
-not require any data set definition from the user and instead create column-families on the fly based on the dimensions output from the user's map.
-
+It'd be interesting to think a bit about how the data store could be fully
+elastic, i.e. not require any data set definition from the user and instead
+create column-families on the fly based on the dimensions output from the
+user's map.
 \subsection{Multiple data sets}

### 2009-05-08 zohmg/process.py
zohmg/process.py: added python version check.

@@ -1,6 +1,6 @@
 from zohmg.config import Config, Environ
 from zohmg.utils import fail
-import os, re
+import os, platform, re
 
 class Process(object):
     def go(self, mapper, input, for_dumbo):
@@ -9,6 +9,7 @@ class Process(object):
         resolver = 'fm.last.darling.HBaseIdentifierResolver'
         outputformat = 'org.apache.hadoop.hbase.mapred.TableOutputFormat'
 
+        python_version = platform.python_version()[:-2]
         opts = [('jobconf',"hbase.mapred.outputtable=" + table),
                 ('jobconf','stream.io.identifier.resolver.class=' + resolver),
                 ('streamoutput','hbase'), # resolved by identifier.resolver
@@ -16,7 +17,7 @@ class Process(object):
                 ('input', input),
                 ('output','/tmp/does-not-matter'),
                 # Push zohmg egg and darling jar.
-                ('libegg','/usr/lib/python2.5/site-packages/zohmg-0.0.1-py2.5.egg'),
+                ('libegg','/usr/lib/python'+python_version+'/site-packages/zohmg-0.0.1-py'+python_version+'.egg'),
                 ('libjar','/usr/local/lib/zohmg/darling-0.0.3.jar'),
                 ('file','lib/usermapper.py') # TODO: handle this more betterer.
                ]

### 2009-05-08 doc/report/msc-report.tex
rounded out introduction, merged subsection on data size.

@@
 
-In this thesis we present a data store called Zohmg. Zohmg is built on top of
-HBase and uses Hadoop to run import jobs in parallel. It solves the problem of
-summing, storing and serving measurements. These measurements may be taken
-from log files or similar sources. Zohmg stores only the summarized data, not
-the atomic measurements. The data store is optimized for speed of data
-retrieval: one of the design goals was to serve data at mouse-click rate and
-promote real-time data exploration.
+We present a data store called Zohmg. It is built on top of HBase and uses
+Hadoop to run import jobs in parallel. Zohmg stores and serves aggregates of
+measurements. The measurements are extracted taken from log files or similar
+sources with user-supplied code.
+
+The data store is optimized for speed of data retrieval: one of the design
+goals was to serve data at mouse-click rate and promote real-time data
+exploration. As such, Zohmg stores only the summarized data, not the atomic
+measurements.
 
@@
 Inevitably, these services and operations will generate a log for later
-analysis. As the services grow in size, so will the data sets they generate grow.
+analysis. As the services grow in size, so will the data sets they generate.
 
 An important property of these data sets is that they generally have a regular
-structure and are represented in independent records. The records can be
-broken up without losing any of their meaning and are as such very well suited
-for parallel analysis. The data sets are not fit to store using traditional
-database techniques, if only for the fact that they are too large to fit in a
-relational database.
+structure and are represented as a number of independent records. The records
+can be broken up without losing any of their meaning and are as such very well
+suited for parallel analysis. The data sets are not fit to store using
+traditional database techniques, if only for the fact that they are too large
+to fit in a relational database.
+
+Each log record typically contains quite a bit of information. For example, a
+log of visits to a web site might contain a timestamp, information about what
+country the visiting user is from, and a string of text that identifies the
+web browser that was used. From these base facts one can collect information
+about what countries generate the most traffic to the site, and which web
+browser is most popular in a certain country, for example.
+
+These bits of information can be thought of as points along an axis; Sweden,
+Norway and Denmark are all points along the country axis, for instance. This
+is an important concept and it will be used throughout the text.
+
+Typically, the size of a data set is in the range of several gigabytes of data
+per day. A single computer is not likely to be able to process that amount of
+data before a new batch arrives, and for very large data sets the single
+computer would simply not be able to fit the intermediate data set in its
+memory at all. Some data sets are many times larger than that and do not even
+fit inside a single computers storage devices.
+
+It is suitable to store the data across many storage nodes in what is called a
+distributed filesystem and to to exploit the parallel nature of the records by
+breaking up the data sets in pieces and have each piece be processed
+separately on a node in a large compute cluster.
+
+% TODO: fixa cites. Luckily, this is a problem that has already been solved.
+Google has published papers on the MapReduce framework \cite{mapreduce}, which
+allows for parallel processing of large data sets, and The Google Filesystem
+\cite{gfs}, a scalable, distributed filesystem. The Apache Hadoop project has
+published open source implementations of a mapreduce framework and a
+distributed filesystem, both of which build on the ideas in the papers
+published by Google.
 
@@
 this restriction does not cripple the system much---if at all.
-
-\subsection{Data size}
-
-The sheer size of the data makes it unfeasible to use a single computer.
-Eventually a single machine would be fed so much data that it cannot cope
-with it. For instance a single web log file for one day can be up to
-several gigabytes. It is problematic for a single machine to handle this
-because of the limitations of memory, loading the entire file in memory.
-Distributed computing solves this by spreading the computations out to
-several machines in parallel. The obvious gain is that the waiting time
-before a computation is done is divided by the amount of nodes used,
-compared to running the computation on one machine. There is also a gain
-related to the actual machines, it is possible to use commodity hardware.
-Because MapReduce and BigTable has high availability and reliability a node
-in the cluster can safely be replaced on the fly if it fails.
-
-Scalability is also very important. Using single machines for computing is
-problematic when the data set grows, more powerful machines has to be
-obtained. More power contained in a single machine results in higher price,
-eventually it is too expensive to perform computations on very large
-data sets. Both MapReduce and BigTable scale close to linearly, adding one
-machine to the cluster adds equal amount of processing power. Since
-MapReduce and BigTable can use commodity hardware it is reasonable to buy
-extra machines to gain extra computing power or storage space.
 

### 2009-05-08 zohmg/process.py
zohmg/process.py: automatic find of zohmg egg.

@@ -1,6 +1,6 @@
 from zohmg.config import Config, Environ
 from zohmg.utils import fail
-import os, platform, re
+import os, re, sys
 
 class Process(object):
     def go(self, mapper, input, for_dumbo):
@@ -9,7 +9,6 @@ class Process(object):
         resolver = 'fm.last.darling.HBaseIdentifierResolver'
         outputformat = 'org.apache.hadoop.hbase.mapred.TableOutputFormat'
 
-        python_version = platform.python_version()[:3]
         opts = [('jobconf',"hbase.mapred.outputtable=" + table),
                 ('jobconf','stream.io.identifier.resolver.class=' + resolver),
                 ('streamoutput','hbase'), # resolved by identifier.resolver
@@ -17,7 +16,7 @@ class Process(object):
                 ('input', input),
                 ('output','/tmp/does-not-matter'),
                 # Push zohmg egg and darling jar.
-                ('libegg','/usr/lib/python'+python_version+'/site-packages/zohmg-0.0.1-py'+python_version+'.egg'),
+                ('libegg',[z for z in sys.path if "zohmg" in z][0]),
                 ('libjar','/usr/local/lib/zohmg/darling-0.0.3.jar'),
                 ('file','lib/usermapper.py') # TODO: handle this more betterer.
                ]

### 2009-05-08 scripts/install_hadoop_hbase.sh
scripts/install_hadoop_hbase.sh: * log to file and output to screen. * changed printf "Action... " to echo on a line of its own.

@@ -26,26 +26,18 @@ install_tmplog=$(mktemp /tmp/zohmg-log.XXXXXXXX)
 # helpers.
 
 # execute $1 and exit if it failed, displaying $2.
-# it is possible to set $3 in order to print to screen (without logging).
 function exec_and_log() {
     # setup variables.
     command=$1
     [ "x" = "x$2" ] && msg="Error: Could not execute $command." || msg=$2
-    screenoutput=$3
 
     # execute and log to intermediate.
-    if [ "x" = "x$screenoutput" ]; then
-        $1 &>"$install_tmplog"
-    else
-        $1
-    fi
+    $1 | tee "$install_tmplog" | cat
     ret=$?
 
-    # log if not requested not to.
-    if [ "x" = "x$3" ]; then
-        cat "$install_log" "$install_tmplog" >"$install_log.new"
-        mv "$install_log.new" "$install_log"
-    fi
+    # log.
+    cat "$install_log" "$install_tmplog" >"$install_log.new"
+    mv "$install_log.new" "$install_log"
 
     # check exit code.
     if [ $ret -ne 0 ]; then
@@ -143,7 +135,7 @@ hbase_conf=$hbase/conf
 echo "Checking for necessary programs..."
 for command in "ant -version" "patch --version" "wget --version"; do
     program=$(echo $command | sed 's/ .*//')
-    printf "Checking for $program... "
+    echo "Checking for $program... "
     exec_and_log "$command" "Error: Missing program: $program not found."
     echo "ok."
 done
@@ -152,7 +144,7 @@ done
 # download or use already existing files.
 if [ "x" = "x$files" ]; then
     # create temporary directories for downloads.
-    printf "Creating temporary directory... "
+    echo "Creating temporary directory... "
     files=$(mktemp -d /tmp/zohmg-deps.XXXXXX)
     mkdir -p $files/patches
     echo "done."
@@ -179,7 +171,7 @@ if [ "x" = "x$files" ]; then
     fi
     echo "Done downloading files."
 else
-    printf "Using previously downloaded files in $files... "
+    echo "Using previously downloaded files in $files... "
     for file in "patches/$patch_1722" "patches/$patch_5450" "$hadoop_tar" "$hbase_tar"; do
         exec_and_log "ls $files/$file" "Error: Could not find file $files/$file."
     done
[…]

### 2009-05-08 doc/report/msc-report.tex
wrapped two lines - no content change.
doc/report/msc-report.tex: * cosmetic changes.

@@
 It is suitable to store the data across many storage nodes in what is called a
-distributed filesystem and to to exploit the parallel nature of the records by
+distributed file system and to to exploit the parallel nature of the records by
 breaking up the data sets in pieces and have each piece be processed
@@
 
-% TODO: fixa cites. Luckily, this is a problem that has already been solved.
-Google has published papers on the MapReduce framework \cite{mapreduce}, which
-allows for parallel processing of large data sets, and The Google Filesystem
-\cite{gfs}, a scalable, distributed filesystem. The Apache Hadoop project has
-published open source implementations of a mapreduce framework and a
-distributed filesystem, both of which build on the ideas in the papers
+% TODO: fixa cites.
+Luckily, this is a problem that has already been solved. Google has published
+papers on the MapReduce framework \cite{mapreduce}, which allows for parallel
+processing of large data sets, and The Google File System \cite{gfs}, a
+scalable, distributed file system. The Apache Hadoop \cite{hadoop} project has
+published free software implementations of a MapReduce framework and a
+distributed file system, both of which build on the ideas in the papers
 published by Google.
@@
 
-How to use the Zohmg system is presented in chapter 5, while chapter 6
+How to use the Zohmg system is presented in chapter~5, while chapter 6
 shows an example work flow.
@@
 The slice operation fixes the values of one or several dimensions in the cube.
-The data omitted from this operation would be any data associated with the values of the dimension that were not fixed.
-The dice operation is a slice on more than two dimensions.
-[source: http://www.cs.ualberta.ca/~zaiane/courses/cmput690/glossary.html]
+The data omitted from this operation would be any data associated with the
+values of the dimension that were not fixed. The dice operation is a slice on
+more than two dimensions. [source:
+http://www.cs.ualberta.ca/~zaiane/courses/cmput690/glossary.html]
 
@@
 
-Time is the most common of the dimensions in the data cube constructed for Zohmg.
-The time stamp is a common property for all data handled by Zohmg; every measurement is expected to have time as one of its dimensions.
-TODO: really, this belongs in the implementation section.
-I find nothing interesting to say about time series in this section.
+Time is the most common of the dimensions in the data cube constructed for
+Zohmg. The time stamp is a common property for all data handled by Zohmg;
+every measurement is expected to have time as one of its dimensions. TODO:
+really, this belongs in the implementation section. I find nothing interesting
+to say about time series in this section.
 
@@
 
- yield parsed.timestamp, dimensions, measurements
+ yield log.timestamp, dimensions, measurements
 \end{verbatim}
@@
 The last line \textit{yields} the tuple. This is a Python construct called a
-generator that lets the map function emit any number of tuples (or none!)
-for a single invocation.
+\textit{generator} that lets the map function emit any number of tuples (or
+none!) for a single invocation.
 
@@
 
-The user writes a mapper which Zohmg wraps, interpreting the output from
-the mapper program and storing it in HBase.
+In order to import data to Zohmg, the user writes a mapper which Zohmg wraps;
+interpreting the output from the mapper program and storing it in HBase.
 
@@
 to the data server via HTTP, an example query could look like (the URL is
-wrapped to fit the page) \\
+wrapped to fit the page)
 
-\texttt{http://localhost:8086/data/?t0=20090101 \\
-\&t1=20090420\&unit=pageviews\&d0=country\&d0v=SE} \\
+\begin{verbatim}
+   http://localhost:8086/data/?t0=20090101
+   &t1=20090420&unit=pageviews
+   &d0=country&d0v=SE
+\end{verbatim}
 
@@
 The data server serves static files from the project's \texttt{clients}
-directory at the server URL appended with \texttt{client/filename}.
+directory at the server URL appended with \texttt{/client/filename}.
 

### 2009-05-11 java/darling/src/fm/last/darling/mapred/MapperWrapper.java
darling: commiting work-in-progress from 2009-04-30

@@ -18,6 +18,7 @@ import fm.last.darling.io.records.NSpacePoint;
 import fm.last.darling.nspace.Dimension;
 import fm.last.darling.nspace.Projectionist;
 import fm.last.darling.user.UserMapper;
+import fm.last.darling.utils.Util;
 
 // the class we wrap around the user's mapper.
 public class MapperWrapper implements Mapper<LongWritable, Text, NSpacePoint, IntWritable> {
@@ -48,7 +49,7 @@ public class MapperWrapper implements Mapper<LongWritable, Text, NSpacePoint, In
 		TreeMap<String,String> points = o.getDimensions();
 		
 		// fan out into projections,
-		ArrayList<ArrayList<Dimension>> rps = readRequestedProjections();
+		ArrayList<ArrayList<Dimension>> rps = Util.readRequestedProjections();
 		for (ArrayList<Dimension> requested : rps) {
 			// reduce down to the dimensions we are interested in. 
 			TreeMap<String,String> projection = Projectionist.dimensionality_reduction(requested, points);
@@ -58,31 +59,5 @@ public class MapperWrapper implements Mapper<LongWritable, Text, NSpacePoint, In
 				collector.collect(point, o.getMeasurement(unit));
 			}
         }
-	}
-
-	private ArrayList<ArrayList<Dimension>> readRequestedProjections() {
-		ArrayList<ArrayList<Dimension>> list = new ArrayList<ArrayList<Dimension>>();
-
-		ArrayList<Dimension> projection0 = new ArrayList<Dimension>();
-		projection0.add(new Dimension("country"));
-		list.add(projection0);
-
-		ArrayList<Dimension> projection1 = new ArrayList<Dimension>();
-		projection1.add(new Dimension("country"));
-		projection1.add(new Dimension("service"));
-		list.add(projection1);
-
-		return list;
-	}
-	
-	// TODO: not used for anything else than illustrative purposes. remove at will.
-	public static void main(String[] args) throws Exception {
-		System.out.println("yeah, that's right.");
-
-		// proof of concept.
-		ZohmgOutputCollector o = new ZohmgOutputCollector();
-		String userclass = "fm.last.darling.user.ExampleMapper";
-		UserMapper usermapper = (UserMapper) Class.forName(userclass).newInstance();
-		usermapper.map(200, "123423423	SE	web	510", o);
 	}
 }

### 2009-05-11 java/darling/src/fm/last/darling/utils/Util.java

@@ -1,9 +1,18 @@
 package fm.last.darling.utils;
 
+import java.io.File;
+import java.io.FileNotFoundException;
+import java.io.IOException;
+import java.util.ArrayList;
 import java.util.Calendar;
 import java.util.GregorianCalendar;
 
+import org.ho.yaml.Yaml;
+
+import fm.last.darling.nspace.Dimension;
+
 public class Util {
+	
 	public static String EpochtoYMD(long epoch) {
         Calendar thence = new GregorianCalendar();
         thence.setTimeInMillis(epoch*1000);
@@ -12,34 +21,25 @@ public class Util {
         int day = thence.get(Calendar.DAY_OF_MONTH);
         return String.format("%04d%02d%02d", year, month, day);
 	}
-
-	// TODO: this is not used no more. remove at will. 
-	// the following was lifted from http://forums.sun.com/thread.jspa?threadID=609364
-	/**
-	 * Returns a byte array containing the two's-complement representation of the integer.<br>
-	 * The byte array will be in big-endian byte-order with a fixes length of 4
-	 * (the least significant byte is in the 4th element).<br>
-	 * <br>
-	 * <b>Example:</b><br>
-	 * <code>intToByteArray(258)</code> will return { 0, 0, 1, 2 },<br>
-	 * <code>BigInteger.valueOf(258).toByteArray()</code> returns { 1, 2 }. 
-	 * @param integer The integer to be converted.
-	 * @return The byte array of length 4.
-	 */
-	public static byte[] intToByteArray (final int integer) {
-		int byteNum = (40 - Integer.numberOfLeadingZeros (integer < 0 ? ~integer : integer)) / 8;
-		byte[] byteArray = new byte[4];
-
-		for (int n = 0; n < byteNum; n++)
-			byteArray[3 - n] = (byte) (integer >>> (n * 8));
-
-		return (byteArray);
-	}
 	
-	public static void main(String[] args) {
-		System.out.println("yeah, utils.");
-		System.out.println(Util.EpochtoYMD(1240300000)); // => Tue Apr 21 08:46:40 +0100 2009
-		System.out.println(Util.EpochtoYMD(1210000000)); // => Mon May 05 16:06:40 +0100 2008
-		System.out.println(Util.EpochtoYMD(123000));     // => Fri Jan 02 11:10:00 +0100 1970
[…]

### 2009-05-11 java/darling/test/java/fm/last/darling/util/UtilTest.java

@@ -1,0 +1,43 @@
+package fm.last.darling.util;
+
+import java.io.File;
+import java.util.ArrayList;
+
+import org.junit.Test;
+import static org.junit.Assert.assertEquals;
+
+import fm.last.darling.nspace.Dimension;
+import fm.last.darling.utils.Util;
+
+public class UtilTest {
+
+	@Test
+	public void testEpochtoYMD() {
+		String expected = "20090421";
+		String ymd = Util.EpochtoYMD(1240300000);
+		assertEquals(ymd, expected);
+	}
+	
+	@Test
+	public void testreadRequestedProjections() {
+		File testDataFolder = new File("test/data"); 
+		File yaml = new File(testDataFolder, "dataset.yaml");
+		ArrayList<ArrayList<Dimension>> ret = Util.readRequestedProjections(yaml);
+		
+		ArrayList<ArrayList<Dimension>> expected = new ArrayList<ArrayList<Dimension>>();
+		ArrayList<Dimension> projection0 = new ArrayList<Dimension>();
+		projection0.add(new Dimension("country"));
+		expected.add(projection0);
+		ArrayList<Dimension> projection1 = new ArrayList<Dimension>();
+		projection1.add(new Dimension("country"));
+		projection1.add(new Dimension("service"));
+		expected.add(projection1);		
+		
+		
+		assertEquals(ret.size(), expected.size());
+		for (int i = 0; i < ret.size(); i++)
+			assertEquals(ret.get(i).toString(), expected.get(i).toString());
+	}
+	
+	
+}

### 2009-05-11 zohmg/config.py
updated TODO: config() is broken.

@@ -9,6 +9,8 @@ import os, re, sys, time
 
 # figure out if we are run inside dumbo.
 # files shipped to dumbo are all put in cwd.
+
+# FIXME: $HADOOP_HOME can and will be set outside of hadoop jobs.
 if "HADOOP_HOME" in os.environ:
     config_path = os.path.abspath("")
 else:

### 2009-05-11 zohmg/config.py
config.py: fixed formatting bug, left two others intact.

@@ -107,14 +107,16 @@ class Environ(object):
         self.read_environ()
 
     def get(self,key):
+        # FIXME: catch KeyError.
         return self.environ[key]
 
     def read_environ(self):
+        # FIXME: think harder about paths; is environment.py really where you think it is?
         sys.path.append(config_path) # add config path so we can import from it.
         try:
             import environment
         except ImportError:
-            msg = "[%s] Error: Could not import %senvironment.py. %s." % (time.asctime(),config_path)
+            msg = "[%s] Error: Could not import environment. %s." % (time.asctime(), config_path)
             fail(msg)
 
         for key in dir(environment):

### 2009-05-11 zohmg/middleware/__init__.py
zohmg/middleware/__init__.py: * added handling of 404 (no application for the path).

@@ -1,1 +1,53 @@
-#
+# used if no application is found (i.e. a 404).
+def not_found_hook(environ,start_response):
+    # user sees this message.
+    message = """<?xml version="1.0" encoding="utf-8"?>
+<!DOCTYPE html PUBLIC "-//W3C//DTD XHTML 1.0 Strict//EN"
+        "http://www.w3.org/TR/xhtml1/DTD/xhtml1-strict.dtd">
+<html xmlns="http://www.w3.org/1999/xhtml" xml:lang="en" lang="en">
+
+<head>
+    <title>Zohmg Error</title>
+    <meta http-equiv="content-type" content="text/html;charset=utf-8" />
+    <meta http-equiv="Content-Type" content="text/html" />
+</head>
+
+<body>
+
+    <h1>404 Not Found</h1>
+
+    <p>
+    There is no application for the path
+    <blockquote>
+    %s
+    </blockquote>
+    </p>
+
+    <p>
+    Available applications
+    <ul>
+        <li>
+            <strong>/client/filename</strong> - serves static files from
+            the project's clients directory.
+        </li>
+        <li>
+            <strong>/data/?query</strong> - serves aggregates from HBase.
+        </li>
+        <li>
+            <strong>/transform/filename/?query</strong> - serves
+            aggregates transformed with a transformer from the project's
+            transformers directory.
+        </li>
+    </ul>
+    </p>
+
+    <p>
+    Zohmg.
+    </p>
+
+</body>
+</html>""" % environ["PATH_INFO"]
+
+    # serve message.
+    start_response("404 Not Found", [("content-type","text/html")])
+    return message

### 2009-05-11 zohmg/mapper.py
fixed order of dimensions in column-family.

@@ -10,11 +10,12 @@ class Mapper(object):
         # from the usermapper: a timestamp, a point in n-space, units and their values.
         for (ts, point, units) in self.usermapper(key, value):
             for pjs in self.projections.values():
-                # we perform dimensionality reduction,
+                # perform dimensionality reduction,
                 reduced = {}
                 for d in pjs:
                     reduced[d] = point[d]
-                # yielding for each requested projection.
+                # yield for each requested projection.
+                # include the projections list to get the order right.
                 for u in units:
                     value = units[u]
-                    yield (ts, reduced, u), value
+                    yield (ts, pjs, reduced, u), value

### 2009-05-11 zohmg/reducer.py

@@ -8,13 +8,20 @@ class Reducer(object):
 
     def __call__(self, key, values):
         #self.reduces += 1
-        ts, dims, unit = key
+        ts, ps, dims, unit = key
         value = sum(values)
 
         # rowkey: "unit-ymd".
         rk = '-'.join([unit, str(ts)])
-        cf = '-'.join(dims.keys())
-        q  = '-'.join(dims.values())
+
+        # construct column-family and qualifier strings.
+        cflist = []
+        qlist  = []
+        for p in ps:
+            cflist.append(p)
+            qlist.append(dims[p])
+        cf = '-'.join(cflist)
+        q  = '-'.join(qlist)
 
         # remember, we'll pass the output of this reducer to HBaseOutputReader.
         yield rk, json.dumps({cf+":"+q : {'value':value}})

### 2009-05-11 zohmg/reducer.py
zohmg/reducer.py: one comment in, two out.

@@ -3,11 +3,9 @@ import simplejson as json
 
 class Reducer(object):
     def __init__(self):
-        #self.reduces = self.counters['reduces']
         self.config = Config()
 
     def __call__(self, key, values):
-        #self.reduces += 1
         ts, ps, dims, unit = key
         value = sum(values)
 
@@ -15,6 +13,7 @@ class Reducer(object):
         rk = '-'.join([unit, str(ts)])
 
         # construct column-family and qualifier strings.
+        # it's important that we get the ordering right.
         cflist = []
         qlist  = []
         for p in ps:

### 2009-05-11 zohmg/config.py
zohmg/config.py: disregards HADOOP_HOME, tries to read config from both config and base dir.

@@ -23,7 +23,7 @@ class Config(object):
         if config_file:
             self.config_file = config_file
         else:
-            self.config_file = config_path + "/dataset.yaml"
+            self.config_file = "dataset.yaml"
 
         self.config = {}
         self.__read_config()
@@ -31,16 +31,25 @@ class Config(object):
 
     def __read_config(self):
         import yaml
-        try:
-            f = open(self.config_file, "r")
-            self.config = yaml.load(f)
-            f.close()
-        except IOError, ioe:
-            msg = "[%s] Error: Could not read %s. %s." % (time.asctime(),self.config_file, ioe.strerror)
-            fail(msg, ioe.errno)
+        possible_configs = [self.config_file, "config/"+self.config_file]
+        config_loaded = False
+        for config_file in possible_configs:
+            if config_loaded:
+                continue
+            try:
+                f = open(config_file, "r")
+                self.config = yaml.load(f)
+                f.close()
+                config_loaded = True
+            except:
+                pass
+
+        if not config_loaded:
+            msg = "[%s] Error: Could not read configuration." % time.asctime()
+            fail(msg)
 
         if not self.sanity_check():
-            msg = "[%s] Error: Could not parse %s." % (time.asctime(),self.config_file)
+            msg = "[%s] Error: Could not parse configuration." % time.asctime()
             fail(msg)
 
         return self.config

### 2009-05-11 zohmg/setup.py
cosmetics in table creation.

@@ -12,11 +12,9 @@ class Setup(object):
             projection = '-'.join(c.config['projections'][p])
             cfs.append(projection)
 
-        print "creating table:"
-        print "  * " + dataset
+        print "creating table '%s'" % dataset
         print " column families:"
-        print "".join((map( lambda cf: "  * "+str(cf)+"\n" , cfs)))
-        print
+        print "".join((map( lambda cf: "  + "+str(cf)+":\n" , cfs)))
 
         try:
             c = setup_transport("localhost")

### 2009-05-11 zohmg/utils.py

@@ -25,7 +25,7 @@ def setup_transport(host):
 
 
 def create_or_bust(c, t, cfs=['fam']):
-    print "creating table %s with %s or so cfs" % (t, len(cfs))
+    print "creating table %s with %s column-families." % (t, len(cfs))
     try:
         cds = []
         for cf in cfs:
@@ -42,6 +42,7 @@ def create_or_bust(c, t, cfs=['fam']):
         print e
         print "create_or_bust => bust"
         exit(3)
+    print "ok."
 
 
 def random_string(size):

### 2009-05-11 zohmg/config.py
fixed Environ.

@@ -1,22 +1,6 @@
 from zohmg.utils import fail
 import os, re, sys, time
 
-# TODO: is it a safe assumption that HADOOP_HOME is in os.environ when
-# running dumbo?
-# ALSO: is it safe to assume that HADOOP_HOME is *not* defined when
-# running locally?
-# TODO: why are we doing this in a global variable?
-
-# figure out if we are run inside dumbo.
-# files shipped to dumbo are all put in cwd.
-
-# FIXME: $HADOOP_HOME can and will be set outside of hadoop jobs.
-if "HADOOP_HOME" in os.environ:
-    config_path = os.path.abspath("")
-else:
-    config_path = os.path.abspath("config/")
-
-
 # TODO: multiple dataset files
 class Config(object):
     def __init__(self, config_file=None):
@@ -115,17 +99,21 @@ class Environ(object):
         self.environ = {}
         self.read_environ()
 
-    def get(self,key):
-        # FIXME: catch KeyError.
-        return self.environ[key]
+    def get(self, key):
+        try:
+            return self.environ[key]
+        except:
+            return ''
 
     def read_environ(self):
-        # FIXME: think harder about paths; is environment.py really where you think it is?
-        sys.path.append(config_path) # add config path so we can import from it.
+        # add config path so we can import from it.
+        sys.path.append(".")
+        sys.path.append("config")
+
         try:
             import environment
         except ImportError:
-            msg = "[%s] Error: Could not import environment. %s." % (time.asctime(), config_path)
+            msg = "[%s] Error: Could not import environment.py" % time.asctime()
             fail(msg)
 
         for key in dir(environment):

### 2009-05-11 zohmg/utils.py
moved import into function body; it triggered import error.

@@ -1,6 +1,5 @@
 import string, sys, time
 from random import Random
-from hbase.ttypes import *
 
 # TODO: move all hbase utils to it's own module.
 
@@ -25,7 +24,8 @@ def setup_transport(host):
 
 
 def create_or_bust(c, t, cfs=['fam']):
-    print "creating table %s with %s column-families." % (t, len(cfs))
+    from hbase.ttypes import *
+    print "creating hbase table %s." % t
     try:
         cds = []
         for cf in cfs:

### 2009-05-11 zohmg/process.py
zohmg import: added support for '--lzo'

@@ -21,9 +21,13 @@ class Process(object):
                 ('file','lib/usermapper.py') # TODO: handle this more betterer.
                ]
 
+        # check for '--lzo' as first extra argument.
+        if len(for_dumbo) > 0 and for_dumbo[0] == '--lzo':
+            opts.append(('inputformat', 'org.apache.hadoop.mapred.LzoTextInputFormat'))
+            for_dumbo.pop(0) # remove '--lzo'.
+
         # read environment and attach.
         env = Environ()
-
         hadoop_home = env.get("HADOOP_HOME")
         if not os.path.isdir(hadoop_home):
             msg = "Error: HADOOP_HOME not set in config/environment.py."
@@ -44,7 +48,6 @@ class Process(object):
             fail(msg)
 
         # pull everything in config and lib.
-        # TODO: why everything in config?
         file_opts = self.__add_files(["config","lib"])
         opts.extend(file_opts)
 

### 2009-05-11 doc/report/msc-report.tex
doc/report/msc-report.tex: only one quote in Implementation section.

@@
 
-The classic trade-off between time and space applys to wether or not
+The classic trade-off between time and space applies to whether or not
 pre-compute an aggregate. Pre-computing to many aggregates might render an
-unfeasible of data amount to store, while pre-computing to few aggregates
+unfeasible amount of data to store, while pre-computing to few aggregates
 renders longer access times for aggregates which are not pre-computed. This
@@
 
-%\chapterquote{A good idea is about ten percent and implementation and hard
-%work, and luck is 90 percent.}{Guy Kawasaki}
+\chapterquote{A good idea is about ten percent and implementation and hard
+work, and luck is 90 percent.}{Guy Kawasaki}
 
-\chapterquote{It is not always what we know or analyzed before we make a
-decision that makes it a great decision. It is what we do after we make the
-decision to implement and execute it that makes it a good
-decision.}{William Pollard}
+%\chapterquote{It is not always what we know or analyzed before we make a
+%decision that makes it a great decision. It is what we do after we make the
+%decision to implement and execute it that makes it a good
+%decision.}{William Pollard}
 
@@
 
-The reducer, then, is very simple: it's a simple sum of the values. The output
+The reducer, then, is very simple: it is a simple sum of the values. The output
 of the reducer will not be stored on disk as is usually the case, but will
@@
 \chapterquote{Example is not the main thing in influencing others, it's
-the only thing}{Albert Schweitzer}
+the only thing.}{Albert Schweitzer}
 

### 2009-05-12 doc/report/msc-report.tex
doc/report/msc-report.tex: small notes for metadata.
doc/report/msc-report.tex: intermediate push for introduction.

@@
 
-\subsection{Scope}
+Last.fm
+
+SQL
+
+EXPLODE
+
+\section{Scope}
 
@@
 this restriction does not cripple the system much---if at all.
+
+\subsubsection*{Related work}
+
+A data warehouse is a repository of an organization's
+electronically stored data. Data warehouses are designed to facilitate
+reporting and analysis. \cite{wikipedia}
+
+The core actions of a data warehouse is to load, transform and extract data.
+The classic thinking in modelling data warehouses is to divide data into facts
+and dimensons. The facts are measures categorized by dimensions.
+
+In OLAP\footnote{On-Line Analytical Processing}, which is a kind of data
+warehouse, the data is thought of as being structured as a cube. The cube is
+typically modelled in a realational database with fact tables and dimension
+tables, upon which the system performs joins to extract data.
+\cite{olap_solutions}
+
+HDW is system which goal is to create a high performance large scale data
+warehouse based on MapReduce and Bigtable. \cite{hdw}
+
+Omniture is used at Last.fm etc.
+
+\section{Goal}
+
+The goal (sv. syfte) with our work.
 
@@
 
-\chapter{Theory}
+\chapter{Analysis}
 
-\chapterquote{Einstein explained his theory to me every day, and on my
-arrival I was fully convinced that he understood it.}{Chaim Weizmann}
+\chapterquote{Arrange whatever pieces come your way.}{Virgina Wolf}
 
-\section{Other data warehouses}
+\section{Task}
 
-A data warehouse is a repository of an organization's
-electronically stored data. Data warehouses are designed to facilitate
-reporting and analysis. \cite{wikipedia}
+\subsection{Import}
 
-The core actions of a data warehouse is to load, transform and extract data.
-The classic thinking in modelling data warehouses is to divide data into facts
-and dimensons. The facts are measures categorized by dimensions.
+\subsection{Store}
 
-In OLAP\footnote{On-Line Analytical Processing}, which is a kind of data
-warehouse, the data is thought of as being structured as a cube. The cube is
-typically modelled in a realational database with fact tables and dimension
-tables, upon which the system performs joins to extract data.
-\cite{olap_solutions}
+\subsection{Export}
+
+\section{Interface}
 
@@
 
-\chapter{Tools}
-
-\chapterquote{Arrange whatever pieces come your way.}{Virgina Wolf}
-
-%\chapterquote{In short, intelligence, considered in what seems to be its
-%original feature, is the faculty of manufacturing artificial objects,
-%especially tools to make tools, and of indefinitely urging the
-%manufacture.}{Henri Bergson}
+\section{System}
 
@@
 
-% TODO> I argue that we should have 'hadoop' and 'hbase' as the main sections,
-% and let mapreduce be a subsection of hadooop, bigtable a dito of hbase.
+\chapter{Method}
 
-\section{MapReduce}
+\section{Apache Hadoop}
+
+Hadoop is a free software implementation of Google's MapReduce-framework
+described in \cite{mapreduce}. Hadoop is written in Java and is part of
+the Apache Software Foundation.  It is a fairly mature piece of software
+and is used in production at numerous companies. \cite{hadoop}
+
+MapReduce is a framework for running jobs in parallel across multiple
+machines. As the name implies, MapReduce is made up of two parts or
+phases: map and reduce. The map phase consumes one piece of input at a
+time (a piece of input can be a line from a log file, for example), and
+emits an intermediate key and value. These keys and values are then
+processed by the reduce phase, which emits the final key-value pairs.
+
+MapReduce works by splitting the input, which commonly is in the size
+range of many gigabytes, into n parts and lets each node of the computing
+cluster work on one or more parts. The mapper on each node emits key-value
+pairs which are fed into the reduce phase. The final output is collated,
+usually on a distributed file system.
+
+\subsection{MapReduce}
 
@@
 performs computations on each row of the data. The mapper then outputs each
-resulting computation on the format \texttt{key,value}. These key-value
-pairs are received by a \textit{reducer} which performs the second stage of
-the computation.
+resulting computation on the format $\langle$\textit{key,value}$\rangle$. These
+key-value pairs are received by a \textit{reducer} which performs the second
+stage of the computation.
+All Zohmg import programs are executed on Hadoop.
 
-\section{BigTable}
+\subsection{Hadoop Distributed File System}
+
+HDFS is a free software implementation of the GFS \cite{gfs}.
+
+The goal of HDFS is to store large data sets while being able to handle
+hardware failure and being deployed in heterogenous hardware and software
+eco systems. Additionally the Hadoop and HDFS interaction is designed with
+the notion that it is cheaper to move the computation, MapReduce programs,
+to the data than the other way around.
+
+Every Hadoop node can be formatted with HDFS, this reserves disk space on
+the node to be used for HDFS. The HDFS container resides on the nodes
+general purpose file system. It is also possible to use HDFS as a
+stand-alone general purpose distributed file system, without Hadoop.
+
+The default block size on HDFS 64 MB, significantly larger than
+file systems used for hard drives. The motivation is that applications
+that use HDFS are not general purpose applications which run on general
+purpose file systems, but batch applications which read, write, or both,
+large amounts of data.
+
+\subsection{Hadoop streaming}
+
+Any program that reads and outputs to the file pointers \texttt{stdin} and
+\texttt{stdout}, respectively, can be used as MapReduce programs with
+Hadoop Streaming. This makes it possible to use any shell script or
+program which inputs and outputs this way as a MapReduce program.
+
+\subsection{Dumbo}
+
+Dumbo is a Python framework for writing Hadoop programs in languages other than
+Java, in a so-called Streaming Mode. It is used extensively at Last.fm for
+writing short prototypes. \cite{dumbo}
+
+All Zohmg programs written in Python use Dumbo to execute on Hadoop.
+
+\section{Apache HBase}
+
+HBase is a free software implementation of Google's BigTable data store
+described in \cite{bigtable}. It is a sub-project of Hadoop, written in
+Java and a part of the Apache Software Foundation. HBase is still in its
+infancy and few organizations use it for mission-critical applications.
+\cite{hbase}
+
+BigTable is a data store designed for traversal of very large data sets.
+Time-stamped cells inside column-familes in rows. The row has a key and
+any number of cells attached to it. The rows in a table are sorted by the
+row key.  The typical use-case of BigTable is to scan a range of rows. It
+is therefor important to set up one's keys so that related data is close
+to each other. For example, if it makes sense to traverse the data in
+chronological order, the key might contain a representation of the data's
+timestamp. The classic example is to have the reversed domain name (i.e.
+com.google.www) as the row key, which means that all sub-domains of any
+domain are next to each other.
+
+Sparseness is a big thing. Null values are free of charge. TODO: *Describe
+how we leverage sparsity*.
+
+TODO: *this should go into implementation.*
+The typical use case for Zohmg is to read one or more column-qualifiers,
+which correspond to one or more points in n-space, over a range of time.
+Therefor, the keys are made up of the time and unit of the measurement,
+and the qualifier queried for represents the point in n-space. More on
+this in the data model section.
+
+All Zohmg data is stored in HBase.
+
+\subsection{BigTable}
 
@@
 
-\subsubsection{Rows}
+\subsubsection*{Rows}
 
@@
 
-\subsubsection{Regions}
+\subsubsection*{Regions}
 
@@
 % Disambiguation needed. -fredrik
-\subsection{Dimensions}
+\subsection*{Dimensions}
 
-\subsubsection{Column-families}
+The degrees of freedom in BigTable.
+
+\subsubsection*{Column-families}
 
@@
 
-\subsubsection{Column-qualifiers}
+\subsubsection*{Column-qualifiers}
 
@@
 
-\subsection{Hierarchies}
+\subsection*{Hierarchies}
 
@@
 used.
-
-\section{Apache Hadoop}
-
-Hadoop is a free software implementation of Google's MapReduce-framework
-described in \cite{mapreduce}. Hadoop is written in Java and is part of
-the Apache Software Foundation.  It is a fairly mature piece of software
-and is used in production at numerous companies. \cite{hadoop}
-
-MapReduce is a framework for running jobs in parallel across multiple
-machines. As the name implies, MapReduce is made up of two parts or
-phases: map and reduce. The map phase consumes one piece of input at a
-time (a piece of input can be a line from a log file, for example), and
-emits an intermediate key and value. These keys and values are then
-processed by the reduce phase, which emits the final key-value pairs.
-
-MapReduce works by splitting the input, which commonly is in the size
-range of many gigabytes, into n parts and lets each node of the computing
-cluster work on one or more parts. The mapper on each node emits key-value
-pairs which are fed into the reduce phase. The final output is collated,
-usually on a distributed file system.
-
-All Zohmg import programs are executed on Hadoop.
-
-\subsection{Hadoop Distributed File System}
-
-HDFS is a free software implementation of the GFS \cite{gfs}.
-
-The goal of HDFS is to store large data sets while being able to handle
-hardware failure and being deployed in heterogenous hardware and software
-eco systems. Additionally the Hadoop and HDFS interaction is designed with
-the notion that it is cheaper to move the computation, MapReduce programs,
-to the data than the other way around.
-
-Every Hadoop node can be formatted with HDFS, this reserves disk space on
-the node to be used for HDFS. The HDFS container resides on the nodes
-general purpose file system. It is also possible to use HDFS as a
-stand-alone general purpose distributed file system, without Hadoop.
-
-The default block size on HDFS 64 MB, significantly larger than
-file systems used for hard drives. The motivation is that applications
-that use HDFS are not general purpose applications which run on general
-purpose file systems, but batch applications which read, write, or both,
-large amounts of data.
-
-\subsection{Hadoop streaming}
-
-Any program that reads and outputs to the file pointers \texttt{stdin} and
-\texttt{stdout}, respectively, can be used as MapReduce programs with
-Hadoop Streaming. This makes it possible to use any shell script or
-program which inputs and outputs this way as a MapReduce program.
-
-\section{Apache HBase}
-
-HBase is a free software implementation of Google's BigTable data store
-described in \cite{bigtable}. It is a sub-project of Hadoop, written in
-Java and a part of the Apache Software Foundation. HBase is still in its
-infancy and few organizations use it for mission-critical applications.
-\cite{hbase}
-
-BigTable is a data store designed for traversal of very large data sets.
-Time-stamped cells inside column-familes in rows. The row has a key and
-any number of cells attached to it. The rows in a table are sorted by the
-row key.  The typical use-case of BigTable is to scan a range of rows. It
-is therefor important to set up one's keys so that related data is close
-to each other. For example, if it makes sense to traverse the data in
-chronological order, the key might contain a representation of the data's
-timestamp. The classic example is to have the reversed domain name (i.e.
-com.google.www) as the row key, which means that all sub-domains of any
-domain are next to each other.
-
-Sparseness is a big thing. Null values are free of charge. TODO: *Describe
-how we leverage sparsity*.
-
-TODO: *this should go into implementation.*
-The typical use case for Zohmg is to read one or more column-qualifiers,
-which correspond to one or more points in n-space, over a range of time.
-Therefor, the keys are made up of the time and unit of the measurement,
-and the qualifier queried for represents the point in n-space. More on
-this in the data model section.
-
-All Zohmg data is stored in HBase.
-
 \subsection{Thrift}
@@
 than Java---for instance Python.
-
-\section{Dumbo}
-
-Dumbo is a framework for writing Hadoop programs in languages other than
-Java, in a so-called Streaming Mode. Dumbo is written in Python. It is used
-extensively at Last.fm for writing short prototypes. \cite{dumbo}
-
-All Zohmg programs written in Python use Dumbo to execute on Hadoop.
-
-\section{Serializing}
-
-Serializing data is the task of converting an object into a sequence of
-bits so that it can be stored on a storage medium or transmitted over a
-network link. Rereading the sequence of bits should restore the original
-state of the object it was in before the serialization.
 
@@
 
-\section{Python}
-
-Python is a general-purpose high-level multi-paradigm---amongst others:
-object-oriented, imperative and functional---scripting language.
-\cite{python}
-
-Zohmg is written mostly in Python. Zohmg uses the following Python modules
-internally; it depends on them to be available on the system.
-
-\subsection{Paste}
-
-Python Paste is a framework for building WSGI\footnote{Web Server Gateway
-Interface, a HTTP extension where a CGI-like environment is passed around.}
-applications and middleware. \cite{definitive_guide_to_pylons}
-
-WSGI applications receive a WSGI request and returns a response with the
-built-in web server.
-
-Middleware is software that acts as an intermediary. WSGI middleware
-receives a WSGI request and then performs logic upon this request, before
-passing the request to a WSGI application. \cite{paste}
-
 \subsection{PyYAML}
@@
 
+\section{Configuration}
+
+Data sets and environment.
+
+\section{Python}
+
+Python is a general-purpose high-level multi-paradigm---amongst others:
+object-oriented, imperative and functional---scripting language.
+\cite{python}
+
+Zohmg is written mostly in Python. Zohmg uses the following Python modules
+internally; it depends on them to be available on the system.
+
+\section{Exporting data}
+
+TODO: More general about middleware, move Paste mentioning to Implementation
+chapter.
+
+Python Paste is a framework for building WSGI\footnote{Web Server Gateway
+Interface, a HTTP extension where a CGI-like environment is passed around.}
+applications and middleware. \cite{definitive_guide_to_pylons}
+
+WSGI applications receive a WSGI request and returns a response with the
+built-in web server.
+
+Middleware is software that acts as an intermediary. WSGI middleware
+receives a WSGI request and then performs logic upon this request, before
+passing the request to a WSGI application. \cite{paste}
+
 \chapter{Implementation}
@@
 work, and luck is 90 percent.}{Guy Kawasaki}
-
-%\chapterquote{It is not always what we know or analyzed before we make a
-%decision that makes it a great decision. It is what we do after we make the
-%decision to implement and execute it that makes it a good
-%decision.}{William Pollard}
 
@@
 
- yield log.timestamp, dimensions, measurements
+    yield log.timestamp,dimensions,measurements
 \end{verbatim}
@@
 
+Typed bytes
+
 \subsubsection{Input formats}
+
+TextInputFormat
+
+LzoInputFormat
 
@@
 
-\chapter{Usage}
+\chapter{Results}
 
-\chapterquote{There comes a time when all people break down and do the
-unthinkable: They read the manual.}{Unknown}
+\chapterquote{When we can't dream any longer we die.}{Emma Goldman}
 
-This is the user manual.
-
-\section{Deployment}
-
-A manage script is used to create tables from a pre-defined schema.
-
-\begin{verbatim}
-   zohmg create project
-\end{verbatim}
-
-\noindent Edit the configuration files, which are text files in YAML and
-Python format.
-
-\begin{verbatim}
-   zohmg setup
-\end{verbatim}
-
-\noindent The above command executes the configuration made and creates the
-infrastructure (HBase tables) for the project.
-
-\section{Import}
-
-The user writes the map part of the MapReduce programs which imports
-log-like data to the data store. The reduce step is taken care of by Zohmg,
-which uses sum reducer that simply sums up values with the same keys.
-
-\subsection{User mappers}
-
-Import jobs use mappers written by the user, either in Python or Java.
-Mappers written by the user import each line from the input file and set
-the arguments to the map function, key and value, as line number and line
-contents, respectively. The user can then perform computations on these
-entities and finally yield the results as a dictionary. The resulting
-dictionary will be interpreted by Zohmg and stored accordingly.
-
-\begin{verbatim}
-   zohmg import mappers/somemapper.py hdfs/path
-\end{verbatim}
-
-\section{Storing data}
-
-% COMMENT: I don't think we should mention the underlying store in the usage section.
-Data is stored in Apache HBase. A natural consequence of this that the
-data is stored sorted by rowkey allowing fast access to an arbitrary
-(sequential) range.
-
-\section{Export}
-
-It is possible to query Zohmg about metadata it has on each of the imported
-data sets.
-
-Data is served to the user either over HTTP in JSON format or over Thrift.
-
-\begin{verbatim}
-   zohmg serve
-\end{verbatim}
-
-\noindent The above serves on default port 8086, with the \texttt{--port}
-argument it is possible to change server port.
-
-The data server have three methods of serving data: raw, transformed, or
-static. The raw and transformed data are aggregates served from HBase,
-transformed in the latter case, while the static method just serve static
-files from the project's \texttt{clients} directory. Aggregates served from
-HBase are returned to the user as JSONP.
-
-\subsection{Data}
-
-Based on the query---which the user submits---the corresponding data is
-served from HBase as JSONP. The query is a simple definition of wanted row
-range, dimensions, and values for these dimensions. The query is submitted
-to the data server via HTTP, an example query could look like (the URL is
-wrapped to fit the page)
-
-\begin{verbatim}
-   http://localhost:8086/data/?t0=20090101
-   &t1=20090420&unit=pageviews
-   &d0=country&d0v=SE
-\end{verbatim}
-
-The above query would return the pageviews for SE between 1 January 2009
-and 4 April 2009.
-
-\subsection{Transformers}
-
-Fetches data from HBase and transforms it with the requested transform
-program from the project's \texttt{transformers} directory, output is
-dumped as JSONP.
-
-\subsection{Clients}
-
-It is possible to build static web pages out of HTML and JavaScript, which
-can interact with the data server. These static pages can for instance
-present some sort of query building mechanism---for instance with drop down
-boxes---and then send this query to the data server and graph the resulting
-data.
-
-The data server serves static files from the project's \texttt{clients}
-directory at the server URL appended with \texttt{/client/filename}.
-
-\chapter{Examples}
-
-\chapterquote{Example is not the main thing in influencing others, it's
-the only thing.}{Albert Schweitzer}
-
-%\chapterquote{There is nothing so annoying as a good example!!}{Mark Twain}
-
-%\chapterquote{Example isn't another way to teach, it is the only way to
-%teach}{Albert Einstein}
-
-\section{Web log analysis}
-
-Configure the dataset.yaml.
-
-Write a custom mapper, which imports data from web logs.
-
-Start data server and extract data from it.
-
-Possibly write a transformer which performs some transformation on the
-aggregated data.
-
-Possibly write a client which will render a graph.
-
-\chapter{Discussion}
-
-\chapterquote{In discussion comes the light!}{Unknown}
-
-\section{Related work}
-
-HDW is system which goal is to create a high performance large scale data
-warehouse based on MapReduce and Bigtable. \cite{hdw}
+\section{Discussion}
 
@@
 user's map.
+
 \subsection{Multiple data sets}
@@
 \subsection{Metadata}
+
+data set introspection
+
+roll-up and drill-down
+
+good when creating apps
 
@@
 
-\chapter{Conclusion}
+\section{Conclusion}
 
-\chapterquote{When we can't dream any longer we die.}{Emma Goldman}
-
-Last.fm is looking at using Zohmg for multiple application areas, such as
-radio stats and web page stats.
+Last.fm is investigating the possibility for using Zohmg for multiple
+application areas, such as radio stats and web page stats.
 
@@
 
+\appendix
+
+\chapter{Usage}
+
+\chapterquote{There comes a time when all people break down and do the
+unthinkable: They read the manual.}{Unknown}
+
+\noindent The following chapter will describe how the Zohmg system is used.
+Project deployment and data import and export will be described briefly.
+
+\section{Deployment}
+
+The Zohmg manage script is used to create a project directory
+
+\begin{verbatim}
+   zohmg create project
+\end{verbatim}
+
+\noindent The next step is editing the configuration files, which are located in
+the project's \texttt{config} directory. The configuration files are text files
+in YAML and Python format. When the files are edited, run the setup command;
+this executes the configuration and creates infrastructure (HBase tables) for
+the project.
+
+\begin{verbatim}
+   zohmg setup
+\end{verbatim}
+
+\section{Import}
+
+In order to import log-like data to the data store, the map part of the
+MapReduce program needs to be created. The reduce step is taken care of by
+Zohmg, which uses sum reducer that simply sums up values with the same keys.
+
+\subsection*{User mappers}
+
+Import jobs use mappers written either in Python or Java. User mappers import
+each line from the input file and set the arguments to the map function, key and
+value, as line number and line contents, respectively. The user can then perform
+computations on these entities and finally yield the results as a dictionary.
+The resulting dictionary will be interpreted by Zohmg and stored accordingly.
+
+\begin{verbatim}
+   zohmg import mappers/somemapper.py hdfs/path
+\end{verbatim}
+
+\section{Export}
+
+It is possible to query Zohmg about metadata it has on each of the imported
+data sets.
+
+Data is served to the user over HTTP in JSONP format or. To start the data
+server execute
+
+\begin{verbatim}
+   zohmg serve
+\end{verbatim}
+
+The command above starts the data server on default port 8086, with the
+\texttt{--port} argument it is possible to change server port.
+
+The data server have three methods of serving data: raw, transformed, or
+static. The raw and transformed data are aggregates served from HBase,
+transformed in the latter case, while the static method just serve static
+files from the project's \texttt{clients} directory. Aggregates served from
+HBase are returned to the user as JSONP.
+
+\subsection*{Data}
+
+Based on the query---which the user submits---the corresponding data is
+served from HBase as JSONP. The query is a simple definition of wanted row
+range, dimensions, projections, or both, and values for these. The query is
+submitted to the data server via HTTP, a query could look like
+
+\begin{verbatim}
+   http://localhost:8086/data/?t0=20090101
+   &t1=20090420&unit=pageviews
+   &d0=country&d0v=SE
+\end{verbatim}
+
+\noindent The above query would return the pageviews for SE between 1 January
+2009 and 4 April 2009.
+
+\subsection*{Transformers}
+
+Fetches data from HBase and transforms it with the requested transform
+program from the project's \texttt{transformers} directory, output is
+dumped as JSONP.
+
+Accessing transformed data is done similarly to the raw data extraction.
+
+\begin{verbatim}
+   http://localhost:8086/transform/transformer/
+   ?t0=20090101&t1=20090420&unit=pageviews
+   &d0=country&d0v=SE
+\end{verbatim}
+
+\noindent The above query extracts the same data as the query from the Data
+section. This data is run through the \texttt{transformer.py} application and
+is, again, served to the user as JSONP.
+
+\subsection*{Clients}
+
+It is possible to build static web pages out of HTML and JavaScript, which
+can interact with the data server. These static pages can for instance
+present some sort of query building mechanism---for instance with drop down
+boxes---and then send this query to the data server and graph the resulting
+data.
+
+The data server serves static files from the project's \texttt{clients} from
+
+\begin{verbatim}
+   http://localhost:8086/clients/filename
+\end{verbatim}
+
+\chapter{Glossary}
+
+\begin{itemize}
+\item \texttt{Serializing}
+
+Serializing data is the task of converting an object into a sequence of
+bits so that it can be stored on a storage medium or transmitted over a
+network link. Rereading the sequence of bits should restore the original
+state of the object it was in before the serialization.
+
 \end{document}

### 2009-05-12 doc/report/msc-report.tex
doc/report/msc-report.tex: added end of itemize in glossary.

@@
 \item \texttt{Serializing}
-
 Serializing data is the task of converting an object into a sequence of
@@
 state of the object it was in before the serialization.
+\end{itemize}
 

### 2009-05-12 doc/report/msc-report.tex
Introduction: background.

@@
 
-Any organization will eventually wish to evaluate and analyze various
-characteristics of the service and operations they manage. These services can
-be anything from web sites to theme parks, and the evaluation might include
-characteristics such as the number of visitors per day, and the average time a
-theme park guest spends queueing for a particular ride.
+Any organization will eventually wish to analyze various characteristics of
+its service and operations. These services can be anything from web sites to
+theme parks. The analysis might include characteristics such as the number of
+visitors per day, or the average time a theme park guest spends queueing for a
+particular ride.
 
-Inevitably, these services and operations will generate a log for later
+Inevitably, the services and operations will generate a log for later
 analysis. As the services grow in size, so will the data sets they generate.
@@
 An important property of these data sets is that they generally have a regular
-structure and are represented as a number of independent records. The records
-can be broken up without losing any of their meaning and are as such very well
-suited for parallel analysis. The data sets are not fit to store using
-traditional database techniques, if only for the fact that they are too large
-to fit in a relational database.
+structure and are broken down into records that are independent of each other.
+The records can be analyzed one at a time without losing any of their meaning
+and are as such very well suited for parallel analysis.
 
-Each log record typically contains quite a bit of information. For example, a
-log of visits to a web site might contain a timestamp, information about what
-country the visiting user is from, and a string of text that identifies the
-web browser that was used. From these base facts one can collect information
-about what countries generate the most traffic to the site, and which web
-browser is most popular in a certain country, for example.
+Each log record typically contains a number of fields. For example, a log of
+visits to a web site might contain a timestamp, information about what country
+the visitor is from and what web browser she used. The values of the fields
+can be thought of as points along an axis; Sweden, Norway and Denmark are all
+points along the country axis, for instance. This is an important concept and
+it will be touched upon repeatedly throughout the text.
 
-These bits of information can be thought of as points along an axis; Sweden,
-Norway and Denmark are all points along the country axis, for instance. This
-is an important concept and it will be used throughout the text.
+After having collected these base facts, one can draw conclusions about what
+countries generate the most traffic to the site, and which web browser is most
+popular in a certain country, and so on.
 
 Typically, the size of a data set is in the range of several gigabytes of data
-per day. A single computer is not likely to be able to process that amount of
-data before a new batch arrives, and for very large data sets the single
-computer would simply not be able to fit the intermediate data set in its
-memory at all. Some data sets are many times larger than that and do not even
-fit inside a single computers storage devices.
+per day. Some data sets are many times larger.
+
+A single computer is not likely to be able to process that amount of data
+before a new batch arrives, and for very large data sets a single computer
+would simply not be able to fit the intermediate data set in its working
+memory.
 
 It is suitable to store the data across many storage nodes in what is called a
-distributed file system and to to exploit the parallel nature of the records by
-breaking up the data sets in pieces and have each piece be processed
+distributed file system and to to exploit the parallel nature of the records
+by breaking up the data sets in pieces and have each piece be processed
 separately on a node in a large compute cluster.
 
-% TODO: fixa cites.
-Luckily, this is a problem that has already been solved. Google has published
-papers on the MapReduce framework \cite{mapreduce}, which allows for parallel
-processing of large data sets, and The Google File System \cite{gfs}, a
-scalable, distributed file system. The Apache Hadoop \cite{hadoop} project has
-published free software implementations of a MapReduce framework and a
-distributed file system, both of which build on the ideas in the papers
-published by Google.
+% TODO: fixa cites. Luckily, this is a problem that has already been solved.
+Google has published papers on the MapReduce framework \cite{mapreduce}, which
+allows for parallel processing of large data sets, and The Google File System
+\cite{gfs}, a scalable, distributed file system. The Apache Hadoop
+\cite{hadoop} project has published free software implementations of a
+MapReduce framework and a distributed file system, both of which build on the
+ideas in the papers published by Google.
 

### 2009-05-12 doc/report/msc-report.tex
Introduction: added Problem description, Goal and Scope.

@@
 
+\section{Background}
+
 Any organization will eventually wish to analyze various characteristics of
@@
 
-% TODO: fixa cites. Luckily, this is a problem that has already been solved.
-Google has published papers on the MapReduce framework \cite{mapreduce}, which
-allows for parallel processing of large data sets, and The Google File System
-\cite{gfs}, a scalable, distributed file system. The Apache Hadoop
-\cite{hadoop} project has published free software implementations of a
-MapReduce framework and a distributed file system, both of which build on the
-ideas in the papers published by Google.
+Luckily, this is a problem that has already been solved. Google has published
+papers on the MapReduce framework \cite{mapreduce}, which allows for parallel
+processing of large data sets, and The Google File System \cite{gfs}, a
+scalable, distributed file system. The Apache Hadoop \cite{hadoop} project has
+published free software implementations of a MapReduce framework and a
+distributed file system, both of which build on the ideas in the papers
+published by Google.
 
 \section{Problem description}
+% Problemsbeskrivning.
 
-Last.fm
+At last.fm there is a central repository where metrics and statistics for
+various services are stored. There is a web site for internal use that
+displays graphs of the services and their statistics. Examples of statistics
+are web site usage broken down by country, scrobbling broken down by client
+and daily subscription counts.
 
-SQL
+The statistics are stored in a relational database. Each measure is stored
+atomically; a summing operation across numerous rows is required to answer
+queries. The database is bursting at the seams because of the size of the
+data. Some of the problems with the current setup is that adding a new data
+source to the database requires manual intervention and that some of the
+questions that analysts ask can not be answered in real-time.
 
-EXPLODE
+Last.fm foresees that its data sets will continue to grow and would like to
+move away from the current setup towards a scalable data store. The data store
+should be able to answers most questions in real time. It should be easy
+to publish new data sets.
+
+\section{Goal}
+% Syftesformulering.
+
+The goal is to build a tool that enables its users to easily turn log files
+into aggregates. The tool should provide a simple mechanism for data analysis
+(import) and querying (export).
+
+The data analysis will require the user to write a MapReduce map. The queries
+will be performed via an export API for consumption by computer programs.
+
+The queries will always be made over a time range and for a single unit. The
+tool will support analytical functions such as drill-down, slicing, etc.
+
+The tool should be aware of dimensions in general and the dimension of time in
+particular. Measurements will be specified in a unit. There will be no limit
+on the number of units.
+
+There should be no theoretical boundary on the number of dimensions. Common
+queries must be answered expediently to allow for analysts to explore data in
+real time.
 
 \section{Scope}
+% Avgransning.
 
-Time-series based analysis is the most important one when analysing web
-logs, it comes natural since all log entries are timestamped.
-\cite{discoveringweb} It is also the most common form we have seen for logs
-in general.
+The data store is just that: a data store with import and export
+functionality. The data store does not interpret the data it store.
 
-Assuming that time-series analysis is the most important, we have imposed
-the restriction on the system that is built so that it only handles
-time-series. Because of this restriction the data store obviously becomes
-less general purpose. Although since most log data is time-series based
-this restriction does not cripple the system much---if at all.
+The data store is time-series-based: all data must include the time dimension.
 
-\subsubsection*{Related work}
+All measures are integers. Sum is the only supported aggregation function. The
+data store does not support reach-like data (i.e. number of unique visitors)
+since it is not summable.
+
+There is no functionality for plotting data sets, etc.
+
+\section{Related work}
 
@@
 Omniture is used at Last.fm etc.
-
-\section{Goal}
-
-The goal (sv. syfte) with our work.
 

### 2009-05-12 doc/report/msc-report.tex
orphan'd text at the end.
doc/report/msc-report.tex: * general work on the Method section. * added to Glossary appendix.
doc/report/references.bib:
* removed Java.
* added OLAP Data Scalability

@@
 
+% chapterquote
 \newcommand{\chapterquote}[2] {
@@
 
-At last.fm there is a central repository where metrics and statistics for
+At Last.fm there is a central repository where metrics and statistics for
 various services are stored. There is a web site for internal use that
@@
 
+% TODO: Why do we do these avgränsningar.
+
 The data store is just that: a data store with import and export
@@
 
-A data warehouse is a repository of an organization's
-electronically stored data. Data warehouses are designed to facilitate
-reporting and analysis. \cite{wikipedia}
-
-The core actions of a data warehouse is to load, transform and extract data.
-The classic thinking in modelling data warehouses is to divide data into facts
-and dimensons. The facts are measures categorized by dimensions.
-
-In OLAP\footnote{On-Line Analytical Processing}, which is a kind of data
-warehouse, the data is thought of as being structured as a cube. The cube is
-typically modelled in a realational database with fact tables and dimension
-tables, upon which the system performs joins to extract data.
-\cite{olap_solutions}
+OLAP is something. It's drawbacks are unknown.
 
@@
 
+Mention data sets here.
+
 \subsection{Import}
@@
 \subsection{Export}
-
-\section{Interface}
 
@@
 values of the dimension that were not fixed. The dice operation is a slice on
-more than two dimensions. [source:
-http://www.cs.ualberta.ca/~zaiane/courses/cmput690/glossary.html]
+more than two dimensions.
 
@@
 
-Time is the most common of the dimensions in the data cube constructed for
-Zohmg. The time stamp is a common property for all data handled by Zohmg;
-every measurement is expected to have time as one of its dimensions. TODO:
-really, this belongs in the implementation section. I find nothing interesting
-to say about time series in this section.
+TODO: Time series is one of the most important measures in log analyzing, it
+is also very natural when the data is time-stamped. \cite{discoveringweb}
 
@@
 
-\section{System}
+\section{Techniques and tools}
 
-Regarding the infrastructure, both BigTable and MapReduce has high
-availability, scalability and reliability. Since BigTable runs on top of
-GFS\footnote{Google File System}, the data stored in BigTable is
-replicated, usually three times, onto other nodes in the cluster.
-\cite{gfs} Likewise, if a node in a MapReduce job goes down or fails to
-complete the job it is reassigned to another node. \cite{mapreduce}
-
-\chapter{Method}
-
-\section{Apache Hadoop}
-
-Hadoop is a free software implementation of Google's MapReduce-framework
-described in \cite{mapreduce}. Hadoop is written in Java and is part of
-the Apache Software Foundation.  It is a fairly mature piece of software
-and is used in production at numerous companies. \cite{hadoop}
-
-MapReduce is a framework for running jobs in parallel across multiple
-machines. As the name implies, MapReduce is made up of two parts or
-phases: map and reduce. The map phase consumes one piece of input at a
-time (a piece of input can be a line from a log file, for example), and
-emits an intermediate key and value. These keys and values are then
-processed by the reduce phase, which emits the final key-value pairs.
-
-MapReduce works by splitting the input, which commonly is in the size
-range of many gigabytes, into n parts and lets each node of the computing
-cluster work on one or more parts. The mapper on each node emits key-value
-pairs which are fed into the reduce phase. The final output is collated,
-usually on a distributed file system.
-
-\subsection{MapReduce}
-
-Analysing the massive amounts of log data raises high demand for computing
-power. The MapReduce \cite{mapreduce} pattern suits the analysing issue
-well. The great benefit of MapReduce is that it runs on a computer cluster
-of commodity hardware where computation takes place independently in
-parallel. All results computed by each node are then reduced into one
-single result.
-
-Basically MapReduce is a two phase computation. First the input data is
-partitioned and each of these partitions is sent to a \textit{mapper} which
-performs computations on each row of the data. The mapper then outputs each
-resulting computation on the format $\langle$\textit{key,value}$\rangle$. These
-key-value pairs are received by a \textit{reducer} which performs the second
-stage of the computation.
-All Zohmg import programs are executed on Hadoop.
-
-\subsection{Hadoop Distributed File System}
-
-HDFS is a free software implementation of the GFS \cite{gfs}.
-
-The goal of HDFS is to store large data sets while being able to handle
-hardware failure and being deployed in heterogenous hardware and software
-eco systems. Additionally the Hadoop and HDFS interaction is designed with
-the notion that it is cheaper to move the computation, MapReduce programs,
-to the data than the other way around.
-
-Every Hadoop node can be formatted with HDFS, this reserves disk space on
-the node to be used for HDFS. The HDFS container resides on the nodes
-general purpose file system. It is also possible to use HDFS as a
-stand-alone general purpose distributed file system, without Hadoop.
-
-The default block size on HDFS 64 MB, significantly larger than
-file systems used for hard drives. The motivation is that applications
-that use HDFS are not general purpose applications which run on general
-purpose file systems, but batch applications which read, write, or both,
-large amounts of data.
-
-\subsection{Hadoop streaming}
-
-Any program that reads and outputs to the file pointers \texttt{stdin} and
-\texttt{stdout}, respectively, can be used as MapReduce programs with
-Hadoop Streaming. This makes it possible to use any shell script or
-program which inputs and outputs this way as a MapReduce program.
-
-\subsection{Dumbo}
-
-Dumbo is a Python framework for writing Hadoop programs in languages other than
-Java, in a so-called Streaming Mode. It is used extensively at Last.fm for
-writing short prototypes. \cite{dumbo}
-
-All Zohmg programs written in Python use Dumbo to execute on Hadoop.
-
-\section{Apache HBase}
-
-HBase is a free software implementation of Google's BigTable data store
-described in \cite{bigtable}. It is a sub-project of Hadoop, written in
-Java and a part of the Apache Software Foundation. HBase is still in its
-infancy and few organizations use it for mission-critical applications.
-\cite{hbase}
-
-BigTable is a data store designed for traversal of very large data sets.
-Time-stamped cells inside column-familes in rows. The row has a key and
-any number of cells attached to it. The rows in a table are sorted by the
-row key.  The typical use-case of BigTable is to scan a range of rows. It
-is therefor important to set up one's keys so that related data is close
-to each other. For example, if it makes sense to traverse the data in
-chronological order, the key might contain a representation of the data's
-timestamp. The classic example is to have the reversed domain name (i.e.
-com.google.www) as the row key, which means that all sub-domains of any
-domain are next to each other.
-
-Sparseness is a big thing. Null values are free of charge. TODO: *Describe
-how we leverage sparsity*.
-
-TODO: *this should go into implementation.*
-The typical use case for Zohmg is to read one or more column-qualifiers,
-which correspond to one or more points in n-space, over a range of time.
-Therefor, the keys are made up of the time and unit of the measurement,
-and the qualifier queried for represents the point in n-space. More on
-this in the data model section.
-
-All Zohmg data is stored in HBase.
-
-\subsection{BigTable}
-
-The data model is a simple one: Rows of data are stored in named tables.
-Each row consists of a sortable row key and an arbitrary number of columns.
-The columns are of the form "column-family:qualifier", where the column
-family is one of a number of fixed column families defined by the table,
-and the qualifier is an arbitrary string specified by the application. In
-effect, column families are pre-announced while qualifiers are not. The
-contents of each column family are stored together, so the user will want
-to store items that have similar characteristics in the same column family.
-Each data point is called a cell. A cell is an uninterpreted array of
-bytes---there are no data types, and it is up to the application to
-interpret the data correctly. The table can store any number of timestamped
-versions of each cell, allowing versioning of data.
-
-The model above can indeed be thought of as a sparse distributed
-multidimensional sorted map. In concept, this multidimensional map is
-identical to a nested hash map from your favourite programming language. The
-dimensions of this multidimensional map is mapped out on the row key, the
-column family and the column qualifier, and possibly also of the version
-timestamp.
+TODO: Why do we chose HBase in favour of other techniques (i.e. SQL)?
 
@@
 
+Sparseness is a big thing. Null values are free of charge. TODO: *Describe
+how we leverage sparsity*.
+
+\chapter{Method}
+
+\section{Procedure}
+
+TODO: Explain our working procedure regarding
+\begin{itemize}
+\item[-] Exploration
+\item[-] Prototyping
+\item[-] Real-world data
+\item[-] Litterature studies
+\item[-] Design phase
+\end{itemize}
+
+\section{Apache Hadoop}
+
+Hadoop is a free software implementation of Google's MapReduce-framework; it is
+written in Java and is part of the Apache Software Foundation. Hadoop is a
+fairly mature piece of software and is used in production at numerous companies,
+among them Last.fm.
+\cite{hadoop}
+
+\subsection{MapReduce}
+
+MapReduce is a framework for running jobs in parallel across multiple
+machines. As the name implies, MapReduce is made up of two parts or
+phases: map and reduce. The map phase consumes one piece of input at a
+time (a piece of input can be a line from a log file, for example), and
+emits an intermediate key and value. These keys and values are then
+processed by the reduce phase, which emits the final key-value pairs.
+
+MapReduce works by splitting the input, which commonly is in the size
+range of many gigabytes, into \textit{n} parts and lets each node of the
+computing cluster work on one or more parts. The mapper on each node emits
+key-value pairs which are fed into the reduce phase. The final output is
+collated, usually on a distributed file system.
+
+\subsection*{Hadoop Distributed File System}
+
+HDFS is a free software implementation of the GFS \cite{gfs}.
+
+The goal of HDFS is to store large data sets while being able to handle
+hardware failure and being deployed in heterogenous hardware and software
+eco systems. Additionally, the Hadoop and HDFS interaction is designed with
+the notion that it is cheaper to move the computation, MapReduce programs,
+to the data than the other way around.
+
+Every Hadoop node can be formatted with HDFS, this reserves disk space on
+the node to be used for HDFS. The HDFS container resides on the nodes
+general purpose file system. It is also possible to use HDFS as a
+stand-alone general purpose distributed file system, without Hadoop.
+
+The default block size on HDFS 64 MB, significantly larger than
+file systems used for hard drives. The motivation is that applications
+that use HDFS are not general purpose applications which run on general
+purpose file systems, but batch applications which read, write, or both,
+large amounts of data.
+
+\subsection*{Hadoop streaming}
+
+Any program that reads and outputs to the file pointers \texttt{stdin} and
+\texttt{stdout}, respectively, can be used as MapReduce programs with
+Hadoop Streaming. This makes it possible to use any shell script or
+program which inputs and outputs this way as a MapReduce program.
+
+\subsubsection*{Dumbo}
+
+Dumbo is a Python framework for writing Hadoop programs in Python; it exploits
+the Hadoop Streaming Mode to do so. It is used extensively at Last.fm for
+writing short prototypes. \cite{dumbo}
+
+\section{Apache HBase}
+
+HBase is a free software implementation of Google's BigTable. \cite{bigtable} It
+is a sub-project of Hadoop, written in Java and a part of the Apache Software
+Foundation. HBase is still in its infancy and few organizations use it for
+mission-critical applications. \cite{hbase}
+
+\subsection{BigTable}
+
+BigTable is a data store designed for traversal of very large data sets.
+Time-stamped cells inside column-families in rows. The row has a key and any
+number of cells attached to it. The rows in a table are sorted by the key.
+The typical use-case of BigTable is to scan a range of rows. It is therefor
+important to set up the row keys so that related data is close to each other.
+For example, if it makes sense to traverse the data in chronological order, the
+key might contain a representation of the data's timestamp. The classic example
+is to have the reversed domain name (i.e. com.google.www) as the row key, which
+means that all sub-domains of any domain are next to each other.
+
 \subsection{Data model}
 
-Sorted distributed multidimensional sparse hashtable.
+The data model is a simple one: Rows of data are stored in labeled tables. Each
+row is indexed with a row key, a column key, and a timestamp.  The columns are
+of the form "column-family:qualifier", where the column-family is one of a
+number of fixed column-families defined by the table, and the qualifier is an
+arbitrary string specified by the application. In effect, column families are
+pre-announced when creating the table while qualifiers are not. The contents of
+each column family are stored together, so the user will want to store items
+that have similar characteristics in the same column family.  Each data point is
+called a cell. A cell is an uninterpreted array of bytes, there are no data
+types, and it is up to the application to interpret the data correctly. The
+table can store any number of timestamped versions of each cell, allowing
+versioning of data.
 
-\subsubsection*{Rows}
+\begin{center}
+    \begin{table}[h]
+    \begin{tabular}{|l|l|l|l|l|}
+    \hline
+    \small \bf Row Key & \small \bf Time & \small \bf Column &
+    \multicolumn{2}{|c|}{\small \bf Column} \\
+     & \small \bf Stamp & \small "contents:" &
+    \multicolumn{2}{|c|}{\small "anchor:"} \\
+    \hline
+     & t9 & & \small "anchor:cnnsi.com" & "CNN" \\
+     & t8 & & \small "anchor:my.lok.ca" & "CNN.com" \\
+    "com.cnn.www" & t6 & \small "{\textless}html\textgreater..." & & \\
+     & t5 & \small "{\textless}html\textgreater..." & & \\
+     & t3 & \small "{\textless}html\textgreater..." & & \\
+    \hline
+    \end{tabular}
+    \caption{Example HBase table.}
+    \end{table}
+\end{center}
 
-Each row in a BigTable is associated with a unique key. All the data is
-stored sorted by rowkey. Because of this it is important to have a thorough
-analysis of how the data is going to be accessed in order to store it
-efficiently.
+\vspace{-24pt}
 
-\subsubsection*{Regions}
-
-BigTable use the concept of tablets, HBase in turn uses regions. In order
-to be consistent with the rest of this paper we henceforth use the term
-region.
-
-A region is a partition of the data stored on a node in the computer
-cluster. A region server (referred to as tablet server in \cite{bigtable})
-holds information on which regions are located on which nodes. More
-precisely the region server knows what range of rowkeys are in each region
-on each node. This behaviour makes it fast to access data while keeping
-reliability and scalability features.
+\noindent The model above can indeed be thought of as a sparse distributed
+multidimensional sorted map. In concept, this multidimensional map is identical
+to a nested hash map. The dimensions of this multidimensional map is mapped out
+on the row key, the column family and the column qualifier, and possibly also of
+the version timestamp.
 
@@
 % Disambiguation needed. -fredrik
-\subsection*{Dimensions}
+\subsection*{Physical storage}
 
-The degrees of freedom in BigTable.
+Each column-family is stored as a single file in HBase; it is then subdivided
+into column-qualifiers within the same file. This is important to consider when
+designing application and schemas, since reading from several column-families
+will open several files.
 
-\subsubsection*{Column-families}
+\subsection*{Sparsity}
 
-Each column-family is stored as a single map-file in HBase.
-
-\subsubsection*{Column-qualifiers}
-
-Each column-family is further divided into one or more column-qualifiers.
+When many cells are empty, the table is said to be sparse. This can potentially
+be a problem, because even sparse cells may cost storage space.  BigTable and
+HBase, however, don't store empty cells; as consequence sparsity is not
+associated with a great storage cost.
+\cite{olap_data_scalability,olap_solutions}
 
@@
 
-Depending on how the rowkeys are formatted different hierarchies are
-created. The order in which the data is stored is important, it defines how
-effective scanning data will be. In the case of having thousand rows and
-requesting ten out of these, then only one por cent of the data is
-interesting. If a scan would be required to visit all the thousand rows a
-lot of rows are skipped. Skipping rows is expensive in that sense that they
-are but the result is thrown away. The goal is to push this cost to a bare
-minimum, using as many of the rows as possible.
+Depending on how the row keys are formatted different, hierarchies are created.
+The order in which the data is stored is important, it defines how effective
+scanning data will be. Sparsity also comes into play concerning effective
+scanning. In the case of having thousand rows and requesting ten out of these,
+then only one por cent of the data is interesting. If a scan would be required
+to visit all the thousand rows a lot of rows are skipped. Skipping rows is
+expensive in that sense that they are but the result is thrown away. The goal is
+to push this cost to a bare minimum, using as many of the rows as possible.
 
@@
 used.
-\subsection{Thrift}
 
-A remote procedure call framework for building scalable cross-language
+\subsection*{Thrift}
+
+HBase has a Thrift server which serves as the interface to languages other than
+Java---for instance Python.
+
+Thrift is a remote procedure call framework for building scalable cross-language
 services. Thrift combines a software stack with a code generation engine to
 build services that work seamlessly between a wide range of programming and
-scripting languages. \cite{thrift}
-
-Thrift was originally developed at Facebook and in April 2007 it was
-released as free software. It entered the Apache Incubator in May 2008.
-
-HBase has a Thrift server which serves as the interface to languages other
-than Java---for instance Python.
-
-\subsection{JSON}
-
-JSON\footnote{JavaScript Object Notation} is a lightweight text-based
-human-readable data interchange format. It is used for representing simple
-data structures and associative arrays (called objects). JSON is often used
-for serializing structured data and transmitting it over network
-connection.
-
-\subsection{YAML}
-
-YAML\footnote{YAML Ain't a Markup Language} is a human-readable data
-serialization format available for a wide range of programming and
 scripting languages.
 
-The outline and markup of YAML documents makes it well suited for data
-which humans are likely to view or edit, such as configuration files.
+Thrift was originally developed at Facebook and in April 2007 it was released as
+free software. It entered the Apache Incubator in May 2008. \cite{thrift}
 
-\section{Java}
+\section{Dependability}
 
-Both Hadoop and HBase are written in Java---a high-level object-oriented
-programming language.
+Regarding the infrastructure, both Hadoop and HBase have high availability,
+reliability, and scalability.
 
-Most Hadoop users write their MapReduce programs in Java, since this is the
-native language of Hadoop. \cite{java}
+Both Hadoop and HBase are distributed systems; making them highly available out
+of the box. Although the Job Tracker\footnote{All client applications are
+submitted to Hadoop's Job Tracker.} is a single point of failure for Hadoop.
+The HDFS NameNode\footnote{A unique server which keeps the directory tree of all
+files in HDFS.} is also a single point of failure. However, a secondary HDFS
+NameNode, which regularly takes a snapshot of the primary NameNode's state, is
+spawned. There exist no automatic fail-over though, if the main the primary
+NameNode crashes manual intervention is needed.
 
-\subsection{PyYAML}
+Both systems have high reliability. Data on HDFS are replicated to---typically
+three---other nodes in the cluster. Jobs submitted to Hadoop also have a high
+reliability: If a node does not succeed with a job it is rescheduled to another
+free node which executes it.
 
-PyYAML is a complete YAML 1.1 compliant parser and emitter for Python.
-\cite{pyyaml}
+Regarding scalability Hadoop and HBase scales close to linearly when nodes are
+added to the cluster.
 
-\subsection{Setuptools}
-
-Python Setuptools is a framework---built on top of the Python module
-distutils---for basic package management of Python software. Amongst other
-things it creates and installs Python Eggs\footnote{A Zip file which
-bundles Python software files.}. \cite{setuptools}
-
-\subsection{simplejson}
-
-A simple, fast, complete and extensible JSON encoder and decoder for
-Python. The Python module simplejson is implemented entirely in Python and
-has no external dependencies. However a C extension which provides a
-serious speed boost is included in simplejson. \cite{simplejson}
-
-\section{Configuration}
-
-Data sets and environment.
+While relational databases can be made highly available and reliable with load
+balancing and so called \textit{Sharding}, the setup is non-trivial. Regarding
+scalability, relational databases are highly dependent on having data in working
+memory; thus they scale poorly.
+% TODO: [citation needed]
 
@@
 
-Zohmg is written mostly in Python. Zohmg uses the following Python modules
-internally; it depends on them to be available on the system.
+Because there existed a Python module for writing MapReduce programs (Dumbo),
+Python was a natural choice during the design and exploration phases. The main
+reason was the rapid development of Python programs, several times shorter than
+their Java counter parts made them faster to develop.
+
+\section{Configuration}
+
+The Zohmg needs configuration of the user's data sets and the environment for
+Hadoop and Dumbo.
+
+The data sets are configured using YAML\footnote{YAML Ain't a Markup Language},
+a human-readable data serializaton format. Any human-readable configure file
+method could have been chosen, e.g. INI files\footnote{A simple configuration
+mechanims where one would have lines of \texttt{variable = value}.}. YAML was
+choosen because it is is very simple for humans to read and edit, and likewise
+for machines.
+
+The environment is a Python script which defines run-time variables, required by
+Hadoop and Dumbo. Using a Python scripts for run-time variables leaves the
+problem of validation to Python.
 
@@
 
-TODO: More general about middleware, move Paste mentioning to Implementation
-chapter.
+In order to join together all the different abilities which were wanted from the
+data server, it was designed as a middleware application. Since the main goal
+was not to build a middleware framework from scratch, an existing such was used.
 
-Python Paste is a framework for building WSGI\footnote{Web Server Gateway
-Interface, a HTTP extension where a CGI-like environment is passed around.}
-applications and middleware. \cite{definitive_guide_to_pylons}
+The data export server is built with a WSGI\footnote{Web Server Gateway
+Interface} framework, it includes both middleware and application parts. The
+middleware applications are responsible for dispatching requests to the
+corresponding application or file.
 
-WSGI applications receive a WSGI request and returns a response with the
-built-in web server.
-
-Middleware is software that acts as an intermediary. WSGI middleware
-receives a WSGI request and then performs logic upon this request, before
-passing the request to a WSGI application. \cite{paste}
+When a user queries the data server, JSON\footnote{JavaScript Object Notation}
+is returned. JSON was chosen because it is i simple text serialization format
+which is widely used in web development.
 
@@
 
-% COMMENT: do we really need to repeat this?
-As stated in chapter 3, Zohmg is built upon existing tools and software. It
-uses Dumbo to run jobs on Apache Hadoop, and uses Apache HBase to persist the
-data.
+TODO: *this should go into implementation.*
+The typical use case for Zohmg is to read one or more column-qualifiers,
+which correspond to one or more points in n-space, over a range of time.
+Therefor, the keys are made up of the time and unit of the measurement,
+and the qualifier queried for represents the point in n-space. More on
+this in the data model section.
 
@@
 depending on the URI it was fed.
+
+Python Paste is a framework for building WSGI\footnote{Web Server Gateway
+Interface, a HTTP extension where a CGI-like environment is passed around.}
+applications and middleware. \cite{definitive_guide_to_pylons}
+
+WSGI applications receive a WSGI request and returns a response with the
+built-in web server.
+
+Middleware is software that acts as an intermediary. WSGI middleware
+receives a WSGI request and then performs logic upon this request, before
+passing the request to a WSGI application. \cite{paste}
 
@@
 
-\begin{itemize}
-\item \texttt{Serializing}
-Serializing data is the task of converting an object into a sequence of
-bits so that it can be stored on a storage medium or transmitted over a
-network link. Rereading the sequence of bits should restore the original
-state of the object it was in before the serialization.
-\end{itemize}
+\begin{tabular}[t]{lp{100mm}}
+\textbf{JSON} &
+JavaScript Object Notation, a lightweight text-based human-readable data
+interchange format. It is used for representing simple data structures and
+associative arrays (called objects). JSON is often used for serializing
+structured data and transmitting it over network connection.
+
+ \\
+
+\textbf{Serializing} &
+Serializing data is the task of converting an object into a sequence of bits,
+making it possible to store on a storage medium or transmitted over a network
+link. Rereading the sequence of bits should restore the original state of the
+object it was in before the serialization. \\
+\end{tabular}
+
+\chapter{Text with no home and no current purpose}
+
+A data warehouse is a repository of an organization's electronically stored
+data. Data warehouses are designed to facilitate reporting and analysis.
+\cite{wikipedia}
+
+The core actions of a data warehouse is to load, transform and extract data.
+The classic thinking in modelling data warehouses is to divide data into facts
+and dimensons. The facts are measures categorized by dimensions.
+
+In OLAP\footnote{On-Line Analytical Processing}, which is a kind of data
+warehouse, the data is thought of as being structured as a cube. The cube is
+typically modelled in a realational database with fact tables and dimension
+tables, upon which the system performs joins to extract data.
+\cite{olap_solutions}
 

### 2009-05-12 scripts/install_hadoop_hbase.sh
plenty-o-fixes in scripts/install_hadoop_hbase.sh: + prints friendly instructions at end of installation + chowns installed files to the sudoing user + uses hbase-0.19.2 instead of 0.19.1 (apache removes old point releases) + changed from sh to bash -- less portable, much more scriptable + (temporarily) disabled tee'ing in exec_and_log -- it fsck'd the return code + makes --prefix work by creating the directory

@@ -1,11 +1,15 @@
-#!/bin/sh
-
+#!/bin/bash
+# install hadoop (patched) and hhbase in /opt
 
 if [ $EUID -ne 0 ]; then
-    echo "Error: Superuser privileges required."
+    echo "you need to be root. please sudo."
     exit 1
 fi
 
+
+# TODO:
+#  use mirrors for ${hadoop,hbase}_tar
+#  also, apache seems to mirror only the very latest point release :-(
 
 # set default variables.
 prefix="/opt"
@@ -13,12 +17,14 @@ hadoop_version="hadoop-0.19.1"
 hadoop_tar="$hadoop_version.tar.gz"
 patch_1722="HADOOP-1722-branch-0.19.patch"
 patch_5450="HADOOP-5450.patch"
-hbase_version="hbase-0.19.1"
+hbase_version="hbase-0.19.2"
 hbase_tar="$hbase_version.tar.gz"
-hadoop_release="http://mirrors.ukfast.co.uk/sites/ftp.apache.org/hadoop/core/hadoop-0.19.1/$hadoop_tar"
-hadoop_1722="https://issues.apache.org/jira/secure/attachment/12401426/$patch_1722"
-hadoop_5450="https://issues.apache.org/jira/secure/attachment/12401846/$patch_5450"
-hbase_release="http://mirrors.ukfast.co.uk/sites/ftp.apache.org/hadoop/hbase/hbase-0.19.1/$hbase_tar"
+#hadoop_release="http://mirrors.ukfast.co.uk/sites/ftp.apache.org/hadoop/core/$hadoop_version/$hadoop_tar"
+hadoop_release="http://apache.mirror.infiniteconflict.com/hadoop//core/$hadoop_version/$hadoop_tar"
+hadoop_1722="http://issues.apache.org/jira/secure/attachment/12401426/$patch_1722"
+hadoop_5450="http://issues.apache.org/jira/secure/attachment/12401846/$patch_5450"
+#hbase_release="http://mirrors.ukfast.co.uk/sites/ftp.apache.org/hadoop/hbase/$hbase_version/$hbase_tar"
+hbase_release="http://apache.mirror.infiniteconflict.com/hadoop/hbase/$hbase_version/$hbase_tar"
 install_log="$(pwd)/install_hadoop_hbase.log"
 install_tmplog=$(mktemp /tmp/zohmg-log.XXXXXXXX)
 
[…]

### 2009-05-12 scripts/install_hadoop_hbase.sh
install_hadoop_hbase.sh> make sure the user knows what's up.

@@ -7,6 +7,7 @@ if [ $EUID -ne 0 ]; then
 fi
 
 
+
 # TODO:
 #  use mirrors for ${hadoop,hbase}_tar
 #  also, apache seems to mirror only the very latest point release :-(
@@ -130,6 +131,10 @@ while [ $1 ]; do
 done
 
 
+# make sure the user knows what's up.
+echo "this script will download, patch and install hadoop & hbase in $prefix"
+echo "any key to continue, CTRL-C to abort."
+read
 
 if [ -f $prefix ]; then
     echo "$prefix is a file; can't install there."

### 2009-05-12 zohmg/create.py
zohmg/create.py:  + disabled 'clients' (not used)  + defaults in environment.py  + TODO: example mapper  + TODO: example dataset  + TODO: example transformer

@@ -1,21 +1,21 @@
 from zohmg.utils import fail
 import os, sys
 
-
-DIRS = ["clients","config","lib","mappers","transformers"]
+# FIXME: temporarily disabled 'clients'
+DIRS = ["config","lib","mappers","transformers"]
 
 # config/environment.py
 ENV_SCRIPT = """# Please define the following environment variables.
 
 # Path to Hadoop directory.
-#HADOOP_HOME = '/opt/hadoop-0.19.1'
+HADOOP_HOME = '/opt/hadoop-0.19.1'
 
 # Jars for Hadoop Core, Hadoop Streaming and HBase
 CLASSPATH = (
   # be sure to set these correctly!
-  #'/opt/hadoop-0.19.1/build/hadoop-0.19.2-dev-core.jar',
-  #'/opt/hadoop-0.19.1/build/contrib/streaming/hadoop-0.19.2-dev-streaming.jar',
-  #'/opt/hbase-0.19.1/build/hbase-0.19.1.jar'
+  '/opt/hadoop-0.19.1/build/hadoop-0.19.2-dev-core.jar',
+  '/opt/hadoop-0.19.1/build/contrib/streaming/hadoop-0.19.2-dev-streaming.jar',
+  '/opt/hbase-0.19.2/build/hbase-0.19.2.jar'
 )
 """
 
@@ -38,7 +38,7 @@ Static files here will be served from the url http://host:port/client/filename.
 
 MAPPER = """# identity mapper.
 def map(key, value):
-    yield key, value
+    # egads!
 """
 
 TRANSFORMER = """# identity transformer.
@@ -67,7 +67,7 @@ class Create(object):
         # Create .zohmg, README, client, environment, mapper and transformer.
         self.__write_to_file('.zohmg')
         self.__write_to_file('README', README)
-        self.__write_to_file("clients/client.html",CLIENT)
+        #self.__write_to_file("clients/client.html",CLIENT)
         self.__write_to_file('config/environment.py', ENV_SCRIPT)
         self.__write_to_file('mappers/identity_mapper.py', MAPPER)
         self.__write_to_file("transformers/identity_transformer.py",TRANSFORMER)

### 2009-05-13 doc/report/msc-report.tex
report style; non-hardcoded paths in Makefile
added TODOs fixed most of Dave's comments

@@
-\documentclass[a4paper,10pt]{book}
-
-\author{Per Andersson, Fredrik M{\"o}llerstrand}
-
-\title{Zohmg---a large scale data store for aggregated time-series-based
-data. [DRAFT]}
-
-% chapterquote
+% chapterquote command.
 \newcommand{\chapterquote}[2] {
@@
 \end{quote}
-
 \vspace{24pt}
@@
 
+\documentclass[a4paper]{report}
+
+\author{Per Andersson, Fredrik M{\"o}llerstrand}
+\title{Zohmg---a large scale data store for aggregated time-series-based
+data. [DRAFT]}
+
 \begin{document}
-
 \maketitle
 
-\noindent \Large{\textbf{Abstract}}
+\noindent {\Large \textbf{Abstract}}
+\vspace{12pt}
+\noindent
 
-\vspace{12pt}
-
-\noindent
+% TODO: use less buzzwords; focus on the problem.
 
@@
 \tableofcontents
-
 \vfill
-
 \pagebreak
@@
 
-Any organization will eventually wish to analyze various characteristics of
-its service and operations. These services can be anything from web sites to
-theme parks. The analysis might include characteristics such as the number of
+% TODO: introduce time-series, data warehousing, mapreduce, bigtable
+% TODO: why is time important? (we query for time, etc.)
+
+Any organization will eventually want to analyze various characteristics of
+its service and operations. The services can be anything from web sites to
+theme park; the analysis might include characteristics such as the number of
 visitors per day, or the average time a theme park guest spends queueing for a
@@
 
-After having collected these base facts, one can draw conclusions about what
-countries generate the most traffic to the site, and which web browser is most
-popular in a certain country, and so on.
+When the records in the data set have been analyzed, one can draw conclusions
+about what countries generate the most traffic to the site, which web browser
+is most popular in a certain country, etc.
 
-Typically, the size of a data set is in the range of several gigabytes of data
-per day. Some data sets are many times larger.
+The size of the data sets at last.fm is in the range of several gigabytes per
+day. Some data sets are many times larger. It is our belief that this is
+typical.
 
@@
 
-Luckily, this is a problem that has already been solved. Google has published
-papers on the MapReduce framework \cite{mapreduce}, which allows for parallel
-processing of large data sets, and The Google File System \cite{gfs}, a
-scalable, distributed file system. The Apache Hadoop \cite{hadoop} project has
-published free software implementations of a MapReduce framework and a
-distributed file system, both of which build on the ideas in the papers
-published by Google.
+% TODO: write about how mapreduce already solves the problem of parallel
+% computation, and how gfs solves the problem of distributed storage.
 
@@
 
+% TODO: define aggregates!
+
 The goal is to build a tool that enables its users to easily turn log files
@@
 
-% TODO: Why do we do these avgränsningar.
+% TODO: argue why we do these avgränsningar.
 
@@
 
+% TODO: this section.
+
 OLAP is something. It's drawbacks are unknown.
 
-HDW is system which goal is to create a high performance large scale data
+HDW is a system whose goal is to create a high performance large scale data
 warehouse based on MapReduce and Bigtable. \cite{hdw}
@@
 This thesis describes the Zohmg system and how to use it.
+
+% TODO: make correct.
+The following overview is entirely incorrect:
 
@@
 
+% TODO: example.
+
 The slice operation fixes the values of one or several dimensions in the cube.
@@
 
-TODO: Time series is one of the most important measures in log analyzing, it
+% TODO: this section.
+
+Time series is one of the most important measures in log analyzing, it
 is also very natural when the data is time-stamped. \cite{discoveringweb}
@@
 \section{Aggregates}
+
+% TODO: be clear about how aggregates are *derived* from one or several vars.
+% TODO: example.
 
@@
 
-The classic trade-off between time and space applies to whether or not
-pre-compute an aggregate. Pre-computing to many aggregates might render an
-unfeasible amount of data to store, while pre-computing to few aggregates
-renders longer access times for aggregates which are not pre-computed. This
-problem is further discussed in the thesis, but left to the user of the
-Zohmg system to solve.
+The classic trade-off between time and space applies to whether or not to
+pre-compute an aggregate. Pre-computing too many aggregates might result in an
+unfeasible amount of data, while pre-computing too few aggregates results in
+longer access times for those aggregates that were not pre-computed.
 
@@
 
-TODO: Why do we chose HBase in favour of other techniques (i.e. SQL)?
+% TODO: Why do we chose HBase in favour of other techniques (i.e. SQL)?
 
@@
 
-\textbf{TODO: -- elaborate on reasons to why starsql does not scale--}
+% TODO: elaborate on reasons to why starsql does not scale
 
@@
 
+\section{Stop making speeches}
 A data warehouse is a repository of an organization's electronically stored
@@
 
+\section{It's not like the movies}
+
+Luckily, this is a problem that has already been solved. Google has published
+papers on the MapReduce framework \cite{mapreduce}, which allows for parallel
+processing of large data sets, and The Google File System \cite{gfs}, a
+scalable, distributed file system. The Apache Hadoop \cite{hadoop} project has
+published free software implementations of a MapReduce framework and a
+distributed file system, both of which build on the ideas in the papers
+published by Google.
+
 \end{document}

### 2009-05-13 doc/report/msc-report.tex
separated data cubes etc into chapter: Concepts.. moved section to bottom.
report: added graphichx package introduction introduces aggregation introduction introduces dimensions by way of Aerodyne Corporation example freshened the goal; we should move parts of it into analysis, other parts to scope. added CONCEPTS IN DATA WAREHOUSING (briefly)

@@
 \documentclass[a4paper]{report}
+\usepackage{graphicx}
 
@@
 visits to a web site might contain a timestamp, information about what country
-the visitor is from and what web browser she used. The values of the fields
-can be thought of as points along an axis; Sweden, Norway and Denmark are all
-points along the country axis, for instance. This is an important concept and
-it will be touched upon repeatedly throughout the text.
+the visitor is from and what web browser she used.
 
@@
 about what countries generate the most traffic to the site, which web browser
-is most popular in a certain country, etc.
+is most popular in a certain country, etc. Summaries of data are generally
+called aggregates. To aggregate data, for example to compute the sum of
+visitors for all European countries, is a common technique for speeding up
+data lookup in data stores.
 
@@
 
-It is suitable to store the data across many storage nodes in what is called a
-distributed file system and to to exploit the parallel nature of the records
-by breaking up the data sets in pieces and have each piece be processed
-separately on a node in a large compute cluster.
+It is suitable to to exploit the parallel nature of the records by breaking up
+the data sets in pieces and to have each piece be processed separately on a
+node in a large compute cluster. It is also suitable to store the data across
+many storage nodes in what is called a distributed file system.
 
@@
 % computation, and how gfs solves the problem of distributed storage.
+
+% introducing variables and dimensions.
+
+Organizations measure their operations by considering many variables. When
+these variables are placed in a spreadsheet, which organizations are wont to
+do, the variables are set on the x and y axes. Each axis represents a logical
+grouping of variables. In the example of a log file of visits to a web site
+above, country of origin is one such variable. A variable is generally called
+a dimension in this paper.
+
+The International Aerodyne Corporation, a leading exporter of woodworking
+nails, uses a spreadsheet to keep track of the number of nails exported to
+each country every month (see figure).
+
+% image of a spreadsheet.
+%\includegraphics[width=70mm]{pictars/aerodyne-nails.png}
+%\imagecaptionplease{crates of nails exported}
+
+In the figure above, the time dimension is set on the x-axis and the country
+dimension is set on the y-axis. If the Aerodyne Corporation were to start
+keeping track of the kind of nails they sell (pins, tacks, spikes, etc.) in
+addition to the country to which the nails were sold, they would effectively
+have added a new dimension to their data.
+
+Many organizations work with data that have more than three dimensions. Such
+data is hard to manage in a spreadsheet, and the organizations need
+specialized tools to analyze and visualize their data.
 
@@
 
-The goal is to build a tool that enables its users to easily turn log files
-into aggregates. The tool should provide a simple mechanism for data analysis
-(import) and querying (export).
-
+The goal is to build a data store that enables its users to perform
+multi-dimensional analysis.
+
+In addition, the secondary goal is to build a tool that allows users to easily
+store their data sets in the data store. The tool should provide a simple – as
+in minimalistic – mechanism for data analysis (import) and querying (export).
+
+% is mapreduce part of the goal? hardly.
 The data analysis will require the user to write a MapReduce map. The queries
@@
 
+% these are more likely part of the scope.
 The queries will always be made over a time range and for a single unit. The
@@
 
-\chapter{Analysis}
-
-\chapterquote{Arrange whatever pieces come your way.}{Virgina Wolf}
-
-\section{Task}
-
-Mention data sets here.
-
-\subsection{Import}
-
-\subsection{Store}
-
-\subsection{Export}
+% CONCEPTS IN DATA WAREHOUSING, by clark de jean
+\chapter{Concepts in Data Warehousing}
+
+This chapter will serve as an introduction to various concepts in data
+warehousing.
+
+\section{Data Warehouses}
+
+A data warehouse is a repository of an organization's electronically stored
+data. Data warehouses are designed to facilitate reporting and analysis; the
+core actions of a data warehouse are to load, transform and extract data.
+
+\section{Data Warehouse implementations}
+
+\subsection{OLAP}
+
+OLAP, which is an acronym for Online analytical processing, is an approach to
+quickly answer multi-dimensional analytical queries.
+
+At the core of any OLAP system is the OLAP cube (also known as a
+multidimensional cube or a hypercube). It consists of numeric facts called
+measures which are categorized by dimensions.
+
+The cube is typically modelled in a relational database with fact tables
+and dimension tables, upon which the system performs joins to extract data.
+\cite{olap_solutions}
 
@@
 
-\section{Techniques and tools}
-
-% TODO: Why do we chose HBase in favour of other techniques (i.e. SQL)?
-
-In actual practice, the main difference between the two models is that
-querying is highly restricted in a BigTable-like data store---there are no
-joins, for example---whereas the relational database was born to be queried.
-On the other hand, the restrictions placed upon the BigTable models are
-what allows it to scale to such a large data set. Indeed, BigTable was born
-to scale.
-
-% TODO: elaborate on reasons to why starsql does not scale
-
-Many applications have the behaviour where they write data once and read it
-a lot more, log analysers for instance. This writing-once-reading-many
-behaviour is precisely what BigTable is designed for. BigTable is designed
-for sequential, as opposed to random, data access. A typical use case is
-the scan, where a large number of rows are read in sequence and a certain
-column or column-family are picked up and returned. Certainly, you will
-want to store those data points that are likely to be needed simultaneously
-close to each other. \cite{bigtable}
-
-Sparseness is a big thing. Null values are free of charge. TODO: *Describe
-how we leverage sparsity*.
+\chapter{Analysis}
+
+\chapterquote{Arrange whatever pieces come your way.}{Virgina Wolf}
+
+\section{Task}
+
+Mention data sets here.
+
+\subsection{Import}
+
+\subsection{Store}
+
+\subsection{Export}
 
@@
 
-\section{Stop making speeches}
-A data warehouse is a repository of an organization's electronically stored
-data. Data warehouses are designed to facilitate reporting and analysis.
-\cite{wikipedia}
-
-The core actions of a data warehouse is to load, transform and extract data.
-The classic thinking in modelling data warehouses is to divide data into facts
-and dimensons. The facts are measures categorized by dimensions.
-
-In OLAP\footnote{On-Line Analytical Processing}, which is a kind of data
-warehouse, the data is thought of as being structured as a cube. The cube is
-typically modelled in a realational database with fact tables and dimension
-tables, upon which the system performs joins to extract data.
-\cite{olap_solutions}
-
 \section{It's not like the movies}
@@
 
+\section{Techniques and tools}
+
+% TODO: Why do we chose HBase in favour of other techniques (i.e. SQL)?
+
+In actual practice, the main difference between the two models is that
+querying is highly restricted in a BigTable-like data store---there are no
+joins, for example---whereas the relational database was born to be queried.
+On the other hand, the restrictions placed upon the BigTable models are
+what allows it to scale to such a large data set. Indeed, BigTable was born
+to scale.
+
+% TODO: elaborate on reasons to why starsql does not scale
+
+Many applications have the behaviour where they write data once and read it
+a lot more, log analysers for instance. This writing-once-reading-many
+behaviour is precisely what BigTable is designed for. BigTable is designed
+for sequential, as opposed to random, data access. A typical use case is
+the scan, where a large number of rows are read in sequence and a certain
+column or column-family are picked up and returned. Certainly, you will
+want to store those data points that are likely to be needed simultaneously
+close to each other. \cite{bigtable}
+
+Sparseness is a big thing. Null values are free of charge. TODO: *Describe
+how we leverage sparsity*.
+
 \end{document}

### 2009-05-13 doc/report/msc-report.tex
concepts in data warehousing +DONE add running example - aerodyne +DONE concepts in data warehousing: write intro, summary. +DONE concepts: is our use of 'hypercube' standard? => yes, common OLAP term. +DONE concepts: explain slice&dice clearer.

@@
 
-This chapter will serve as an introduction to various concepts in data
-warehousing.
+This chapter introduces data warehouses and the concept of data cubes. It then
+introduces common operations on data cubes: projections, slicing and dicing,
+and aggregation.
 
@@
 
-\section{Data Warehouse implementations}
-
 \subsection{OLAP}
@@
 At the core of any OLAP system is the OLAP cube (also known as a
-multidimensional cube or a hypercube). It consists of numeric facts called
-measures which are categorized by dimensions.
-
-The cube is typically modelled in a relational database with fact tables
-and dimension tables, upon which the system performs joins to extract data.
-\cite{olap_solutions}
+multi-dimensional cube or a hypercube). The cube consists of numeric facts
+called measures which are categorized by dimensions. For example, the number
+of nails sold is a measure while country, month and type of nail are
+dimensions. It is from OLAP that we have appropriated the concept of a data
+cube.
 
@@
 
-A data cube is a conceptual n-dimensional array of values, in which the values
-are numerical facts belonging to some dimension. The data cube is also known as
-a hypercube, relating to multidimensional analogues of squares and cubes. In
-this context, each dimension of the cube corresponds to an attribute of the
-data in question and each point along the dimension's axis is a value of the
-corresponding attribute. The mental model of a data cube aids in reasoning
-about dimensions. Data cubes are traditionally modeled as star schemas in
-relational database systems. \cite{olap_solutions}
-
-\subsection{Dimensions}
-
-The school book example of a data cube is three dimensional where each of
-the three dimensions correspond to one of the attributes time, products,
-and stores. Adding another attribute, for instance sales, adds another
-dimension to the data cube, because of the one to one relationship of
-attributes and dimensions.
+A data cube is a conceptual n-dimensional array of values. It aids in
+reasoning about dimensions. In the context of this paper, each dimension of
+the cube corresponds to an attribute of the data in question and each point
+along the dimension's axis is a value of the corresponding attribute. At any
+point where the n dimensions intersect, a measure can be found. For example,
+if the dimensions are country and nail type, the point in the cube where
+country is equal to Sweden and nail type is equal to tack will contain a
+measurement of the units of tack nails sold in Sweden.
+
+Data cubes are traditionally modeled as star schemas in relational database
+systems. \cite{olap_solutions}
 
@@
 One of the most common operations on a data cube is the projection. A
-projection is a dimensionality reduction, where one or more dimensions are
-discarded from the cube with the measurements along each axis summed up,
-leaving a data cube of a lesser dimension.
-
-A typical projection is that from an n-dimensional cube down to two
-dimensions, which is the case when plotting an object---with three or more
-dimensions---on screen or paper.
+projection is a function to another data cube, where one or more dimensions
+are discarded from the cube and the measurements along each dimensions are
+summed up, leaving a data cube of a lesser dimension.
+
+For example, consider the three-dimensional data cube that represents sales of
+nails; the dimensions are nail type, country and month. A projection from that
+cube to one with two dimensions -- month and nail type -- would flatten the
+country dimension: the measurements in the new cube represent the global sales
+of each nail type for each month.
 
 \subsection{Slice and dice}
-
-% TODO: example.
 
@@
 
-\section{Time series}
-
-% TODO: this section.
-
-Time series is one of the most important measures in log analyzing, it
-is also very natural when the data is time-stamped. \cite{discoveringweb}
+For example, consider again the three-dimensional data cube that represents
+sales of nails; the dimensions are nail type, country and month. A slice of
+that cube with the country dimension fixed to Germany would contains sales
+figures for each nail type and month for Germany only.
 
@@
 
-% TODO: be clear about how aggregates are *derived* from one or several vars.
-% TODO: example.
-
-An aggregate is a composite of several values. Typical examples of
-aggregate functions are average, count, and sum.
+An aggregate is a composite value that is derived from several variables.
+Typical examples of aggregate functions are average, count, and sum.
+
+For example, sales figures for a whole year may be summed up and stored along
+the year dimension.
 
@@
 longer access times for those aggregates that were not pre-computed.
+
+\section{Summary}
+
+This chapter introduced Data Warehouses in general and OLAP in particular. The
+concept of data cubes was introduced by OLAP. The chapter was rounded off with
+a discussion of dimensionality reducing operations and aggregations of data
+cubes.
 

### 2009-05-13 doc/report/msc-report.tex
doc/report/msc-report.tex: * reduced vspace at bottom of \chapterquote. * added chapter 3, system analysis.

--- 2009-05-13 doc/report/msc-report.tex
@@
 \end{quote}
-\vspace{24pt}
+\vspace{12pt}
 }
@@
 
-\chapter{Analysis}
-
-\chapterquote{Arrange whatever pieces come your way.}{Virgina Wolf}
-
-\section{Task}
-
-Mention data sets here.
-
-\subsection{Import}
-
-\subsection{Store}
-
-\subsection{Export}
+\chapter{Concepts in Data Warehousing}
 
@@
 
-\section{Techniques and tools}
-
-% TODO: Why do we chose HBase in favour of other techniques (i.e. SQL)?
-
-In actual practice, the main difference between the two models is that
-querying is highly restricted in a BigTable-like data store---there are no
-joins, for example---whereas the relational database was born to be queried.
-On the other hand, the restrictions placed upon the BigTable models are
-what allows it to scale to such a large data set. Indeed, BigTable was born
-to scale.
-
-% TODO: elaborate on reasons to why starsql does not scale
-
-Many applications have the behaviour where they write data once and read it
-a lot more, log analysers for instance. This writing-once-reading-many
-behaviour is precisely what BigTable is designed for. BigTable is designed
-for sequential, as opposed to random, data access. A typical use case is
-the scan, where a large number of rows are read in sequence and a certain
-column or column-family are picked up and returned. Certainly, you will
-want to store those data points that are likely to be needed simultaneously
-close to each other. \cite{bigtable}
-
-Sparseness is a big thing. Null values are free of charge. TODO: *Describe
-how we leverage sparsity*.
+\chapter{System Analysis}
+
+\chapterquote{Methods and means cannot be separated from the ultimate aim.}{Emma
+Goldman}
+
+\section*{Introduction}
+
+This chapter will discuss and analyze the system that is to be built. First the
+function wanted from the system is presented; while discussing the input and
+output characteristics, and possible challenges. Secondly, a rough sketch and
+discussion of the system's different parts will be presented. Thirdly, the
+requirements of the system will be featured.
+
+Throughout this chapter distinction is made between the developer: who develops
+the system, and the user: who uses the product which the developer
+creates. Although, in reality, they might be the same person.
+
+\section{System function}
+
+Last.fm has a central statistics and metrics tool built on a relational
+database. It is possible to import analyzed data to this tool; the data is
+presented to the user as graphs. Drawbacks of the system include problems with
+new data source additions: they can sometime not be loaded because they are to
+big. If a user wants to perform a custom query, manual intervention by the
+developer is necessary. There is no way for the system to handle custom queries.
+
+The data source is typically a log file; it is often a large file, up to
+gigabytes in size, of structured data. The data is structured the same way on
+each line, this makes writing a program for analyzing the entire data set as
+simple as analyzing a single line.
+
+Reporting the analyzed is the data is the final step. This is what the user is
+interested in: Viewing the analyzed data. The tool at Last.fm presents the user
+with a dashboard that has a selection of graphs on primary metrics; it also has
+a list of all available graphs. All the graphs are pre-aggregated, no support
+exist for custom queries. Enabling the user to query the aggregated data---for
+instance viewing the most popular web browser in a country---would open
+possibilities to explore the data. During exploration interesting details could
+possibly be discovered; details which may be excluded from the pre-rendered
+aggregates.
+
+Since the data sources at Last.fm are time-stamped logs, the constraint to only
+handle, time-series was brought upon the system. This decision was reached
+mainly to keep the implementation simple.
+
+\subsection*{System parts}
+
+The following three components form the outline of the system: importing,
+storing, and exporting data.
+
+At Last.fm there are several different facilities which produce logs, among
+them: web servers, scrobbling\footnote{Last.fm users can submit what music they
+listen to, this is called scrobbling.}, radio etc. This data reside on a
+commodity hardware cluster used for parallel computation. The system needs to be
+able to easily import data from this source.
+
+% storage.
+Because the data sources are typically large, the storage need to be efficient.
+Storing the analyzed data can be done in different ways, for instance storing
+the raw data or just aggregates. Storing the raw data is very expensive, since
+it would basically copy the data to the data store. Because of this the decision
+was made to only store aggregates.
+
+There are several different ways of modelling multidimensional data.
+
+TODO: How to discuss modelling of data in a general sense, without mentioning
+MapReduce and BigTable?
+
+% export.
+Due to the several different data sources, there exist several different types
+of applications. Therefor the system does not implement any user interface, this
+is left for the developer to implement.
+
+\section{System requirements}
+
+The four key requirements of the system are: ease of use, speed, scalability,
+and being able to fit in the existing software and hardware eco-system.
+
+% ease of use.
+The system should have an easy facility for importing and analyzing new data
+sources, while still being fast and efficient. A simple script---or tweak of an
+already existing data import script---should be enough for importing and
+analyzing a new data source.
+
+% speed.
+Time is of essence when exporting data to the user. The user should be able to
+explore the data at mouse-click rate, not having to wait for computation. This
+is usually solved by pre-rendering the data which can be viewed or queried.
+
+Importing data requires computing power, which is a limited resource. This calls
+for the need to make the import phase efficient. Using a desktop workstation to
+process several gigabytes of data is simply impossible; it might require more
+time than the actual worth of data it is importing---for instance taking more
+than one day to import a days worth of data. There are a number of ways to solve
+this, among them: more powerful machines and parallelizing.
+
+Acquiring more powerful machines quickly becomes expensive. Even if more
+powerful machines are acquired they might not have enough power for processing
+several gigabytes of data: the data to be processed might not fit in the working
+memory.
+
+Another technique is parallelizing, using a cluster of commodity hardware
+machines. Using commodity hardware for computation significantly reduces the
+cost, since commodity hardware is cheap; thusly preferred in most situations.
+
+% scaling.
+In order for a computational setup, cluster or otherwise, to be efficient it
+needs to scale well, this means adding more hardware should add as close as
+possible to the hardware's potential addition to the computational power.
+
+% fit in existing eco system.
+The system needs to co-exist and integrate seamlessly with the already existing
+hardware and software eco-system at Last.fm. Thus, building on top of the
+already existing software a natural choice.
+
+\section*{Summary}
+
+This chapter has analyzed and discussed the function and requirements of the
+system to be built.
+
+The system function part discussed the system's goals and parts. The goals are
+to have a system which enables the developer to easily import and analyze new
+data sources and to be fast enough to serve the analyzed data at mouse-click
+rate to the user. The system's three core parts---importing, storing, and
+extracting data---have also been discussed.
+
+Furthermore the requirements on the system were discussed; the four key
+requirements are: ease of use, speed, scaling, and fitting into the existing
+hardware and software eco-system.
 
@@
 
+\chapterquote{Arrange whatever pieces come your way.}{Virgina Wolf}
+
 \section{Procedure}
@@
 TODO: Explain our working procedure regarding
+
 \begin{itemize}

### 2009-05-13 doc/report/msc-report.tex

--- 2009-05-13 doc/report/msc-report.tex
@@
 
-\chapter{Analysis}
+\chapter{System Analysis}
+
+\chapterquote{Methods and means cannot be separated from the ultimate aim.}{Emma
+Goldman}
+
+\section*{Introduction}
+
+This chapter will discuss and analyze the system that is to be built. First the
+function wanted from the system is presented; while discussing the input and
+output characteristics, and possible challenges. Secondly, a rough sketch and
+discussion of the system's different parts will be presented. Thirdly, the
+requirements of the system will be featured.
+
+Throughout this chapter distinction is made between the developer: who develops
+the system, and the user: who uses the product which the developer
+creates. Although, in reality, they might be the same person.
+
+\section{System function}
+
+Last.fm has a central statistics and metrics tool built on a relational
+database. It is possible to import analyzed data to this tool; the data is
+presented to the user as graphs. Drawbacks of the system include problems with
+new data source additions: they can sometime not be loaded because they are to
+big. If a user wants to perform a custom query, manual intervention by the
+developer is necessary. There is no way for the system to handle custom queries.
+
+The data source is typically a log file; it is often a large file, up to
+gigabytes in size, of structured data. The data is structured the same way on
+each line, this makes writing a program for analyzing the entire data set as
+simple as analyzing a single line.
+
+Reporting the analyzed is the data is the final step. This is what the user is
+interested in: Viewing the analyzed data. The tool at Last.fm presents the user
+with a dashboard that has a selection of graphs on primary metrics; it also has
+a list of all available graphs. All the graphs are pre-aggregated, no support
+exist for custom queries. Enabling the user to query the aggregated data---for
+instance viewing the most popular web browser in a country---would open
+possibilities to explore the data. During exploration interesting details could
+possibly be discovered; details which may be excluded from the pre-rendered
+aggregates.
+
+Since the data sources at Last.fm are time-stamped logs, the constraint to only
+handle, time-series was brought upon the system. This decision was reached
+mainly to keep the implementation simple.
+
+\subsection*{System parts}
+
+The following three components form the outline of the system: importing,
+storing, and exporting data.
+
+At Last.fm there are several different facilities which produce logs, among
+them: web servers, scrobbling\footnote{Last.fm users can submit what music they
+listen to, this is called scrobbling.}, radio etc. This data reside on a
+commodity hardware cluster used for parallel computation. The system needs to be
+able to easily import data from this source.
+
+% storage.
+Because the data sources are typically large, the storage need to be efficient.
+Storing the analyzed data can be done in different ways, for instance storing
+the raw data or just aggregates. Storing the raw data is very expensive, since
+it would basically copy the data to the data store. Because of this the decision
+was made to only store aggregates.
+
+There are several different ways of modelling multidimensional data.
+
+TODO: How to discuss modelling of data in a general sense, without mentioning
+MapReduce and BigTable?
+
+% export.
+Due to the several different data sources, there exist several different types
+of applications. Therefor the system does not implement any user interface, this
+is left for the developer to implement.
+
+\section{System requirements}
+
+The four key requirements of the system are: ease of use, speed, scalability,
+and being able to fit in the existing software and hardware eco-system.
+
+% ease of use.
+The system should have an easy facility for importing and analyzing new data
+sources, while still being fast and efficient. A simple script---or tweak of an
+already existing data import script---should be enough for importing and
+analyzing a new data source.
+
+% speed.
+Time is of essence when exporting data to the user. The user should be able to
+explore the data at mouse-click rate, not having to wait for computation. This
+is usually solved by pre-rendering the data which can be viewed or queried.
+
+Importing data requires computing power, which is a limited resource. This calls
+for the need to make the import phase efficient. Using a desktop workstation to
+process several gigabytes of data is simply impossible; it might require more
+time than the actual worth of data it is importing---for instance taking more
+than one day to import a days worth of data. There are a number of ways to solve
+this, among them: more powerful machines and parallelizing.
+
+Acquiring more powerful machines quickly becomes expensive. Even if more
+powerful machines are acquired they might not have enough power for processing
+several gigabytes of data: the data to be processed might not fit in the working
+memory.
+
+Another technique is parallelizing, using a cluster of commodity hardware
+machines. Using commodity hardware for computation significantly reduces the
+cost, since commodity hardware is cheap; thusly preferred in most situations.
+
+% scaling.
+In order for a computational setup, cluster or otherwise, to be efficient it
+needs to scale well, this means adding more hardware should add as close as
+possible to the hardware's potential addition to the computational power.
+
+% fit in existing eco system.
+The system needs to co-exist and integrate seamlessly with the already existing
+hardware and software eco-system at Last.fm. Thus, building on top of the
+already existing software a natural choice.
+
+\section*{Summary}
+
+This chapter has analyzed and discussed the function and requirements of the
+system to be built.
+
+The system function part discussed the system's goals and parts. The goals are
+to have a system which enables the developer to easily import and analyze new
+data sources and to be fast enough to serve the analyzed data at mouse-click
+rate to the user. The system's three core parts---importing, storing, and
+extracting data---have also been discussed.
+
+Furthermore the requirements on the system were discussed; the four key
+requirements are: ease of use, speed, scaling, and fitting into the existing
+hardware and software eco-system.
+
+\chapter{Method}
 
@@
 
-\section{Task}
-
-Mention data sets here.
-
-\subsection{Import}
-
-\subsection{Store}
-
-\subsection{Export}
-
-\chapter{Method}
-
 \section{Procedure}

### 2009-05-13 doc/report/msc-report.tex
doc/report/msc-report.tex: * minor corrections and changes to System analysis chapter.

@@
 
-This chapter will discuss and analyze the system that is to be built. First the
-function wanted from the system is presented; while discussing the input and
-output characteristics, and possible challenges. Secondly, a rough sketch and
-discussion of the system's different parts will be presented. Thirdly, the
-requirements of the system will be featured.
+This chapter will discuss and analyze the Zohmg system, that is to be built.
+
+First the application area and the wanted system functionality are presented;
+while discussing the input and output characteristics, and possible challenges.
+Furthermore, a rough sketch and discussion of the Zohmg systems different
+components will be presented. Secondly, the Zohmg system requirements will be
+featured.
 
 Throughout this chapter distinction is made between the developer: who develops
-the system, and the user: who uses the product which the developer
+the application, and the user: who uses the product which the developer
 creates. Although, in reality, they might be the same person.
@@
 database. It is possible to import analyzed data to this tool; the data is
-presented to the user as graphs. Drawbacks of the system include problems with
-new data source additions: they can sometime not be loaded because they are to
-big. If a user wants to perform a custom query, manual intervention by the
-developer is necessary. There is no way for the system to handle custom queries.
+presented to the user as graphs. Disadvantages of this system include problems
+with new data source additions: they can sometimes not be loaded because they
+are to big. If a user wants to perform a custom query, manual intervention by
+the developer is necessary. There is no way for this system to handle custom
+queries.
 
@@
 
-Reporting the analyzed is the data is the final step. This is what the user is
+Reporting the analyzed data is the final step. This is what the user is
 interested in: Viewing the analyzed data. The tool at Last.fm presents the user
@@
 a list of all available graphs. All the graphs are pre-aggregated, no support
-exist for custom queries. Enabling the user to query the aggregated data---for
+exists for custom queries. Enabling the user to query the aggregated data---for
 instance viewing the most popular web browser in a country---would open
@@
 Since the data sources at Last.fm are time-stamped logs, the constraint to only
-handle, time-series was brought upon the system. This decision was reached
-mainly to keep the implementation simple.
-
-\subsection*{System parts}
-
-The following three components form the outline of the system: importing,
+handle, time-series was brought upon the Zohmg system. This decision was
+reached mainly to keep the implementation simple.
+
+\subsection*{System components}
+
+The following three components form the outline of the Zohmg system: importing,
 storing, and exporting data.
@@
 them: web servers, scrobbling\footnote{Last.fm users can submit what music they
-listen to, this is called scrobbling.}, radio etc. This data reside on a
-commodity hardware cluster used for parallel computation. The system needs to be
-able to easily import data from this source.
+listen to, this is called scrobbling.}, radio streaming etc. This data reside
+on a commodity hardware cluster used for parallel computation. The system needs
+to be able to easily import data from this source.
 
@@
 Due to the several different data sources, there exist several different types
-of applications. Therefor the system does not implement any user interface, this
-is left for the developer to implement.
+of applications. Therefor the Zohmg system does not implement any user
+interface, this is left for the developer to implement.
 
@@
 
-The four key requirements of the system are: ease of use, speed, scalability,
-and being able to fit in the existing software and hardware eco-system.
+The four key requirements of the Zohmg system are: ease of use, speed,
+scalability, and being able to fit in the existing software and hardware
+eco-system.
 
@@
 
-Another technique is parallelizing, using a cluster of commodity hardware
+The other technique, parallelizing, uses a cluster of commodity hardware
 machines. Using commodity hardware for computation significantly reduces the
@@
 % scaling.
-In order for a computational setup, cluster or otherwise, to be efficient it
-needs to scale well, this means adding more hardware should add as close as
-possible to the hardware's potential addition to the computational power.
+In order for a computational setup, cluster or other, to be efficient it needs
+to scale well, this means adding more hardware should add as close as possible
+to the hardware's potential addition to the computational power.
 
@@
 
-The system function part discussed the system's goals and parts. The goals are
-to have a system which enables the developer to easily import and analyze new
-data sources and to be fast enough to serve the analyzed data at mouse-click
-rate to the user. The system's three core parts---importing, storing, and
-extracting data---have also been discussed.
-
-Furthermore the requirements on the system were discussed; the four key
+The system function section discussed the Zohmg system's goal and its
+components. The goal is to have a system which enables the developer to easily
+import and analyze new data sources and to be fast enough to serve the analyzed
+data at mouse-click rate to the user. The system's three core
+components---importing, storing, and extracting data---have also been
+discussed.
+
+Furthermore the requirements on the Zohmg system were discussed; the four key
 requirements are: ease of use, speed, scaling, and fitting into the existing

### 2009-05-13 doc/report/msc-report.tex
doc/report/msc-report.tex: * included Aerodyne nail crate image as table. * change "paper" to "thesis". * corrected table centering. * changed hash map to hash table. * minor change in OLAP section. * removed numbering of some sections.

@@
 above, country of origin is one such variable. A variable is generally called
-a dimension in this paper.
+a dimension in this thesis.
 
@@
 
-% image of a spreadsheet.
-%\includegraphics[width=70mm]{pictars/aerodyne-nails.png}
-%\imagecaptionplease{crates of nails exported}
+\begin{table}[h]
+    \begin{center}
+        \begin{tabular}{|l|l|l|l|}
+        \hline
+         & January & February & March \\
+        \hline
+        Canada & 120 & 140 & 120 \\
+        \hline
+        Great Britain & 60 & 50 & 50 \\
+        \hline
+        Germany & 80 & 100 & 110 \\
+        \hline
+        Sweden & 40 & 30 & 30 \\
+        \hline
+        \end{tabular}
+        \caption{Spreadsheet showing crates of nails exported.}
+    \end{center}
+\end{table}
+
+\vspace{-12pt}
 
@@
 
+\chapterquote{Some random quote might be suitable here.}{Per}
+\section*{Introduction}
+
 This chapter introduces data warehouses and the concept of data cubes. It then
@@
 
-\subsection{OLAP}
-
-OLAP, which is an acronym for Online analytical processing, is an approach to
-quickly answer multi-dimensional analytical queries.
+\subsection*{OLAP}
+
+OLAP\footnote{On-Line Analytical Processing} is an approach to quickly answer
+multi-dimensional analytical queries.
 
@@
 A data cube is a conceptual n-dimensional array of values. It aids in
-reasoning about dimensions. In the context of this paper, each dimension of
+reasoning about dimensions. In the context of this thesis, each dimension of
 the cube corresponds to an attribute of the data in question and each point
@@
 
-\subsection{Projections}
+\subsection*{Projections}
 
@@
 
-\subsection{Slice and dice}
+\subsection*{Slice and dice}
 
@@
 
+\begin{table}[h]
     \begin{center}
-    \begin{table}[h]
         \begin{tabular}{|l|l|l|l|l|}
@@
         \caption{Example HBase table.}
+    \end{center}
 \end{table}
-\end{center}
-
-\vspace{-24pt}
-
-\noindent The model above can indeed be thought of as a sparse distributed
+
+\vspace{-12pt}
+
+The model above can indeed be thought of as a sparse distributed
 multidimensional sorted map. In concept, this multidimensional map is identical
-to a nested hash map. The dimensions of this multidimensional map is mapped out
-on the row key, the column family and the column qualifier, and possibly also of
-the version timestamp.
+to a nested hash table. The dimensions of this multidimensional map is mapped
+out on the row key, the column family and the column qualifier, and possibly
+also of the version timestamp.
 

### 2009-05-13 doc/report/msc-report.tex

@@
 
-It is suitable to to exploit the parallel nature of the records by breaking up
-the data sets in pieces and to have each piece be processed separately on a
+It is suitable to exploit the parallel nature of the records by breaking up
+the data sets in pieces and to have each piece be processed on a separate
 node in a large compute cluster. It is also suitable to store the data across
@@
 
-OLAP\footnote{On-Line Analytical Processing} is an approach to quickly answer
+OLAP\footnote{On-Line Analytical Processing} is an umbrella term for systems
+built in a certain manner and that are optimized quickly answer
 multi-dimensional analytical queries.
@@
 multi-dimensional cube or a hypercube). The cube consists of numeric facts
-called measures which are categorized by dimensions. For example, the number
+-- called measures -- which are grouped by dimensions. For example, the number
 of nails sold is a measure while country, month and type of nail are
-dimensions. It is from OLAP that we have appropriated the concept of a data
+dimensions. It is from OLAP that [we] have appropriated the concept of a data
 cube.
@@
 
+% looking for a mathier explanation here.
 A data cube is a conceptual n-dimensional array of values. It aids in
@@
 with new data source additions: they can sometimes not be loaded because they
-are to big. If a user wants to perform a custom query, manual intervention by
+are too big. If a user wants to perform a custom query, manual intervention by
 the developer is necessary. There is no way for this system to handle custom
@@
 Reporting the analyzed data is the final step. This is what the user is
-interested in: Viewing the analyzed data. The tool at Last.fm presents the user
+interested in: viewing the analyzed data. The tool at Last.fm presents the user
 with a dashboard that has a selection of graphs on primary metrics; it also has
@@
 Since the data sources at Last.fm are time-stamped logs, the constraint to only
-handle, time-series was brought upon the Zohmg system. This decision was
-reached mainly to keep the implementation simple.
+handle time-series was brought upon the system. This decision was reached
+mainly to keep the implementation simple.
 
@@
 listen to, this is called scrobbling.}, radio streaming etc. This data reside
-on a commodity hardware cluster used for parallel computation. The system needs
-to be able to easily import data from this source.
+on a commodity hardware cluster used for parallel computation. The system needs to
+be able to easily import data from these logs, taking in to account both the
+locality of the data as well as the structure of the individual data records.
 
 % storage.
-Because the data sources are typically large, the storage need to be efficient.
+Because the data sources are typically large, the storage needs to be efficient.
 Storing the analyzed data can be done in different ways, for instance storing
@@
 MapReduce and BigTable?
+% thinking in simple data structures,
+% ..
 
 % export.
-Due to the several different data sources, there exist several different types
-of applications. Therefor the Zohmg system does not implement any user
-interface, this is left for the developer to implement.
+Due to the several different data sources, there exist several
+different types of applications. Therefor the system does not implement any user
+interface other than the REST API\footnote{The REST API is detailed in appendix
+X}. Making sense of the data is left for the developer.
 
@@
 The four key requirements of the Zohmg system are: ease of use, speed,
-scalability, and being able to fit in the existing software and hardware
-eco-system.
+scalability, and adoption, where adoption means being able to fit in the
+existing software and hardware eco-system.
+% dave probably knows a better word than 'fit'.
 

### 2009-05-14 doc/report/msc-report.tex
problem-oriented abstract.

@@
 
-This thesis describes a data store for multi-dimensional time-series-based
-data. The data is aggregated over many units of measurements and across
-multiple dimensions. One of these dimensions is always time.
-
-We present a data store called Zohmg. It is built on top of HBase and uses
-Hadoop to run import jobs in parallel. Zohmg stores and serves aggregates of
-measurements. The measurements are extracted taken from log files or similar
-sources with user-supplied code.
-
-The data store is optimized for speed of data retrieval: one of the design
-goals was to serve data at mouse-click rate and promote real-time data
-exploration. As such, Zohmg stores only the summarized data, not the atomic
-measurements.
-
-Similar data stores exist: they generally use a relational database system as
-their underlying data store. The novelty of our approach is to model OLAP
-cubes on top of a bigtable-like data store.
+Making sense of data at a massive scale is one of the biggest challenges we face
+at Last.fm. Interpreting patterns in user behaviour becomes a challenge when
+millions of users interact in billions of combinations; the data sets must be
+analyzed, summarized and presented visually in order to convey their meaning.
+
+This thesis describes a data store for multi-dimensional time-series-based data.
+Measurements are summarized across multiple dimensions. The data store is
+optimized for speed of data retrieval: one of the design goals is to serve data
+at mouse-click rate to promote real-time data exploration.
+
+Similar data stores exist; they generally use relational database systems as
+their backing database. The novelty of our approach is to model multidimensional
+data cubes on top of a column-oriented database, reaping the scalability
+benefits of such databases.
 

### 2009-05-14 doc/report/msc-report.tex
Procedure. sort of sucks, sort of not happy with this. please refactor at will.

@@
 
-TODO: Explain our working procedure regarding
-
-\begin{itemize}
-\item[-] Exploration
-\item[-] Prototyping
-\item[-] Real-world data
-\item[-] Litterature studies
-\item[-] Design phase
-\end{itemize}
+This section describes how we structured our work over the course of the thesis
+project. It describes how we, by working in an exploratory fashion, reached a
+good understanding of the problem at hand, and how we successively inched closer
+to a working solution by building early prototypes and iteratively improving
+upon them.
+
+The design phase, which we entered once our understanding of the problem scope
+and its theoretical solution was satisfactory, and the following implementation
+are described, followed by a summary of our literature studies.
+
+\subsection*{Beginning}
+
+% not so sure I like this section much..
+
+We entered this project with a basic understanding of the problem Last.fm wanted
+to solve but with little formal knowledge of the problem domain. We were tasked
+with building a data store that would hopefully replace the store that was
+presently in use.
+
+We met the relevant people, were introduced to the current system and were
+presented with literature in the area, namely OLAP Solutions and Whatever in
+Data Warehousing.
+
+\subsection*{Exploration by prototyping}
+
+We explored the problem domain by asking questions about the environment, the
+system we were to build and its requirements.
+
+To explore the environment, for example, we asked what data sources there are
+(web logs, scrobbling logs and radio streaming logs) and where that data is
+stored (on HDFS, it turned out). We asked how much data the system was supposed
+to handle (lots) and what the output format ought to be (any format, within
+reason).
+
+Most of the exploration was straight forward: many questions could be answered
+from observations or by logical reasoning. Others were more open-ended, such as
+deciding on a suitable output format. For the open-ended questions we would list
+reasonable options and pitt them against each other, first on the drawing board
+and eventually as prototypes.
+
+% this is a little thin. The prototypes are small pieces of code that would
+answer questions such as Is it at all possible to.. ? and What is the most
+efficient way to..?
+
+\subsection*{Design Phase}
+
+After having become familiar with the problem and its theoretical solution, it
+was time to design the software.
+
+Our initial design was ambitious: it contained, amongst other things, a query
+language and sub-components for query analyzation, to identity the most common
+queries, and for query optimization. These are likely advantageous features in a
+large data store but one can do without them.
+
+In the end, we much simplified our initial design and ended up with a design
+that was simpler and more to-the-point.
+
+\subsection*{Implementation}
+
+The implementation phase was rather small: we would simply stitch together our evolved prototypes and arrive at a solid ground to stand on: we had a system for running MapReduce jobs, storing data in HBase, etc.
+
+Writing code on top of this that performed data cube manipulation was a kinch!
+
+\subsection*{Literature studies}
+% vet inte riktigt vad jag vill saga om det har.
+
+Studying previous work in related fields brought OLAP systems to our attention.
+It was to become very influential in our theoretical understanding of the
+subject.
+
+\subsection*{Summary}
+
+This section has talked about the procedure we followed during our thesis work.
+It described our exploratory model and the prototyping we did, as well as the design phase that led up to the implementation.
 

### 2009-05-14 doc/report/msc-report.tex
doc/report/msc-report.tex: * added listings package, used for code listing. * WIP on Implementation chapter.

@@
 
-\documentclass[a4paper]{report}
+\documentclass[a4paper,10pt]{report}
 \usepackage{graphicx}
+\usepackage{listings}
 
@@
 
-\section{Configuration}
-
-The Zohmg needs configuration of the user's data sets and the environment for
-Hadoop and Dumbo.
-
-The data sets are configured using YAML\footnote{YAML Ain't a Markup Language},
-a human-readable data serializaton format. Any human-readable configure file
-method could have been chosen, e.g. INI files\footnote{A simple configuration
-mechanims where one would have lines of \texttt{variable = value}.}. YAML was
-choosen because it is is very simple for humans to read and edit, and likewise
-for machines.
+\section*{Summary}
+
+Summarize the method chapter.
+
+Working methods.
+
+Tools.
+
+Something else-or-another.
+
+\chapter{Implementation}
+
+\chapterquote{A good idea is about ten percent and implementation and hard
+work, and luck is 90 percent.}{Guy Kawasaki}
+
+\section*{Introduction}
+
+This chapter will present the implementation of the Zohmg system. For a usage
+manual see the appendix.
+
+% TODO: Do we need to talk about the storage implementation?
+
+In the first section the implementation of project creation, configuration, and
+setup, will be presented. The second section features the data import
+implementation. The third section presents the implementation of data
+interchange between Zohmg and other software but also throughout the Zohmg
+system. The fourth, and final, section presents the implementation details of
+the Zohmg system's exporting compontent.
+
+\section{Project management}
+
+A Zohmg project is necessary for using Zohmg; they are created with a management
+script, which is also used for interacting with the Zohmg system. When the user
+has created a Zohmg project directory further actions can be performed.  The
+
+Zohmg projects reside within a project directory, which contains five
+directories: \texttt{clients}, \texttt{config}, \texttt{lib}, \texttt{mappers},
+and \texttt{transformers}. The details of these directories are presented
+throughout this chapter.
+
+\subsection*{Configuration}
+
+The Zohmg system needs configuration of the user's data sets and the environment
+for Hadoop and Dumbo. When the data set files are configured, the user runs the
+management script, which executes the configuration and creates tables in HBase
+according to the configuration.
+
+\subsubsection*{Data sets}
+
+% TODO: define what a data set is, and what makes it different from a project.
+
+In this context a data set is the schema that the user defines for wanted data
+sets to be handled by Zohmg. The configuration of a data set is simply a list
+of dimensions and projections, which will layout the HBase tables according to
+the schema. This configuration will also be read during the import phase, for
+handling data to be imported.
+
+The data set files reside in the project's \texttt{config} directory. The data
+sets are configured using YAML\footnote{YAML Ain't a Markup Language}, a
+human-readable data serializaton format. Any human-readable configure file
+method could have been chosen. YAML was choosen because it is is very simple
+for humans to read and edit, and likewise for machines. The Python module
+PyYAML was used to implement the YAML configuration reading part. \cite{pyyaml}
+
+\subsubsection*{Environment}
 
@@
 Hadoop and Dumbo. Using a Python scripts for run-time variables leaves the
-problem of validation to Python.
-
-\section{Exporting data}
-
-In order to join together all the different abilities which were wanted from the
-data server, it was designed as a middleware application. Since the main goal
-was not to build a middleware framework from scratch, an existing such was used.
-
-The data export server is built with a WSGI\footnote{Web Server Gateway
-Interface} framework, it includes both middleware and application parts. The
-middleware applications are responsible for dispatching requests to the
-corresponding application or file.
-
-When a user queries the data server, JSON\footnote{JavaScript Object Notation}
-is returned. JSON was chosen because it is i simple text serialization format
-which is widely used in web development.
-
-\chapter{Implementation}
-
-\chapterquote{A good idea is about ten percent and implementation and hard
-work, and luck is 90 percent.}{Guy Kawasaki}
-
-TODO: *this should go into implementation.*
-The typical use case for Zohmg is to read one or more column-qualifiers,
-which correspond to one or more points in n-space, over a range of time.
-Therefor, the keys are made up of the time and unit of the measurement,
-and the qualifier queried for represents the point in n-space. More on
-this in the data model section.
-
-% TODO: not sure this is precisely a story about the implementation.
-% putting it here for now.
-
--- A truly magnetic data store - the idea of enabling the user --
-
-From a user's perspective, storing aggregates in Zohmg is as simple as writing
-a map function (in the MapReduce sense) and letting Zohmg run that map function
-over the data set. The user's map function is called \textit{usermap function}
-from here on. In order to let Zohmg know what output to expect from the mapper,
-the user also defines the dimensions and units of his data. After all, the user
-most likely knows his own data better than anyone else.
+problem of validation to Python. The environment configuration file resides in
+the project's \texttt{config} directory.
+
+\section{Importing}
+
+In order to import data to Zohmg, the user writes a map function in Python which
+Zohmg wraps; interpreting the output from the mapper program and storing it in
+HBase.
 
@@
 
-\begin{verbatim}
+\vspace{12pt}
+
+\begin{lstlisting}[language=Python]
    def map(key, value):
@@
       yield log.timestamp,dimensions,measurements
-\end{verbatim}
+\end{lstlisting}
 
@@
 
-\section{Configuration}
-
-\subsection{Data sets}
-
-% TODO: define what a data set is, and what makes it different from a project.
-
-The user's data sets are defined using YAML. The data set files reside in
-the project's \texttt{config} directory.
-
-\subsection{Environment}
-
-Certain paths need to be known by Zohmg in order to function, these are set
-in the config/environment.py file.
-
-\section{Importing}
-
-In order to import data to Zohmg, the user writes a mapper which Zohmg wraps;
-interpreting the output from the mapper program and storing it in HBase.
-
-Support exists for writing the mapper in Java or Python.
-
 \section{Data interchange}
@@
 
+\section{Exporting data}
+
+In order to join together all the different abilities wanted from the data
+server, it was designed as a middleware application. Middleware is software
+that acts as an intermediary.
+
+The data export server is built with a WSGI\footnote{Web Server Gateway
+Interface, a HTTP extension where a CGI-like environment is passed around.}
+framework, which includes both middleware and application parts. The middleware
+applications are responsible for dispatching requests to the corresponding
+application or file.
+WSGI middleware receives a
+WSGI request and then performs logic upon this request, before passing the
+request to a WSGI application. \cite{definitive_guide_to_pylons,paste}
+Since the main goal was not
+to build a middleware framework from scratch, an existing such was used.
+
+When a user queries the data server, JSON\footnote{JavaScript Object Notation}
+is returned. JSON was chosen because it is i simple text serialization format
+which is widely used in web development.
+
 \subsection{External}
@@
 framework---the data server dispatches to the requested application
-depending on the URI it was fed.
-
-Python Paste is a framework for building WSGI\footnote{Web Server Gateway
-Interface, a HTTP extension where a CGI-like environment is passed around.}
-applications and middleware. \cite{definitive_guide_to_pylons}
+depending on the URI it was fed. \cite{python_paste}
 
@@
 built-in web server.
-
-Middleware is software that acts as an intermediary. WSGI middleware
-receives a WSGI request and then performs logic upon this request, before
-passing the request to a WSGI application. \cite{paste}
 
@@
 
+\section*{Summary}
+
+Summary here.
+
 \chapter{Results}
@@
 \section{Discussion}
+
+Discussion here.
 
@@
 Project deployment and data import and export will be described briefly.
+
+\section{Installation}
+
+During the installation phase superuser (root) privileges are required.
+
+Install Hadoop and HBase.
+
+\begin{verbatim}
+   sh scripts/install_hadoop_hbase.sh
+\end{verbatim}
+
+The above command will install Hadoop and HBase to \texttt{/opt} by default,
+this is customizable with script arguments. Both Hadoop and HBase will be
+configured for local (standalone) mode when the installation is done. Installing
+Hadoop and HBase on a cluster requires additional attention, for
+instance passphraseless SSH\footnote{Secure Shell, an encrypted data transport
+protocol; it is commonly used for remote access.} keys and additional program
+configuration.
+
+When Hadoop and HBase is installed install Zohmg by invoking the following
+command in the root of the source tree
+
+\begin{verbatim}
+   python install.py
+\end{verbatim}
+
+Creates eggs of Zohmg system and puts them in system-wide directory for python
+packages.
+
+Zohmg files copied, scripts, documentation, library files (other's software).
 
@@
 
-% TODO: Why do we chose HBase in favour of other techniques (i.e. SQL)?
+TODO: Why do we chose HBase in favour of other techniques (i.e. SQL)?
 
@@
 
-% TODO: elaborate on reasons to why starsql does not scale
+TODO: elaborate on reasons to why starsql does not scale
 
@@
 
+\section{Electromagnetic field theory}
+
+-- A truly magnetic data store - the idea of enabling the user --
+
+From a user's perspective, storing aggregates in Zohmg is as simple as writing
+a map function (in the MapReduce sense) and letting Zohmg run that map function
+over the data set. The user's map function is called \textit{usermap function}
+from here on. In order to let Zohmg know what output to expect from the mapper,
+the user also defines the dimensions and units of his data. After all, the user
+most likely knows his own data better than anyone else.
+
+\section{Typical use case}
+
+The typical use case for Zohmg is to read one or more column-qualifiers,
+which correspond to one or more points in n-space, over a range of time.
+Therefor, the keys are made up of the time and unit of the measurement,
+and the qualifier queried for represents the point in n-space. More on
+this in the data model section.
+
 \end{document}

### 2009-05-14 doc/report/msc-report.tex
"The thesis is presented in three major sections: problem statement, theoretical solution and implementation details." - it's really not.
doc/report/msc-report.tex: * minor additions and placeholders.

@@
 benefits of such databases.
-
-The thesis is presented in three major sections: problem statement,
-theoretical solution and implementation details.
 
@@
 
+\subsection*{Last.fm}
+
+% TODO: renskriv these thoughts.
+
+Last.fm is an internet radio station focusing on content, being a musical
+encyclopedia. The web site.
+
+Last.fm lets users submit what music they listen to, called \textit{scrobbling}.
+
+Radio streams to users.
+
+The web site, radio streaming and submissions (scrobbling) generates gigabytes
+of data every day.
+
 \section{Problem description}
@@
 \chapterquote{Some random quote might be suitable here.}{Per}
+
 \section*{Introduction}
@@
 
-\section{Summary}
+\section*{Summary}
 
@@
 \chapterquote{Arrange whatever pieces come your way.}{Virgina Wolf}
+
+\section*{Introduction}
+
+Introduce the chapter here.
 

### 2009-05-15 doc/report/msc-report.tex
doc/report/msc-report.tex: WIP before pub.

@@
 
-\section{Background}
+\section*{Background}
 
@@
 
-\section{Problem description}
+\section*{Problem description}
 % Problemsbeskrivning.
@@
 
-\section{Goal}
+\section*{Goal}
 % Syftesformulering.
@@
 
-\section{Scope}
+\section*{Scope}
 % Avgransning.
@@
 
-\section{Related work}
+\section*{Related work}
 
@@
 
-\section{Overview}
+\section*{Overview}
 
@@
 % CONCEPTS IN DATA WAREHOUSING, by clark de jean
-\chapter{Concepts in Data Warehousing}
-
-\chapterquote{Some random quote might be suitable here.}{Per}
+\chapter{Concepts}
+
+\chapterquote{ I don't let go of concepts - I meet them with understanding.
+Then they let go of me.}{Byron Katie}
 
@@
 
-This chapter introduces data warehouses and the concept of data cubes. It then
-introduces common operations on data cubes: projections, slicing and dicing,
-and aggregation.
+This chapters first section presents data warehouses and the concept of data
+cubes. It then introduces common operations on data cubes: projections, slicing
+and dicing, and aggregation.
+
+Secondly, this chapter features new techniques in distributed storage and
+computing; these techniques are Google's BigTable and MapReduce.
 
@@
 At the core of any OLAP system is the OLAP cube (also known as a
-multi-dimensional cube or a hypercube). The cube consists of numeric facts
--- called measures -- which are grouped by dimensions. For example, the number
-of nails sold is a measure while country, month and type of nail are
+multi-dimensional cube or a hypercube). The cube consists of numeric
+facts---called measures---which are grouped by dimensions. For example, the
+number of nails sold is a measure while country, month and type of nail are
 dimensions. It is from OLAP that [we] have appropriated the concept of a data
@@
 
-\section{Data cubes}
+\subsection*{Data cubes}
 
@@
 
-\subsection*{Projections}
+\subsubsection*{Projections}
 
@@
 nails; the dimensions are nail type, country and month. A projection from that
-cube to one with two dimensions -- month and nail type -- would flatten the
+cube to one with two dimensions---month and nail type---would flatten the
 country dimension: the measurements in the new cube represent the global sales
@@
 
-\subsection*{Slice and dice}
+\subsubsection*{Slice and dice}
+
+% TODO: this is actually exactly the same thing as projection.
 
@@
 
-\section{Aggregates}
+\subsection*{Aggregates}
 
@@
 
-\section*{Summary}
-
-This chapter introduced Data Warehouses in general and OLAP in particular. The
-concept of data cubes was introduced by OLAP. The chapter was rounded off with
-a discussion of dimensionality reducing operations and aggregations of data
-cubes.
-
-\chapter{System Analysis}
-
-\chapterquote{Methods and means cannot be separated from the ultimate aim.}{Emma
-Goldman}
-
-\section*{Introduction}
-
-This chapter will discuss and analyze the Zohmg system, that is to be built.
-
-First the application area and the wanted system functionality are presented;
-while discussing the input and output characteristics, and possible challenges.
-Furthermore, a rough sketch and discussion of the Zohmg systems different
-components will be presented. Secondly, the Zohmg system requirements will be
-featured.
-
-Throughout this chapter distinction is made between the developer: who develops
-the application, and the user: who uses the product which the developer
-creates. Although, in reality, they might be the same person.
-
-\section{System function}
-
-Last.fm has a central statistics and metrics tool built on a relational
-database. It is possible to import analyzed data to this tool; the data is
-presented to the user as graphs. Disadvantages of this system include problems
-with new data source additions: they can sometimes not be loaded because they
-are too big. If a user wants to perform a custom query, manual intervention by
-the developer is necessary. There is no way for this system to handle custom
-queries.
-
-The data source is typically a log file; it is often a large file, up to
-gigabytes in size, of structured data. The data is structured the same way on
-each line, this makes writing a program for analyzing the entire data set as
-simple as analyzing a single line.
-
-Reporting the analyzed data is the final step. This is what the user is
-interested in: viewing the analyzed data. The tool at Last.fm presents the user
-with a dashboard that has a selection of graphs on primary metrics; it also has
-a list of all available graphs. All the graphs are pre-aggregated, no support
-exists for custom queries. Enabling the user to query the aggregated data---for
-instance viewing the most popular web browser in a country---would open
-possibilities to explore the data. During exploration interesting details could
-possibly be discovered; details which may be excluded from the pre-rendered
-aggregates.
-
-Since the data sources at Last.fm are time-stamped logs, the constraint to only
-handle time-series was brought upon the system. This decision was reached
-mainly to keep the implementation simple.
-
-\subsection*{System components}
-
-The following three components form the outline of the Zohmg system: importing,
-storing, and exporting data.
-
-At Last.fm there are several different facilities which produce logs, among
-them: web servers, scrobbling\footnote{Last.fm users can submit what music they
-listen to, this is called scrobbling.}, radio streaming etc. This data reside
-on a commodity hardware cluster used for parallel computation. The system needs to
-be able to easily import data from these logs, taking in to account both the
-locality of the data as well as the structure of the individual data records.
-
-% storage.
-Because the data sources are typically large, the storage needs to be efficient.
-Storing the analyzed data can be done in different ways, for instance storing
-the raw data or just aggregates. Storing the raw data is very expensive, since
-it would basically copy the data to the data store. Because of this the decision
-was made to only store aggregates.
-
-There are several different ways of modelling multidimensional data.
-
-TODO: How to discuss modelling of data in a general sense, without mentioning
-MapReduce and BigTable?
-% thinking in simple data structures,
-% ..
-
-% export.
-Due to the several different data sources, there exist several
-different types of applications. Therefor the system does not implement any user
-interface other than the REST API\footnote{The REST API is detailed in appendix
-X}. Making sense of the data is left for the developer.
-
-\section{System requirements}
-
-The four key requirements of the Zohmg system are: ease of use, speed,
-scalability, and adoption, where adoption means being able to fit in the
-existing software and hardware eco-system.
-% dave probably knows a better word than 'fit'.
-
-% ease of use.
-The system should have an easy facility for importing and analyzing new data
-sources, while still being fast and efficient. A simple script---or tweak of an
-already existing data import script---should be enough for importing and
-analyzing a new data source.
-
-% speed.
-Time is of essence when exporting data to the user. The user should be able to
-explore the data at mouse-click rate, not having to wait for computation. This
-is usually solved by pre-rendering the data which can be viewed or queried.
-
-Importing data requires computing power, which is a limited resource. This calls
-for the need to make the import phase efficient. Using a desktop workstation to
-process several gigabytes of data is simply impossible; it might require more
-time than the actual worth of data it is importing---for instance taking more
-than one day to import a days worth of data. There are a number of ways to solve
-this, among them: more powerful machines and parallelizing.
-
-Acquiring more powerful machines quickly becomes expensive. Even if more
-powerful machines are acquired they might not have enough power for processing
-several gigabytes of data: the data to be processed might not fit in the working
-memory.
-
-The other technique, parallelizing, uses a cluster of commodity hardware
-machines. Using commodity hardware for computation significantly reduces the
-cost, since commodity hardware is cheap; thusly preferred in most situations.
-
-% scaling.
-In order for a computational setup, cluster or other, to be efficient it needs
-to scale well, this means adding more hardware should add as close as possible
-to the hardware's potential addition to the computational power.
-
-% fit in existing eco system.
-The system needs to co-exist and integrate seamlessly with the already existing
-hardware and software eco-system at Last.fm. Thus, building on top of the
-already existing software a natural choice.
-
-\section*{Summary}
-
-This chapter has analyzed and discussed the function and requirements of the
-system to be built.
-
-The system function section discussed the Zohmg system's goal and its
-components. The goal is to have a system which enables the developer to easily
-import and analyze new data sources and to be fast enough to serve the analyzed
-data at mouse-click rate to the user. The system's three core
-components---importing, storing, and extracting data---have also been
-discussed.
-
-Furthermore the requirements on the Zohmg system were discussed; the four key
-requirements are: ease of use, speed, scaling, and fitting into the existing
-hardware and software eco-system.
-
-\chapter{Method}
-
-\chapterquote{Arrange whatever pieces come your way.}{Virgina Wolf}
-
-\section*{Introduction}
-
-Introduce the chapter here.
-
-\section{Procedure}
-
-This section describes how we structured our work over the course of the thesis
-project. It describes how we, by working in an exploratory fashion, reached a
-good understanding of the problem at hand, and how we successively inched closer
-to a working solution by building early prototypes and iteratively improving
-upon them.
-
-The design phase, which we entered once our understanding of the problem scope
-and its theoretical solution was satisfactory, and the following implementation
-are described, followed by a summary of our literature studies.
-
-\subsection*{Beginning}
-
-% not so sure I like this section much..
-
-We entered this project with a basic understanding of the problem Last.fm wanted
-to solve but with little formal knowledge of the problem domain. We were tasked
-with building a data store that would hopefully replace the store that was
-presently in use.
-
-We met the relevant people, were introduced to the current system and were
-presented with literature in the area, namely OLAP Solutions and Whatever in
-Data Warehousing.
-
-\subsection*{Exploration by prototyping}
-
-We explored the problem domain by asking questions about the environment, the
-system we were to build and its requirements.
-
-To explore the environment, for example, we asked what data sources there are
-(web logs, scrobbling logs and radio streaming logs) and where that data is
-stored (on HDFS, it turned out). We asked how much data the system was supposed
-to handle (lots) and what the output format ought to be (any format, within
-reason).
-
-Most of the exploration was straight forward: many questions could be answered
-from observations or by logical reasoning. Others were more open-ended, such as
-deciding on a suitable output format. For the open-ended questions we would list
-reasonable options and pitt them against each other, first on the drawing board
-and eventually as prototypes.
-
-% this is a little thin. The prototypes are small pieces of code that would
-answer questions such as Is it at all possible to.. ? and What is the most
-efficient way to..?
-
-\subsection*{Design Phase}
-
-After having become familiar with the problem and its theoretical solution, it
-was time to design the software.
-
-Our initial design was ambitious: it contained, amongst other things, a query
-language and sub-components for query analyzation, to identity the most common
-queries, and for query optimization. These are likely advantageous features in a
-large data store but one can do without them.
-
-In the end, we much simplified our initial design and ended up with a design
-that was simpler and more to-the-point.
-
-\subsection*{Implementation}
-
-The implementation phase was rather small: we would simply stitch together our evolved prototypes and arrive at a solid ground to stand on: we had a system for running MapReduce jobs, storing data in HBase, etc.
-
-Writing code on top of this that performed data cube manipulation was a kinch!
-
-\subsection*{Literature studies}
-% vet inte riktigt vad jag vill saga om det har.
-
-Studying previous work in related fields brought OLAP systems to our attention.
-It was to become very influential in our theoretical understanding of the
-subject.
-
-\subsection*{Summary}
-
-This section has talked about the procedure we followed during our thesis work.
-It described our exploratory model and the prototyping we did, as well as the design phase that led up to the implementation.
-
-\section{Apache Hadoop}
-
-Hadoop is a free software implementation of Google's MapReduce-framework; it is
-written in Java and is part of the Apache Software Foundation. Hadoop is a
-fairly mature piece of software and is used in production at numerous companies,
-among them Last.fm.
-\cite{hadoop}
-
-\subsection{MapReduce}
-
-MapReduce is a framework for running jobs in parallel across multiple
-machines. As the name implies, MapReduce is made up of two parts or
-phases: map and reduce. The map phase consumes one piece of input at a
-time (a piece of input can be a line from a log file, for example), and
-emits an intermediate key and value. These keys and values are then
-processed by the reduce phase, which emits the final key-value pairs.
-
-MapReduce works by splitting the input, which commonly is in the size
-range of many gigabytes, into \textit{n} parts and lets each node of the
-computing cluster work on one or more parts. The mapper on each node emits
-key-value pairs which are fed into the reduce phase. The final output is
-collated, usually on a distributed file system.
-
-\subsection*{Hadoop Distributed File System}
-
-HDFS is a free software implementation of the GFS \cite{gfs}.
-
-The goal of HDFS is to store large data sets while being able to handle
-hardware failure and being deployed in heterogenous hardware and software
-eco systems. Additionally, the Hadoop and HDFS interaction is designed with
-the notion that it is cheaper to move the computation, MapReduce programs,
-to the data than the other way around.
-
-Every Hadoop node can be formatted with HDFS, this reserves disk space on
-the node to be used for HDFS. The HDFS container resides on the nodes
-general purpose file system. It is also possible to use HDFS as a
-stand-alone general purpose distributed file system, without Hadoop.
-
-The default block size on HDFS 64 MB, significantly larger than
-file systems used for hard drives. The motivation is that applications
-that use HDFS are not general purpose applications which run on general
-purpose file systems, but batch applications which read, write, or both,
-large amounts of data.
-
-\subsection*{Hadoop streaming}
-
-Any program that reads and outputs to the file pointers \texttt{stdin} and
-\texttt{stdout}, respectively, can be used as MapReduce programs with
-Hadoop Streaming. This makes it possible to use any shell script or
-program which inputs and outputs this way as a MapReduce program.
-
-\subsubsection*{Dumbo}
-
-Dumbo is a Python framework for writing Hadoop programs in Python; it exploits
-the Hadoop Streaming Mode to do so. It is used extensively at Last.fm for
-writing short prototypes. \cite{dumbo}
-
-\section{Apache HBase}
-
-HBase is a free software implementation of Google's BigTable. \cite{bigtable} It
-is a sub-project of Hadoop, written in Java and a part of the Apache Software
-Foundation. HBase is still in its infancy and few organizations use it for
-mission-critical applications. \cite{hbase}
+\section{New techniques}
+
+Classic data warehouses use relational databases as storage. A star or snowflake
+schema would be used to modell the dimensions.
+
+New techniques that have appeared include column-oriented, schema free
+databases. Google's BigTable is such a database. Another new technique for
+distributed processing is Google's MapReduce.
 
@@
 
-\subsection{Data model}
+\subsubsection{Data model}
 
@@
 
+% TODO: fix positioning.
 \begin{table}[h]
@@
 
+\vspace{12pt}
+
+\begin{lstlisting}[caption=BigTable data model illustration.,captionpos=b]
+   (row key,column key,timestamp) -> value
+\end{lstlisting}
+
+\subsection{MapReduce}
+
+MapReduce is a framework for running jobs in parallel across multiple
+machines. As the name implies, MapReduce is made up of two parts or
+phases: map and reduce. The map phase consumes one piece of input at a
+time (a piece of input can be a line from a log file, for example), and
+emits an intermediate key and value. These keys and values are then
+processed by the reduce phase, which emits the final key-value pairs.
+
+MapReduce works by splitting the input, which commonly is in the size
+range of many gigabytes, into \textit{n} parts and lets each node of the
+computing cluster work on one or more parts. The mapper on each node emits
+key-value pairs which are fed into the reduce phase. The final output is
+collated, usually on a distributed file system.
+
+\section*{Summary}
+
+This chapter introduced Data Warehouses in general and OLAP in particular. The
+concept of data cubes was introduced by OLAP. Dimensionality reducing operations
+and aggregations of data cubes was also introduced in this chapter.
+
+The chapter rounded of with a presentation of new techniques for distributed
+storage and processing; namely Google's BigTable and MapReduce techniques.
+
+\chapter{System Analysis}
+
+\chapterquote{Methods and means cannot be separated from the ultimate aim.}{Emma
+Goldman}
+
+\section*{Introduction}
+
+This chapter will discuss and analyze the Zohmg system, that is to be built.
+
+First the application area and the wanted system functionality are presented;
+while discussing the input and output characteristics, and possible challenges.
+Furthermore, a rough sketch and discussion of the Zohmg systems different
+components will be presented. Secondly, the Zohmg system requirements will be
+featured.
+
+Throughout this chapter distinction is made between the developer: who develops
+the application, and the user: who uses the product which the developer
+creates. Although, in reality, they might be the same person.
+
+\section{System function}
+
+Last.fm has a central statistics and metrics tool built on a relational
+database. It is possible to import analyzed data to this tool; the data is
+presented to the user as graphs. Disadvantages of this system include problems
+with new data source additions: they can sometimes not be loaded because they
+are too big. If a user wants to perform a custom query, manual intervention by
+the developer is necessary. There is no way for this system to handle custom
+queries.
+
+The data source is typically a log file; it is often a large file, up to
+gigabytes in size, of structured data. The data is structured the same way on
+each line, this makes writing a program for analyzing the entire data set as
+simple as analyzing a single line.
+
+Reporting the analyzed data is the final step. This is what the user is
+interested in: viewing the analyzed data. The tool at Last.fm presents the user
+with a dashboard that has a selection of graphs on primary metrics; it also has
+a list of all available graphs. All the graphs are pre-aggregated, no support
+exists for custom queries. Enabling the user to query the aggregated data---for
+instance viewing the most popular web browser in a country---would open
+possibilities to explore the data. During exploration interesting details could
+possibly be discovered; details which may be excluded from the pre-rendered
+aggregates.
+
+Since the data sources at Last.fm are time-stamped logs, the constraint to only
+handle time-series was brought upon the system. This decision was reached
+mainly to keep the implementation simple.
+
+\subsection*{System components}
+
+The following three components form the outline of the Zohmg system: importing,
+storing, and exporting data.
+
+At Last.fm there are several different facilities which produce logs, among
+them: web servers, scrobbling\footnote{Last.fm users can submit what music they
+listen to, this is called scrobbling.}, radio streaming etc. This data reside
+on a commodity hardware cluster used for parallel computation. The system needs to
+be able to easily import data from these logs, taking in to account both the
+locality of the data as well as the structure of the individual data records.
+
+% storage.
+Because the data sources are typically large, the storage needs to be efficient.
+Storing the analyzed data can be done in different ways, for instance storing
+the raw data or just aggregates. Storing the raw data is very expensive, since
+it would basically copy the data to the data store. Because of this the decision
+was made to only store aggregates.
+
+There are several different ways of modelling multidimensional data.
+
+TODO: How to discuss modelling of data in a general sense, without mentioning
+MapReduce and BigTable?
+% thinking in simple data structures,
+% ..
+
+% export.
+Due to the several different data sources, there exist several
+different types of applications. Therefor the system does not implement any user
+interface other than the REST API\footnote{The REST API is detailed in appendix
+X}. Making sense of the data is left for the developer.
+
+\section{System requirements}
+
+The four key requirements of the Zohmg system are: ease of use, speed,
+scalability, and adoption, where adoption means being able to fit in the
+existing software and hardware eco-system.
+% dave probably knows a better word than 'fit'.
+
+% ease of use.
+The system should have an easy facility for importing and analyzing new data
+sources, while still being fast and efficient. A simple script---or tweak of an
+already existing data import script---should be enough for importing and
+analyzing a new data source.
+
+% speed.
+Time is of essence when exporting data to the user. The user should be able to
+explore the data at mouse-click rate, not having to wait for computation. This
+is usually solved by pre-rendering the data which can be viewed or queried.
+
+Importing data requires computing power, which is a limited resource. This calls
+for the need to make the import phase efficient. Using a desktop workstation to
+process several gigabytes of data is simply impossible; it might require more
+time than the actual worth of data it is importing---for instance taking more
+than one day to import a days worth of data. There are a number of ways to solve
+this, among them: more powerful machines and parallelizing.
+
+Acquiring more powerful machines quickly becomes expensive. Even if more
+powerful machines are acquired they might not have enough power for processing
+several gigabytes of data: the data to be processed might not fit in the working
+memory.
+
+The other technique, parallelizing, uses a cluster of commodity hardware
+machines. Using commodity hardware for computation significantly reduces the
+cost, since commodity hardware is cheap; thusly preferred in most situations.
+
+% scaling.
+In order for a computational setup, cluster or other, to be efficient it needs
+to scale well, this means adding more hardware should add as close as possible
+to the hardware's potential addition to the computational power.
+
+% fit in existing eco system.
+The system needs to co-exist and integrate seamlessly with the already existing
+hardware and software eco-system at Last.fm. Thus, building on top of the
+already existing software a natural choice.
+
+\section*{Summary}
+
+This chapter has analyzed and discussed the function and requirements of the
+system to be built.
+
+The system function section discussed the Zohmg system's goal and its
+components. The goal is to have a system which enables the developer to easily
+import and analyze new data sources and to be fast enough to serve the analyzed
+data at mouse-click rate to the user. The system's three core
+components---importing, storing, and extracting data---have also been
+discussed.
+
+Furthermore the requirements on the Zohmg system were discussed; the four key
+requirements are: ease of use, speed, scaling, and fitting into the existing
+hardware and software eco-system.
+
+\chapter{Method}
+
+\chapterquote{Arrange whatever pieces come your way.}{Virgina Wolf}
+
+\section*{Introduction}
+
+The first section in this chapter will present the work procedure during the
+master thesis. The following chapters will discuss selected tools and their
+impact on design and implementation choices for the Zohmg system.
+
+\section{Procedure}
+
+% TODO: generally use passive tense.
+% TODO: renskriv.
+
+This section describes how we structured our work over the course of the thesis
+project. It describes how we, by working in an exploratory fashion, reached a
+good understanding of the problem at hand, and how we successively inched closer
+to a working solution by building early prototypes and iteratively improving
+upon them.
+
+The design phase, which we entered once our understanding of the problem scope
+and its theoretical solution was satisfactory, and the following implementation
+are described, followed by a summary of our literature studies.
+
+\subsection*{Beginning}
+
+% not so sure I like this section much..
+
+We entered this project with a basic understanding of the problem Last.fm wanted
+to solve but with little formal knowledge of the problem domain. We were tasked
+with building a data store that would hopefully replace the store that was
+presently in use.
+
+We met the relevant people, were introduced to the current system and were
+presented with literature in the area, namely OLAP Solutions and Whatever in
+Data Warehousing.
+
+\subsection*{Exploration by prototyping}
+
+The problem domain was explored by asking, the system to be built and the
+requirements upon it.
+
+To explore the environment, for example, we asked what data sources there are
+(web logs, scrobbling logs and radio streaming logs) and where that data is
+stored (on HDFS, it turned out). We asked how much data the system was supposed
+to handle (lots) and what the output format ought to be (any format, within
+reason).
+
+Most of the exploration was straight forward: many questions could be answered
+from observations or by logical reasoning. Others were more open-ended, such as
+deciding on a suitable output format. For the open-ended questions we would list
+reasonable options and pitt them against each other, first on the drawing board
+and eventually as prototypes.
+
+% this is a little thin. The prototypes are small pieces of code that would
+TODO: answer questions such as Is it at all possible to.. ? and What is the most
+efficient way to..?
+
+\subsubsection*{Literature studies}
+% vet inte riktigt vad jag vill saga om det har.
+
+Studying previous work in related fields brought OLAP systems to our attention.
+It was to become very influential in our theoretical understanding of the
+subject.
+
+\subsubsection*{Python}
+
+Python is a general-purpose high-level multi-paradigm---amongst others:
+object-oriented, imperative and functional---scripting language. \cite{python}
+
+Because there existed a Python module for writing MapReduce programs (Dumbo,
+further later in this chapter), Python was a natural choice during the design
+and exploration phases. The main reason was the rapid development of Python
+programs, several times shorter than their Java counter parts made them faster
+to develop.
+
+\subsection*{Design Phase}
+
+After having become familiar with the problem and its theoretical solution, it
+was time to design the software.
+
+Our initial design was ambitious: it contained, amongst other things, a query
+language and sub-components for query analyzation, to identity the most common
+queries, and for query optimization. These are likely advantageous features in a
+large data store but one can do without them.
+
+In the end, we much simplified our initial design and ended up with a design
+that was simpler and more to-the-point.
+
+\subsection*{Implementation}
+
+The implementation phase was rather small: we would simply stitch together our evolved prototypes and arrive at a solid ground to stand on: we had a system for running MapReduce jobs, storing data in HBase, etc.
+
+Writing code on top of this that performed data cube manipulation was a kinch!
+
+\section{Apache Hadoop}
+
+Hadoop is a free software implementation of Google's MapReduce-framework; it is
+written in Java, and is part of the Apache Software Foundation. Hadoop is a
+fairly mature piece of software and is used in production at numerous companies,
+among them Last.fm. \cite{hadoop}
+
+\subsection*{Hadoop Distributed File System}
+
+HDFS is a free software implementation of the GFS\footnote{Google File System}.
+\cite{gfs}
+
+The goal of HDFS is to store large data sets while being able to handle
+hardware failure and being deployed in heterogenous hardware and software
+eco systems. Additionally, the Hadoop and HDFS interaction is designed with
+the notion that it is cheaper to move the computation, MapReduce programs,
+to the data than the other way around.
+
+Every Hadoop node can be formatted with HDFS, this reserves disk space on
+the node to be used for HDFS. The HDFS container resides on the nodes
+general purpose file system. It is also possible to use HDFS as a
+stand-alone general purpose distributed file system, without Hadoop.
+
+The default block size on HDFS 64 MB, significantly larger than
+file systems used for hard drives. The motivation is that applications
+that use HDFS are not general purpose applications which run on general
+purpose file systems, but batch applications which read, write, or both,
+large amounts of data.
+
+\subsection*{Hadoop streaming}
+
+Any program that reads and outputs to the file pointers \texttt{stdin} and
+\texttt{stdout}, respectively, can be used as MapReduce programs with
+Hadoop Streaming. This makes it possible to use any shell script or
+program which inputs and outputs this way as a MapReduce program.
+
+\subsubsection*{Dumbo}
+
+Dumbo is a Python framework for writing Hadoop programs in Python; it exploits
+the Hadoop Streaming Mode to do so. It is used extensively at Last.fm for
+writing short prototypes. \cite{dumbo}
+
+\section{Apache HBase}
+
+HBase is a free software implementation of Google's BigTable. \cite{bigtable} It
+is a sub-project of Hadoop, written in Java and a part of the Apache Software
+Foundation. HBase is still in its infancy and few organizations use it for
+mission-critical applications. \cite{hbase}
+
 % TODO> why is this Tool & BigTable-subsection called Dimensions?
@@
 
-\section{Python}
-
-Python is a general-purpose high-level multi-paradigm---amongst others:
-object-oriented, imperative and functional---scripting language.
-\cite{python}
-
-Because there existed a Python module for writing MapReduce programs (Dumbo),
-Python was a natural choice during the design and exploration phases. The main
-reason was the rapid development of Python programs, several times shorter than
-their Java counter parts made them faster to develop.
-
 \section*{Summary}
 
-Summarize the method chapter.
-
-Working methods.
-
-Tools.
-
-Something else-or-another.
+The first section of this chapter talked about the procedure during the thesis
+work. It described the exploratory model and the prototyping performed, as well
+as the design phase that led up to the implementation.
+
+The following sections presented and discussed the selected tools; they also
+discussed the impact on design and implementation choices.
 
@@
 
-\begin{lstlisting}[language=Python]
+\begin{lstlisting}[language=Python,caption=An example map function.,captionpos=b]
    def map(key, value):
@@
 
+% TODO: analysis.
 - Something about cardinality -
@@
 
+% TODO: method.
 -- Limits on cardinality --
@@
 
+% TODO: implementation.
 -- The implementation of all that --
@@
 
+% TODO: data interchange.
 The reducer needs to communicate to the OutputReader under what rowkey,
@@
 framework---the data server dispatches to the requested application
-depending on the URI it was fed. \cite{python_paste}
+depending on the URL it was fed. \cite{python_paste}
 
@@
 
-Middleware which dispatches to applications based on the URI. The URLParser
-removes the first part of the URI and tries to find a matching application
-to relay the request to. For instance, consider the URI
-\texttt{/client/file.html}. The first part---\texttt{client}---will be
-removed and the request \texttt{/file.html} will be sent to the application
-named \texttt{client.py}.
+The URLParser class is middleware that comes with Python Paste; its purpose is
+to dispatch to applications based on the URL. The URLParser removes the first
+part of the URL and tries to find a matching application to relay the request
+to. For instance, consider the URL \texttt{/client/file.html}. The first
+part---\texttt{client}---will be removed and the request \texttt{/file.html}
+will be sent to the application named \texttt{client.py}.
 
@@
 
-Dispatches request to serve static files from the project's
-\texttt{clients} directory. Custom clients---which interact with Zohmg---can
-be built with HTML and JavaScript, see the examples.
+This is a custom application built for Zohmg. It dispatches request to serve
+static files from the project's \texttt{clients} directory. Custom
+clients---which interact with Zohmg---can be built with HTML and JavaScript, see
+the examples.
 
@@
 
-% ?
-Fetches data from HBase via its Thrift interface. The data is served to the
-user as JSONP\footnote{JSON with padding: JSON extension which adds name of
-callback function as an argument to the function itself.}.
+The data serving part of Zohmg fetches data from HBase via its Thrift interface.
+The user submits a scan query with row range, unit, and dimensions, and all the
+wanted values associated with these variables.
+
+\begin{verbatim}
+   http://localhost:8086/data/?t0=20090101
+   &t1=20090420&unit=pageviews
+   &d0=country&d0v=SE
+\end{verbatim}
+
+The above query would return the pageviews for SE between 1 January
+2009 and 4 April 2009. The results from the scan is served to the user as
+JSONP\footnote{JSON with padding: JSON extension which adds name of callback
+function as an argument to the function itself.}.
 
@@
 
+Proper support for multiple data sets in a Zohmg project.
+
 \subsection{Metadata}
@@
 
-\section{Installation}
+\section*{Installation}
 
@@
 
-\section{Deployment}
+\section*{Deployment}
 
@@
 
-\section{Import}
+\section*{Import}
 
@@
 
-\section{Export}
+\section*{Export}
 
@@
 
-\noindent The above query would return the pageviews for SE between 1 January
+The above query would return the pageviews for SE between 1 January
 2009 and 4 April 2009.
@@
 structured data and transmitting it over network connection.
+
+ \\
+
+\textbf{OLAP} &
+An acronym for On-Line Analytical Processing. It is a subdivision of the broader
+category \textit{business logic}. OLAP is an approach to quickly find answer to
+user queries.
 

### 2009-05-18 doc/report/msc-report.tex
doc/report/msc-report.tex: * checked a lot of dave's comments. * added apache log example, maybe needs appropriate placing. * work on implementation chapter (WIP).

@@
 \chapter{Introduction}
+
+% TODO: placeholder quote.
+\chapterquote{One day some of the kids from my neighbourhood carried my mother's
+groceries all the way home---you know why? It was out of respect!}{Travis
+Bickle}
 
@@
 
-\section{New techniques}
+\section{New Techniques}
 
@@
 BigTable is a data store designed for traversal of very large data sets.
-Time-stamped cells inside column-families in rows. The row has a key and any
-number of cells attached to it. The rows in a table are sorted by the key.
-The typical use-case of BigTable is to scan a range of rows. It is therefor
-important to set up the row keys so that related data is close to each other.
-For example, if it makes sense to traverse the data in chronological order, the
-key might contain a representation of the data's timestamp. The classic example
-is to have the reversed domain name (i.e. com.google.www) as the row key, which
-means that all sub-domains of any domain are next to each other.
-
-\subsubsection{Data model}
+BigTable's cells are basically indexed by row key, column key, and a time-stamp.
+Each row has a key and any number of cells attached to it. The rows in a table
+are sorted by their keys. The typical use-case of BigTable is to scan a range of
+rows. It is therefor important to set up the row keys so that related data is
+close to each other. For example, if it makes sense to traverse the data in
+chronological order, the key might contain a representation of the data's
+timestamp. The classic example is to have the reversed domain name (i.e.
+com.google.www) as the row key, which means that all sub-domains of any domain
+are next to each other (see table 2.1).
+
+\subsubsection{Data Model}
+
+BigTable's data model can indeed be thought of as a sparse distributed
+multidimensional sorted map. In concept, this multidimensional map is identical
+to a nested hash table. The dimensions of this multidimensional map is mapped
+out on the row key, the column family and the column qualifier, and possibly
+also of the version timestamp.
+
+\vspace{12pt}
+
+\begin{lstlisting}[caption=BigTable data model illustration.,captionpos=b]
+   (row key,column key,timestamp) -> value
+\end{lstlisting}
+
+\vspace{12pt}
 
 The data model is a simple one: Rows of data are stored in labeled tables. Each
-row is indexed with a row key, a column key, and a timestamp. The columns are
-of the form "column-family:qualifier", where the column-family is one of a
-number of fixed column-families defined by the table, and the qualifier is an
-arbitrary string specified by the application. In effect, column families are
+row is indexed with a row key, a column key, and a timestamp. The columns have
+the form "column-family:qualifier", where the column-family is one of a number
+of fixed column-families defined by the table's schema, and the qualifier is an
+arbitrary string specified by the application. In effect, column-families are
 pre-announced when creating the table while qualifiers are not. The contents of
@@
 
-\vspace{-12pt}
-
-The model above can indeed be thought of as a sparse distributed
-multidimensional sorted map. In concept, this multidimensional map is identical
-to a nested hash table. The dimensions of this multidimensional map is mapped
-out on the row key, the column family and the column qualifier, and possibly
-also of the version timestamp.
-
-\vspace{12pt}
-
-\begin{lstlisting}[caption=BigTable data model illustration.,captionpos=b]
-   (row key,column key,timestamp) -> value
-\end{lstlisting}
-
 \subsection{MapReduce}
@@
 
-\section{System function}
+\section{System Function}
 
@@
 % thinking in simple data structures,
-% ..
-
-% export.
-Due to the several different data sources, there exist several
-different types of applications. Therefor the system does not implement any user
-interface other than the REST API\footnote{The REST API is detailed in appendix
-X}. Making sense of the data is left for the developer.
-
-\section{System requirements}
-
-The four key requirements of the Zohmg system are: ease of use, speed,
-scalability, and adoption, where adoption means being able to fit in the
-existing software and hardware eco-system.
-% dave probably knows a better word than 'fit'.
-
-% ease of use.
-The system should have an easy facility for importing and analyzing new data
-sources, while still being fast and efficient. A simple script---or tweak of an
-already existing data import script---should be enough for importing and
-analyzing a new data source.
-
-% speed.
-Time is of essence when exporting data to the user. The user should be able to
-explore the data at mouse-click rate, not having to wait for computation. This
-is usually solved by pre-rendering the data which can be viewed or queried.
-
-Importing data requires computing power, which is a limited resource. This calls
-for the need to make the import phase efficient. Using a desktop workstation to
-process several gigabytes of data is simply impossible; it might require more
-time than the actual worth of data it is importing---for instance taking more
-than one day to import a days worth of data. There are a number of ways to solve
-this, among them: more powerful machines and parallelizing.
-
-Acquiring more powerful machines quickly becomes expensive. Even if more
-powerful machines are acquired they might not have enough power for processing
-several gigabytes of data: the data to be processed might not fit in the working
-memory.
-
-The other technique, parallelizing, uses a cluster of commodity hardware
-machines. Using commodity hardware for computation significantly reduces the
-cost, since commodity hardware is cheap; thusly preferred in most situations.
-
-% scaling.
-In order for a computational setup, cluster or other, to be efficient it needs
-to scale well, this means adding more hardware should add as close as possible
-to the hardware's potential addition to the computational power.
-
-% fit in existing eco system.
-The system needs to co-exist and integrate seamlessly with the already existing
-hardware and software eco-system at Last.fm. Thus, building on top of the
-already existing software a natural choice.
-
-\section*{Summary}
-
-This chapter has analyzed and discussed the function and requirements of the
-system to be built.
-
-The system function section discussed the Zohmg system's goal and its
-components. The goal is to have a system which enables the developer to easily
-import and analyze new data sources and to be fast enough to serve the analyzed
-data at mouse-click rate to the user. The system's three core
-components---importing, storing, and extracting data---have also been
-discussed.
-
-Furthermore the requirements on the Zohmg system were discussed; the four key
-requirements are: ease of use, speed, scaling, and fitting into the existing
-hardware and software eco-system.
-
-\chapter{Method}
-
-\chapterquote{Arrange whatever pieces come your way.}{Virgina Wolf}
-
-\section*{Introduction}
-
-The first section in this chapter will present the work procedure during the
-master thesis. The following chapters will discuss selected tools and their
-impact on design and implementation choices for the Zohmg system.
-
-\section{Procedure}
-
-% TODO: generally use passive tense.
-% TODO: renskriv.
-
-This section describes how we structured our work over the course of the thesis
-project. It describes how we, by working in an exploratory fashion, reached a
-good understanding of the problem at hand, and how we successively inched closer
-to a working solution by building early prototypes and iteratively improving
-upon them.
-
-The design phase, which we entered once our understanding of the problem scope
-and its theoretical solution was satisfactory, and the following implementation
-are described, followed by a summary of our literature studies.
-
-\subsection*{Beginning}
-
-% not so sure I like this section much..
-
-We entered this project with a basic understanding of the problem Last.fm wanted
-to solve but with little formal knowledge of the problem domain. We were tasked
-with building a data store that would hopefully replace the store that was
-presently in use.
-
-We met the relevant people, were introduced to the current system and were
-presented with literature in the area, namely OLAP Solutions and Whatever in
-Data Warehousing.
-
-\subsection*{Exploration by prototyping}
-
-The problem domain was explored by asking, the system to be built and the
-requirements upon it.
-
-To explore the environment, for example, we asked what data sources there are
-(web logs, scrobbling logs and radio streaming logs) and where that data is
-stored (on HDFS, it turned out). We asked how much data the system was supposed
-to handle (lots) and what the output format ought to be (any format, within
-reason).
-
-Most of the exploration was straight forward: many questions could be answered
-from observations or by logical reasoning. Others were more open-ended, such as
-deciding on a suitable output format. For the open-ended questions we would list
-reasonable options and pitt them against each other, first on the drawing board
-and eventually as prototypes.
-
-% this is a little thin. The prototypes are small pieces of code that would
-TODO: answer questions such as Is it at all possible to.. ? and What is the most
-efficient way to..?
-
-\subsubsection*{Literature studies}
-% vet inte riktigt vad jag vill saga om det har.
-
-Studying previous work in related fields brought OLAP systems to our attention.
-It was to become very influential in our theoretical understanding of the
-subject.
-
-\subsubsection*{Python}
-
-Python is a general-purpose high-level multi-paradigm---amongst others:
-object-oriented, imperative and functional---scripting language. \cite{python}
-
-Because there existed a Python module for writing MapReduce programs (Dumbo,
-further later in this chapter), Python was a natural choice during the design
-and exploration phases. The main reason was the rapid development of Python
-programs, several times shorter than their Java counter parts made them faster
-to develop.
-
-\subsection*{Design Phase}
-
-After having become familiar with the problem and its theoretical solution, it
-was time to design the software.
-
-Our initial design was ambitious: it contained, amongst other things, a query
-language and sub-components for query analyzation, to identity the most common
-queries, and for query optimization. These are likely advantageous features in a
-large data store but one can do without them.
-
-In the end, we much simplified our initial design and ended up with a design
-that was simpler and more to-the-point.
-
-\subsection*{Implementation}
-
-The implementation phase was rather small: we would simply stitch together our evolved prototypes and arrive at a solid ground to stand on: we had a system for running MapReduce jobs, storing data in HBase, etc.
-
-Writing code on top of this that performed data cube manipulation was a kinch!
-
-\section{Apache Hadoop}
-
-Hadoop is a free software implementation of Google's MapReduce-framework; it is
-written in Java, and is part of the Apache Software Foundation. Hadoop is a
-fairly mature piece of software and is used in production at numerous companies,
-among them Last.fm. \cite{hadoop}
-
-\subsection*{Hadoop Distributed File System}
-
-HDFS is a free software implementation of the GFS\footnote{Google File System}.
-\cite{gfs}
-
-The goal of HDFS is to store large data sets while being able to handle
-hardware failure and being deployed in heterogenous hardware and software
-eco systems. Additionally, the Hadoop and HDFS interaction is designed with
-the notion that it is cheaper to move the computation, MapReduce programs,
-to the data than the other way around.
-
-Every Hadoop node can be formatted with HDFS, this reserves disk space on
-the node to be used for HDFS. The HDFS container resides on the nodes
-general purpose file system. It is also possible to use HDFS as a
-stand-alone general purpose distributed file system, without Hadoop.
-
-The default block size on HDFS 64 MB, significantly larger than
-file systems used for hard drives. The motivation is that applications
-that use HDFS are not general purpose applications which run on general
-purpose file systems, but batch applications which read, write, or both,
-large amounts of data.
-
-\subsection*{Hadoop streaming}
-
-Any program that reads and outputs to the file pointers \texttt{stdin} and
-\texttt{stdout}, respectively, can be used as MapReduce programs with
-Hadoop Streaming. This makes it possible to use any shell script or
-program which inputs and outputs this way as a MapReduce program.
-
-\subsubsection*{Dumbo}
-
-Dumbo is a Python framework for writing Hadoop programs in Python; it exploits
-the Hadoop Streaming Mode to do so. It is used extensively at Last.fm for
-writing short prototypes. \cite{dumbo}
-
-\section{Apache HBase}
-
-HBase is a free software implementation of Google's BigTable. \cite{bigtable} It
-is a sub-project of Hadoop, written in Java and a part of the Apache Software
-Foundation. HBase is still in its infancy and few organizations use it for
-mission-critical applications. \cite{hbase}
-
-% TODO> why is this Tool & BigTable-subsection called Dimensions?
-% Disambiguation needed. -fredrik
-\subsection*{Physical storage}
-
-Each column-family is stored as a single file in HBase; it is then subdivided
-into column-qualifiers within the same file. This is important to consider when
-designing application and schemas, since reading from several column-families
-will open several files.
-
-\subsection*{Sparsity}
-
-When many cells are empty, the table is said to be sparse. This can potentially
-be a problem, because even sparse cells may cost storage space.  BigTable and
-HBase, however, don't store empty cells; as consequence sparsity is not
-associated with a great storage cost.
-\cite{olap_data_scalability,olap_solutions}
-
-\subsection*{Hierarchies}
-
-Depending on how the row keys are formatted different, hierarchies are created.
-The order in which the data is stored is important, it defines how effective
-scanning data will be. Sparsity also comes into play concerning effective
-scanning. In the case of having thousand rows and requesting ten out of these,
-then only one por cent of the data is interesting. If a scan would be required
-to visit all the thousand rows a lot of rows are skipped. Skipping rows is
-expensive in that sense that they are but the result is thrown away. The goal is
-to push this cost to a bare minimum, using as many of the rows as possible.
-
-The minimum cost of scanning data is achieved by ordering the data so that
-a scan uses every visited row. In the case with thousand rows of which ten
-are wanted, would mean that ten rows scan are scanned of which all are
-used.
-
-\subsection*{Thrift}
-
-HBase has a Thrift server which serves as the interface to languages other than
-Java---for instance Python.
-
-Thrift is a remote procedure call framework for building scalable cross-language
-services. Thrift combines a software stack with a code generation engine to
-build services that work seamlessly between a wide range of programming and
-scripting languages.
-
-Thrift was originally developed at Facebook and in April 2007 it was released as
-free software. It entered the Apache Incubator in May 2008. \cite{thrift}
-
-\section{Dependability}
-
-Regarding the infrastructure, both Hadoop and HBase have high availability,
-reliability, and scalability.
-
-Both Hadoop and HBase are distributed systems; making them highly available out
-of the box. Although the Job Tracker\footnote{All client applications are
-submitted to Hadoop's Job Tracker.} is a single point of failure for Hadoop.
-The HDFS NameNode\footnote{A unique server which keeps the directory tree of all
-files in HDFS.} is also a single point of failure. However, a secondary HDFS
-NameNode, which regularly takes a snapshot of the primary NameNode's state, is
-spawned. There exist no automatic fail-over though, if the main the primary
-NameNode crashes manual intervention is needed.
-
-Both systems have high reliability. Data on HDFS are replicated to---typically
-three---other nodes in the cluster. Jobs submitted to Hadoop also have a high
-reliability: If a node does not succeed with a job it is rescheduled to another
-free node which executes it.
-
-Regarding scalability Hadoop and HBase scales close to linearly when nodes are
-added to the cluster.
-
-While relational databases can be made highly available and reliable with load
-balancing and so called \textit{Sharding}, the setup is non-trivial. Regarding
-scalability, relational databases are highly dependent on having data in working
-memory; thus they scale poorly.
-% TODO: [citation needed]
-
-\section*{Summary}
-
-The first section of this chapter talked about the procedure during the thesis
-work. It described the exploratory model and the prototyping performed, as well
-as the design phase that led up to the implementation.
-
-The following sections presented and discussed the selected tools; they also
-discussed the impact on design and implementation choices.
-
-\chapter{Implementation}
-
-\chapterquote{A good idea is about ten percent and implementation and hard
-work, and luck is 90 percent.}{Guy Kawasaki}
-
-\section*{Introduction}
-
-This chapter will present the implementation of the Zohmg system. For a usage
-manual see the appendix.
-
-% TODO: Do we need to talk about the storage implementation?
-
-In the first section the implementation of project creation, configuration, and
-setup, will be presented. The second section features the data import
-implementation. The third section presents the implementation of data
-interchange between Zohmg and other software but also throughout the Zohmg
-system. The fourth, and final, section presents the implementation details of
-the Zohmg system's exporting compontent.
-
-\section{Project management}
-
-A Zohmg project is necessary for using Zohmg; they are created with a management
-script, which is also used for interacting with the Zohmg system. When the user
-has created a Zohmg project directory further actions can be performed.  The
-
-Zohmg projects reside within a project directory, which contains five
-directories: \texttt{clients}, \texttt{config}, \texttt{lib}, \texttt{mappers},
-and \texttt{transformers}. The details of these directories are presented
-throughout this chapter.
-
-\subsection*{Configuration}
-
-The Zohmg system needs configuration of the user's data sets and the environment
-for Hadoop and Dumbo. When the data set files are configured, the user runs the
-management script, which executes the configuration and creates tables in HBase
-according to the configuration.
-
-\subsubsection*{Data sets}
-
-% TODO: define what a data set is, and what makes it different from a project.
-
-In this context a data set is the schema that the user defines for wanted data
-sets to be handled by Zohmg. The configuration of a data set is simply a list
-of dimensions and projections, which will layout the HBase tables according to
-the schema. This configuration will also be read during the import phase, for
-handling data to be imported.
-
-The data set files reside in the project's \texttt{config} directory. The data
-sets are configured using YAML\footnote{YAML Ain't a Markup Language}, a
-human-readable data serializaton format. Any human-readable configure file
-method could have been chosen. YAML was choosen because it is is very simple
-for humans to read and edit, and likewise for machines. The Python module
-PyYAML was used to implement the YAML configuration reading part. \cite{pyyaml}
-
-\subsubsection*{Environment}
-
-The environment is a Python script which defines run-time variables, required by
-Hadoop and Dumbo. Using a Python scripts for run-time variables leaves the
-problem of validation to Python. The environment configuration file resides in
-the project's \texttt{config} directory.
-
-\section{Importing}
-
-In order to import data to Zohmg, the user writes a map function in Python which
-Zohmg wraps; interpreting the output from the mapper program and storing it in
-HBase.
-
-A standard MapReduce map function emits key-value pairs. Not so in the case of
-the Zohmg mapper: the usermap function emits triples (that is, a tuple of three
-values) consisting of a timestamp, a list of dimensions and their respective
-values, and a list of measurements. In effect, the dimensions and the timestamp
-specify a point in n-space and the measurements represent the observations made
-at that point.
-
-Going back to the running example of apache access logs, a usermap function
-might look something like this:
-
-\vspace{12pt}
-
-\begin{lstlisting}[language=Python,caption=An example map function.,captionpos=b]
-   def map(key, value):
-      log = parse_apache_log(value)
-      dimensions = {'useragent' : log.useragent,
-                    'path'      : log.path,
-                    'status'    : log.statuscode}
-      values = {'pageviews' : 1,
-                'bytes'     : log.size}
-
-      yield log.timestamp,dimensions,measurements
-\end{lstlisting}
-
-\vspace{12pt}
-
-The actual parsing of the log is abstracted; in the above example the
-emitting of dimensionality is showcased. It is assumed that
-\texttt{parse\_apache\_log()} accepts a single line from the log file and
-returns an object with the appropriate fields, such as useragent and
-statuscode.
-
-For every line of the log file, a point in n-space (the dimensions and the
-timestamp) is emitted together with a count of pageviews and bytes; one
-pageview per request and the number of bytes served in the request. It is
-possible to add business logic here, perhaps only counting requests from
-useragents that are not web spiders and that resulted in a HTTP 200 status
-code.
-
-The last line \textit{yields} the tuple. This is a Python construct called a
-\textit{generator} that lets the map function emit any number of tuples (or
-none!) for a single invocation.
-
-The user will import a day worth of logs like so:
-
-\begin{verbatim}
-   zohmg import mappers/apache.py \
-   /data/weblogs/2009/05/10
-\end{verbatim}
-
-This will instruct zohmg to apply the map function found in \texttt{apache.py}
-to all logs that it can find in the directory
-\texttt{/data/weblogs/2009/05/10} on HDFS. The emitted output is duly
-collected and projections are carefully stored in the data store.
 
@@
 
+% export.
+Due to the several different data sources, there exist several
+different types of applications. Therefor the system does not implement any user
+interface other than the REST API\footnote{The REST API is detailed in appendix
+X}. Making sense of the data is left for the developer.
+
+\section{System Requirements}
+
+The four key requirements of the Zohmg system are: ease of use, speed,
+scalability, and adoption, where adoption means being able to fit in the
+existing software and hardware eco-system.
+% dave probably knows a better word than 'fit'.
+
+% ease of use.
+The system should have an easy facility for importing and analyzing new data
+sources, while still being fast and efficient. A simple script---or tweak of an
+already existing data import script---should be enough for importing and
+analyzing a new data source.
+
+% speed.
+Time is of essence when exporting data to the user. The user should be able to
+explore the data at mouse-click rate, not having to wait for computation. This
+is usually solved by pre-rendering the data which can be viewed or queried.
+
+Importing data requires computing power, which is a limited resource. This calls
+for the need to make the import phase efficient. Using a desktop workstation to
+process several gigabytes of data is simply impossible; it might require more
+time than the actual worth of data it is importing---for instance taking more
+than one day to import a days worth of data. There are a number of ways to solve
+this, among them: more powerful machines and parallelizing.
+
+Acquiring more powerful machines quickly becomes expensive. Even if more
+powerful machines are acquired they might not have enough power for processing
+several gigabytes of data: the data to be processed might not fit in the working
+memory.
+
+The other technique, parallelizing, uses a cluster of commodity hardware
+machines. Using commodity hardware for computation significantly reduces the
+cost, since commodity hardware is cheap; thusly preferred in most situations.
+
+% scaling.
+In order for a computational setup, cluster or other, to be efficient it needs
+to scale well, this means adding more hardware should add as close as possible
+to the hardware's potential addition to the computational power.
+
+% fit in existing eco system.
+The system needs to co-exist and integrate seamlessly with the already existing
+hardware and software eco-system at Last.fm. Thus, building on top of the
+already existing software a natural choice.
+
+\section*{Summary}
+
+This chapter has analyzed and discussed the function and requirements of the
+system to be built.
+
+The system function section discussed the Zohmg system's goal and its
+components. The goal is to have a system which enables the developer to easily
+import and analyze new data sources and to be fast enough to serve the analyzed
+data at mouse-click rate to the user. The system's three core
+components---importing, storing, and extracting data---have also been
+discussed.
+
+Furthermore the requirements on the Zohmg system were discussed; the four key
+requirements are: ease of use, speed, scaling, and fitting into the existing
+hardware and software eco-system.
+
+\chapter{Method}
+
+\chapterquote{Arrange whatever pieces come your way.}{Virgina Wolf}
+
+\section*{Introduction}
+
+The first section in this chapter will present the work procedure during the
+master thesis. The following chapters will discuss selected tools and their
+impact on design and implementation choices for the Zohmg system.
+
+\section{Procedure}
+
+% TODO: generally use passive tense.
+% TODO: renskriv.
+
+This section describes how we structured our work over the course of the thesis
+project. It describes how we, by working in an exploratory fashion, reached a
+good understanding of the problem at hand, and how we successively inched closer
+to a working solution by building early prototypes and iteratively improving
+upon them.
+
+The design phase, which we entered once our understanding of the problem scope
+and its theoretical solution was satisfactory, and the following implementation
+are described, followed by a summary of our literature studies.
+
+\subsection*{Beginning}
+
+% not so sure I like this section much..
+
+We entered this project with a basic understanding of the problem Last.fm wanted
+to solve but with little formal knowledge of the problem domain. We were tasked
+with building a data store that would hopefully replace the store that was
+presently in use.
+
+We met the relevant people, were introduced to the current system and were
+presented with literature in the area, namely OLAP Solutions and Whatever in
+Data Warehousing.
+
+\subsection*{Exploration}
+
+The possible tools were explored by testing and benchmarking them.
+
+The problem area was explored by prototyping solutions to a given subproblem.
+
+\subsubsection*{Software}
+
+% TODO: renskriv.
+MapReduce frameworks: Hadoop, disco.
+
+Hadoop is used in production at Last.fm, it made sense to use it because the
+work done would integrate with the Last.fm's infrastructure easily.
+
+There exists a few databases which inspired by BigTable; they aim to be
+distributed and have column-oriented properties: HBase, Cassandra, MonetDB,
+MongoDB, Hypertable. CouchDB is an document-oriented database.
+
+These databases were explored and tried out. Sadly most of these BigTable-like
+databases were in a very early phase of development and hard to even install.
+The experience was that HBase was the most mature of them. Maturity was the
+compound experience with the software: community, codebase, documentation,
+stability etc. HBase is also tightly coupled with Hadoop which is used in
+production at Last.fm.
+
+CouchDB was experienced to be mature as well but lacked integration with
+MapReduce, all database communication would be necessary to do over HTTP.
+Setting up CouchDB with replication required some hands-on, it lacks sorting of
+documents which BigTable-like databases utilize.
+
+\subsubsection*{Benchmarks}
+
+Only HBase made it this far, the other databases in the roster was hard or
+impossible to either compile, install, setup, or all of them.
+
+HBase was benchmarked (the benchmarks was similar to the ones found in the
+BigTable paper?).
+
+\subsubsection*{Prototyping}
+
+The problem domain was explored by asking, the system to be built and the
+requirements upon it.
+
+To explore the environment, for example, we asked what data sources there are
+(web logs, scrobbling logs and radio streaming logs) and where that data is
+stored (on HDFS, it turned out). We asked how much data the system was supposed
+to handle (lots) and what the output format ought to be (any format, within
+reason).
+
+Most of the exploration was straight forward: many questions could be answered
+from observations or by logical reasoning. Others were more open-ended, such as
+deciding on a suitable output format. For the open-ended questions we would list
+reasonable options and pitt them against each other, first on the drawing board
+and eventually as prototypes.
+
+% this is a little thin. The prototypes are small pieces of code that would
+TODO: answer questions such as Is it at all possible to.. ? and What is the most
+efficient way to..?
+
+\subsection*{Literature studies}
+% vet inte riktigt vad jag vill saga om det har.
+
+Studying previous work in related fields brought OLAP systems to our attention.
+It was to become very influential in our theoretical understanding of the
+subject.
+
+\subsection*{Python}
+
+Python is a general-purpose high-level multi-paradigm---amongst others:
+object-oriented, imperative and functional---scripting language. \cite{python}
+
+Because there existed a Python module for writing MapReduce programs (Dumbo for
+Hadoop, both further later in this chapter), Python was a natural choice during
+the design and exploration phases. Python was also a good choice because of the
+rapid prototype development it enables.
+
+\subsection*{Design Phase}
+
+After having become familiar with the problem and its theoretical solution, it
+was time to design the software.
+
+Our initial design was ambitious: it contained, amongst other things, a query
+language and sub-components for query analyzation, to identity the most common
+queries, and for query optimization. These are likely advantageous features in a
+large data store but one can do without them.
+
+In the end, we much simplified our initial design and ended up with a design
+that was simpler and more to-the-point.
+
 % TODO: method.
--- Limits on cardinality --
+TODO: -- Limits on cardinality --
 
@@
 \textit{normalized\_path} as one of his dimensions, we might run into trouble.]
+
+\subsection*{Implementation}
+
+The implementation phase was rather small: we would simply stitch together our evolved prototypes and arrive at a solid ground to stand on: we had a system for running MapReduce jobs, storing data in HBase, etc.
+
+Writing code on top of this that performed data cube manipulation was a kinch!
+
+\section{Apache Hadoop}
+
+Hadoop is a free software implementation of Google's MapReduce-framework; it is
+written in Java, and is part of the Apache Software Foundation. Hadoop is a
+fairly mature piece of software and is used in production at numerous companies,
+among them Last.fm. \cite{hadoop}
+
+\subsection*{Hadoop Distributed File System}
+
+HDFS is a free software implementation of the GFS\footnote{Google File System}.
+\cite{gfs}
+
+The goal of HDFS is to store large data sets while being able to handle
+hardware failure and being deployed in heterogenous hardware and software
+eco systems. Additionally, the Hadoop and HDFS interaction is designed with
+the notion that it is cheaper to move the computation, MapReduce programs,
+to the data than the other way around.
+
+Every Hadoop node can be formatted with HDFS, this reserves disk space on the
+node to be used for HDFS. The HDFS partition\footnote{The HDFS partition is a
+directory containing HDFS files on a general-purpose file system.} resides on
+the node's general purpose file system. It is also possible to use HDFS as a
+stand-alone general purpose distributed file system, without Hadoop.
+
+The default block size on HDFS 64 MB, significantly larger than
+file systems used for hard drives. The motivation is that applications
+that use HDFS are not general purpose applications which run on general
+purpose file systems, but batch applications which read, write, or both,
+large amounts of data.
+
+\subsection*{Hadoop streaming}
+
+Any program that reads and outputs to the file pointers \texttt{stdin} and
+\texttt{stdout}, respectively, can be used as MapReduce programs with
+Hadoop Streaming. This makes it possible to use any shell script or
+program which inputs and outputs this way as a MapReduce program.
+
+\subsubsection*{Dumbo}
+
+Dumbo is a Python framework for writing Hadoop programs in Python; it exploits
+the Hadoop Streaming Mode to do so. It is used extensively at Last.fm for
+writing short prototypes. \cite{dumbo}
+
+\section{Apache HBase}
+
+HBase is a free software implementation of Google's BigTable. \cite{bigtable} It
+is a sub-project of Hadoop, written in Java and a part of the Apache Software
+Foundation. HBase is still in its infancy and few organizations use it for
+mission-critical applications. \cite{hbase}
+
+\subsection*{Physical storage}
+
+Each column-family is stored as a single file in HBase; it is then subdivided
+into column-qualifiers within this file. This is important to consider when
+designing application and schemas, since reading from several column-families
+will open several files.
+
+\subsection*{Sparsity}
+
+When many cells are empty, the table is said to be sparse. This can potentially
+be a problem, because even sparse cells may cost storage space.  BigTable and
+HBase, however, don't store empty cells; as consequence sparsity is not
+associated with a great storage cost.
+\cite{olap_data_scalability,olap_solutions}
+
+\subsection*{Hierarchies}
+
+Depending on how the row keys are formatted different, hierarchies are created.
+The order in which the data is stored is important, it defines how effective
+scanning data will be. Sparsity also comes into play during effective scanning.
+In the case of having thousand rows and requesting ten out of these, then only
+one per cent of the data is interesting. If a scan would be required to visit
+all the thousand rows a lot of rows are skipped. Skipping rows is expensive in
+that sense that they are visited but the result is thrown away. The goal is to
+push this cost to a bare minimum, using as many of the rows as possible.
+
+The minimum cost of scanning data is achieved by ordering the data so that
+a scan uses every visited row. In the case with thousand rows of which ten
+are wanted, would mean that ten rows scan are scanned of which all are
+used.
+
+\subsection*{Thrift}
+
+HBase has a Thrift server which serves as the interface to languages other than
+Java---for instance Python.
+
+Thrift is a remote procedure call framework for building scalable cross-language
+services. Thrift combines a software stack with a code generation engine to
+build services that work seamlessly between a wide range of programming and
+scripting languages. \cite{thrift}
+
+Thrift was originally developed at Facebook and in April 2007 it was released as
+free software. It entered the Apache Incubator in May 2008.
+
+\section{Dependability}
+
+Regarding the infrastructure, both Hadoop and HBase have high availability,
+reliability, and scalability.
+
+Both Hadoop and HBase are distributed systems; making them highly available out
+of the box. Although the Job Tracker\footnote{All client applications are
+submitted to Hadoop's Job Tracker.} is a single point of failure for Hadoop.
+The HDFS NameNode\footnote{A unique server which keeps the directory tree of all
+files in HDFS.} is also a single point of failure. However, a secondary HDFS
+NameNode, which regularly takes a snapshot of the primary NameNode's state, is
+spawned. There exist no automatic fail-over though, if the main the primary
+NameNode crashes manual intervention is needed.
+
+Both systems have high reliability. Data on HDFS are replicated to---typically
+three---other nodes in the cluster. Jobs submitted to Hadoop also have a high
+reliability: If a node does not succeed with a job it is rescheduled to another
+free node which executes it.
+
+Regarding scalability Hadoop and HBase scales close to linearly when nodes are
+added to the cluster.
+
+While relational databases can be made highly available and reliable with load
+balancing and so called \textit{sharding}, the setup is non-trivial. Regarding
+scalability, relational databases are highly dependent on having data in working
+memory; thus they scale poorly.
+% TODO: [citation needed]
+
+\section*{Summary}
+
+The first section of this chapter talked about the procedure during the thesis
+work. It described the exploratory model and the prototyping performed, as well
+as the design phase that led up to the implementation.
+
+The following sections presented and discussed the selected tools; they also
+discussed the impact on design and implementation choices.
+
+\chapter{Implementation}
+
+\chapterquote{A good idea is about ten percent and implementation and hard
+work, and luck is 90 percent.}{Guy Kawasaki}
+
+\section*{Introduction}
+
+This chapter will present the implementation of the Zohmg system. For a usage
+manual see the appendix.
+
+% TODO: Do we need to talk about the storage implementation?
+
+In the first section the implementation of project creation, configuration, and
+setup, will be presented. The second section features the data import
+implementation. The third section presents the implementation of data
+interchange between Zohmg and other software but also throughout the Zohmg
+system. The fourth, and final, section presents the implementation details of
+the Zohmg system's exporting compontent.
+
+\section{Project Management}
+
+A Zohmg project is necessary for using Zohmg; they are created with a management
+script, which is also used for interacting with the Zohmg system. When the user
+has created a Zohmg project directory further actions can be performed.  The
+
+Zohmg projects reside within a project directory, which contains five
+directories: \texttt{clients}, \texttt{config}, \texttt{lib}, \texttt{mappers},
+and \texttt{transformers}. The details of these directories are presented
+throughout this chapter.
+
+\subsection*{Configuration}
+
+The Zohmg system needs configuration of the user's data sets and the environment
+for Hadoop and Dumbo. When the data set files are configured, the user runs the
+management script, which executes the configuration and creates tables in HBase
+according to the configuration.
+
+\subsubsection*{Data sets}
+
+% TODO: define what a data set is, and what makes it different from a project.
+
+In this context a data set is the schema that the user defines for wanted data
+sets to be handled by Zohmg. The configuration of a data set is simply a list
+of dimensions and projections, which will layout the HBase tables according to
+the schema. This configuration will also be read during the import phase, for
+handling data to be imported.
+
+The data set files reside in the project's \texttt{config} directory. The data
+sets are configured using YAML\footnote{YAML Ain't a Markup Language}, a
+human-readable data serializaton format. Any human-readable configure file
+method could have been chosen. YAML was choosen because it is is very simple
+for humans to read and edit, and likewise for machines. The Python module
+PyYAML was used to implement the YAML configuration reading part. \cite{pyyaml}
+
+\subsubsection*{Environment}
+
+The environment is a Python script which defines run-time variables, required by
+Hadoop and Dumbo. Using a Python script for run-time variables leaves the
+problem of validation to Python. The environment configuration file resides in
+the project's \texttt{config} directory.
+
+\section{Importing}
+
+In order to import data to Zohmg, the user writes a map function in Python which
+Zohmg wraps; interpreting the output from the mapper program and storing it in
+HBase.
+
+A standard MapReduce map function emits key-value pairs. Not so in the case of
+the Zohmg mapper: the usermap function emits triples (that is, a tuple of three
+values) consisting of a timestamp, a list of dimensions and their respective
+values, and a list of measurements. In effect, the dimensions and the timestamp
+specify a point in n-space and the measurements represent the observations made
+at that point.
+
+% TODO: move apache logs to introduction perhaps.
+An example data source is the Apache HTTP Server access logs. The Apache HTTP
+Server Combined Log Format is commonly used to log requests. The log format is
+configurable, an example log line might have the following format
+
+\vspace{12pt}
+
+\begin{lstlisting}[caption=Example log line entry for the Apache HTTP Server's
+Combined Log Format.,captionpos=b]
+   127.0.0.1 - emma [20/Apr/2009:13:37:00 +0100]
+   "GET /music HTTP/1.0" 200 2326 "http://last.fm/"
+   "Mozilla/5.0 (X11; U; Linux i686)"
+\end{lstlisting}
+
+\vspace{12pt}
+
+The above example log entry is line-broken due to space constraints, normally
+the entry would be on one line. The example log line contains access info for a
+specific request; it contains the key entries: the client's IP address,
+timestamp, requested resource, HTTP status code for the request, the client's
+user agent etc.
+
+Revisiting the running example of Apache access logs, a usermap function might
+look something like this:
+
+\vspace{12pt}
+
+\begin{lstlisting}[language=Python,caption=An example map function.,captionpos=b]
+   def map(key, value):
+      log = parse_apache_log(value)
+      dimensions = {'useragent' : log.useragent,
+                    'path'      : log.path,
+                    'status'    : log.statuscode}
+      values = {'pageviews' : 1,
+                'bytes'     : log.size}
+
+      yield log.timestamp,dimensions,measurements
+\end{lstlisting}
+
+\vspace{12pt}
+
+The above example showcases the emitting of dimensionality. In the example the
+actual parsing of the log is abstracted out by the function
+\texttt{parse\_apache\_log()}; it is assumed that this function accepts a single
+line from an Apache HTTP Server log file and returns an object with the
+appropriate fields, such as useragent and statuscode.
+
+For every line of the log file, a point in n-space (the dimensions and the
+timestamp) is emitted together with a count of pageviews and bytes; one
+pageview per request and the number of bytes served in the request. It is
+possible to add business logic here, perhaps only counting requests from
+useragents that are not web spiders which resulted in a HTTP 200 status code.
+
+The last line \textit{yields} the tuple. This is a Python construct, similar to
+an iterator, called a \textit{generator} that lets the map function emit any
+number of tuples for a single invocation.
+
+The user will import a day worth of logs by invoking
+
+\begin{verbatim}
+   zohmg import mappers/apache.py \
+   /data/weblogs/2009/05/10
+\end{verbatim}
+
+This will instruct zohmg to apply the map function found in \texttt{apache.py}
+to all logs that it can find in the directory
+\texttt{/data/weblogs/2009/05/10} on HDFS. The emitted output is duly
+collected and projections are carefully stored in the data store.
 
@@
 
+\section{Data Interchange}
+
 % TODO: data interchange.
@@
 
-\section{Data interchange}
-
-\subsection{Internal}
-
-Typed bytes
-
-\subsubsection{Input formats}
+\subsection*{Internal}
+
+Using JSON for data interchange between map and reduce parts in Zohmg import
+jobs.
+
+For better performance typed bytes may be used. This is what Dumbo uses
+internally.
+
+\subsubsection{Input Formats}
+
+An input format is a Java class which Hadoop uses for interpreting the input
+data.
+
+TableInputFormat
 
@@
 
-\subsubsection{Output formats}
+\subsubsection{Output Formats}
 
 % TODO: not truly correct
-The Darling output format for HBase. Between Zohmg and Hadoop JSON is used
-for serializing data.
-
-\section{Exporting data}
+The Darling output format for HBase. Between Dumbo, which Zohmg uses for data
+import, and Hadoop JSON is used for serializing data.
+
+The Darling output format increased import speed by a factor ten, compared to
+using Thrift.
+
+\section{Exporting Data}
 
@@
 applications are responsible for dispatching requests to the corresponding
-application or file.
-WSGI middleware receives a
-WSGI request and then performs logic upon this request, before passing the
-request to a WSGI application. \cite{definitive_guide_to_pylons,paste}
-Since the main goal was not
-to build a middleware framework from scratch, an existing such was used.
+application or file. WSGI middleware receives a WSGI request and then performs
+logic upon this request, before passing the request to a WSGI application.
+\cite{definitive_guide_to_pylons,paste} Since the main goal was not to build a
+middleware framework from scratch, an existing such was used.
 
@@
 is returned. JSON was chosen because it is i simple text serialization format
-which is widely used in web development.
-
-\subsection{External}
+which is widely used in web development. The Python module simplejson was
+selected for the this implementation detail. \cite{simplejson}
+
+\subsection*{External}
 
@@
 
-The URLParser class is middleware that comes with Python Paste; its purpose is
-to dispatch to applications based on the URL. The URLParser removes the first
-part of the URL and tries to find a matching application to relay the request
-to. For instance, consider the URL \texttt{/client/file.html}. The first
-part---\texttt{client}---will be removed and the request \texttt{/file.html}
-will be sent to the application named \texttt{client.py}.
+The URLParser class is a middleware application that comes with Python Paste;
+its purpose is to dispatch to applications based on the URL. The URLParser
+removes the first part of the URL and tries to find a matching application to
+relay the request to. For instance, consider the URL \texttt{/client/file.html}.
+The first part---\texttt{client}---will be removed and the request
+\texttt{/file.html} will be sent to the application named \texttt{client.py}.
 
@@
 
-Fetches data from HBase and transforms it with the requested transform
-program from the project's transformers directory, output is
-dumped as JSONP.
+The transformers application fetches data from HBase with the same mechanism as
+the raw data application. The data is transformed with a transformer---a Python
+program in the project's \texttt{transformers} directory---requested by the
+user. Similar to the raw data serving, the output is served as JSONP.
 
@@
 
-\section{Future work}
+\section{Future Work}
 
@@
 
-\subsection{Multiple data sets}
+\subsection*{Multiple Data Sets}
 
@@
 
-\subsection{Metadata}
+\subsection*{Metadata}
 
@@
 
-\subsection{Updating data set files}
+\subsection*{Updating Data Set Files}
 
@@
 
-\subsection{Importing without MapReduce}
+\subsection*{Importing Without MapReduce}
 
@@
 
-Install Hadoop and HBase.
+Hadoop and HBase is required to run Zohmg, they form the foundation of data
+processing and storage which Zohmg builds upon. Zohmg is built with Python and
+also requires a few Python modules, on a Debian based system these are installed
+by the install script (see below).
+
+To install Hadoop and HBase automatically a script is provided.
 
@@
 configuration.
+
+The Zohmg installation depends on Python Setuptools \cite{setuptools}, which
+needs to be installed prior to the installation.
 

### 2009-05-19 doc/report/msc-report.tex
doc/report/msc-report.tex: * implementation chapter.

@@
 setup, will be presented. The second section features the data import
-implementation. The third section presents the implementation of data
-interchange between Zohmg and other software but also throughout the Zohmg
-system. The fourth, and final, section presents the implementation details of
-the Zohmg system's exporting compontent.
+implementation; it will present example data as well as an example user map
+function to import this. The third section presents the implementation of data
+interchange between Zohmg and other software, such as Hadoop, Dumbo, and HBase,
+but also throughout the Zohmg system. The third section also presents the
+implementation details of the Zohmg system's exporting compontent.
 
@@
 
-A standard MapReduce map function emits key-value pairs. Not so in the case of
-the Zohmg mapper: the usermap function emits triples (that is, a tuple of three
-values) consisting of a timestamp, a list of dimensions and their respective
-values, and a list of measurements. In effect, the dimensions and the timestamp
+A standard MapReduce map function emits key-value pairs. The Zohmg usermap
+function, instead, emits triples (tuples with three values), which consist of a
+timestamp, a hash table of dimensions and their respective values, and a hash
+table with measurements (values). In effect, the dimensions and the timestamp
 specify a point in n-space and the measurements represent the observations made
@@
 % TODO: move apache logs to introduction perhaps.
-An example data source is the Apache HTTP Server access logs. The Apache HTTP
-Server Combined Log Format is commonly used to log requests. The log format is
+An example data source is the Apache HTTP Server access logs. The Combined Log
+Format is commonly used to log Apache HTTP Server requests. The log format is
 configurable, an example log line might have the following format
@@
 
-The above example log entry is line-broken due to space constraints, normally
-the entry would be on one line. The example log line contains access info for a
-specific request; it contains the key entries: the client's IP address,
-timestamp, requested resource, HTTP status code for the request, the client's
-user agent etc.
-
-Revisiting the running example of Apache access logs, a usermap function might
-look something like this:
+The above example log entry is line broken due to space constraints, normally
+the entire entry would be on a single line. The example log line contains access
+info for a specific request; it contains the key entries: the client's IP
+address, timestamp, requested resource, HTTP status code for the request, the
+client's user agent etc.
+
+Processing---which means: read, analyze, and store---logs of the format
+mentioned in Listing 5.1, requires the user to write a map function. This user
+map function might look something like
 
@@
 
-For every line of the log file, a point in n-space (the dimensions and the
-timestamp) is emitted together with a count of pageviews and bytes; one
-pageview per request and the number of bytes served in the request. It is
-possible to add business logic here, perhaps only counting requests from
-useragents that are not web spiders which resulted in a HTTP 200 status code.
+For every line of the log file (the value argument to the map function), a point
+in n-space (the dimensions and the timestamp) is emitted together with a count
+of pageviews and bytes; one pageview per request and the number of bytes served
+in the request. It is possible to add logic here, perhaps only counting requests
+from useragents that are not web spiders which resulted in a HTTP 200 status
+code.
 
@@
 
+Since the wish is to store projections rather than all combinations of all
+dimensions, the data needs to be mangled slightly before sending it down
+the pipe to the underlying data store. Dumbo expects a key-value pair from the
+mapper. Hadoop will collect all keys and give the reducer a key and an iterator
+of values.
+
+% TODO: ok to remove this?
+%If projections are to be stored, both the desired dimensions and the unit
+%in question needs to be encoded in the key, and the value of the unit
+%emitted in the \textit{value} field.
+%
+%By doing this it is possible to use a simple SumReducer to sum the values
+%for each unit at any point in n-space.
+
+Remember, the usermapper gives us a point in n-space and a list of
+measurements. We need to break out the measurements and sum over each such
+measurement. Zohmg solves this by wrapping the user map function.
+
+The wrapper takes each tuple emitted by the user map function and performs a
+dimensionality reduction, ending up with the projections requested by the user.
+For every projection and unit, the dimensions remaining after the reduction are
+appended to the string representation of the unit and emitted as the key. The
+value is simply the value of the unit. The Hadoop framework will do the heavy
+work of collecting all keys and summing the values.
+
+% TODO: Move this to DISCUSSION.
+%An alternative approach would have been to perform the dimensionality
+%reduction at a later stage, namely in the reducer of a multi-pass MapReduce
+%job.
+
+The reducer, then, is very simple: it is a simple sum of the values. The output
+of the reducer will not be stored on disk as is usually the case, but will
+instead be interpreted by a custom OutputReader (see Data Interchange section)
+that persists the data to HBase.
+
 The user will import a day worth of logs by invoking
@@
    zohmg import mappers/apache.py \
-   /data/weblogs/2009/05/10
+   /data/weblogs/2009/04/20
 \end{verbatim}
@@
 This will instruct zohmg to apply the map function found in \texttt{apache.py}
-to all logs that it can find in the directory
-\texttt{/data/weblogs/2009/05/10} on HDFS. The emitted output is duly
-collected and projections are carefully stored in the data store.
-
-% TODO: implementation.
--- The implementation of all that --
-
-Since the wish is to store projections rather than all combinations of all
-dimensions, the data needs to be mangled slightly before sending it down
-the pipe to the underlying data store.
-
-Dumbo expects a key-value pair from the mapper. Hadoop will collect all keys
-and give the reducer a key and an iterator of values.
-
-If projections are to be stored, both the desired dimensions and the unit
-in question needs to be encoded in the key, and the value of the unit
-emitted in the \textit{value} field.
-
-By doing this it is possible to use a simple SumReducer to sum the values
-for each unit at any point in n-space.
-
-Remember, the usermapper gives us a point in n-space and a list of
-measurements. We need to break out the measurements and sum over each such
-measurement. Zohmg solves this by wrapping the usermapper in a cleverly named
-devicie, namely a mapperwrapper.
-
-% TODO: this para sounds as if it was written by a drunken monkey.
-The mapperwrapper takes each tuple emitted by the usermapper and performs a
-dimensionality reduction, ending up with the projections requested by the
-user. For every projection and unit, the dimensions remaining after the
-reduction are appended to the string representation of the unit and emitted
-as the key. The value is simply the value of the unit. The Hadoop framework
-will do the heavy work of collecting all keys and summing the values.
-
-An alternative approach would have been to perform the dimensionality
-reduction at a later stage, namely in the reducer of a multi-pass MapReduce
-job.
-
-The reducer, then, is very simple: it is a simple sum of the values. The output
-of the reducer will not be stored on disk as is usually the case, but will
-instead be interpreted by a custom OutputReader that persists the data to
-HBase.
+to all logs that it can find in the directory \texttt{/data/weblogs/2009/04/20}
+on HDFS. The emitted output is duly collected and projections are carefully
+stored in the data store.
 
@@
 column-family and qualifier to store the computed sum. It does this by
-emitting the row key as the key and a JSON-serialized map of the form
+emitting the row key as the key and a JSON-serialized map of the form \\
 \texttt{\{"column-family:qualifier" : value\}}.
@@
 
-TableInputFormat
-
-TextInputFormat
-
-LzoInputFormat
+The simplest input format would be the TextInputFormat, this is used to parse
+flat text files which reside on HDFS.
+
+Hadoop is able to read LZO compressed files with the LzoInputFormat. In order to
+be able to do so effectively the compressed files first have to be indexed with
+a tool that is provided with Hadoop.
+
+% TODO: When is the TableInputFormat used.
+The TableInputFormat is able read input to a Hadoop program from an HBase table.
 
@@
 % TODO: not truly correct
+% TODO: A detailed explanation of generally everything in the darling dir.
 The Darling output format for HBase. Between Dumbo, which Zohmg uses for data
@@
 The Darling output format increased import speed by a factor ten, compared to
-using Thrift.
+using Thrift. The HBase Thrift interface is slow because it first spawns a Java
+HBase client which interprets everything from the HBase Thrift server.
+This actually adds an extra step to the interaction with HBase making it slower
+than talking to it natively with Java.
 
@@
 
-The transformers application fetches data from HBase with the same mechanism as
-the raw data application. The data is transformed with a transformer---a Python
-program in the project's \texttt{transformers} directory---requested by the
-user. Similar to the raw data serving, the output is served as JSONP.
+The transformer middleware application fetches data from HBase with the same
+mechanism as the raw data server (via the Thrift interface); it then uses a user
+transformer to alter the data and return it to the user. The user transformer is
+a Python program in the project's \texttt{transformers} directory. See the
+appendix A for more information.
+
+The data is received in the transformer application as a hash table, which is
+then available for manipulation. The result is returned from the transformer
+application as a hash table and the transformer middleware returns the
+transformed result to the user as JSONP.
 
@@
 
-Summary here.
+This chapter's first section has presented the implementation of the Zohmg
+system. The concept of Zohmg project's and management of thesame were
+introduced; as well as how these are configured. The second section presented
+how to import data with Zohmg, giving examples of data and the user map
+function. In the third, and final, section Zohmg's data interchange was
+featured. Both the internal data interchange, which is used together with
+Hadoop, Dumbo, and HBase, and the external data interchange, which is used by
+the user to query Zohmg.
 
@@
 
-Fetches data from HBase and transforms it with the requested transform
-program from the project's \texttt{transformers} directory, output is
-dumped as JSONP.
+It is possible to transform the raw data in Zohmg by using a so called
+\textit{transformer}. The transformer application receives a query---of the same
+format as the raw data server's---and then transforms the data with the
+requested transformer program from the project's \texttt{transformers}
+directory, output is dumped as JSONP.
 
@@
 
-\noindent The above query extracts the same data as the query from the Data
+The above query extracts the same data as the query from the Data
 section. This data is run through the \texttt{transformer.py} application and

### 2009-05-20 doc/report/msc-report.tex
doc/report/msc-report.tex: * cosmetic changes. * general shaping. * added Storage / Data Model section to Implementation chapter. needs work. * additions to Introduction chapter.
doc/report/references.bib:
* changed OLAP data scalability to book, for looks.
* added Omniture.

@@
 
-Making sense of data at a massive scale is one of the biggest challenges we face
-at Last.fm. Interpreting patterns in user behaviour becomes a challenge when
+Analyzing data at a massive scale is one of the biggest challenges we face at
+Last.fm. Interpreting patterns in user behaviour becomes a challenge when
 millions of users interact in billions of combinations; the data sets must be
@@
 
-Inevitably, the services and operations will generate a log for later
-analysis. As the services grow in size, so will the data sets they generate.
+Generally, the services and operations will generate a log for later analysis.
+As the services grow in size, so will the data sets they generate.
 
@@
 
-The size of the data sets at last.fm is in the range of several gigabytes per
+The size of the data sets at Last.fm is in the range of several gigabytes per
 day. Some data sets are many times larger. It is our belief that this is
@@
 nails, uses a spreadsheet to keep track of the number of nails exported to
-each country every month (see figure).
+each country every month (see table 1.1).
 
@@
 Last.fm is an internet radio station focusing on content, being a musical
-encyclopedia. The web site.
-
-Last.fm lets users submit what music they listen to, called \textit{scrobbling}.
-
-Radio streams to users.
+encyclopedia. Last.fm lets users submit what music they listen to, called
+\textit{scrobbling}, this user submitted data is analyzed and Last.fm offer
+their users personalized charts, music and event recommendations, based ther
+listening. Last.fm also streams content, via internet, to users. The Last.fm web
+site ties together all these areas and presents it to the users.
 
@@
 
-OLAP is something. It's drawbacks are unknown.
+OLAP\footnote{On-Line Analytical Processing} is a broad category of techniques
+for processing data, and visualizing the analysis. It is called online because
+the data store will be queried, and the user user expects a, at least close to,
+direct response.
 
@@
 
-Omniture is used at Last.fm etc.
+Omniture is a data warehouse used at Last.fm. Criticism against Omniture is
+that it is cumbersome and slow to use. \cite{omniture}
 
@@
 
-% TODO: make correct.
-The following overview is entirely incorrect:
-
-Chapter 2 establish the theoretical base of the problem statement.
-
-In chapter 3, the tools used for the implementation are declared, moving
-over to chapter 4 where the details of the implementation are presented.
-
-How to use the Zohmg system is presented in chapter~5, while chapter 6
-shows an example work flow.
-
-Chapters 7 and 8 discuss the results, related, and future work, and
-presents the conclusion, respectively.
+In chapter 2 general data warehous concepts are introduced; it also introduces
+new techniques that the reader might not know.
+
+Chapter 3 establish the theoretical base of the problem statement, and the
+expected system's function and requirements.
+
+Chapter 4 presents the work method during the thesis. It also presents tools
+possible to use for solving the problem statement.
+
+In chapter 5, the implementation choices and details of the Zohmg system are
+presented.
+
+Chapters 6 discuss the results, related, future work, and presents the
+conclusion.
+
+In appendix A an example work flow is presented. Appendix B contains a glossary.
 
@@
 
-Secondly, this chapter features new techniques in distributed storage and
-computing; these techniques are Google's BigTable and MapReduce.
+The last section features new techniques in distributed storage and computing;
+these techniques are Google's BigTable and MapReduce.
 
@@
 
-OLAP\footnote{On-Line Analytical Processing} is an umbrella term for systems
-built in a certain manner and that are optimized quickly answer
-multi-dimensional analytical queries.
+OLAP is an umbrella term for systems built in a certain manner and that are
+optimized quickly answer multi-dimensional analytical queries.
 
@@
          & t9 & & \small "anchor:cnnsi.com" & "CNN" \\
-         & t8 & & \small "anchor:my.lok.ca" & "CNN.com" \\
+         & t8 & & \small "anchor:my.look.ca" & "CNN.com" \\
         "com.cnn.www" & t6 & \small "{\textless}html\textgreater..." & & \\
@@
 distributed and have column-oriented properties: HBase, Cassandra, MonetDB,
-MongoDB, Hypertable. CouchDB is an document-oriented database.
+MongoDB, Hypertable. CouchDB is a document-oriented database.
 
@@
    127.0.0.1 - emma [20/Apr/2009:13:37:00 +0100]
-   "GET /music HTTP/1.0" 200 2326 "http://last.fm/"
+   "GET /music HTTP/1.0" 200 2324 "http://last.fm/"
    "Mozilla/5.0 (X11; U; Linux i686)"
@@
 
-\begin{lstlisting}[language=Python,caption=An example map function.,captionpos=b]
+\begin{lstlisting}[language=Python,caption={An example map function, parsing
+Apache access logs.},captionpos=b]
    def map(key, value):
@@
 
-Since the wish is to store projections rather than all combinations of all
+Since the goal is to store projections rather than all combinations of all
 dimensions, the data needs to be mangled slightly before sending it down
@@
 
-Remember, the usermapper gives us a point in n-space and a list of
-measurements. We need to break out the measurements and sum over each such
-measurement. Zohmg solves this by wrapping the user map function.
+Remember, the user map function gives a point in n-space and a list of
+measurements. The measurements needs to be extracted and each such measurement
+is summed over. Zohmg solves this by wrapping the user map function.
 
@@
 
+\section{Storage / Data Model}
+
+% TODO: renskriv.
+The typical use case for Zohmg is to read one or more column-qualifiers,
+which correspond to one or more points in n-space, over a range of time.
+Therefor, the keys are made up of the time and unit of the measurement,
+and the qualifier queried for represents the point in n-space. More on
+this in the data model section.
+
+Explain our data model, the relation between column-families, column-qualifiers
+and projections and dimensions.
+
 \section{Data Interchange}
@@
 logic upon this request, before passing the request to a WSGI application.
-\cite{definitive_guide_to_pylons,paste} Since the main goal was not to build a
+\cite{paste,definitive_guide_to_pylons} Since the main goal was not to build a
 middleware framework from scratch, an existing such was used.
@@
 framework---the data server dispatches to the requested application
-depending on the URL it was fed. \cite{python_paste}
+depending on the URL it was fed. \cite{paste}
 
@@
 
+\textbf{Replication} &
+.
+
+ \\
+
 \textbf{Serializing} &
@@
 object it was in before the serialization. \\
+
+ \\
+
+\textbf{Sharding} &
+.
 \end{tabular}
@@
 
-\section{Typical use case}
-
-The typical use case for Zohmg is to read one or more column-qualifiers,
-which correspond to one or more points in n-space, over a range of time.
-Therefor, the keys are made up of the time and unit of the measurement,
-and the qualifier queried for represents the point in n-space. More on
-this in the data model section.
-
 \end{document}

### 2009-07-02 src/setup.py
version => 0.2.1

@@ -19,7 +19,7 @@ from setuptools import setup, find_packages
 
 setup(
     name = 'zohmg',
-    version = '0.2.0',
+    version = '0.2.1',
     author = 'Fredrik Mollerstrand, Per Andersson',
     author_email = '{fredrik,per}@last.fm',
     license = 'Apache License 2.0',

### 2009-07-02 src/zohmg/cmd.py

@@ -29,7 +29,7 @@ for (dir, dirnames, files) in os.walk(eggpath):
             sys.path.append(dir+"/"+file)
 
 # TODO: read version from $somewhere.
-version = '0.2.0'
+version = '0.2.1'
 
 def usage(reason = None):
     zohmg = os.path.basename(sys.argv[0])

### 2009-07-02 src/setup.py
version => 0.2.2-dev

@@ -19,7 +19,7 @@ from setuptools import setup, find_packages
 
 setup(
     name = 'zohmg',
-    version = '0.2.1',
+    version = '0.2.2-dev',
     author = 'Fredrik Mollerstrand, Per Andersson',
     author_email = '{fredrik,per}@last.fm',
     license = 'Apache License 2.0',

### 2009-07-02 src/zohmg/cmd.py

@@ -29,7 +29,7 @@ for (dir, dirnames, files) in os.walk(eggpath):
             sys.path.append(dir+"/"+file)
 
 # TODO: read version from $somewhere.
-version = '0.2.1'
+version = '0.2.2-dev'
 
 def usage(reason = None):
     zohmg = os.path.basename(sys.argv[0])

### 2009-07-02 src/zohmg/data.py
factored out rowkey interpreting code into method; added test.

@@ -175,7 +175,7 @@ def rowkey_formatter(projection, d0, d0v, filters, t0, t1):
     for d in projection:
         rowkeyarray.append(d)
         if d == d0:
-            if d0v == '':
+            if d0v == '' or d0v == ['']:
                 rowkeyarray.append('all')
             else:
                 rowkeyarray.append(d0v)
@@ -205,6 +205,13 @@ def rowkey_formatter(projection, d0, d0v, filters, t0, t1):
 
     return startrow, stoprow
     
+# rowkey => (timestamp, dimension+attribute pairs)
+def rowkey_interpreter(rowkey):
+    timestamp = rowkey[-8:]
+    x = rowkey[:-9]
+    it = iter(x.split('-'))
+    return timestamp, dict(zip(it, it))
+
 
 def scan(table, columns, startrow, stoprow, basedimension, range, filters, data):
     t0, t1 = range
@@ -222,13 +229,7 @@ def scan(table, columns, startrow, stoprow, basedimension, range, filters, data)
         if next == []:
             break
         r = next[0]
-        # extract date from row key.
-        rowkey = r.row
-        ymd = rowkey[-8:]
-        x   = rowkey[:-9]
-        # wha?
-        it = iter(x.split('-'))
-        ds = dict(zip(it, it))
+        timestamp, ds = rowkey_interpreter(r.row)
         dval = ds.get(basedimension)
         del ds[basedimension]
 
@@ -237,13 +238,13 @@ def scan(table, columns, startrow, stoprow, basedimension, range, filters, data)
             if not k in filters: continue
             if not ds[k] in filters[k]: filter_accepts = False
 
-        if filter_accepts and (ymd >= t0 and ymd <= t1):
+        if filter_accepts and (timestamp >= t0 and timestamp <= t1):
             # read possible old values, add.
             for column in r.columns:
                 t[dval] = t.get(dval, 0)
                 t[dval] += int(r.columns[column].value)
             # and save.
-            data[ymd] = dict_addition(t, data.get(ymd, {}))
+            data[timestamp] = dict_addition(t, data.get(timestamp, {}))
 
     return data, numrows
 
@@ -278,5 +279,5 @@ def hbase_get(table, projections, query, filters):
         rows += r
     print "scanned a total of %d rows." % rows
 
-    # returns a list of dicts sorted by ymd.
-    return [ {ymd:data[ymd]} for ymd in sorted(data) ]
[…]

### 2009-07-02 tests/test_data.py

@@ -52,6 +52,11 @@ class TestData(unittest.TestCase):
         # sometimes jsonp.
 
 
+    def test_scan(self):
+        # TODO: mock scanner, assert correctness of scan().
+        pass
+
+
     def test_hbase_get(self):
         table = 'test' # must there be test data, then?
         projections = [['user']]
@@ -78,6 +83,11 @@ class TestData(unittest.TestCase):
 
         self.assertEquals(startrow, expected_startrow)
         self.assertEquals(stoprow, expected_stoprow)
+
+    def test_rowkey_interpreter(self):
+        rowkey = 'artist-97930-track-102203-20090601'
+        expected = ('20090601', {'artist': '97930', 'track': '102203'})
+        self.assertEqual(expected, zohmg.data.rowkey_interpreter(rowkey))
 
 
     def test_dict_addition(self):

### 2009-07-02 src/zohmg/data.py
refactored rowkey_interpreter() to allow for time units of any length.

@@ -207,9 +207,9 @@ def rowkey_formatter(projection, d0, d0v, filters, t0, t1):
     
 # rowkey => (timestamp, dimension+attribute pairs)
 def rowkey_interpreter(rowkey):
-    timestamp = rowkey[-8:]
-    x = rowkey[:-9]
-    it = iter(x.split('-'))
+    rk = rowkey.split('-')
+    timestamp = rk.pop()
+    it = iter(rk)
     return timestamp, dict(zip(it, it))
 
 

### 2009-07-02 tests/test_data.py

@@ -89,6 +89,10 @@ class TestData(unittest.TestCase):
         expected = ('20090601', {'artist': '97930', 'track': '102203'})
         self.assertEqual(expected, zohmg.data.rowkey_interpreter(rowkey))
 
+        rowkey = 'artist-97930-track-102203-20090601223000'
+        expected = ('20090601223000', {'artist': '97930', 'track': '102203'})
+        self.assertEqual(expected, zohmg.data.rowkey_interpreter(rowkey))
+
 
     def test_dict_addition(self):
         a = {'x': 1, 'y': 1}; aprim = a.copy()

### 2009-07-02 doc/report/msc-report.tex
fixed some comments on the report.

@@
 
-It is suitable to exploit the parallel nature of the records by breaking up
-the data sets in pieces and to have each piece be processed on a separate
-node in a large compute cluster. It is also suitable to store the data across
+It is appropriate to exploit the parallel nature of the records by breaking up
+the data sets in pieces and to have each piece be processed on a separate node
+in a large compute cluster. It is also appropriate to store the data across
 many storage nodes in what is called a distributed file system.
@@
 
-Throughout this chapter distinction is made between the developer: who develops
-the application, and the user: who uses the product which the developer
+Throughout this chapter distinction is made between the developer who develops
+the application, and the user who uses the product which the developer
 creates. Although, in reality, they might be the same person.
@@
 database. It is possible to import analyzed data to this tool; the data is
-presented to the user as graphs. Disadvantages of this system include problems
-with new data source additions: they can sometimes not be loaded because they
-are too big. If a user wants to perform a custom query, manual intervention by
-the developer is necessary. There is no way for this system to handle custom
-queries.
+presented to the user as graphs. Disadvantages of this system include
+\begin{itemize}
+\item Problems with new data source additions: they can sometimes not be loaded
+because they are too big.
+\item If a user wants to perform a custom query, manual intervention by the
+developer is necessary.
+\item There is no way for this system to handle custom queries.
+\end{itemize}
 
@@
 them: web servers, scrobbling\footnote{Last.fm users can submit what music they
-listen to, this is called scrobbling.}, radio streaming etc. This data reside
+listen to, this is called scrobbling.}, radio streaming etc. This data resides
 on a commodity hardware cluster used for parallel computation. The system needs to
@@
 higher cardinality: there are some 300 countries in the world. Certainly,
-there could be dimensions whose cardinality is infinite: the \textit{path}
+there could be dimensions whose cardinality is unbound: the \textit{path}
 dimension in the web log example, for example, could in theory have a
@@
 project. It describes how we, by working in an exploratory fashion, reached a
-good understanding of the problem at hand, and how we successively inched closer
+good understanding of the problem at hand, and how we successively moved closer
 to a working solution by building early prototypes and iteratively improving
@@
 and its theoretical solution was satisfactory, and the following implementation
-are described, followed by a summary of our literature studies.
+are described, followed by a summary of our tools studies.
 
@@
 
-Only HBase made it this far, the other databases in the roster was hard or
+Only HBase made it this far, the other databases in the roster were hard or
 impossible to either compile, install, setup, or all of them.

### 2009-07-13 doc/report/msc-report.tex
minor changes to thesis.
thesis improvements.

@@
 
-Any organization will eventually want to analyze various characteristics of
-its service and operations. The services can be anything from web sites to
-theme park; the analysis might include characteristics such as the number of
+Any organization will eventually want to analyze various characteristics of its
+service and operations. The services can be anything from web sites to theme
+parks; the analysis might include characteristics such as the number of
 visitors per day, or the average time a theme park guest spends queueing for a
-particular ride.
-
-Generally, the services and operations will generate a log for later analysis.
-As the services grow in size, so will the data sets they generate.
+particular ride. Generally, the services and operations will generate a log for
+later analysis. As the services grow in size, so will the data sets they
+generate. The size of the data sets at Last.fm is in the range of several
+gigabytes per day. Some data sets are many times larger. It is our belief that
+this is typical.
 
@@
 structure and are broken down into records that are independent of each other.
-The records can be analyzed one at a time without losing any of their meaning
-and are as such very well suited for parallel analysis.
+The records can be analyzed independently of each other without losing any of
+their meaning, as such they are very well suited for parallel analysis.
 
 Each log record typically contains a number of fields. For example, a log of
-visits to a web site might contain a timestamp, information about what country
-the visitor is from and what web browser she used.
-
-When the records in the data set have been analyzed, one can draw conclusions
+visits to a web site might contain a time-stamp, information about what country
+the visitor originates from, and what web browser used by the user.
+
+When the records in the data set have been analyzed, conclusions can be made
 about what countries generate the most traffic to the site, which web browser
@@
 called aggregates. To aggregate data, for example to compute the sum of
-visitors for all European countries, is a common technique for speeding up
-data lookup in data stores.
-
-The size of the data sets at Last.fm is in the range of several gigabytes per
-day. Some data sets are many times larger. It is our belief that this is
-typical.
+visitors for all European countries, is a common technique for speeding up data
+lookup in data stores.
 
@@
 Organizations measure their operations by considering many variables. When
-these variables are placed in a spreadsheet, which organizations are wont to
-do, the variables are set on the x and y axes. Each axis represents a logical
-grouping of variables. In the example of a log file of visits to a web site
-above, country of origin is one such variable. A variable is generally called
-a dimension in this thesis.
-
-The International Aerodyne Corporation, a leading exporter of woodworking
-nails, uses a spreadsheet to keep track of the number of nails exported to
-each country every month (see table 1.1).
+these variables are placed in a spreadsheet, the variables are set on the x and
+y axes. Each axis represents a logical grouping of variables. In the example of
+a log file of visits to a web site above, country of origin is one such
+variable. A variable is generally called a dimension in this thesis.
+
+As an example: The International Aerodyne Corporation, a leading exporter of
+woodworking nails, uses a spreadsheet to keep track of the number of nails
+exported to each country every month (see table 1.1).
 
@@
 
-% TODO: renskriv these thoughts.
-
-Last.fm is an internet radio station focusing on content, being a musical
-encyclopedia. Last.fm lets users submit what music they listen to, called
-\textit{scrobbling}, this user submitted data is analyzed and Last.fm offer
-their users personalized charts, music and event recommendations, based ther
-listening. Last.fm also streams content, via internet, to users. The Last.fm web
-site ties together all these areas and presents it to the users.
-
-The web site, radio streaming and submissions (scrobbling) generates gigabytes
-of data every day.
+Last.fm is an internet radio station that also has a large focus on metadata.
+Their users can submit what music they listen to, this is called
+\textit{scrobbling}. Data submitted by users is analyzed and the user is offered
+personalized charts, music, and event recommendations, based on their listening.
+Users can also edit metadata about artists, albums, events etc; this forms a
+musical encyclopedia and community. The Last.fm web site ties together the above
+mentioned areas and presents it to the users.
 
@@
 
-% TODO: define aggregates!
-
 The goal is to build a data store that enables its users to perform
@@
 In addition, the secondary goal is to build a tool that allows users to easily
-store their data sets in the data store. The tool should provide a simple – as
-in minimalistic – mechanism for data analysis (import) and querying (export).
-
-% is mapreduce part of the goal? hardly.
-The data analysis will require the user to write a MapReduce map. The queries
-will be performed via an export API for consumption by computer programs.
-
-% these are more likely part of the scope.
-The queries will always be made over a time range and for a single unit. The
-tool will support analytical functions such as drill-down, slicing, etc.
-
-The tool should be aware of dimensions in general and the dimension of time in
-particular. Measurements will be specified in a unit. There will be no limit
-on the number of units.
-
-There should be no theoretical boundary on the number of dimensions. Common
-queries must be answered expediently to allow for analysts to explore data in
-real time.
-
-\section*{Scope}
-% Avgransning.
-
-% TODO: argue why we do these avgränsningar.
-
-The data store is just that: a data store with import and export
-functionality. The data store does not interpret the data it store.
-
-The data store is time-series-based: all data must include the time dimension.
-
-All measures are integers. Sum is the only supported aggregation function. The
-data store does not support reach-like data (i.e. number of unique visitors)
-since it is not summable.
-
-There is no functionality for plotting data sets, etc.
-
-\section*{Related work}
+store their data sets in this data store. The tool should provide a simple---as
+in minimalistic---mechanism for data analysis (import) and querying (export).
+
+The data analysis will require the user to write a custom import program. The
+queries will be performed via an export API for consumption by computer
+programs. Common queries must be answered expediently to allow for analysts to
+explore data in real time.
+
+The tool will support analytical functions such as drill-down, slicing, etc; it
+should also be aware of dimensions in general and the time dimension in
+particular. Each dimension contains a measurement in a specified unit.  Example
+units are plays, page views, space travels, or such. There should be no upper
+bound on either on the number of units nor on the number of dimensions.
+
+\section*{Existing Software}
 
@@
 chronological order, the key might contain a representation of the data's
-timestamp. The classic example is to have the reversed domain name (i.e.
+time-stamp. The classic example is to have the reversed domain name (i.e.
 com.google.www) as the row key, which means that all sub-domains of any domain
@@
 out on the row key, the column family and the column qualifier, and possibly
-also of the version timestamp.
+also of the version time-stamp.
 
@@
 \begin{lstlisting}[caption=BigTable data model illustration.,captionpos=b]
-   (row key,column key,timestamp) -> value
+   (row key,column key,time-stamp) -> value
 \end{lstlisting}
@@
 The data model is a simple one: Rows of data are stored in labeled tables. Each
-row is indexed with a row key, a column key, and a timestamp. The columns have
+row is indexed with a row key, a column key, and a time-stamp. The columns have
 the form "column-family:qualifier", where the column-family is one of a number
@@
 types, and it is up to the application to interpret the data correctly. The
-table can store any number of timestamped versions of each cell, allowing
+table can store any number of time-stamped versions of each cell, allowing
 versioning of data.
@@
 
-\subsection*{System components}
+\subsection*{System Components}
 
@@
 
+% import.
+The Zohmg import component concists of a framework which enables the user to
+write import programs. The import component handles half of the MapReduce
+pattern, the reduce part; it also adds an interface for relaying the imported
+data to the storage component. The import programs written by the user include
+the other half of the MapReduce pattern, namely the map part, and uses the
+interface for storing the imported data.
+
 % storage.
-Because the data sources are typically large, the storage needs to be efficient.
-Storing the analyzed data can be done in different ways, for instance storing
-the raw data or just aggregates. Storing the raw data is very expensive, since
-it would basically copy the data to the data store. Because of this the decision
-was made to only store aggregates.
-
-There are several different ways of modelling multidimensional data.
-
-TODO: How to discuss modelling of data in a general sense, without mentioning
-MapReduce and BigTable?
-% thinking in simple data structures,
-
-% TODO: analysis.
-- Something about cardinality -
-
-TODO?: When reasoning about dimensions, the issue of cardinality will come
-up sooner or later.
-
+The storage component is built upon a BigTable-like database. Although BigTable
+is schema less, the user imposes a loose structure by configuring the layout of
+the wanted data sets. Basically, this configuration defines the table layout and
+defines what structure the imported data must have.
+
+% analysis.
 The cardinality of a dimension is the number of distinct values of the
@@
 % export.
-Due to the several different data sources, there exist several
-different types of applications. Therefor the system does not implement any user
-interface other than the REST API\footnote{The REST API is detailed in appendix
-X}. Making sense of the data is left for the developer.
+Due to the several different data sources, there exist several different types
+of applications one could build on top of Zohmg. Therefore the system does not
+implement any user interface other than the HTTP API; which takes a query with
+time range, unit, and selected units to extract.
+
+When extracting data it is important to remember that data handled by and stored
+in Zohmg is not typed in any way. The developer is responsible for interpreting
+the data in a correct manner.
 
@@
 scalability, and adoption, where adoption means being able to fit in the
-existing software and hardware eco-system.
-% dave probably knows a better word than 'fit'.
+existing software and hardware ecosystem.
 
@@
 In order for a computational setup, cluster or other, to be efficient it needs
-to scale well, this means adding more hardware should add as close as possible
-to the hardware's potential addition to the computational power.
+to scale well. In this sense, scaling means that adding more hardware should, in
+a best case scenario, add all of the computational power of the new hardware.
 
@@
 The system needs to co-exist and integrate seamlessly with the already existing
-hardware and software eco-system at Last.fm. Thus, building on top of the
-already existing software a natural choice.
+hardware and software eco-system at Last.fm. Thus, building on top of already
+existing software is a natural choice.
 
@@
 
+% TODO
 The first section in this chapter will present the work procedure during the
-master thesis. The following chapters will discuss selected tools and their
-impact on design and implementation choices for the Zohmg system.
+master thesis. The following sections will discuss selected tools and their
+impact on design and implementation choices for the Zohmg system. The last
+section will describe the scope of the thesis.
 
@@
 
-% TODO: renskriv.
-MapReduce frameworks: Hadoop, disco.
-
-Hadoop is used in production at Last.fm, it made sense to use it because the
-work done would integrate with the Last.fm's infrastructure easily.
+There exists several MapReduce frameworks, for example Apache Hadoop and
+Disco\cite{disco} which are both free software. As mentioned in the scope
+section Hadoop was selected to be built upon because it already had an
+preeminent role in production at Last.fm.
 
@@
 
-These databases were explored and tried out. Sadly most of these BigTable-like
-databases were in a very early phase of development and hard to even install.
-The experience was that HBase was the most mature of them. Maturity was the
-compound experience with the software: community, codebase, documentation,
-stability etc. HBase is also tightly coupled with Hadoop which is used in
-production at Last.fm.
+These databases were explored and tested, but most of them were in a very early
+phase of development and hard to even install. The experience was that HBase was
+the most mature of them. Maturity was the compound experience with the software:
+community, codebase, documentation, stability etc. HBase is also tightly coupled
+with Hadoop which is used in production at Last.fm.
 
 CouchDB was experienced to be mature as well but lacked integration with
-MapReduce, all database communication would be necessary to do over HTTP.
-Setting up CouchDB with replication required some hands-on, it lacks sorting of
-documents which BigTable-like databases utilize.
+MapReduce, resulting in that all database communication had to be done over
+HTTP. This would use vast resources to keep all HTTP connections open, working
+around this adds additional work compared to BigTable-like databases. CouchDB
+also lacks sorting of documents over a computer cluster, which BigTable-like
+databases utilize. The result was that setting up CouchDB with replication,
+which would spread the work, required hands-on work.
 
@@
 
+% TODO: need data on this.
 HBase was benchmarked (the benchmarks was similar to the ones found in the
@@
 
+\section{Scope}
+% Avgransning.
+
+% TODO:
+% * Live here or live in Introduction chapter? (Lots of concepts are
+%   presented after the introduction chapter.)
+% * Live at an earlier stage in this chapter?
+\subsection*{Tools}
+
+The MapReduce framework Hadoop is used in production at Last.fm. Hadoop was
+selected to be used for distributed computing, because of the need for simple
+integration with already existing infrastructure and developer competence at
+Last.fm.
+
+The BigTable-implementation HBase was selected to be used as the underlying data
+storage. It was selected because of its integration with Hadoop, and also, out of
+the box, solved several Zohmg related distributed storage problems. Several
+other distributed databases were tested, but HBase turned out to be the best
+alternative because of the easy integration with Last.fm's existing
+infrastructure, code maturity, as well as its vibrant community.
+
+\subsection*{Data Store}
+
+The Zohmg data store should just store data, it should have basic import
+and export functionality. Zohmg shall not interpret the stored data, this is
+left to the user to do in a correct manner. This decision was reached because of
+time constraints when implementing Zohmg.
+
+Because BigTable stores data sorted on row keys, it was decided that Zohmg only
+supports time-series-based data. Having data sorted between nodes in a computer
+cluster balances the network, computational, and storage loads, resulting in
+more even utilization of the cluster. Since Zohmg only supports time-series,
+each query has to be made over a time range.
+
+It was decided that the row keys should consist of unit and time-stamp. This
+would keep related data within close range, e.g. all rows with the same unit
+would be stored next to each other. This means reading data with different units
+would imply expensive jumps within the stored data. Because of this it was
+decided that queries only support a single unit to be queried.
+
+All measures are integers. Sum is the only supported aggregation function. The
+data store does not support reach-like data (i.e. number of unique visitors)
+since it is not summable.
+
+Zohmg is designed as a data store, which simply imports, stores, and exports
+data. Although presenting the data is important, this is left to the user to
+solve. This restriction was made because of keeping the implementation simple,
+and more general.
+
+Because the data sources are typically large, the storage needs to be efficient.
+Storing the analyzed data can be done in different ways, for instance storing
+the raw data or just aggregates. Storing the raw data is very expensive, since
+it would basically copy the data to the data store. Because of this the decision
+was made to only store aggregates.
+
 \section*{Summary}
 
+% TODO
 The first section of this chapter talked about the procedure during the thesis
@@
 discussed the impact on design and implementation choices.
+
+The last section describes the scope of the thesis.
 
@@
 
-A Zohmg project is necessary for using Zohmg; they are created with a management
-script, which is also used for interacting with the Zohmg system. When the user
-has created a Zohmg project directory further actions can be performed.  The
-
-Zohmg projects reside within a project directory, which contains five
-directories: \texttt{clients}, \texttt{config}, \texttt{lib}, \texttt{mappers},
-and \texttt{transformers}. The details of these directories are presented
-throughout this chapter.
+Zohmg uses the concept of projects. These projects are self contained
+application data directories which are used with Zohmg. A Zohmg project is
+necessary for using Zohmg; they are created with a management script, which is
+also used for interacting with the Zohmg system. When the user has created a
+Zohmg project directory further actions can be performed. The Zohmg projects
+reside within a project directory, which contains five directories:
+\texttt{clients}, \texttt{config}, \texttt{lib}, \texttt{mappers}, and
+\texttt{transformers}. The details of these directories are presented throughout
+this chapter.
 
@@
 
-\subsubsection*{Data sets}
+\subsubsection*{Data Sets}
 
@@
 function, instead, emits triples (tuples with three values), which consist of a
-timestamp, a hash table of dimensions and their respective values, and a hash
-table with measurements (values). In effect, the dimensions and the timestamp
+time-stamp, a hash table of dimensions and their respective values, and a hash
+table with measurements (values). In effect, the dimensions and the time-stamp
 specify a point in n-space and the measurements represent the observations made
@@
 info for a specific request; it contains the key entries: the client's IP
-address, timestamp, requested resource, HTTP status code for the request, the
+address, time-stamp, requested resource, HTTP status code for the request, the
 client's user agent etc.
@@
 For every line of the log file (the value argument to the map function), a point
-in n-space (the dimensions and the timestamp) is emitted together with a count
+in n-space (the dimensions and the time-stamp) is emitted together with a count
 of pageviews and bytes; one pageview per request and the number of bytes served
@@
 transformer to alter the data and return it to the user. The user transformer is
-a Python program in the project's \texttt{transformers} directory. See the
-appendix A for more information.
+a Python program in the project's \texttt{transformers} directory. See appendix
+A for more information.
 

### 2009-07-14 doc/report/msc-report.tex
report: method.

@@
 distributed processing is Google's MapReduce.
+
+TODO:
+Other techniques that have been frequently used to build data warehouses
+include: A, B, and C, amongst others.
+
+A solves distributed computing by ...
+B is blablabla ...
+The framework C was ...
 
@@
 
+TODO: expand.
+
 \section*{Summary}
@@
 
-% TODO
-The first section in this chapter will present the work procedure during the
-master thesis. The following sections will discuss selected tools and their
-impact on design and implementation choices for the Zohmg system. The last
-section will describe the scope of the thesis.
+The first section describes how the thesis project work was begun and later
+structured. The design and implementation phases are also described in this
+section.
+
+The following two sections presents the tools which were selected to build Zohmg
+upon.
+
+The last section presents the scope of the thesis.
 
@@
 
-% TODO: generally use passive tense.
-% TODO: renskriv.
-
-This section describes how we structured our work over the course of the thesis
-project. It describes how we, by working in an exploratory fashion, reached a
-good understanding of the problem at hand, and how we successively moved closer
-to a working solution by building early prototypes and iteratively improving
-upon them.
-
-The design phase, which we entered once our understanding of the problem scope
-and its theoretical solution was satisfactory, and the following implementation
-are described, followed by a summary of our tools studies.
-
-\subsection*{Beginning}
-
-% not so sure I like this section much..
-
-We entered this project with a basic understanding of the problem Last.fm wanted
-to solve but with little formal knowledge of the problem domain. We were tasked
-with building a data store that would hopefully replace the store that was
-presently in use.
-
-We met the relevant people, were introduced to the current system and were
-presented with literature in the area, namely OLAP Solutions and Whatever in
-Data Warehousing.
-
-\subsection*{Exploration}
-
-The possible tools were explored by testing and benchmarking them.
-
-The problem area was explored by prototyping solutions to a given subproblem.
-
-\subsubsection*{Software}
+This project started with an introduction to existing tools and infrastructure
+at Last.fm. The introduction also included specifying requirements on the final
+product, and introduction to several important concepts in data warehousing
+(such as OLAP). The concepts were presented with related literature and research
+papers.
+
+The general procedure of the thesis project was done as exploratory work, which
+involved rapid prototyping on sub problems and iteratively improving these
+prototypes. A working system solution successively moved closer when more of
+these sub problems was solved in a satisfactory manner.
+
+Once a thourough understanding of the problem scope and its theoretical solutian
+the design were entered. Finally a short implementation phase was executed.
+
+\subsection*{Literature Studies}
+
+To establish good understanding of the problem in general and new concepts,
+literature and research papers were studied. The concepts studied are data
+warehousing, OLAP, distributed storage with column-oriented databases such
+as BigTable, and distributed computing with the MapReduce pattern and
+frameworks.
+
+Also a lot of related and surronding, possibly useful, concepts were studied;
+although some of these never made it to the final solution.
+\cite{pig,mapreducemerge}
+
+\subsection*{Software}
 
@@
 
-\subsubsection*{Benchmarks}
-
-Only HBase made it this far, the other databases in the roster were hard or
+\subsubsection{Benchmarks}
+
+In order to evaluate software which candidated for taking responsibility for the
+data storage, benchmarked were planned. Finally, only HBase made it to the
+benchmarking part, the other databases in the roster were difficult or
 impossible to either compile, install, setup, or all of them.
 
-% TODO: need data on this.
-HBase was benchmarked (the benchmarks was similar to the ones found in the
-BigTable paper?).
-
-\subsubsection*{Prototyping}
-
-The problem domain was explored by asking, the system to be built and the
-requirements upon it.
-
-To explore the environment, for example, we asked what data sources there are
-(web logs, scrobbling logs and radio streaming logs) and where that data is
-stored (on HDFS, it turned out). We asked how much data the system was supposed
-to handle (lots) and what the output format ought to be (any format, within
-reason).
-
-Most of the exploration was straight forward: many questions could be answered
-from observations or by logical reasoning. Others were more open-ended, such as
-deciding on a suitable output format. For the open-ended questions we would list
-reasonable options and pitt them against each other, first on the drawing board
-and eventually as prototypes.
-
-% this is a little thin. The prototypes are small pieces of code that would
-TODO: answer questions such as Is it at all possible to.. ? and What is the most
-efficient way to..?
-
-\subsection*{Literature studies}
-% vet inte riktigt vad jag vill saga om det har.
-
-Studying previous work in related fields brought OLAP systems to our attention.
-It was to become very influential in our theoretical understanding of the
-subject.
-
-\subsection*{Python}
+TODO: HBase was benchmarked (the benchmarks was similar to the ones found in the
+BigTable paper?). NEEDS DATA!!
+
+\subsection*{Prototyping}
+
+As mentioned above, the problem domain was explored by prototyping solutions to
+sub problems of the greater system. The prototypes were for example used to find
+out if a problem was solvable at all, or what the most efficient implementation
+of an idea was.
+
+Most of the exploration was straight forward. Many questions could be answered
+from observations or by logical reasoning, other questions were more open ended.
+Such an open ended question was deciding on a suitable output format\footnote{An
+output format is a concept in Hadoop; it is a format which the reduce step uses
+to output the results.}. For open ended questions the approach was to first
+reason about suitable solutions, compare them against each other on the drawing
+board and then eventually as prototypes.
+
+The first prototypes handled data sources. The different data sources had
+different formats and locations, the related prototypes dealt with how these
+data sources should be analyzed and imported with MapReduce. The succeeding
+prototypes explored export and data storage functionality.
+
+\subsubsection{Python}
 
@@
 
+TODO: ALL OF THIS.
 After having become familiar with the problem and its theoretical solution, it
@@
 
-The implementation phase was rather small: we would simply stitch together our evolved prototypes and arrive at a solid ground to stand on: we had a system for running MapReduce jobs, storing data in HBase, etc.
-
-Writing code on top of this that performed data cube manipulation was a kinch!
-
-\section{Apache Hadoop}
+The implementation phase was rather small. The already existing evolved
+prototypes, which solved sub problems, was simply stitched together. The result
+gave a solid ground to stand on: A system which could run import programs
+(MapReduce jobs), shuffle data from the import programs into HBase, and export
+data from HBase through a HTTP API. Data cube implementation was added to this
+resulting system.
+
+\section{Tools}
+
+\subsection{Apache Hadoop}
 
@@
 
-\subsection*{Hadoop Distributed File System}
+\subsubsection{Hadoop Distributed File System}
 
@@
 
-\subsection*{Hadoop streaming}
+\subsubsection{Hadoop streaming}
 
@@
 
-\subsubsection*{Dumbo}
+\subsubsection{Dumbo}
 
@@
 
-\section{Apache HBase}
+\subsection{Apache HBase}
 
@@
 
-\subsection*{Physical storage}
+\subsubsection{Physical storage}
 
@@
 
-\subsection*{Sparsity}
+\subsubsection{Sparsity}
 
@@
 
-\subsection*{Hierarchies}
+\subsubsection{Hierarchies}
 
@@
 
-\subsection*{Thrift}
+\subsubsection{Thrift}
 
@@
 
-\section{Dependability}
+\subsection{Dependability}
 
@@
 cluster balances the network, computational, and storage loads, resulting in
-more even utilization of the cluster. Since Zohmg only supports time-series,
-each query has to be made over a time range.
+more even utilization of the cluster. Time-series analysis is also the most
+common for logs, since each entry is time-stamped. \cite{discoveringweb} Since
+Zohmg only supports time-series, each query has to be made over a time range.
 
@@
 
-% TODO
 The first section of this chapter talked about the procedure during the thesis
 work. It described the exploratory model and the prototyping performed, as well
-as the design phase that led up to the implementation.
+as the design phase which led to the implementation.
 
 The following sections presented and discussed the selected tools; they also
-discussed the impact on design and implementation choices.
-
-The last section describes the scope of the thesis.
+discussed the tools' impact on design and implementation choices.
+
+The last section described the scope of the thesis.
 

### 2009-07-14 doc/report/msc-report.tex
report: method chapter.

@@
 
+TODO: -- Limits on cardinality --
+
+[What are the limits of our system? How high cardinality do we support? (This
+is closely tied to hbase and the qualifier-limit.) If Martin is going to have
+\textit{normalized\_path} as one of his dimensions, we might run into trouble.]
+
 % export.
@@
 
-TODO: ALL OF THIS.
-After having become familiar with the problem and its theoretical solution, it
-was time to design the software.
-
-Our initial design was ambitious: it contained, amongst other things, a query
-language and sub-components for query analyzation, to identity the most common
-queries, and for query optimization. These are likely advantageous features in a
-large data store but one can do without them.
-
-In the end, we much simplified our initial design and ended up with a design
-that was simpler and more to-the-point.
-
-% TODO: method.
-TODO: -- Limits on cardinality --
-
-[What are the limits of our system? How high cardinality do we support? (This
-is closely tied to hbase and the qualifier-limit.) If Martin is going to have
-\textit{normalized\_path} as one of his dimensions, we might run into trouble.]
+After familiarity with the problem and its theoretical solution was reached,
+the design phase was entered. During this phase a grand design for the system
+was layed out.
+
+The initial design was ambitious; it contained, amongst other things, components
+to import, store, and exporte data, a query language and sub-components for
+query analyzation. The intended tasks for these components were to identify the
+most common queries and optimization. These features are likely advantageous in
+a large data store but it is possible to do without them.
+
+In the end, it was decided to use a simplified version of the initial design.
+As mentioned in chapter 3 in the section System Components, the final design
+only included components to import, store, and export data.
 
@@
 data. Although presenting the data is important, this is left to the user to
-solve. This restriction was made because of keeping the implementation simple,
-and more general.
+solve. The section Design Phase in this chapter mentions an original plan where
+a query language and sub-components for optimization were included. In the end,
+these were left out. These restrictions were made because of keeping the
+implementation simple, and more general.
 
@@
 was made to only store aggregates.
+
+TODO: add summary on cardinality limits on cardinality here.
 

### 2009-07-15 doc/report/msc-report.tex
report: added references.
report: speling corrections and other minorities.

@@
 HDW is a system whose goal is to create a high performance large scale data
-warehouse based on MapReduce and Bigtable. \cite{hdw}
+warehouse based on MapReduce and BigTable. \cite{hdw}
 
@@
 
-In chapter 2 general data warehous concepts are introduced; it also introduces
+In chapter 2 general data warehouse concepts are introduced; it also introduces
 new techniques that the reader might not know.
@@
 
-% CONCEPTS IN DATA WAREHOUSING, by clark de jean
 \chapter{Concepts}
@@
 Classic data warehouses use relational databases as storage. A star or snowflake
-schema would be used to modell the dimensions.
+schema would be used to model the dimensions.
 
@@
 % import.
-The Zohmg import component concists of a framework which enables the user to
+The Zohmg import component consists of a framework which enables the user to
 write import programs. The import component handles half of the MapReduce
@@
 % fit in existing eco system.
-The system needs to co-exist and integrate seamlessly with the already existing
-hardware and software eco-system at Last.fm. Thus, building on top of already
+The system needs to coexist and integrate seamlessly with the already existing
+hardware and software eco system at Last.fm. Thus, building on top of already
 existing software is a natural choice.
@@
 requirements are: ease of use, speed, scaling, and fitting into the existing
-hardware and software eco-system.
+hardware and software eco system.
 
@@
 
-Once a thourough understanding of the problem scope and its theoretical solutian
+Once a thorough understanding of the problem scope and its theoretical solution
 the design were entered. Finally a short implementation phase was executed.
@@
 
-Also a lot of related and surronding, possibly useful, concepts were studied;
+Also a lot of related and surrounding, possibly useful, concepts were studied;
 although some of these never made it to the final solution.
-\cite{pig,mapreducemerge}
+\cite{piglatin,mapreducemerge}
 
@@
 the most mature of them. Maturity was the compound experience with the software:
-community, codebase, documentation, stability etc. HBase is also tightly coupled
+community, code base, documentation, stability etc. HBase is also tightly coupled
 with Hadoop which is used in production at Last.fm.
@@
 the design phase was entered. During this phase a grand design for the system
-was layed out.
+was laid out.
 
 The initial design was ambitious; it contained, amongst other things, components
-to import, store, and exporte data, a query language and sub-components for
-query analyzation. The intended tasks for these components were to identify the
+to import, store, and export data, a query language and sub-components for
+query analysis. The intended tasks for these components were to identify the
 most common queries and optimization. These features are likely advantageous in
@@
 The goal of HDFS is to store large data sets while being able to handle
-hardware failure and being deployed in heterogenous hardware and software
+hardware failure and being deployed in heterogeneous hardware and software
 eco systems. Additionally, the Hadoop and HDFS interaction is designed with
@@
 data store does not support reach-like data (i.e. number of unique visitors)
-since it is not summable.
+since it is not sumable.
 
@@
 but also throughout the Zohmg system. The third section also presents the
-implementation details of the Zohmg system's exporting compontent.
+implementation details of the Zohmg system's exporting component.
 
@@
 sets are configured using YAML\footnote{YAML Ain't a Markup Language}, a
-human-readable data serializaton format. Any human-readable configure file
-method could have been chosen. YAML was choosen because it is is very simple
+human-readable data serialization format. Any human-readable configure file
+method could have been chosen. YAML was chosen because it is is very simple
 for humans to read and edit, and likewise for machines. The Python module
@@
 
-A standard MapReduce map function emits key-value pairs. The Zohmg usermap
+A standard MapReduce map function emits key-value pairs. The Zohmg user map
 function, instead, emits triples (tuples with three values), which consist of a
@@
 in n-space (the dimensions and the time-stamp) is emitted together with a count
-of pageviews and bytes; one pageview per request and the number of bytes served
+of page views and bytes; one page view per request and the number of bytes served
 in the request. It is possible to add logic here, perhaps only counting requests
-from useragents that are not web spiders which resulted in a HTTP 200 status
+from user agents that are not web spiders which resulted in a HTTP 200 status
 code.
@@
 
-This will instruct zohmg to apply the map function found in \texttt{apache.py}
+This will instruct Zohmg to apply the map function found in \texttt{apache.py}
 to all logs that it can find in the directory \texttt{/data/weblogs/2009/04/20}
@@
 % TODO: data interchange.
-The reducer needs to communicate to the OutputReader under what rowkey,
+The reducer needs to communicate to the OutputReader under what row key,
 column-family and qualifier to store the computed sum. It does this by
@@
 Interface, a HTTP extension where a CGI-like environment is passed around.}
-framework, which includes both middleware and application parts. The middleware
-applications are responsible for dispatching requests to the corresponding
+framework, which includes both middleware and application parts. The middle
+ware applications are responsible for dispatching requests to the corresponding
 application or file. WSGI middleware receives a WSGI request and then performs
@@
 
-The above query would return the pageviews for SE between 1 January
+The above query would return the page views for SE between 1 January
 2009 and 4 April 2009. The results from the scan is served to the user as
@@
 This chapter's first section has presented the implementation of the Zohmg
-system. The concept of Zohmg project's and management of thesame were
+system. The concept of Zohmg project's and management of the same were
 introduced; as well as how these are configured. The second section presented
@@
 
-The above query would return the pageviews for SE between 1 January
+The above query would return the page views for SE between 1 January
 2009 and 4 April 2009.
@@
 a map function (in the MapReduce sense) and letting Zohmg run that map function
-over the data set. The user's map function is called \textit{usermap function}
+over the data set. The user's map function is called \textit{user map function}
 from here on. In order to let Zohmg know what output to expect from the mapper,

### 2009-08-11 doc/report/msc-report.tex
little fixes and rewrites.

@@
 
-Analyzing data at a massive scale is one of the biggest challenges we face at
-Last.fm. Interpreting patterns in user behaviour becomes a challenge when
-millions of users interact in billions of combinations; the data sets must be
-analyzed, summarized and presented visually in order to convey their meaning.
+Analyzing data at a massive scale is one of the biggest challenges that
+Last.fm is facing. Interpreting patterns in user behaviour becomes a challenge
+when millions of users interact in billions of combinations; the data sets
+must be analyzed, summarized and presented visually.
 
@@
 
-Similar data stores exist; they generally use relational database systems as
-their backing database. The novelty of our approach is to model multidimensional
-data cubes on top of a column-oriented database, reaping the scalability
-benefits of such databases.
+Similar data stores do exist but they generally use relational database systems
+as their backing database. The novelty of our approach is to model
+multidimensional data cubes on top of a distributed, column-oriented database to
+reap the scalability benefits of such databases.
 
@@
 
-This chapters first section presents data warehouses and the concept of data
-cubes. It then introduces common operations on data cubes: projections, slicing
-and dicing, and aggregation.
+This first section of this chapter presents data warehouses and the concept of
+data cubes. It then introduces common operations on data cubes: projections,
+slicing and dicing, and aggregation.
 
@@
 
-MapReduce is a framework for running jobs in parallel across multiple
-machines. As the name implies, MapReduce is made up of two parts or
-phases: map and reduce. The map phase consumes one piece of input at a
-time (a piece of input can be a line from a log file, for example), and
-emits an intermediate key and value. These keys and values are then
-processed by the reduce phase, which emits the final key-value pairs.
+MapReduce is a framework for distributed computations on large data sets in
+parallel across multiple clustered computers. It was inspired by the map and
+reduce functions commonly found in functional programming languages.
+
+As the name implies, MapReduce is made up of two parts or phases: map and
+reduce. The map phase consumes one piece of input at a time (a piece of input
+can be a line from a log file, for example), and emits an intermediate key and
+value. These keys and values are then processed by the reduce phase, which emits
+the final key-value pairs.
 
@@
 TODO: expand.
+TODO: graph.
 

### 2009-08-11 doc/report/msc-report.tex
minor fixes in bigtable section.

@@
 
-BigTable is a data store designed for traversal of very large data sets.
-BigTable's cells are basically indexed by row key, column key, and a time-stamp.
-Each row has a key and any number of cells attached to it. The rows in a table
-are sorted by their keys. The typical use-case of BigTable is to scan a range of
-rows. It is therefor important to set up the row keys so that related data is
-close to each other. For example, if it makes sense to traverse the data in
-chronological order, the key might contain a representation of the data's
-time-stamp. The classic example is to have the reversed domain name (i.e.
-com.google.www) as the row key, which means that all sub-domains of any domain
-are next to each other (see table 2.1).
+BigTable is a data store designed for storing very large data sets. BigTable's
+cells are basically indexed by row key, column key, and a time-stamp. Each row
+has a key and any number of cells attached to it. The rows in a table are sorted
+by their keys. The typical use-case of BigTable is to scan a range of rows. It
+is therefor important to set up the row keys so that related data is close to
+each other. For example, if it makes sense to traverse the data in chronological
+order, the key might contain a representation of the data's time-stamp. The
+classic example is to have the reversed domain name (i.e. com.google.www) as the
+row key, which means that all sub-domains of any domain are next to each other
+(see table 2.1).
 
@@
 
-BigTable's data model can indeed be thought of as a sparse distributed
-multidimensional sorted map. In concept, this multidimensional map is identical
-to a nested hash table. The dimensions of this multidimensional map is mapped
-out on the row key, the column family and the column qualifier, and possibly
-also of the version time-stamp.
+BigTable's data model can indeed be thought of as a sparse, distributed, sorted,
+multidimensional map. In concept, this multidimensional map is identical to a
+nested hash table. The dimensions of the map are: row key, column family and
+column qualifier, and a version time-stamp.
 
@@
 The data model is a simple one: Rows of data are stored in labeled tables. Each
-row is indexed with a row key, a column key, and a time-stamp. The columns have
+row is indexed with a row key, a column key, and a time-stamp. The column keys have
 the form "column-family:qualifier", where the column-family is one of a number

### 2009-08-11 doc/report/msc-report.tex
updated mapreduce section.

@@
 The data model is a simple one: Rows of data are stored in labeled tables. Each
-row is indexed with a row key, a column key, and a time-stamp. The column keys have
-the form "column-family:qualifier", where the column-family is one of a number
-of fixed column-families defined by the table's schema, and the qualifier is an
-arbitrary string specified by the application. In effect, column-families are
-pre-announced when creating the table while qualifiers are not. The contents of
-each column family are stored together, so the user will want to store items
+row is indexed with a row key, a column key, and a time-stamp. The column keys
+have the form "column-family:qualifier", where the column-family is one of a
+number of fixed column-families defined by the table's schema, and the qualifier
+is an arbitrary string specified by the application. In effect, column-families
+are pre-announced when creating the table while qualifiers are not. The contents
+of each column family are stored together, so the user will want to store items
 that have similar characteristics in the same column family. Each data point is
@@
 
-MapReduce is a framework for distributed computations on large data sets in
-parallel across multiple clustered computers. It was inspired by the map and
-reduce functions commonly found in functional programming languages.
+MapReduce is a software framework for distributed computations on very large
+data sets. The computations are run in parallel across a network of computers.
 
 As the name implies, MapReduce is made up of two parts or phases: map and
-reduce. The map phase consumes one piece of input at a time (a piece of input
-can be a line from a log file, for example), and emits an intermediate key and
-value. These keys and values are then processed by the reduce phase, which emits
-the final key-value pairs.
-
-MapReduce works by splitting the input, which commonly is in the size
-range of many gigabytes, into \textit{n} parts and lets each node of the
-computing cluster work on one or more parts. The mapper on each node emits
-key-value pairs which are fed into the reduce phase. The final output is
-collated, usually on a distributed file system.
-
-TODO: expand.
-TODO: graph.
+reduce. The user specifies a map function that consumes a single key/value pair
+to generate a set of intermediate key/value pairs, and a reduce function that
+merges all intermediate values that belongs to the same key. The abstraction of
+two functions to process skmall parts of the large data set was inspired by the
+map and reduce functions commonly found in functional programming languages such
+as Lisp. The use of this model allows the framework to easily parallelize the
+computations. In the inevitable case of a failure at a computational node, the
+computations that failed can simply be re-executed on another node.
+
+MapReduce splits the input into \textit{n} pieces and assigns each computational
+node of the cluster to work on one or more piece. The map function runs on each
+node and emits key/value pairs. These are passed to the reduce function where
+the values are collated by the reduce function.
+
+[graphical element]
 

### 2009-08-12 doc/report/msc-report.tex
Report: abstract, preface, glossary, homeless text
Added swedish abstract, preface, and glossary words.
Removed Appendix C, homeless text.

@@
 \vspace{12pt}
-\noindent
 
@@
 
-Analyzing data at a massive scale is one of the biggest challenges that
-Last.fm is facing. Interpreting patterns in user behaviour becomes a challenge
-when millions of users interact in billions of combinations; the data sets
-must be analyzed, summarized and presented visually.
+\noindent Analyzing data at a massive scale is one of the biggest challenges
+that Last.fm is facing. Interpreting patterns in user behaviour becomes a
+challenge when millions of users interact in billions of combinations; the data
+sets must be analyzed, summarized and presented visually.
 
@@
 reap the scalability benefits of such databases.
+
+\vspace{48pt}
+
+\noindent {\Large \textbf{Sammanfattning}}
+\vspace{12pt}
+
+% TODO: sync with english abstract.
+
+\noindent Att analysera data p{\aa} en massiv skala {\"a}r en av de st{\"o}rsta
+utmaningarna som Last.fm st{\aa}r inf{\"o}r. Att tolka m{\"o}nster i
+anv{\"a}ndarbeteende blir en utmaning n{\"a}r miljoner anv{\"a}ndare samspelar i
+miljarder kombinationer. Datam{\"a}ngderna m{\aa}ste analyseras, summeras och
+presenteras visuellt.
+
+Detta examensarbete beskriver ett datalager f{\"o}r multidimensionell
+tidsseriebaserad data. M{\aa}tt {\"a}r summerade {\"o}ver multipla dimensioner.
+Datalagret {\"a}r optimerat f{\"o}r dataextraheringshastighet: Ett av
+designm{\aa}len {\"a}r att servera data i musklickshastighet f{\"o}r att
+fr{\"a}mja utforskning av data i realtid.
+
+Liknande datalager existerar men de anv{\"a}nder oftast relationella
+databassystem som databas f{\"o}r back-end. Nyheten i v{\aa}rt angripss{\"a}tt
+{\"a}r att modellera multidimensionella datakuber ovanp{\aa} en distribuerad,
+kolumnorienterad databas f{\"o}r att utnyttja skalbarhetsf{\"o}rdelarna av
+s{\aa}dana databaser.
+
+\pagebreak
+\phantom{ }
+\pagebreak
+
+\noindent {\Large \textbf{Preface}}
+\vspace{12pt}
+
+\noindent Per Andersson and Fredrik M{\"o}llerstrand both study Computer Science
+at Chalmers University of Technology. This report is part of their master thesis
+degree. The work was performed at Last.fm's office in London during spring 2009.
+
+The authors would like to specially thank the supervisors---David Sands at
+Chalmers and Johan Oskarsson and Martin Dittus at Last.fm---for their help and
+support.
+
+The authors would also like to thank the staff at Last.fm and their family and
+friends for moral support before, during, and after the thesis work.
+
+\pagebreak
+\phantom{ }
+\pagebreak
 
@@
 \begin{tabular}[t]{lp{100mm}}
+\textbf{Column-oriented} &
+A column-oriented database stores its content by column rather than by row.
+This is an advantage for databases such as data warehouses, where aggregates
+computed over large numbers of similar data items.
+
+ \\
+
 \textbf{JSON} &
@@
 \textbf{Replication} &
-.
+The process of sharing information for ensurance of consistency between
+redundant sources. For instance multiple databases on different machines share
+the same content for sake of redundancy, the information is not lost if one
+machine fails.
 
@@
 \textbf{Sharding} &
-.
+A method for horizontal partitioning in a database. Horizontal partinioning is a
+design principle whereby rows of a database table are held seperately. Each
+partition forms a so called \textit{shard}, which may be located on a separate
+database server. The advantage is the number of rows in each table is reduced,
+which reduces index size and thus performance.
 \end{tabular}
 
-\chapter{Text with no home and no current purpose}
-
-\section{It's not like the movies}
-
-Luckily, this is a problem that has already been solved. Google has published
-papers on the MapReduce framework \cite{mapreduce}, which allows for parallel
-processing of large data sets, and The Google File System \cite{gfs}, a
-scalable, distributed file system. The Apache Hadoop \cite{hadoop} project has
-published free software implementations of a MapReduce framework and a
-distributed file system, both of which build on the ideas in the papers
-published by Google.
-
-\section{Techniques and tools}
-
-TODO: Why do we chose HBase in favour of other techniques (i.e. SQL)?
-
-In actual practice, the main difference between the two models is that
-querying is highly restricted in a BigTable-like data store---there are no
-joins, for example---whereas the relational database was born to be queried.
-On the other hand, the restrictions placed upon the BigTable models are
-what allows it to scale to such a large data set. Indeed, BigTable was born
-to scale.
-
-TODO: elaborate on reasons to why starsql does not scale
-
-Many applications have the behaviour where they write data once and read it
-a lot more, log analysers for instance. This writing-once-reading-many
-behaviour is precisely what BigTable is designed for. BigTable is designed
-for sequential, as opposed to random, data access. A typical use case is
-the scan, where a large number of rows are read in sequence and a certain
-column or column-family are picked up and returned. Certainly, you will
-want to store those data points that are likely to be needed simultaneously
-close to each other. \cite{bigtable}
-
-Sparseness is a big thing. Null values are free of charge. TODO: *Describe
-how we leverage sparsity*.
-
-\section{Electromagnetic field theory}
-
--- A truly magnetic data store - the idea of enabling the user --
-
-From a user's perspective, storing aggregates in Zohmg is as simple as writing
-a map function (in the MapReduce sense) and letting Zohmg run that map function
-over the data set. The user's map function is called \textit{user map function}
-from here on. In order to let Zohmg know what output to expect from the mapper,
-the user also defines the dimensions and units of his data. After all, the user
-most likely knows his own data better than anyone else.
-
 \end{document}

### 2009-08-12 doc/report/msc-report.tex
report: spelling correction.

@@
 merges all intermediate values that belongs to the same key. The abstraction of
-two functions to process skmall parts of the large data set was inspired by the
+two functions to process small parts of the large data set was inspired by the
 map and reduce functions commonly found in functional programming languages such

### 2009-08-12 doc/report/msc-report.tex
fixed first part of implementation chapter.

--- 2009-08-11 doc/report/msc-report.tex
@@
 
-This chapter will present the implementation of the Zohmg system. For a usage
-manual see the appendix.
+This chapter will present the implementation of Zohmg. For a usage manual, see
+the appendix.
 
@@
 In the first section the implementation of project creation, configuration, and
-setup, will be presented. The second section features the data import
-implementation; it will present example data as well as an example user map
-function to import this. The third section presents the implementation of data
-interchange between Zohmg and other software, such as Hadoop, Dumbo, and HBase,
-but also throughout the Zohmg system. The third section also presents the
-implementation details of the Zohmg system's exporting component.
+setup will be presented. The second section features the data import
+implementation. It will present example data as well as an example user map
+function to show this. The third section presents the implementation of data
+interchange between Zohmg and other software such as Hadoop, Dumbo, and HBase.
+It also shows an overall of Zohmg. The third section also presents the
+implementation details of Zohmg's data export component.
 
@@
 
-Zohmg uses the concept of projects. These projects are self contained
-application data directories which are used with Zohmg. A Zohmg project is
-necessary for using Zohmg; they are created with a management script, which is
-also used for interacting with the Zohmg system. When the user has created a
-Zohmg project directory further actions can be performed. The Zohmg projects
-reside within a project directory, which contains five directories:
-\texttt{clients}, \texttt{config}, \texttt{lib}, \texttt{mappers}, and
-\texttt{transformers}. The details of these directories are presented throughout
-this chapter.
+Zohmg employes the concept of projects. A project is a self contained
+application data directory. A Zohmg project is necessary for using Zohmg; they
+are created with a management script, which is also used for interacting with
+the Zohmg system. When the user has created a Zohmg project directory further
+actions can be performed. The Zohmg projects reside within a project directory,
+which contains five directories: \texttt{clients}, \texttt{config},
+\texttt{lib}, and \texttt{mappers}. The details of these directories are
+presented throughout this chapter.
 
@@
 
-The Zohmg system needs configuration of the user's data sets and the environment
-for Hadoop and Dumbo. When the data set files are configured, the user runs the
-management script, which executes the configuration and creates tables in HBase
-according to the configuration.
+Zohmg needs configuration of the user's data sets and the environment for Hadoop
+and Dumbo. When the data set files are configured, the user runs the management
+script, which executes the configuration and creates tables in HBase according
+to the configuration.
 
@@
 
-% TODO: define what a data set is, and what makes it different from a project.
-
-In this context a data set is the schema that the user defines for wanted data
-sets to be handled by Zohmg. The configuration of a data set is simply a list
-of dimensions and projections, which will layout the HBase tables according to
-the schema. This configuration will also be read during the import phase, for
-handling data to be imported.
-
-The data set files reside in the project's \texttt{config} directory. The data
-sets are configured using YAML\footnote{YAML Ain't a Markup Language}, a
-human-readable data serialization format. Any human-readable configure file
-method could have been chosen. YAML was chosen because it is is very simple
-for humans to read and edit, and likewise for machines. The Python module
-PyYAML was used to implement the YAML configuration reading part. \cite{pyyaml}
+A data set is any set of data that the user has. Zohmg requires the user to
+write a configuration that describes the attributes of the data. The
+configuration of a data set is a list of dimensions and projections. This
+configuration will be read during the import phase to match analysed data
+against the schema definition.
+
+The data set configuration files reside in the project's \texttt{config}
+directory. The data sets are configured using YAML\footnote{YAML Ain't a Markup
+Language}, a human-readable data serialization format. Any human-readable
+configure file method could have been chosen. YAML was chosen because it is is
+very simple for humans to read and edit, and likewise for machines. The Python
+module PyYAML was used to implement the YAML configuration reading part.
+\cite{pyyaml}
+
+A typical configuration looks like this:
+
+\begin{lstlisting}[caption=A data set configuration example.,captionpos=b]
+dataset: submissions
+
+dimensions:
+ - artist
+ - track
+ - user
+
+projections:
+ - artist
+ - user
+ - user-artist
+ - user-track
+ - artist-track
+
+units:
+ - scrobbles
+ - loves
+ - bans
+ \end{lstlisting}
 
@@
 
-The environment is a Python script which defines run-time variables, required by
-Hadoop and Dumbo. Using a Python script for run-time variables leaves the
-problem of validation to Python. The environment configuration file resides in
-the project's \texttt{config} directory.
+The execution environment is described by a Python script. In it are defined
+run-time variables which are required by Hadoop and Dumbo, namely the Java class
+path (CLASSPATH) and the absolute path to the Hadoop directory (HADOOP_HOME).
+
+By using a Python script for run-time variables leaves the problem of validation
+to Python. The environment configuration file resides in the project's
+\texttt{config} directory.
 
@@
 
-In order to import data to Zohmg, the user writes a map function in Python which
-Zohmg wraps; interpreting the output from the mapper program and storing it in
-HBase.
+In order to import data to Zohmg, the user writes a map function in Python.
+Zohmg employs Dumbo to run the map function on the Hadoop instance defined by
+the HADOOP_HOME variable.
 
 A standard MapReduce map function emits key-value pairs. The Zohmg user map
-function, instead, emits triples (tuples with three values), which consist of a
+function emits triples (tuples with three values), which consist of a
 time-stamp, a hash table of dimensions and their respective values, and a hash
-table with measurements (values). In effect, the dimensions and the time-stamp
-specify a point in n-space and the measurements represent the observations made
-at that point.
+table with units and their respective measurements (values). In effect, the
+dimensions and the time-stamp specify a point in n-space (the key) and the
+measurements represent the observations made at that point (the value(s)).
+
+A reducer function that is specific to Zohmg is employed to perform aggregation
+on the output from the map phase. The reducer sums the measurements for each
+point in n-space and for each unit and emits the resulting aggregates.
+
+The output from the reducer is interpreted by a custom output reader. Commonly,
+an output reader stores the output of the reduce phase on disk. The output
+reader that Zohmg employs instead stores the aggregates in HBase. It is written
+in Java.
+
+\subsection*{Example of Importing}
 
@@
                     'status'    : log.statuscode}
-      values = {'pageviews' : 1,
+      measurements = {'pageviews' : 1,
                       'bytes'     : log.size}
 
-      yield log.timestamp,dimensions,measurements
+      yield log.timestamp, dimensions, measurements
 \end{lstlisting}
@@
 \begin{verbatim}
-   zohmg import mappers/apache.py \
-   /data/weblogs/2009/04/20
+   zohmg import mappers/apache.py /data/weblogs/2009/04/20
 \end{verbatim}

### 2009-08-12 doc/report/msc-report.tex
data model.

@@
 
-% TODO: renskriv.
-The typical use case for Zohmg is to read one or more column-qualifiers,
-which correspond to one or more points in n-space, over a range of time.
-Therefor, the keys are made up of the time and unit of the measurement,
-and the qualifier queried for represents the point in n-space. More on
-this in the data model section.
-
-Explain our data model, the relation between column-families, column-qualifiers
-and projections and dimensions.
+The underlying data store is HBase, which is a column-oriented database. Column
+families are pre-defined, but everything else, such as column qualifiers and
+rowkeys, can assume arbitrary values.
+
+Zohmg models the aggregated data with this in mind. An HBase table is created
+for each data set, and in the table is a single column family called 'unit'. For
+every unit that the data set contains there will be a column qualifier. For
+example, 'unit:pageviews' would be one such column qualifier in the example
+above. Employing the qualifier to encode arbitrary unit names means that the
+user can add new units to his data set configuration without having to change
+the schema of the HBase table.
+
+Zohmg stores one or more projections of the data, and each projection is made up
+of one or more dimensions in addition to the ever-present time dimension. Every
+dimension is a tuple of dimension name and dimension value. The dimension tuples
+of the projection are stored in the rowkey. The rowkey is a string; the
+dimension names and values are separated by a dash. As such, the dash is one of
+the few disallowed characters in dimension names and values.
+
+For example, a rowkey for the projection of the user agent and status code
+dimensions might look like this:
+
+\begin{lstlisting}[caption=Example of rowkeys,captionpos=b]
+    useragent-firefox-statuscode-200-20090420
+    useragent-firefox-statuscode-200-20090421
+    useragent-firefox-statuscode-200-20090422
+    ..
+    useragent-firefox-statuscode-404-20090420
+    useragent-firefox-statuscode-404-20090421
+    useragent-firefox-statuscode-404-20090422
+    ..
+    useragent-safari-statuscode-200-20090420
+    useragent-safari-statuscode-200-20090421
+    ..
+\end{lstlisting}
+
+The rowkeys are sorted lexicographically. This means that all entries for the
+user agent 'firefox' will be found close to each other. If the data analyst is
+more interested in finding all entries with status code 200, it would make sense
+to set up the projection so that status code is the first dimension.
 
@@
 
-% TODO: data interchange.
 The reducer needs to communicate to the OutputReader under what row key,
@@
 The Darling output format for HBase. Between Dumbo, which Zohmg uses for data
-import, and Hadoop JSON is used for serializing data.
+import, and JSON is used for serializing data.
 

### 2009-08-12 doc/report/msc-report.tex
subsection{Internal Data Interchange}

@@
 
-\section{Data Interchange}
-
-The reducer needs to communicate to the OutputReader under what row key,
-column-family and qualifier to store the computed sum. It does this by
-emitting the row key as the key and a JSON-serialized map of the form \\
+\subsection{Internal Data Interchange}
+
+The reducer needs to communicate to the output reader under what row and column
+the computed aggregate is to be stored. The reducer does this by emitting the
+row key as key, and a JSON-serialized hash table, which is called a dict in
+python, as value. The structure of the map is of the form \\
 \texttt{\{"column-family:qualifier" : value\}}.
 
-The OutputReader interprets the JSON and persists the data. Et voila!
-
-\subsection*{Internal}
-
-Using JSON for data interchange between map and reduce parts in Zohmg import
-jobs.
-
-For better performance typed bytes may be used. This is what Dumbo uses
-internally.
-
-\subsubsection{Input Formats}
-
-An input format is a Java class which Hadoop uses for interpreting the input
-data.
-
-The simplest input format would be the TextInputFormat, this is used to parse
-flat text files which reside on HDFS.
-
-Hadoop is able to read LZO compressed files with the LzoInputFormat. In order to
-be able to do so effectively the compressed files first have to be indexed with
-a tool that is provided with Hadoop.
-
-% TODO: When is the TableInputFormat used.
-The TableInputFormat is able read input to a Hadoop program from an HBase table.
-
-\subsubsection{Output Formats}
-
-% TODO: not truly correct
-% TODO: A detailed explanation of generally everything in the darling dir.
-The Darling output format for HBase. Between Dumbo, which Zohmg uses for data
-import, and JSON is used for serializing data.
-
-The Darling output format increased import speed by a factor ten, compared to
-using Thrift. The HBase Thrift interface is slow because it first spawns a Java
-HBase client which interprets everything from the HBase Thrift server.
-This actually adds an extra step to the interaction with HBase making it slower
-than talking to it natively with Java.
+The output reader, which is a Java class that is executed within the Hadoop
+framework, de-serializes the JSON structure and constructs an HBase object that
+is persisted to the underlying HBase data store.
 

### 2009-08-12 doc/report/msc-report.tex
more on cardinality, and removed an incorrect passage on the row key.

@@
 
-TODO: -- Limits on cardinality --
-
-[What are the limits of our system? How high cardinality do we support? (This
-is closely tied to hbase and the qualifier-limit.) If Martin is going to have
-\textit{normalized\_path} as one of his dimensions, we might run into trouble.]
+Because of the distributed nature of HBase and its ability to store a very large
+number of rows, there is no actual upper bound on the cardinality supported by
+Zohmg. There may however be complex queries that have filters on
+high-cardinality that Zohmg will be unable to answer. This can be remedied in
+part by setting up projections that cater especially to those queries, but the
+issue can not be fully avoided. The authors have successfully stored data where
+the cardinality of the dimensions exceeded millions.
 
@@
 
-It was decided that the row keys should consist of unit and time-stamp. This
-would keep related data within close range, e.g. all rows with the same unit
-would be stored next to each other. This means reading data with different units
-would imply expensive jumps within the stored data. Because of this it was
-decided that queries only support a single unit to be queried.
-
 All measures are integers. Sum is the only supported aggregation function. The
@@
 was made to only store aggregates.
-
-TODO: add summary on cardinality limits on cardinality here.
 

### 2009-08-12 doc/report/msc-report.tex
report: updated preface.

--- 2009-08-12 doc/report/msc-report.tex
@@
 reap the scalability benefits of such databases.
+
+\vspace{48pt}
+
+\noindent {\Large \textbf{Sammanfattning}}
+\vspace{12pt}
+
+% TODO: sync with english abstract.
+
+\noindent Att analysera data p{\aa} en massiv skala {\"a}r en av de st{\"o}rsta
+utmaningarna som Last.fm st{\aa}r inf{\"o}r. Att tolka m{\"o}nster i
+anv{\"a}ndarbeteende blir en utmaning n{\"a}r miljoner anv{\"a}ndare samspelar i
+miljarder kombinationer. Datam{\"a}ngderna m{\aa}ste analyseras, summeras och
+presenteras visuellt.
+
+Detta examensarbete beskriver ett datalager f{\"o}r multidimensionell
+tidsseriebaserad data. M{\aa}tt {\"a}r summerade {\"o}ver multipla dimensioner.
+Datalagret {\"a}r optimerat f{\"o}r dataextraheringshastighet: Ett av
+designm{\aa}len {\"a}r att servera data i musklickshastighet f{\"o}r att
+fr{\"a}mja utforskning av data i realtid.
+
+Liknande datalager existerar men de anv{\"a}nder oftast relationella
+databassystem som databas f{\"o}r back-end. Nyheten i v{\aa}rt angripss{\"a}tt
+{\"a}r att modellera multidimensionella datakuber ovanp{\aa} en distribuerad,
+kolumnorienterad databas f{\"o}r att utnyttja skalbarhetsf{\"o}rdelarna av
+s{\aa}dana databaser.
+
+\pagebreak
+\phantom{ }
+\pagebreak
+
+\noindent {\Large \textbf{Preface}}
+\vspace{12pt}
+
+\noindent Per Andersson and Fredrik M{\"o}llerstrand both study Computer
+Science at Chalmers University of Technology in Gothenburg, Sweden. This thesis
+is part of their master's degree. The work was performed at Last.fm's
+office in London during spring 2009.
+
+The authors would like to especially thank the supervisors---David Sands at
+Chalmers and Johan Oskarsson and Martin Dittus at Last.fm---for their help and
+support.
+
+Per would like to especially thank Hedvig Kamp for help and guidance with the
+thesis and general matters in life.
+
+The authors would also like to thank their respective families and friends, and
+the staff at Last.fm for their moral support before, during, and after the
+thesis work.
+
+\vspace{24pt}
+
+\noindent \textbf{Acknowledgements}
+\vspace{12pt}
+
+\noindent Figure 2.1 is Copyright {\copyright} 2008 Robert Burrell Donkin,
+licensed under Creative Commons Attribution 3.0 License.
+
+\pagebreak
+\phantom{ }
+\pagebreak
 
@@
 
-[graphical element]
+\vspace{12pt}
+
+\begin{figure}[h]
+    \begin{center}
+        \includegraphics[scale=0.50]{map-reduce.eps}
+        \caption{Overview of MapReduce dataflow.}
+    \end{center}
+\end{figure}
 
@@
 
-This chapter will present the implementation of Zohmg. For a usage manual, see
-the appendix.
+This chapter will present the implementation of the Zohmg system. For a usage
+manual see the appendix.
 
@@
 In the first section the implementation of project creation, configuration, and
-setup will be presented. The second section features the data import
-implementation. It will present example data as well as an example user map
-function to show this. The third section presents the implementation of data
-interchange between Zohmg and other software such as Hadoop, Dumbo, and HBase.
-It also shows an overall of Zohmg. The third section also presents the
-implementation details of Zohmg's data export component.
+setup, will be presented. The second section features the data import
+implementation; it will present example data as well as an example user map
+function to import this. The third section presents the implementation of data
+interchange between Zohmg and other software, such as Hadoop, Dumbo, and HBase,
+but also throughout the Zohmg system. The third section also presents the
+implementation details of the Zohmg system's exporting component.
 
@@
 
-Zohmg employes the concept of projects. A project is a self contained
-application data directory. A Zohmg project is necessary for using Zohmg; they
-are created with a management script, which is also used for interacting with
-the Zohmg system. When the user has created a Zohmg project directory further
-actions can be performed. The Zohmg projects reside within a project directory,
-which contains five directories: \texttt{clients}, \texttt{config},
-\texttt{lib}, and \texttt{mappers}. The details of these directories are
-presented throughout this chapter.
+Zohmg uses the concept of projects. These projects are self contained
+application data directories which are used with Zohmg. A Zohmg project is
+necessary for using Zohmg; they are created with a management script, which is
+also used for interacting with the Zohmg system. When the user has created a
+Zohmg project directory further actions can be performed. The Zohmg projects
+reside within a project directory, which contains five directories:
+\texttt{clients}, \texttt{config}, \texttt{lib}, \texttt{mappers}, and
+\texttt{transformers}. The details of these directories are presented throughout
+this chapter.
 
@@
 
-Zohmg needs configuration of the user's data sets and the environment for Hadoop
-and Dumbo. When the data set files are configured, the user runs the management
-script, which executes the configuration and creates tables in HBase according
-to the configuration.
+The Zohmg system needs configuration of the user's data sets and the environment
+for Hadoop and Dumbo. When the data set files are configured, the user runs the
+management script, which executes the configuration and creates tables in HBase
+according to the configuration.
 
@@
 
-A data set is any set of data that the user has. Zohmg requires the user to
-write a configuration that describes the attributes of the data. The
-configuration of a data set is a list of dimensions and projections. This
-configuration will be read during the import phase to match analysed data
-against the schema definition.
-
-The data set configuration files reside in the project's \texttt{config}
-directory. The data sets are configured using YAML\footnote{YAML Ain't a Markup
-Language}, a human-readable data serialization format. Any human-readable
-configure file method could have been chosen. YAML was chosen because it is is
-very simple for humans to read and edit, and likewise for machines. The Python
-module PyYAML was used to implement the YAML configuration reading part.
-\cite{pyyaml}
-
-A typical configuration looks like this:
-
-\begin{lstlisting}[caption=A data set configuration example.,captionpos=b]
-dataset: submissions
-
-dimensions:
- - artist
- - track
- - user
-
-projections:
- - artist
- - user
- - user-artist
- - user-track
- - artist-track
-
-units:
- - scrobbles
- - loves
- - bans
- \end{lstlisting}
+% TODO: define what a data set is, and what makes it different from a project.
+
+In this context a data set is the schema that the user defines for wanted data
+sets to be handled by Zohmg. The configuration of a data set is simply a list
+of dimensions and projections, which will layout the HBase tables according to
+the schema. This configuration will also be read during the import phase, for
+handling data to be imported.
+
+The data set files reside in the project's \texttt{config} directory. The data
+sets are configured using YAML\footnote{YAML Ain't a Markup Language}, a
+human-readable data serialization format. Any human-readable configure file
+method could have been chosen. YAML was chosen because it is is very simple
+for humans to read and edit, and likewise for machines. The Python module
+PyYAML was used to implement the YAML configuration reading part. \cite{pyyaml}
 
@@
 
-The execution environment is described by a Python script. In it are defined
-run-time variables which are required by Hadoop and Dumbo, namely the Java class
-path (CLASSPATH) and the absolute path to the Hadoop directory (HADOOP_HOME).
-
-By using a Python script for run-time variables leaves the problem of validation
-to Python. The environment configuration file resides in the project's
-\texttt{config} directory.
+The environment is a Python script which defines run-time variables, required by
+Hadoop and Dumbo. Using a Python script for run-time variables leaves the
+problem of validation to Python. The environment configuration file resides in
+the project's \texttt{config} directory.
 
@@
 
-In order to import data to Zohmg, the user writes a map function in Python.
-Zohmg employs Dumbo to run the map function on the Hadoop instance defined by
-the HADOOP_HOME variable.
+In order to import data to Zohmg, the user writes a map function in Python which
+Zohmg wraps; interpreting the output from the mapper program and storing it in
+HBase.
 
 A standard MapReduce map function emits key-value pairs. The Zohmg user map
-function emits triples (tuples with three values), which consist of a
+function, instead, emits triples (tuples with three values), which consist of a
 time-stamp, a hash table of dimensions and their respective values, and a hash
-table with units and their respective measurements (values). In effect, the
-dimensions and the time-stamp specify a point in n-space (the key) and the
-measurements represent the observations made at that point (the value(s)).
-
-A reducer function that is specific to Zohmg is employed to perform aggregation
-on the output from the map phase. The reducer sums the measurements for each
-point in n-space and for each unit and emits the resulting aggregates.
-
-The output from the reducer is interpreted by a custom output reader. Commonly,
-an output reader stores the output of the reduce phase on disk. The output
-reader that Zohmg employs instead stores the aggregates in HBase. It is written
-in Java.
-
-\subsection*{Example of Importing}
+table with measurements (values). In effect, the dimensions and the time-stamp
+specify a point in n-space and the measurements represent the observations made
+at that point.
 
@@
                     'status'    : log.statuscode}
-      measurements = {'pageviews' : 1,
+      values = {'pageviews' : 1,
                 'bytes'     : log.size}
 
-      yield log.timestamp, dimensions, measurements
+      yield log.timestamp,dimensions,measurements
 \end{lstlisting}
@@
 \begin{verbatim}
-   zohmg import mappers/apache.py /data/weblogs/2009/04/20
+   zohmg import mappers/apache.py \
+   /data/weblogs/2009/04/20
 \end{verbatim}
@@
 
-The underlying data store is HBase, which is a column-oriented database. Column
-families are pre-defined, but everything else, such as column qualifiers and
-rowkeys, can assume arbitrary values.
-
-Zohmg models the aggregated data with this in mind. An HBase table is created
-for each data set, and in the table is a single column family called 'unit'. For
-every unit that the data set contains there will be a column qualifier. For
-example, 'unit:pageviews' would be one such column qualifier in the example
-above. Employing the qualifier to encode arbitrary unit names means that the
-user can add new units to his data set configuration without having to change
-the schema of the HBase table.
-
-Zohmg stores one or more projections of the data, and each projection is made up
-of one or more dimensions in addition to the ever-present time dimension. Every
-dimension is a tuple of dimension name and dimension value. The dimension tuples
-of the projection are stored in the rowkey. The rowkey is a string; the
-dimension names and values are separated by a dash. As such, the dash is one of
-the few disallowed characters in dimension names and values.
-
-For example, a rowkey for the projection of the user agent and status code
-dimensions might look like this:
-
-\begin{lstlisting}[caption=Example of rowkeys,captionpos=b]
-    useragent-firefox-statuscode-200-20090420
-    useragent-firefox-statuscode-200-20090421
-    useragent-firefox-statuscode-200-20090422
-    ..
-    useragent-firefox-statuscode-404-20090420
-    useragent-firefox-statuscode-404-20090421
-    useragent-firefox-statuscode-404-20090422
-    ..
-    useragent-safari-statuscode-200-20090420
-    useragent-safari-statuscode-200-20090421
-    ..
-\end{lstlisting}
-
-The rowkeys are sorted lexicographically. This means that all entries for the
-user agent 'firefox' will be found close to each other. If the data analyst is
-more interested in finding all entries with status code 200, it would make sense
-to set up the projection so that status code is the first dimension.
+% TODO: renskriv.
+The typical use case for Zohmg is to read one or more column-qualifiers,
+which correspond to one or more points in n-space, over a range of time.
+Therefor, the keys are made up of the time and unit of the measurement,
+and the qualifier queried for represents the point in n-space. More on
+this in the data model section.
+
+Explain our data model, the relation between column-families, column-qualifiers
+and projections and dimensions.
 
@@
 
+% TODO: data interchange.
 The reducer needs to communicate to the OutputReader under what row key,
@@
 The Darling output format for HBase. Between Dumbo, which Zohmg uses for data
-import, and JSON is used for serializing data.
+import, and Hadoop JSON is used for serializing data.
 

### 2009-08-12 doc/report/msc-report.tex
report: intentionally added blank pages.
added README to appendix as verbatim

--- 2009-08-12 doc/report/msc-report.tex
@@
 run-time variables which are required by Hadoop and Dumbo, namely the Java class
-path (CLASSPATH) and the absolute path to the Hadoop directory (HADOOP_HOME).
+path and the absolute path to the Hadoop directory.
 
@@
 Zohmg employs Dumbo to run the map function on the Hadoop instance defined by
-the HADOOP_HOME variable.
+the hadoop home variable.
 
@@
 
+\chapter{Zohmg README}
+
+\begin{verbatim}
+INTRODUCING: ZOHMG.
+-------------------
+
+Zohmg is a data store for aggregation of multi-dimensional time series data. It
+is built on top of Hadoop, Dumbo and HBase. The core idea is to pre-compute
+aggregates and store them in a read-efficient manner – Zohmg is
+wasteful with storage in order to answer queries faster.
+
+This README assumes a working installation of zohmg. Please see INSTALL for
+installation instructions.
+
+Zohmg is alpha software. Be gentle.
+
+CONTACT
+-------
+
+IRC:           #zohmg at freenode
+Code:          http://github.com/zohmg/zohmg/tree/master
+User list:     http://groups.google.com/group/zohmg-user
+Dev list:      http://groups.google.com/group/zohmg-dev
+
+RUN THE ZOHMG
+-------------
+
+Zohmg is installed as a command line script. Try it out:
+
+  $> zohmg help
+
+AN EXAMPLE: TELEVISION!
+-----------------------
+
+Imagine that you are the director of operations of a web-based
+television channel. Your company streams videos, called "clips".
+
+Every time a clip is streamed across the intertubes, your software
+makes a note of it in a log file. The log contains information about
+the clip (id, length, producer) and about the user watching it
+(country of origin, player).
+
+Your objective is to analyze the logs to find out how many people
+watch a certain clip over time, broken down by country and player,
+etc.
+
+The rest of this text will show you by example how to make sense of
+logs using Zohmg.
+
+THE ANATOMY OF YOUR DATA SET
+----------------------------
+
+Each line of the logfile has the following space-delimited format:
+
+timestamp clipid producerid length usercountry player love
+
+Examples:
+1245775664 1440 201 722 GB VLC 0
+1245775680 1394 710 2512 GB VLC 1
+1245776010 1440 201 722 DE QUICKTIME 0
+
+The timestamp is a UNIX timestamp, the length is counted in seconds
+and the id's are all integers. Usercountry is a two-letter ISO
+standard code while the player is an arbitrary string.
+
+The last field -- "love" -- is a boolean that indicates whether the user
+clicked the heart shaped icon, meaning that she was truly moved by the
+clip.
+
+DO YOU SPEAK ZOHMG?
+-------------------
+
+In the parlance of Zohmg, clip and producer as well as country and
+player are dimensions. 1440, 'GB', 'VLC', etc, are called attributes of
+those dimensions.
+
+The length of a clip, whether it was loved, etc, are called
+measurements. In the configuration, which we will take a look at
+shortly, we define the units of measurements. Measurements must be
+integers so that they can be summed.
+
+Simple enough!
+
+CREATE YOUR FIRST 'PROJECT'
+---------------------------
+
+Every Zohmg project lives in its own directory. A project is created
+like so:
+
+  $> zohmg create television
+
+This creates a project directory named 'television'. Its contents are:
+
+ config     - environment and dataset configuration.
+ lib        - eggs or jars that will be automatically included in job jar.
+ mappers    - mapreduce mappers (you will write these!)
+
+CONFIGURE YOUR 'PROJECT'
+------------------------
+
+The next step is to configure environment.py and dataset.yaml.
+
+config/environment.py:
+
+Define HADOOP_HOME and set the paths for all three jars (hadoop, hadoop-streaming,
+hbase). You might need to run 'ant package' in $HADOOP_HOME to have
+the streaming jar built for you.
+
+config/dataset.yaml:
+
+Defining your data set means defining dimensions, projections and
+units.
+
+The dimensions lets Zohmg know what your data looks like while the
+projections hints at what queries you will want to ask. Once you get
+more comfortable with Zohmg you will want to optimize your projections
+for efficiency.
+
+For example, if you wish to find out how many times a certain clip has
+been watched broken down by country, you would want to set up a
+projection where clip is the first dimension and country is the second
+one.
+
+For the television station, something like the following will do just
+fine.
+
+## config/dataset.yaml
+
+dataset: television
+
+dimensions:
+ - clip
+ - producer
+ - country
+ - player
+
+projections:
+ - clip
+ - player
+ - country
+ - clip-country
+ - producer-country
+ - producer-clip-player
+
+units:
+ - plays
+ - loves
+ - seconds
+
+After you've edited environment.py and dataset.yaml:
+
+  $> zohmg setup
+
+This command creates an HBase table with the same name as your dataset.
+
+Verify that the table was created:
+
+  $> hbase shell
+  hbase(main):001:0> list
+  television
+  1 row(s) in 0.1337 seconds
+
+Brilliant!
+
+DATA IMPORT
+-----------
+
+After the project is created and setup correctly it is time to import some
+data.
+
+Data import is a process in two steps: write a map function that analyzes
+the data line for line, and run that function over the data. The data
+is normally stored on HDFS, but it is also possible to run on local data.
+
+WRITE A  MAPPER
+----------------
+
+First we'll write the mapper. It will have this signature:
+
+  def map(key, value)
+
+The 'key' argument defines the line number of the input data and is
+usually (but not always!) not very interesting. The 'value' argument is a
+string - it represents a single line of input.
+
+Analyzing a single line of input is straightforward: split the line
+on spaces and collect the debris.
+
+## mappers/mapper.py
+
+import time
+
+def map(key, value):
+    # split on space; make sure there are 7 parts.
+    parts = value.split(' ')
+    if len(parts) < 7: return
+
+    # extract values.
+    epoch = parts[0]
+    clipid, producerid, length = parts[1:4]
+    country, player, love      = parts[4:7]
+
+    # format timestamp as yyyymmdd.
+    ymd = "%d%02d%02d" % time.localtime(float(epoch))[0:3]
+
+    # dimension attributes are strings.
+    dimensions = {}
+    dimensions['clip']     = str(clipid)
+    dimensions['producer'] = str(producerid)
+    dimensions['country']  = country
+    dimensions['player']   = player
+
+    # measurements are integers.
+    measurements = {}
+    measurements['plays']   = 1
+    measurements['seconds'] = int(length)
+    measurements['loves']   = int(love)
+
+    yield ymd, dimensions, measurements
+
+The output of the mapper is a three-tuple: the first element is a
+string of format yyyymmdd (i.e. "20090601") and the other two elements
+are dictionaries.
+
+The mapper's output is fed to a reducer that sums the values of the
+units and passes the data on to the underlying data store.
+
+RUN THE MAPPER
+--------------
+
+Populate a file with a small data sample:
+
+  $> cat > data/short.log
+  1245775664 1440 201 722 GB VLC 0
+  1245775680 1394 710 2512 GB VLC 1
+  1245776010 1440 201 722 DE QUICKTIME 0
+  ^D
+
+Perform a local test-run:
+
+  $> zohmg import mappers/mapper.py data/short.log --local
+
+The first argument to import is the path to the python file containing the map
+function, the second is a path on the local file system.
+
+The local run will direct its output to a file instead of writing to
+the data store. Inspect it like so:
+
+  $> cat /tmp/zohmg-output
+  'clip-1440-country-DE-20090623' '{"unit:length": {"value": 722}}'
+  'clip-1440-country-DE-20090623' '{"unit:plays": {"value": 1}}'
+  'clip-all-country-DE-20090623'  '{"unit:length": {"value": 722}}'
+  [..]
+
+If Hadoop and HBase are up, run the mapper on the cluster:
+
+  $> zohmg import mappers/mapper.py /data/television/20090620.log
+
+Assuming the mapper finished successfully there is now some data in
+the HBase table. Verify this by firing up the HBase shell:
+
+  $> hbase shell
+  hbase(main):001:0> scan 'television'
+  [..]
+
+Lots of data scrolling by? Good! (Interrupt with CTRL-C at your leisure.)
+
+SERVE DATA
+----------
+
+Start the Zohmg server:
+
+  $> zohmg server
+
+The Zohmg server listens on port 8086 at localhost by default. Browse
+http://localhost:8086/ and have a look!
+
+THE API
+-------
+
+Zohmg's data server exposes the data store through an HTTP API. Every
+request returns a JSON-formatted string.
+
+The JSON looks something like this:
+
+[{"20090601": {"DE": 270, "US": 21, "SE": 5547}}, {"20090602": {"DE":
+9020, "US": 109, "SE": 11497}}, {"20090603": {"DE": 10091, "US": 186,
+"SE": 8863}}]
+
+The API is extremely simple: it supports a single type of GET request
+and there is no authentication.
+
+There are four required parameters: t0, t1, unit and d0.
+
+The parameters t0 and t1 define the time span. They are strings of the
+format "yyyymmdd", i.e. "t0=20090601&t1=20090630".
+
+The unit parameter defines the one unit for which you query, for
+example 'unit=plays'.
+
+The parameter d0 defines the base dimension, for example 'd0=country'.
+
+Values for d0 can be defined by setting the parameter d0v to a
+comma-separated string of values, for example "d0v=US,SE,DE". If d0v is
+empty, all values of the base dimension will be returned.
+
+A typical query string looks like this:
+
+ http://localhost:8086/?t0=20090601&t1=20090630&unit=plays&d0=country&d0v=DE,SE,US
+
+This example query would return the number of clips streamed for the
+time span between the first and last of June broken down by the
+countries Sweden, Germany and the United States.
+
+The API supports JSONP via the jsonp parameter. By setting this
+parameter, the returned JSON is wrapped in a function call.
+
+PLOTTING THE DATA
+-----------------
+
+There is an example client bundled with Zohmg. It is served by the Zohmg server
+at http://localhost:8086/graph/ and is quite useful for exploring your dataset.
+
+You may also want to peek at the javascript code to gain some inspiration and
+insight into how the beast works.
+
+The graphing tool uses Javascript to query the data server, and plots
+graphs with the help of Google Charts.
+
+KNOWN ISSUES
+------------
+
+Zohmg is alpha software and is still learning how to behave properly in mixed
+company. You will spot more than a few minor quirks while using it and might
+perhaps even run into the odd show-stopper. If you do, please let us know!
+
+Zohmg currently only supports mappers written in Python. Eventually, you will
+also be able to write mappers in Java.
+\end{verbatim}
+
 \chapter{Text with no home and no current purpose}

### 2009-08-12 doc/report/msc-report.tex
report: hypenation of swedish abstract.
report: added lists of figures and tables.
report: added front and first pages.
* chalmers-header.*: header image used on front page.
* msc-report.tex: content nad formatting of front and first pages.

--- 2009-08-12 doc/report/msc-report.tex
@@
 \usepackage{listings}
-
-\author{Per Andersson, Fredrik M{\"o}llerstrand}
-\title{Zohmg---a large scale data store for aggregated time-series-based
-data. [DRAFT]}
+\usepackage{chngpage}
 
 \begin{document}
-\maketitle
+
+\newpage
+\changepage{60mm}{}{}{-39mm}{}{-45mm}{}{}{}
+
+\renewcommand\thepage{}
+
+\begin{center}
+    \includegraphics{chalmers-header.eps}
+\end{center}
+
+\vspace{60mm}
+
+\begin{adjustwidth}[]{20mm}{-40mm}
+    {\huge \textbf{Zohmg---a large scale data store for \\
+    aggregated time-series-based data. [DRAFT]}}
+
+    \vspace{6pt}
+
+    \noindent {\large \textit{Master of Science Thesis}}
+
+    \vspace{36pt}
+
+    \noindent {\Large \textbf{PER ANDERSSON}}
+
+    \noindent {\Large \textbf{FREDRIK M{\"O}LLERSTRAND}}
+
+    \vfill
+
+    \noindent Department of Computer Science and Engineering
+
+    \noindent CHALMERS UNIVERSITY OF TECHNOLOGY
+
+    \noindent G{\"o}teborg, Sweden, 2009
+\end{adjustwidth}
+
+\newpage
+\renewcommand\thepage{\arabic{page}}
+\changepage{-60mm}{}{}{39mm}{}{47mm}{}{}{}
+
+\pagebreak
+
+\setcounter{page}{1}
+
+\noindent The Author grants to Chalmers University of Technology and University
+of Gothenburg the non-exclusive right to publish the Work electronically and in
+a non-commercial purpose make it accessible on the Internet.
+
+\vspace{12pt}
+
+\noindent The Author warrants that he/she is the author to the Work, and
+warrants that the Work does not contain text, pictures or other material that
+violates copyright law.
+
+\vspace{12pt}
+
+\noindent The Author shall, when transferring the rights of the Work to a third
+party (for example a publisher or a company), acknowledge the third party about
+this agreement. If the Author has signed a copyright agreement with a third
+party regarding the Work, the Author warrants hereby that he/she has obtained
+any necessary permission from this third party to let Chalmers University of
+Technology and University of Gothenburg store the Work electronically and make
+it accessible on the Internet.
+
+\vspace{64pt}
+
+\noindent Zohmg---a large scale data store for aggregated time-series-based
+data.
+
+\vspace{24pt}
+
+\noindent PER ANDERSSON
+
+\noindent FREDRIK M{\"O}LLERSTRAND
+
+\vspace{12pt}
+
+\noindent {\copyright} PER ANDERSSON, August 2009
+
+\noindent {\copyright} FREDRIK M{\"O}LLERSTRAND, August 2009
+
+\vspace{12pt}
+
+\noindent Examiner: DAVID SANDS
+
+\vspace{72pt}
+
+\noindent Department of Computer Science and Engineering
+
+\noindent Chalmers University of Technology
+
+\noindent SE-412 96 Göteborg
+
+\noindent Sweden
+
+\noindent Telephone + 46 (0)31-772 1000
+
+\vfill
+
+\noindent Department of Computer Science and Engineering
+
+\noindent G{\"o}teborg, Sweden, 2009
+
+\pagebreak
 
@@
 Detta examensarbete beskriver ett datalager f{\"o}r multidimensionell
-tidsseriebaserad data. M{\aa}tt {\"a}r summerade {\"o}ver multipla dimensioner.
+tidsserie-baserad data. M{\aa}tt {\"a}r summerade {\"o}ver multipla dimensioner.
 Datalagret {\"a}r optimerat f{\"o}r dataextraheringshastighet: Ett av
@@
 \pagebreak
-\phantom{ }
+This page is intentionally left blank.
 \pagebreak
@@
 \vfill
+
+\pagebreak
+
+\begin{minipage}{116mm}
+    \listoffigures
+\end{minipage}
+
+\begin{minipage}{116mm}
+    \listoftables
+\end{minipage}
+
 \pagebreak
@@
 
+\pagebreak
+This page is intentionally left blank.
+\pagebreak
+
 \chapter{System Analysis}
@@
 
+\pagebreak
+This page is intentionally left blank.
+\pagebreak
+
 \chapter{Implementation}
@@
 
+\pagebreak
+This page is intentionally left blank.
+\pagebreak
+
 \chapter{Glossary}

### 2009-08-12 doc/report/msc-report.tex
report: added map-reduce diagram.
* Makefile: now builds with dvips -Ppdf instead of pdflatex.
* TODO: removed map-reduce expand note.
* map-reduce.*: diagram of map-reduce data flow.
* msc-report.tex: included map-reduce diagram.

--- 2009-08-12 doc/report/msc-report.tex
@@
 
-\noindent Per Andersson and Fredrik M{\"o}llerstrand both study Computer
-Science at Chalmers University of Technology in Gothenburg, Sweden. This thesis
-is part of their master's degree. The work was performed at Last.fm's
-office in London during spring 2009.
-
-The authors would like to especially thank the supervisors---David Sands at
+\noindent Per Andersson and Fredrik M{\"o}llerstrand both study Computer Science
+at Chalmers University of Technology. This report is part of their master thesis
+degree. The work was performed at Last.fm's office in London during spring 2009.
+
+The authors would like to specially thank the supervisors---David Sands at
 Chalmers and Johan Oskarsson and Martin Dittus at Last.fm---for their help and
@@
 
-Per would like to especially thank Hedvig Kamp for help and guidance with the
-thesis and general matters in life.
-
-The authors would also like to thank their respective families and friends, and
-the staff at Last.fm for their moral support before, during, and after the
-thesis work.
+The authors would also like to thank the staff at Last.fm and their family and
+friends for moral support before, during, and after the thesis work.
 

### 2009-08-12 doc/report/msc-report.tex

--- 2009-08-12 doc/report/msc-report.tex
@@
 \textbf{Sharding} &
-.
+A method for horizontal partitioning in a database. Horizontal partinioning is a
+design principle whereby rows of a database table are held seperately. Each
+partition forms a so called \textit{shard}, which may be located on a separate
+database server. The advantage is the number of rows in each table is reduced,
+which reduces index size and thus performance.
 \end{tabular}
@@
 
-\chapter{Text with no home and no current purpose}
-
-\section{It's not like the movies}
-
-Luckily, this is a problem that has already been solved. Google has published
-papers on the MapReduce framework \cite{mapreduce}, which allows for parallel
-processing of large data sets, and The Google File System \cite{gfs}, a
-scalable, distributed file system. The Apache Hadoop \cite{hadoop} project has
-published free software implementations of a MapReduce framework and a
-distributed file system, both of which build on the ideas in the papers
-published by Google.
-
-\section{Techniques and tools}
-
-TODO: Why do we chose HBase in favour of other techniques (i.e. SQL)?
-
-In actual practice, the main difference between the two models is that
-querying is highly restricted in a BigTable-like data store---there are no
-joins, for example---whereas the relational database was born to be queried.
-On the other hand, the restrictions placed upon the BigTable models are
-what allows it to scale to such a large data set. Indeed, BigTable was born
-to scale.
-
-TODO: elaborate on reasons to why starsql does not scale
-
-Many applications have the behaviour where they write data once and read it
-a lot more, log analysers for instance. This writing-once-reading-many
-behaviour is precisely what BigTable is designed for. BigTable is designed
-for sequential, as opposed to random, data access. A typical use case is
-the scan, where a large number of rows are read in sequence and a certain
-column or column-family are picked up and returned. Certainly, you will
-want to store those data points that are likely to be needed simultaneously
-close to each other. \cite{bigtable}
-
-Sparseness is a big thing. Null values are free of charge. TODO: *Describe
-how we leverage sparsity*.
-
-\section{Electromagnetic field theory}
-
--- A truly magnetic data store - the idea of enabling the user --
-
-From a user's perspective, storing aggregates in Zohmg is as simple as writing
-a map function (in the MapReduce sense) and letting Zohmg run that map function
-over the data set. The user's map function is called \textit{user map function}
-from here on. In order to let Zohmg know what output to expect from the mapper,
-the user also defines the dimensions and units of his data. After all, the user
-most likely knows his own data better than anyone else.
-
 \end{document}

### 2009-08-12 doc/report/msc-report.tex

@@
 
-\noindent Per Andersson and Fredrik M{\"o}llerstrand both study Computer Science
-at Chalmers University of Technology. This report is part of their master thesis
-degree. The work was performed at Last.fm's office in London during spring 2009.
-
-The authors would like to specially thank the supervisors---David Sands at
+\noindent Per Andersson and Fredrik M{\"o}llerstrand both study Computer
+Science at Chalmers University of Technology in Gothenburg, Sweden. This thesis
+is part of their master's degree. The work was performed at Last.fm's
+office in London during spring 2009.
+
+The authors would like to especially thank the supervisors---David Sands at
 Chalmers and Johan Oskarsson and Martin Dittus at Last.fm---for their help and
@@
 
-The authors would also like to thank the staff at Last.fm and their family and
-friends for moral support before, during, and after the thesis work.
+Per would like to especially thank Hedvig Kamp for help and guidance with the
+thesis and general matters in life.
+
+The authors would also like to thank their respective families and friends, and
+the staff at Last.fm for their moral support before, during, and after the
+thesis work.
+
+\vspace{24pt}
+
+\noindent \textbf{Acknowledgements}
+\vspace{12pt}
+
+\noindent Figure 2.1 is Copyright {\copyright} 2008 Robert Burrell Donkin,
+licensed under Creative Commons Attribution 3.0 License.
 
 \pagebreak
-\phantom{ }
+This page is intentionally left blank.
 \pagebreak
@@
 storage and processing; namely Google's BigTable and MapReduce techniques.
+
+\pagebreak
+This page is intentionally left blank.
+\pagebreak
 
@@
 The last section described the scope of the thesis.
+
+\pagebreak
+This page is intentionally left blank.
+\pagebreak
 

### 2009-08-15 doc/report/msc-report.tex
added 6.1 - results

@@
 s{\aa}dana databaser.
+
+\vfill
+
+Keywords: Data warehouse, MapReduce, BigTable, Python, distributed
 
@@
 
+\section{Results}
+
+The goal of this thesis was to build a tool that allows its user to perform
+multi-dimensional analysis of a large corpus of data. That goal has been
+achieved by implementing a data store that leverages HBase as a backing store
+and utilizes the Hadoop MapReduce framework as its import functionality.
+
+The resulting software is a command line application written in Python. The
+command line application is the sole interface to the data store. From here the
+user can initiate import jobs, start a server for exporting data, and manage
+projects folders.
+
+The import jobs are executed on the Hadoop MapReduce framework. The job
+execution is delegated to Dumbo, a python module for running Hadoop jobs that
+are not written in Java. The job consist of a map function that wraps around the
+map function written by the user, and a reducer function that sums together
+measurements for each point in the multi-dimensional measurement space. The
+output from the MapReduce job is stored in an HBase table.
+
+Data is exported from the data store through an HTTP interface. Queries to the
+data store are made with HTTP GET requests. The response is encoded as JSON, a
+lightweight computer interchange format. The exports are not parallelized; they
+are performed as one or more scans over an HBase table.
+
+The data store is time series based, meaning that all data points are specified
+along the time axis, as well as any number of other dimensional axes. A single
+level of granularity is used throughout each data set. The values of dimensions
+are entirely user-definable.
+
 \section{Discussion}

### 2009-08-15 doc/report/msc-report.tex
6.2 - discussion

@@
 
-Discussion here.
+The decision to implement the import functionality on Hadoop was driven by both
+practical and environmental reasons: it is unfeasible to analyse large data sets
+on a non-distributed system so some sort of distributed computation model was
+needed to reach a satisfiable performance, and Last.fm already has a large
+Hadoop cluster and a technology culture in which Hadoop is used extensively. The
+decision to use Hadoop was in the end a non-decision, an already made
+conclusion.
+
+The reasons behind using Dumbo and writing Zohmg in Python are similar. Last.fm
+was already moving towards writing a larger set of its analysis tools in Python
+thanks to Dumbo, which is a product of another intern at the company. Python is
+a language which allows for rapid prototyping and is as such well suited for
+writing both the user's map function and indeed the whole of Zohmg.
+
+The main point of contention was which backing store to use. A number of
+different systems that fitted the requirements were tried out but none of them
+proved to be either very dissimilar from the BigTable apporach of HBase, or
+indeed ready for production. HBase has the advantage of being closely tied to
+Hadoop which means that there was already tools available for writing
+HBase-centric MapReduce jobs in Hadoop. The technology culture at Last.fm again
+proved to be the main reason: there were already projects within the company
+that were using HBase, so the needed infrastructure and know-how was there.
+
+The import facility makes use of MapReduce and a custom mapper and reducer.
+There is still the unresolved issue of whether it would be more efficient to not
+use a reducer at all and write to HBase directly from the mapper. The argument
+is that HBase is better at handling writes to random locations in the key space.
+The downside of this would be that summing for aggregation would have to be done
+at every cell update.
+
+The data export uses a simple HTTP server. The query and response fit well into
+the HTTP primitives and the ubiquity of HTTP clients and libraries makes it a
+good choice for a simple and quick interface. The HTTP server queries HBase over
+Thrift, which is quick enough but less than optimal. A more advanced solution
+would be to implement a Thrift server in Java and offer a Thrift interface to
+Zohmg. That would presumably be more efficient. There could also be situations
+where a MapReduce job would be well-served to use the aggregates as input. That
+situation would be resolved by providing an custom input format that reads
+straight from the HBase table.
+
+The decision to use JSON was based on the fact that JSON is lighweight and
+efficient to parse, and has support for the few primitives that are needed to
+describe the exported data. Also, JSON parsers are ubiquitious and libraries are
+available in almost any language.
 

### 2009-08-15 doc/report/msc-report.tex
cleaned 6.

@@
 
-Says Wikipedia: "Organizations generally start off with relatively simple use
-of data warehousing. Over time, more sophisticated use of data warehousing
-evolves." Zohmg is that relatively simple data warehouse; the seed from which
-a more sophisticated data warehouse may evolve.
+Says Wikipedia: "Organizations generally start off with relatively simple use of
+data warehousing. Over time, more sophisticated use of data warehousing
+evolves." Zohmg is a simple data warehouse; the seed from which a more
+sophisticated data warehouse may evolve.
 
 There is as of yet no support for specifying granularities. The only supported
-resolution of time is the day.
-
-The single aggregation performed is summing of the measurements. Future work
-may include average, max and min, etc.
-
-There is no metadata store. It would be very helpful for the application
-writer to know what the known values are for each dimension.
-
-It'd be interesting to think a bit about how the data store could be fully
-elastic, i.e. not require any data set definition from the user and instead
-create column-families on the fly based on the dimensions output from the
-user's map.
-
-\subsection*{Multiple Data Sets}
-
-Proper support for multiple data sets in a Zohmg project.
-
-\subsection*{Metadata}
-
-data set introspection
-
-roll-up and drill-down
-
-good when creating apps
-
-\subsection*{Updating Data Set Files}
-
-Update tables on changes in configuration---data set files.
-
-\subsection*{Importing Without MapReduce}
-
-Importing content without firing of a MapReduce job, e.g. via a Thrift
-service.
+resolution of time is the day. Future work could include adding support for
+querying data at any level of granularity.
+
+The single act of aggregation performed is the summing of the measurements.
+Future work may include average, max and min, etcetera.
+
+There is no metadata store. It would be very helpful for the application writer
+to know what the known values are for each dimension. Future work could include
+a metadata service with a simple Thrift interface at which the user can query
+for dimensions, their respective cardinalities and values, etcetera.
 
@@
 
-Last.fm is investigating the possibility for using Zohmg for multiple
-application areas, such as radio stats and web page stats.
+Last.fm will use for Zohmg in multiple instances of data analysis, most
+prominently in the field of web log analysis.
 

### 2009-08-16 doc/report/msc-report.tex
cleaned TODO's and bla's
tagged as RC1
report: expanded OLAP explanation and moved to OLAP section.

@@
 
+Modelling OLAP cubes can be done in different ways, using different techniques.
+The two main techniques are: MOLAP\footnote{Multidimensional On-Line Analytical
+Processing} (or just OLAP), and ROLAP\footnote{Relational On-Line Analytical
+Processing}. Data warehouses designed after the MOLAP principle use optimized
+multi-dimensional array storage whereas ROLAP data warehouses use relational
+databases as storage. When designing a ROLAP data warehouse a star or snowflake
+schema would be used to model the dimensions.
+
 \subsection*{Data cubes}
@@
 
-Classic data warehouses use relational databases as storage. A star or snowflake
-schema would be used to model the dimensions.
-
-New techniques that have appeared include column-oriented, schema free
-databases. Google's BigTable is such a database. Another new technique for
-distributed processing is Google's MapReduce.
-
-TODO:
-Other techniques that have been frequently used to build data warehouses
-include: A, B, and C, amongst others.
-
-A solves distributed computing by ...
-B is blablabla ...
-The framework C was ...
+One new techniques which can be used for OLAP data warehouses include Google's
+BigTable, which is a distributed column-oriented schema-less database.
+
+Google's MapReduce is a new technique for distributed parallel computations.
 
@@
 
-TODO: HBase was benchmarked (the benchmarks was similar to the ones found in the
-BigTable paper?). NEEDS DATA!!
-
 \subsection*{Prototyping}

### 2009-08-16 doc/report/msc-report.tex
report: new techniques section.

@@
 
-One new techniques which can be used for OLAP data warehouses include Google's
-BigTable, which is a distributed column-oriented schema-less database.
-
-Google's MapReduce is a new technique for distributed parallel computations.
+The OLAP field contains much research, for instance complete techniques such as
+Dwarf \cite{dwarf}, which is an OLAP engine, the Stanford Data Warehousing
+Project \cite{thestanford}, and DROLAP \cite{drolap}.
+
+Google is the originator of two recent techniques: BigTable---a distributed
+column-oriented schema-less database---and MapReduce---a framework for
+distributed parallel computations.
+
+The following sections will present and review these two techniques.
 

### 2009-08-17 doc/report/msc-report.tex
includegraphics - needs size.
little fixes.
report: fixed alignment of Zohmg diagrams.
report: de-indeed-ification

@@
 
-BigTable's data model can indeed be thought of as a sparse, distributed, sorted,
+BigTable's data model can be thought of as a sparse, distributed, sorted,
 multidimensional map. In concept, this multidimensional map is identical to a
@@
 to generate a set of intermediate key/value pairs, and a reduce function that
-merges all intermediate values that belongs to the same key. The abstraction of
-two functions to process small parts of the large data set was inspired by the
-map and reduce functions commonly found in functional programming languages such
-as Lisp. The use of this model allows the framework to easily parallelize the
-computations. In the inevitable case of a failure at a computational node, the
-computations that failed can simply be re-executed on another node.
+merges all intermediate values that belongs to the same key.
+
+MapReduce is not a new concept. The abstraction of two functions to process
+small parts of the large data set was inspired by the map and reduce functions
+commonly found in functional programming languages such as Lisp.
 
@@
 the values are collated by the reduce function.
+
+The use of this model allows the framework to easily parallelize the
+computations. In the inevitable case of a failure at a computational node, the
+computations that failed can simply be re-executed on another node.
 
@@
 While relational databases can be made highly available and reliable with load
-balancing and so called \textit{sharding}, the setup is non-trivial. Regarding
-scalability, relational databases are highly dependent on having data in working
-memory; thus they scale poorly.
-% TODO: [citation needed]
+balancing and so called \textit{sharding}, the setup is non-trivial.
 
@@
 in Java.
+
+\begin{figure}[h]
+    \begin{center}
+        \includegraphics[scale=0.40]{zohmg-import.eps}
+        \caption{Diagram over data import to Zohmg.}
+    \end{center}
+\end{figure}
 
@@
 
+\begin{figure}[h]
+    \begin{center}
+        \includegraphics[scale=0.35]{zohmg-export.eps}
+        \caption{Diagram over data export from Zohmg.}
+    \end{center}
+\end{figure}
+
 \subsection*{External}
@@
 a language which allows for rapid prototyping and is as such well suited for
-writing both the user's map function and indeed the whole of Zohmg.
+writing both the user's map function and the whole of Zohmg.
 
@@
 proved to be either very dissimilar from the BigTable apporach of HBase, or
-indeed ready for production. HBase has the advantage of being closely tied to
+ready for production. HBase has the advantage of being closely tied to
 Hadoop which means that there was already tools available for writing

### 2009-08-19 doc/report/msc-report.tex
report: fixed pagination.
report: fixed data set configuration example.
report: capitalization of headers.
report: conclusion is not needed.
report: tag RC2.
report: fixed Göteborg.
report: added little bit on performance to results.

@@
 
-\noindent SE-412 96 Göteborg
+\noindent SE-412 96 G{\"o}teborg
 
@@
 
-\subsubsection*{Slice and dice}
+\subsubsection*{Slice and Dice}
 
@@
 
-As the name implies, MapReduce is made up of two parts or phases: map and
-reduce. The user specifies a map function that consumes a single key/value pair
-to generate a set of intermediate key/value pairs, and a reduce function that
-merges all intermediate values that belongs to the same key.
-
-MapReduce is not a new concept. The abstraction of two functions to process
-small parts of the large data set was inspired by the map and reduce functions
-commonly found in functional programming languages such as Lisp.
-
-MapReduce splits the input into \textit{n} pieces and assigns each computational
-node of the cluster to work on one or more piece. The map function runs on each
-node and emits key/value pairs. These are passed to the reduce function where
-the values are collated by the reduce function.
-
-The use of this model allows the framework to easily parallelize the
-computations. In the inevitable case of a failure at a computational node, the
-computations that failed can simply be re-executed on another node.
-
 \vspace{12pt}
@@
 \end{figure}
+
+As the name implies, MapReduce is made up of two parts or phases: map and
+reduce. The user specifies a map function that consumes a single key/value pair
+to generate a set of intermediate key/value pairs, and a reduce function that
+merges all intermediate values that belongs to the same key.
+
+MapReduce is not a new concept. The abstraction of two functions to process
+small parts of the large data set was inspired by the map and reduce functions
+commonly found in functional programming languages such as Lisp.
+
+MapReduce splits the input into \textit{n} pieces and assigns each computational
+node of the cluster to work on one or more piece. The map function runs on each
+node and emits key/value pairs. These are passed to the reduce function where
+the values are collated by the reduce function.
+
+The use of this model allows the framework to easily parallelize the
+computations. In the inevitable case of a failure at a computational node, the
+computations that failed can simply be re-executed on another node.
 
@@
 
-\subsubsection{Physical storage}
+\subsubsection{Physical Storage}
 
@@
 
-\pagebreak
-This page is intentionally left blank.
-\pagebreak
-
 \chapter{Implementation}
@@
 
-A typical configuration looks like this:
+Listing 5.1 shows an example data set configuration.
+
+\vspace{12pt}
 
@@
 
-\subsection{Internal Data Interchange}
+\subsection*{Internal Data Interchange}
 
@@
 are entirely user-definable.
+
+Zohmg functionality and performance was tested with two data sets. The first one
+being Last.fm's web logs for the entirety of 2006, the second one being scrobble
+logs. Both of these data sets are several Gigabytes big and both sets have a
+cardinality of millions or more. Zohmg imported managed to import these two data
+sets and then serve the aggregates at mouse-click rate.
 
@@
 
-\section{Conclusion}
-
-Last.fm will use for Zohmg in multiple instances of data analysis, most
-prominently in the field of web log analysis.
-
 \pagebreak
@@
 \appendix
+
+\pagebreak
+This page is intentionally left blank.
+\pagebreak
 
@@
 
-\subsection*{User mappers}
+\subsection*{User Mappers}
 
@@
 
+\pagebreak
+This page is intentionally left blank.
+\pagebreak
+
 \chapter{Zohmg README}

### 2009-08-19 doc/report/msc-report.tex
report: tagged FINAL
report: tagged PRINT.0

@@
 \begin{adjustwidth}[]{20mm}{-40mm}
-    {\huge \textbf{Zohmg---a large scale data store for \\
-    aggregated time-series-based data. [DRAFT]}}
+    {\huge \textbf{Zohmg---A Large Scale Data Store for \\
+    Aggregated Time-series-based Data}}
 
@@
 
-\noindent Zohmg---a large scale data store for aggregated time-series-based
-data.
+\noindent Zohmg---A Large Scale Data Store for Aggregated Time-series-based
+Data
 

### 2009-11-13 scripts/copy.py
removed scripts/copy.py

@@ -1,21 +1,0 @@
-# Licensed to the Apache Software Foundation (ASF) under one
-# or more contributor license agreements.  See the NOTICE file
-# distributed with this work for additional information
-# regarding copyright ownership.  The ASF licenses this file
-# to you under the Apache License, Version 2.0 (the
-# "License"); you may not use this file except in compliance
-# with the License.  You may obtain a copy of the License at
-#
-#   http://www.apache.org/licenses/LICENSE-2.0
-#
-# Unless required by applicable law or agreed to in writing,
-# software distributed under the License is distributed on an
-# "AS IS" BASIS, WITHOUT WARRANTIES OR CONDITIONS OF ANY
-# KIND, either express or implied.  See the License for the
-# specific language governing permissions and limitations
-# under the License.
-
-# convenience script for the 'manual' install, hay.
-import zohmg.install
-if __name__ == "__main__":
-    zohmg.copy_files()

### 2009-11-16 src/zohmg/cmd.py
version bump: 0.2.3-dev

@@ -29,7 +29,7 @@ for (dir, dirnames, files) in os.walk(eggpath):
             sys.path.append(dir+"/"+file)
 
 # TODO: read version from $somewhere.
-version = '0.2.2-dev'
+version = '0.2.3-dev'
 
 def usage(reason = None):
     zohmg = os.path.basename(sys.argv[0])

### 2009-11-16 examples/apache/lib/apachelogparser.py
added mappers/apache.py and lib/apachelogparser.py with tests.

@@ -1,0 +1,47 @@
+import calendar, time, re
+
+class ApacheLogParser:
+    def __init__(self):
+        # regex for the 'combined' log format.
+        self.regex = re.compile(r'''
+                                ^([0-9\.]+)\          # IP address
+                                [^ ]+\                # ignore
+                                [^ ]+\                # ignore
+                                \[([^ ]+)[^\]]+\]\    # timestamp
+                                "([A-Z]+)\            # method
+                                ([^ ]+)[^"]+"\        # path
+                                ([0-9\-]+)\           # status
+                                ([0-9\-]+)\           # size
+                                "([^"]+)"\            # referrer
+                                "([^"]+)"             # agent
+                                ''', re.VERBOSE)
+        # example of log line:
+        # 85.229.87.106 - - [15/Nov/2009:18:01:25 +0000]
+        # "GET / HTTP/1.1" 200 883 "-"
+        # "Mozilla/5.0 (Macintosh; U; Intel Mac OS X 10.5; en-GB; rv:1.9.1.5)
+        # Gecko/20091102 Firefox/3.5.5"
+
+    def parse(self, line):
+        """parses log line, returns dict."""
+        parsed = {}
+        try:
+            mo = self.regex.search(line)
+            parsed['host'] = mo.group(1)
+            parsed['timestamp'], parsed['epoch'] = self.timestamps(mo.group(2))
+            parsed['method']   = mo.group(3)
+            parsed['path']     = mo.group(4)
+            parsed['status']   = mo.group(5)
+            parsed['size']     = mo.group(6)
+            parsed['referrer'] = mo.group(7)
+            parsed['agent']    = mo.group(8)
+        except AttributeError: raise ValueError()
+        return parsed
+
+    def timestamps(self, timestamp):
+        """parses timestamp of format 15/Nov/2009:18:01:25
+        assumes timestamp is in UTC
+        returns tuple of (formatted_timestamp, epoch)"""
+        time_struct = time.strptime(timestamp, "%d/%b/%Y:%H:%M:%S")
+        formatted = time.strftime("%Y-%m-%d %H:%M:%S", time_struct)
+        epoch = calendar.timegm(time_struct)
+        return formatted, epoch

### 2009-11-16 examples/apache/tests/helper.py

@@ -1,0 +1,3 @@
+import sys
+sys.path.append("../lib")
+sys.path.append("../mappers")

### 2009-11-16 examples/apache/tests/test_apachelogparser.py

@@ -1,0 +1,21 @@
+import unittest
+import helper
+from apachelogparser import ApacheLogParser
+
+class TestApacheLogParser(unittest.TestCase):
+    def test_parse(self):
+        log = '85.229.87.106 - - [15/Nov/2009:18:01:25 +0000] '\
+              '"GET / HTTP/1.1" 200 883 "-" "firefox"'
+        parsed = ApacheLogParser().parse(log)
+        self.assertEqual(parsed['host'], '85.229.87.106')
+        self.assertEqual(parsed['timestamp'], '2009-11-15 18:01:25')
+        self.assertEqual(parsed['epoch'], 1258308085)
+        self.assertEqual(parsed['method'], 'GET')
+        self.assertEqual(parsed['path'], '/')
+        self.assertEqual(parsed['status'], '200')
+        self.assertEqual(parsed['size'], '883')
+        self.assertEqual(parsed['referrer'], '-')
+        self.assertEqual(parsed['agent'], 'firefox')
+
+if __name__ == '__main__':
+    unittest.main()

### 2009-11-16 examples/apache/tests/test_mapper.py

@@ -1,0 +1,24 @@
+import unittest
+import helper
+import apache
+
+class TestMap(unittest.TestCase):
+    def test_map(self):
+        logline = '85.229.87.106 - - [15/Nov/2009:18:01:25 +0000] '\
+                  '"GET / HTTP/1.1" 200 883 "-" "firefox"'
+
+        ts, dimensions, measurements = apache.map(0, logline).next()
+
+        # ts => 20091115
+        # ds => {'status': '200', 'referrer': '-', 'agent': 'firefox',
+        #        'host': '85.229.87.106', 'path': '/', 'method': 'GET'}
+        # ms => {'requests': 1, 'bytes': 883}
+
+        self.assertEqual(ts, "20091115")
+        for d in ['host', 'method', 'path', 'status', 'referrer', 'agent']:
+            self.assertTrue(d in dimensions)
+        self.assertEqual(measurements['bytes'], 883)
+        self.assertEqual(measurements['requests'], 1)
+
+if __name__ == '__main__':
+    unittest.main()

### 2009-11-16 examples/apache/lib/useragent.py
added test for UserAgent

@@ -1,0 +1,10 @@
+class UserAgent:
+    def __init__(self, user_agent_string):
+        self.user_agent_string = user_agent_string
+        # compile regexes.
+    def classify(self):
+        pass
+    def is_browser(self):
+        pass
+    def is_agent(self):
+        pass

### 2009-11-16 examples/apache/tests/test_useragent.py

@@ -1,0 +1,14 @@
+import unittest
+import helper
+from useragent import UserAgent
+
+class TestUserAgent(unittest.TestCase):
+    def test_classify(self):
+        user_agent_string = 'Mozilla/5.0 (Macintosh; U; Intel Mac OS X 10.5; en-GB; rv:1.9.1.5) Gecko/20091102 Firefox/3.5.5'
+        ua = UserAgent(user_agent_string)
+        self.assertEqual(ua.classify(), "firefox")
+        self.assertTrue(ua.is_browser())
+        self.assertFalse(ua.is_robot())
+
+if __name__ == '__main__':
+    unittest.main()

### 2009-11-16 examples/apache/lib/useragent.py
UserAgent passes tests.

@@ -1,10 +1,45 @@
+import re
+
 class UserAgent:
+    browsers = [("Mozilla.*MSIE ([^;]+);.*"           , 'ie'),
+                ("Opera/([^ ]+).*|.*Opera ([^ ;]+).*" , 'opera'),
+                (".*Firefox/([^ ,;/()']+).*"          , 'firefox'),
+                (".* Chrome/([^ ,;/()']+).*"          , 'chrome'),
+                (".*Safari/([^ ,)]+).*"               , 'safari'),
+                (".*AppleWebKit/([^ )]+).*"           , 'webkit'), # generic webkit
+                (".*rv:([^ ;)]+)\\) Gecko/\\d+.*"     , 'gecko')]  # generic gecko
+    crawlers = [("Mozilla.*Googlebot/.*", 'google'),
+                ("Mozilla.*Ask Jeeves/Teoma.*", 'askjeeves'),
+                ("Mozilla.*Yahoo! Slurp.*", 'yahooslurp'),
+                ("facebookplatform/.*", 'facebook'),
+                ("Baiduspider.*", 'baidu'),
+                ("msnbot.*", 'msn')]
+
     def __init__(self, user_agent_string):
         self.user_agent_string = user_agent_string
         # compile regexes.
+        self.browser_patterns = self.compile_patterns(self.browsers)
+        self.crawler_patterns = self.compile_patterns(self.crawlers)
+
     def classify(self):
-        pass
+        for pattern, agent in self.browser_patterns + self.crawler_patterns:
+            if pattern.search(self.user_agent_string): return agent
+        return 'other'
+
     def is_browser(self):
-        pass
-    def is_agent(self):
-        pass
+        for pattern, agent in self.browser_patterns:
+            if pattern.search(self.user_agent_string): return True
+        return False
+
+    def is_robot(self):
+        for pattern, agent in self.crawler_patterns:
+            if pattern.search(self.user_agent_string): return True
+        return False
+
+
+    def compile_patterns(self, patterns):
+        compiled = []
+        for p, i in patterns:
+            compiled.append((re.compile(p), i))
+        return compiled
+

### 2009-11-16 examples/apache/mappers/apache.py
mappers/apache.py: added user agent classification.

@@ -2,6 +2,7 @@
 # log lines are in the 'combined' format.
 
 from apachelogparser import ApacheLogParser
+from useragent import UserAgent
 import time
 
 def ymd(epoch):
@@ -23,7 +24,7 @@ def map(key, logline):
     dimensions['path']     = parsed['path']
     dimensions['status']   = parsed['status']
     dimensions['referrer'] = parsed['referrer']
-    dimensions['agent']    = parsed['agent']
+    dimensions['agent']    = UserAgent(parsed['agent']).classify()
 
     # measurements are integers.
     measurements = {}

### 2009-11-16 examples/apache/tests/test_mapper.py

@@ -5,7 +5,8 @@ import apache
 class TestMap(unittest.TestCase):
     def test_map(self):
         logline = '85.229.87.106 - - [15/Nov/2009:18:01:25 +0000] '\
-                  '"GET / HTTP/1.1" 200 883 "-" "firefox"'
+                  '"GET / HTTP/1.1" 200 883 '\
+                  '"-" "Mozilla/5.0 (Macintosh; U; Intel Mac OS X 10.5; en-GB; rv:1.9.1.5) Gecko/20091102 Firefox/3.5.5"'
 
         ts, dimensions, measurements = apache.map(0, logline).next()
 
@@ -17,6 +18,7 @@ class TestMap(unittest.TestCase):
         self.assertEqual(ts, "20091115")
         for d in ['host', 'method', 'path', 'status', 'referrer', 'agent']:
             self.assertTrue(d in dimensions)
+        self.assertEqual(dimensions['agent'], 'firefox')
         self.assertEqual(measurements['bytes'], 883)
         self.assertEqual(measurements['requests'], 1)
 

### 2009-11-16 src/zohmg/__init__.py
removed license from empty file.

@@ -1,18 +1,0 @@
-# Licensed to the Apache Software Foundation (ASF) under one
-# or more contributor license agreements.  See the NOTICE file
-# distributed with this work for additional information
-# regarding copyright ownership.  The ASF licenses this file
-# to you under the Apache License, Version 2.0 (the
-# "License"); you may not use this file except in compliance
-# with the License.  You may obtain a copy of the License at
-#
-#   http://www.apache.org/licenses/LICENSE-2.0
-#
-# Unless required by applicable law or agreed to in writing,
-# software distributed under the License is distributed on an
-# "AS IS" BASIS, WITHOUT WARRANTIES OR CONDITIONS OF ANY
-# KIND, either express or implied.  See the License for the
-# specific language governing permissions and limitations
-# under the License.
-
-#

### 2009-11-16 src/zohmg/hbase_thrift/ttypes.py
adding auto-generated thrift code to zohmg/hbase_thrift, removing lib/egg/hbase-0.20-py2.5.egg
generated with:
$> thrift --gen py:new_style /usr/local/hbase-0.20.1/src/java/org/apache/hadoop/hbase/thrift/Hbase.thrift

@@ -1,0 +1,780 @@
+#
+# Autogenerated by Thrift
+#
+# DO NOT EDIT UNLESS YOU ARE SURE THAT YOU KNOW WHAT YOU ARE DOING
+#
+
+from thrift.Thrift import *
+
+from thrift.transport import TTransport
+from thrift.protocol import TBinaryProtocol
+try:
+  from thrift.protocol import fastbinary
+except:
+  fastbinary = None
+
+
+class TCell(object):
+  """
+  TCell - Used to transport a cell value (byte[]) and the timestamp it was
+  stored with together as a result for get and getRow methods. This promotes
+  the timestamp of a cell to a first-class value, making it easy to take
+  note of temporal data. Cell is used all the way from HStore up to HTable.
+  
+  Attributes:
+   - value
+   - timestamp
+  """
+
+  thrift_spec = (
+    None, # 0
+    (1, TType.STRING, 'value', None, None, ), # 1
+    (2, TType.I64, 'timestamp', None, None, ), # 2
+  )
+
+  def __init__(self, value=None, timestamp=None,):
+    self.value = value
+    self.timestamp = timestamp
+
+  def read(self, iprot):
+    if iprot.__class__ == TBinaryProtocol.TBinaryProtocolAccelerated and isinstance(iprot.trans, TTransport.CReadableTransport) and self.thrift_spec is not None and fastbinary is not None:
+      fastbinary.decode_binary(self, iprot.trans, (self.__class__, self.thrift_spec))
+      return
+    iprot.readStructBegin()
+    while True:
+      (fname, ftype, fid) = iprot.readFieldBegin()
+      if ftype == TType.STOP:
+        break
+      if fid == 1:
+        if ftype == TType.STRING:
+          self.value = iprot.readString();
+        else:
+          iprot.skip(ftype)
+      elif fid == 2:
+        if ftype == TType.I64:
+          self.timestamp = iprot.readI64();
+        else:
+          iprot.skip(ftype)
+      else:
+        iprot.skip(ftype)
+      iprot.readFieldEnd()
+    iprot.readStructEnd()
+
+  def write(self, oprot):
+    if oprot.__class__ == TBinaryProtocol.TBinaryProtocolAccelerated and self.thrift_spec is not None and fastbinary is not None:
+      oprot.trans.write(fastbinary.encode_binary(self, (self.__class__, self.thrift_spec)))
[…]

### 2009-11-16 src/zohmg/hbase.py
introducing ZohmgHBase

@@ -1,0 +1,49 @@
+from thrift import Thrift
+from thrift.transport import TSocket
+from thrift.transport import TTransport
+from thrift.protocol import TBinaryProtocol
+from hbase_thrift import Hbase
+from hbase_thrift.ttypes import ColumnDescriptor
+from hbase_thrift.ttypes import AlreadyExists, IOError, IllegalArgument
+
+class ZohmgHBase:
+
+    @classmethod
+    def transport(self, host='localhost'):
+        try:
+            transport = TSocket.TSocket(host, 9090)
+            transport = TTransport.TBufferedTransport(transport)
+            protocol  = TBinaryProtocol.TBinaryProtocol(transport)
+            client = Hbase.Client(protocol)
+            transport.open()
+            return client
+        except ImportError, e:
+            sys.stderr.write(str(e)+'\n')
+            sys.exit(8)
+        except Exception, e:
+            print "e: " + str(e.__class__) + " ^ " + str(e)
+            sys.stderr.write("could not setup thrift transport.\n")
+            sys.stderr.write("is the thrift server switched on?\n")
+            sys.exit(16)
+
+
+    @classmethod
+    def create_table(self, client, table_name, families=[]):
+        """."""
+        try:
+            columns = []
+            for family in families:
+                column = ColumnDescriptor(family+":")
+                columns.append(column)
+            client.createTable(table_name, columns)
+
+        except AlreadyExists:
+            sys.stderr.write("oh noes, %s already exists.\n" % t)
+            exit(2)
+        except IOError, e:
+            sys.stderr.write("bust: IOError: "+ str(e) + "\n")
+            exit(3)
+        except IllegalArgument, e:
+            sys.stderr.write("error: " + str(e) + "\n")
+            sys.stderr.write(" => bust\n")
+            exit(4)

### 2009-11-16 src/zohmg/setup.py

@@ -15,8 +15,9 @@
 # specific language governing permissions and limitations
 # under the License.
 
+import sys
+from zohmg.hbase import ZohmgHBase
 from zohmg.config import Config
-from utils import *
 
 # reads conf, creates table.
 class Setup(object):
@@ -26,11 +27,5 @@ class Setup(object):
 
         print "creating table '%s'" % dataset
 
-        try:
-            c = setup_transport("localhost")
-        except:
-            sys.stderr.write("could not setup thrift transport.\n")
-            sys.stderr.write("is the thrift server turned on?\n")
-            sys.exit(1)
-
-        create_or_bust(c, dataset, column_family)
+        client = ZohmgHBase.transport("localhost")
+        ZohmgHBase.create_table(client, dataset, column_family)

### 2009-11-16 src/zohmg/utils.py

@@ -18,47 +18,6 @@
 import string, sys, time
 from random import Random
 
-# TODO: move all hbase utils to it's own module.
-
-#
-# HBase helpers
-#
-def setup_transport(host):
-    from thrift import Thrift
-    from thrift.transport import TSocket
-    from thrift.transport import TTransport
-    from thrift.protocol import TBinaryProtocol
-    from hbase import Hbase
-    try:
-        transport = TSocket.TSocket(host, 9090)
-        transport = TTransport.TBufferedTransport(transport)
-        protocol = TBinaryProtocol.TBinaryProtocol(transport)
-        client = Hbase.Client(protocol)
-        transport.open()
-        return client
-    except:
-        raise
-
-
-def create_or_bust(c, t, cfs=['fam']):
-    from hbase.ttypes import ColumnDescriptor, AlreadyExists, IOError, IllegalArgument
-    try:
-        cds = []
-        for cf in cfs:
-            cd = ColumnDescriptor({'name' : str(cf)+":" })
-            cds.append(cd)
-        c.createTable(t, cds)
-    except AlreadyExists:
-        sys.stderr.write("oh noes, %s already exists.\n" % t)
-        exit(2)
-    except IOError:
-        sys.stderr.write("bust: IOError\n")
-        exit(3)
-    except IllegalArgument, e:
-        sys.stderr.write("error: " + str(e) + "\n")
-        sys.stderr.write("create_or_bust => bust\n")
-        exit(3)
-
 def random_string(size):
     # subopt for larger sizes.
     if size > len(string.letters):

### 2009-11-16 src/zohmg/setup.py
I always liked hearing an 'ok.'

@@ -29,3 +29,4 @@ class Setup(object):
 
         client = ZohmgHBase.transport("localhost")
         ZohmgHBase.create_table(client, dataset, column_family)
+        print 'ok.'

### 2009-11-16 src/zohmg/hbase.py
'zohmg reset' is stellar; added ZohmgHBase#delete_table &c.
:$> zohmg reset
reset will *wipe all data* in dataset 'apache'.
ARE YOU QUITE SURE? ('yes' to confirm.)
^Chyorgh!

@@ -1,3 +1,4 @@
+import sys
 from thrift import Thrift
 from thrift.transport import TSocket
 from thrift.transport import TTransport
@@ -28,8 +29,12 @@ class ZohmgHBase:
 
 
     @classmethod
-    def create_table(self, client, table_name, families=[]):
+    def create_table(self, table_name, families=[], client=None):
         """."""
+        if not client:
+            # default to localhost.
+            client = ZohmgHBase.transport("localhost")
+
         try:
             columns = []
             for family in families:
@@ -38,7 +43,7 @@ class ZohmgHBase:
             client.createTable(table_name, columns)
 
         except AlreadyExists:
-            sys.stderr.write("oh noes, %s already exists.\n" % t)
+            sys.stderr.write("oh noes, %s already exists.\n" % table_name)
             exit(2)
         except IOError, e:
             sys.stderr.write("bust: IOError: "+ str(e) + "\n")
@@ -47,3 +52,31 @@ class ZohmgHBase:
             sys.stderr.write("error: " + str(e) + "\n")
             sys.stderr.write(" => bust\n")
             exit(4)
+
+    @classmethod
+    def delete_table(self, table_name, client=None):
+        if not client:
+            # default to localhost.
+            client = ZohmgHBase.transport("localhost")
+        ZohmgHBase.disable_table(table_name, client) and \
+        ZohmgHBase.drop_table(table_name, client)
+
+    @classmethod
+    def disable_table(self, table_name, client=None):
+        try:
+            client.disableTable(table_name)
+            print "%s disabled." % table_name
+        except IOError, e:
+            print "error: %s" % e
+            raise
+        return True
+
+    @classmethod
+    def drop_table(self, table_name, client=None):
+        try:
+            client.deleteTable(table_name)
+            print "%s dropped." % table_name
+        except IOError, e:
+            print "IOError: %s" % e
+            raise
+        return True

### 2009-11-16 src/zohmg/reset.py

@@ -18,8 +18,8 @@
 import sys
 from zohmg.setup import Setup
 from zohmg.config import Config
-from zohmg.utils import setup_transport, disable, drop
-from hbase.ttypes import IOError
+from zohmg.hbase import ZohmgHBase
+from hbase_thrift.ttypes import IOError
 
 class Reset(object):
     def please(self):
@@ -30,31 +30,27 @@ class Reset(object):
         print "reset will *wipe all data* in dataset '%s'." % table
         print "ARE YOU QUITE SURE? ('yes' to confirm.)"
 
-        response = sys.stdin.readline().strip()
-        if response.lower() not in ["yes", "yes!"]:
-            print 'phew!'
+        try:
+            response = sys.stdin.readline().strip()
+            if response.lower() not in ["yes", "yes!"]:
+                print 'phew!'
+                sys.exit(0)
+        except KeyboardInterrupt:
+            print 'hyorgh!'
             sys.exit(0)
 
         # disable+drop.
-        print "ok, wiping!"
         try:
-            c = setup_transport(host)
-            disable(c, table)
-            drop(c, table)
+            print "ok, wiping!"
+            ZohmgHBase.delete_table(table)
+            # recreate.
+            print
+            print "recreating."
+            Setup().go()
         except Exception, e:
             print 'reset failed :-('
             print 'error: ' + str(e)
             sys.exit(1)
-        except IOError:
-            # problem is, hbase.ttypes.IOError isn't very expressive.
-            print 'reset failed :-('
-            print '(does the table perhaps not exist?)'
-            sys.exit(1)
-
-        # recreate (with the help of our dear friend setup).
-        print
-        print "recreating."
-        Setup().go()
 
         print
         print "%s reset'd" % table

### 2009-11-16 src/zohmg/setup.py

@@ -26,7 +26,5 @@ class Setup(object):
         column_family = ["unit"]
 
         print "creating table '%s'" % dataset
-
-        client = ZohmgHBase.transport("localhost")
-        ZohmgHBase.create_table(client, dataset, column_family)
+        ZohmgHBase.create_table(dataset, column_family)
         print 'ok.'

### 2009-11-16 src/zohmg/utils.py
drop'd hbase methods from utils.py

@@ -43,38 +43,6 @@ def timing_p(func):
         return elapsed
     return wrapper
 
-
-def disable(c, table):
-    try:
-        c.disableTable(table)
-        print "%s disabled." % table
-    except IOError, e:
-        print "error: %s" % e
-        return False
-    return True
-
-
-def enable(c, table):
-    try:
-        c.enableTable(table)
-        print "%s enabled." % table
-    except IOError, e:
-        print "error: %s" % e
-        return False
-    return True
-
-
-def drop(c, table):
-    try:
-        c.deleteTable(table)
-        print "%s dropped." % table
-    except IOError, e:
-        print "IOError: %s" % e
-        exit(128)
-    except NotFound, e:
-        print "NotFound: %s" % e
-
-
 #
 # General helpers
 #

### 2009-11-16 examples/apache/config/environment.py
added config/environment.py

@@ -1,0 +1,10 @@
+HADOOP_HOME = '/usr/local/hadoop'
+HBASE_HOME  = '/usr/local/hbase'
+
+# jars for hadoop, hadoop streaming and hbase
+# (you may need to 'ant package' in $HADOOP_HOME to build the streaming jar.)
+CLASSPATH = (
+  HADOOP_HOME + '/' + 'hadoop-0.20.1-core.jar',
+  HADOOP_HOME + '/' + 'build/contrib/streaming/hadoop-0.20.2-dev-streaming.jar',
+  HBASE_HOME  + '/' + 'hbase-0.20.1.jar'
+)

### 2009-11-16 src/zohmg/cmd.py
dfs or not.

@@ -39,7 +39,7 @@ def usage(reason = None):
     print "usage:"
     print zohmg + " create <dir>"
     print zohmg + " setup"
-    print zohmg + " import <mapper> <hdfs-input-dir> [--local] [--lzo]"
+    print zohmg + " import <mapper> <input-dir> [--local] [--lzo]"
     print zohmg + " server [--host=<host>] [--port=<port>]"
     print zohmg + " reset"
     print zohmg + " help"
